### 2016-11-05 manuscript.tex
new files

\documentclass{article}
\usepackage[margin=0.5in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{natbib}

\begin{document}
\title{Relative Fitness vs Absolute Fitness - 2d Travelling Wave}

\section*{Introduction}
\section*{Methods}
\section*{Results}
\section*{Discussion}

\section*{Appendices}
The following notes provide some basic calculations and formulas needed for constructing and running the two dimensional model.

\subsection*{A1: Deterministic Equations}
We define the quantities $D(t) = \sum_{i} N_i(t)/K=N(t)/K$ is the "measure of density dependence" (check), $N(t) = \sum_i N_i(t)$ is the total population size, $\bar{c}(t)=\sum_i c_i N_i(t)/N(t)$ is the mean relative fitness of the population, $b$ is the per-capita birth rate, and $K$ is the maximum sustainably achievable population size for any fitness class $i$.  Let $g$ be the number of genotypes.  The deterministic equations for the two dimensional model can be written as,
\begin{equation}
\begin{aligned}
\frac{dN_i}{dt}&=b N_i (1-D+{c_i \over \bar{c}}D)(1-D)-\mu_i N_i\\
&=\left(b+(1-{c_i \over \bar{c}})bD^2- \mu_i\right)N_i -b(2-{c_i \over \bar{c}})DN_i\\
\end{aligned}
\end{equation}
The coefficients $c_i = i c$ are measures of relative fitness following the model proposed by Desai and Fisher \citep{DesFish07}, and the death rates $\mu_i = i s$ follow the model proposed by \citep{BertMas15}.  The system can then be written as
\begin{equation}
\begin{aligned}
\frac{dN_i}{dt}&=\sum_{j=1}^g\left[\left(b+(1-{c_i \over \bar{c}}){bN^2\over K^2}- \mu_i\right)\delta_{ij} - {b \over K}(2-{c_i \over \bar{c}})N_i \right]N_j\\
\end{aligned}
\end{equation}
\textbf{Remark}: These equations will be modified later, but two things are worth noting.  In the previous 1d model, the steady state solutions play a role in calculating the probability of fixation $p_f$ at the nose.  Mainly, in the module pfix[] takes the assumption that the system is near steady state.  This is easy to calculate for the case of just one measure of fitness.  In the 2d situation, we can generate a mutant class at the boundary, but the steady state solution is still not known and must be computed.  It would also be good to get a rough idea of how long it takes for the solutions to approximate the steady state values (\textit{some readings of asymptotics}).
\subsection*{A2: Data Structures}
List of new data structures needed in code:
\begin{itemize}
\item{genotypes = $\{\{s_1,c_1\},...,\{s_g,c_g\}\}$ is a two dimensional array.  The first element $s_i$ isthe  absolute fitness "unscaled" and the second element $c_i$ is relative fitness "uncscaled"; both have to be multiplied by $s$ and $c$.}
\item{newmutantgenotypes = $\{\{fc_1,fk_1,\{s_1,c_1\}\},...,\{fc_g,fk_g,\{s_g,c_g\}\}\}$}.  The list includes information of the candidate new mutant genotypes at the boundary of the current set of fitness classes.  These are the $\{s_i,c_i\}\}$, while the first element $fc_i$ is the index of the fitness class that generated the mutant class at the boundary, the second element $fk_i$ indicates if it was generated at on the boundary for absolute fitness $fk_i=1$ or relative fitness $fk_i=2$.  
\end{itemize}
\subsection*{Module Prototypes}
\begin{itemize}
\item getUniqueGenotype[genotypeElem]\\
\item testGenoType[genotype, fitnesscriteria, indx]\\
\item SolveDiffEqs[$r_0$, $\mu_0$, s, c, K, genotypeabundances, genotypes, timeuntilnextenvchange]\\
\item PruneUnfitClasses[genotypes, genotypeabundances]\\
\item getNextMutants[genotypes]
\end{itemize}
\newpage
\section*{Notes From Meeting No. 1}
Program modules in this order
\begin{enumerate}
\item{Modify NDSolve to accommodate the 2 dimensional fitness space.  Define data structures for holding genotype information and modify parameters to include separate selection coefficients.}
\item{Modify existing pruning modules to prune two dimensional sets of genotypes.}
\item{Test new modules with following scenarios.}
\begin{itemize}
\item{Run system until for lengthy period to check intuition that all fitness class disappear except fittest one.}
\item{Run code up to certain time, then add new fitness class and check that new fitness class is appropriately incorporated into system, then run again until fittest class remains.}
\item{Run code along individual dimensions to test that they follow expected behavior.}
\item{Run code that simulate environmental shift and test that it follows expected behavior.}
\end{itemize}
\end{enumerate}

\section{Substitional Load}
Here we define the substitutional load for the 2D model.  The total selective deaths for each class $i=1,...,k$ are given by the quantity,
\[

\]
\begin{equation}

\end{equation}

\bibliographystyle{plain}
\bibliography{eebprojs2016bib}

\end{document}

### 2016-11-05 plots.py

@@ -1,0 +1,113 @@
+# -*- coding: utf-8 -*-
+"""
+Spyder Editor
+
+This temporary script file is located here:
+/home/dirge3141/.spyder2/.temp.py
+"""
+
+# The main purpose of this code is to get summary statistics for the 
+# simulation results as functions of the ratios of key parameters
+
+# IMPORT PACKAGES AND DEFINE LOCAL FUNCTIONS----------------------------------
+#-----------------------------------------------------------------------------
+import pickle
+import scipy as sp
+import numpy as np
+import matplotlib.pyplot as plt
+
+#reference of uploaded data types:
+getp=dict(Ur=0,Ua=1,sr=2,sa=3,K=4,b=5,d=6,alldata=7)
+
+# the first thing to do is the get some visualizations and store them somwhere
+
+# GETTING PROCESSED DATA - can only due one data set at a time due to size
+[amin, amax, aavg, times, data,select_data]=[[],[],[],[],[],[]]
+
+for expNo in range(30):
+    for i in range(16):
+        fd = open('./AllData/PythonData/simdata'+str(i)+'.data')
+        data = pickle.load(fd)
+        select_data=[data[getp['Ur']],data[getp['Ua']],data[getp['sr']],data[getp['sa']], \
+            data[getp['alldata']][expNo]]
+        amin=[24]+[val[1] for val in select_data[4]]
+        amax=[24]+[val[2] for val in select_data[4]]
+        aavg=[24]+[val[3] for val in select_data[4]]
+        times=[0]+[val[0] for val in select_data[4]]
+        fig1,axes1=plt.subplots(1,1,figsize=[8,8])
+        subfig1=plt.subplot(111)
+        plt.plot(times,amin,c='blue')
+        plt.plot(times,amax,c='blue')
+        plt.plot(times,aavg,c='red')
+        plt.ylim([0,60])
+        plt.xlim([0,90000])    
+        plt.title("Sim: "+str(i)+", Exp: "+str(expNo+1)+", Ur: "+str(select_data[0]) \
+                +", Ua: "+str(select_data[1])+", sr: "+str(select_data[2]) \
+                +", sa: "+str(select_data[3]))
+        plt.xlabel("generations")
+        plt.ylabel("absolute fitness")
+        plt.show()
[…]

### 2016-12-01 manuscript.tex
Outlines updated with more sections and comments.
more updates to outline and writeup, and some changes to testcode file.

@@
 \begin{document}
-\title{Relative Fitness vs Absolute Fitness - 2d Travelling Wave}
+\title{The Speed of Evolution in a Two Dimensional Trait Space}
 
 \section*{Introduction}
+
+%-----------------------------------------------------------------------------------------
 \section*{Methods}
+Our model describes the simultaneous evolution of two traits in asexual populations with no recombination, whose size $N$ is large but fixed.
+We consider only beneficial mutations in our framework which occur at a rate $U_i$, for i=1,2.
+A beneficial mutation modifying trait $i$ provides a fixed incremental gain $\Delta c_i$, in the relative fitness of a mutant.
+
+Our model describes the simultaneous evolution of two traits (or loci) in an asexual population with no recombination.
+The population size is taken to be large but fixed and finite.
+Each trait evolves in the concurrent mutations regime, which
+
+for the simultaneous evolution of two traits follows the construction of Desai and Fisher (2007), which permits the partitioning of the population subpopulations into deterministic and stochastic dominated.
+we make a number of assumptions for the construction of our model.
+Our model describes the simultaneous evolution of two traits (or loci) in an asexual population, whose size is large but fixed.
+We assume that there is no recombination
+
+adapting simultaneously, both evolving in the concurrent mutations regime.
+We consider only asexual populations with no recombination, whose population size is large but constant.
+Adaptation in each trait dimension is denoted $i$ for trait 1
+
+Following the work of Desai and Fisher, we designate fitness classes whose relative frequencies are given by $p_{i,j}$.  
+, which are represented by points in the $i$-$j$ trait space represented by the product space $\mathbb{N}\times \mathbb{N}$.
+
+The parameters of the model are the population size $N$, the mutation rate $U_k$ for each trait, and the traits selective advantage $\Delta c_i$.  
+\subsection*{Selection in Two Dimensions}
+The population is divided into subpopulations, called fitness classes, which 
+\subsection*{Mutation-Selection Balance in a Two Dimensional Trait Space}
+
 \section*{Results}
+
 \section*{Discussion}
@@
 \[
-
+L(t)=\max_{i,j}\left \{ d \ln(p_ij) \over dt \right \}
 \]
-\begin{equation}
-
-\end{equation}
 
 \bibliographystyle{plain}
-\bibliography{eebprojs2016bib}
+\bibliography{biblio}
 

### 2016-12-02 manuscript.tex
more changes made to methods section

@@
 \section*{Methods}
-Our model describes the simultaneous evolution of two traits in asexual populations with no recombination, whose size $N$ is large but fixed.
-We consider only beneficial mutations in our framework which occur at a rate $U_i$, for i=1,2.
-A beneficial mutation modifying trait $i$ provides a fixed incremental gain $\Delta c_i$, in the relative fitness of a mutant.
+Our model describes the simultaneous evolution of two traits (or loci) in an asexual population with no recombination.
+The population size $N$ is assumed to large but fixed.
+We consider only beneficial mutations in our framework which occur at a rate $U_k$, where $k=1,2$ designates the trait.
+A beneficial mutation modifying trait $k$ provides a fixed incremental gain $\Delta c_k$, in the relative fitness of a mutant.
+No distinction is made between the types of beneficial mutations, and fitness is strictly determined by the number of benenficial mutations in the genotype of an individual.
+Its relative fitness is given as its growth rate in excess of the population's mean growth rate $r_{i,j}-\bar{r}$, where $r_{i,j}=1+\Delta c_1 i + \Delta c_2 j$.
 
-Our model describes the simultaneous evolution of two traits (or loci) in an asexual population with no recombination.
-The population size is taken to be large but fixed and finite.
-Each trait evolves in the concurrent mutations regime, which
+The populutation is divided into subpopulations consisting of fitness classes, whose frequencies are designated by $p_{i,j}$ with $i$ consists
 

### 2016-12-06 manuscript.tex
Simulations done in test code to show joanna and jason results, and last weekend updates to the manuscript.

--- 2016-11-05 manuscript.tex
@@
 \begin{document}
-\title{Relative Fitness vs Absolute Fitness - 2d Travelling Wave}
+\title{The Speed of Evolution in a Two Dimensional Trait Space}
+\author{Kevin Gomez, Joanna Masel, Jason Bertram}
+\date{2017}
+
+\maketitle
+\newpage
 
 \section*{Introduction}
+
+%-----------------------------------------------------------------------------------------
 \section*{Methods}
+\subsection*{Model Description}
+The construction of our model employs much of the theoretical framework developed by Desai and Fisher \citep{DesFish07}, but with the key distinction of including two traits that evolve through selective and mutation forces.
+We consider asexual populations with no recombination, and assume
+
+\subsection*{Parameters, Model Basics, and Definitions}
+The population size $N$ is assumed to large but fixed.
+Beneficial mutations occur at a rate $U_k$, where $k=1,2$ designates the trait.
+A beneficial mutation modifying trait $k$ provides a fixed incremental gain $\Delta c_k$, in the relative growth rate $r_{i,j}$ of a mutant, where $r_{i,j}=1+\Delta c_1 i + \Delta c_2 j$.
+We adopt the convention of using an index $i$ and $j$ to indicate the number of beneficial mutations in the first and second trait, respectively.
+The fitness of an individual or a fitness class can be represented as the point $(i,j) \in \mathbb{N}^2$.
+
+Relative fitness is defined as its growth rate in excess of the population's mean growth rate $r_{i,j}-\bar{r}$, .
+The mean fitness of a population can also be written as $\bar{r}=1+\Delta c_1 \bar{i} + \Delta c_2 \bar{j}$, where the means are marginal with respect to the joint distribution $p_{i,j}$.
+
+The frequency of a fitness classes which grows deterministically is denoted as $p_{i,j}$, and its abundance is governed by the coupled equation defined in (1).
+\begin{equation}
+\frac{dp_{i,j}}{dt}=(r_{i,j}-\bar{r})p_{i,j}
+\end{equation}
+From this equation it follows that the relative fitness of an individual from the fitness class $(i,j)$ equals $d(\ln(p_{i,j}))/dt$.
+
+\subsection*{Adaptation in Two Dimensions}
+%
+The population is divided into subpopulations, called fitness classes, which
+
+\subsection*{Mutation-Selection Balance in a Two Dimensional Trait Space}
+
 \section*{Results}
+
 \section*{Discussion}
@@
 
-\subsection*{A1: Deterministic Equations}
-We define the quantities $D(t) = \sum_{i} N_i(t)/K=N(t)/K$ is the "measure of density dependence" (check), $N(t) = \sum_i N_i(t)$ is the total population size, $\bar{c}(t)=\sum_i c_i N_i(t)/N(t)$ is the mean relative fitness of the population, $b$ is the per-capita birth rate, and $K$ is the maximum sustainably achievable population size for any fitness class $i$.
-Let $g$ be the number of genotypes.
-The deterministic equations for the two dimensional model can be written as,
-\begin{equation}
-\begin{aligned}
-\frac{dN_i}{dt}&=b N_i (1-D+{c_i \over \bar{c}}D)(1-D)-\mu_i N_i\\
-&=\left(b+(1-{c_i \over \bar{c}})bD^2- \mu_i\right)N_i -b(2-{c_i \over \bar{c}})DN_i\\
-\end{aligned}
-\end{equation}
-The coefficients $c_i = i c$ are measures of relative fitness following the model proposed by Desai and Fisher \citep{DesFish07}, and the death rates $\mu_i = i s$ follow the model proposed by \citep{BertMas15}.
-The system can then be written as
-\begin{equation}
-\begin{aligned}
-\frac{dN_i}{dt}&=\sum_{j=1}^g\left[\left(b+(1-{c_i \over \bar{c}}){bN^2\over K^2}- \mu_i\right)\delta_{ij} - {b \over K}(2-{c_i \over \bar{c}})N_i \right]N_j\\
-\end{aligned}
-\end{equation}
-\textbf{Remark}: These equations will be modified later, but two things are worth noting.
-In the previous 1d model, the steady state solutions play a role in calculating the probability of fixation $p_f$ at the nose.
-Mainly, in the module pfix[] takes the assumption that the system is near steady state.
-This is easy to calculate for the case of just one measure of fitness.
-In the 2d situation, we can generate a mutant class at the boundary, but the steady state solution is still not known and must be computed.
-It would also be good to get a rough idea of how long it takes for the solutions to approximate the steady state values (\textit{some readings of asymptotics}).
-\subsection*{A2: Data Structures}
-List of new data structures needed in code:
-\begin{itemize}
-\item{genotypes = $\{\{s_1,c_1\},...,\{s_g,c_g\}\}$ is a two dimensional array.
-The first element $s_i$ isthe  absolute fitness "unscaled" and the second element $c_i$ is relative fitness "uncscaled"; both have to be multiplied by $s$ and $c$.}
-\item{newmutantgenotypes = $\{\{fc_1,fk_1,\{s_1,c_1\}\},...,\{fc_g,fk_g,\{s_g,c_g\}\}\}$}.
-The list includes information of the candidate new mutant genotypes at the boundary of the current set of fitness classes.
-These are the $\{s_i,c_i\}\}$, while the first element $fc_i$ is the index of the fitness class that generated the mutant class at the boundary, the second element $fk_i$ indicates if it was generated at on the boundary for absolute fitness $fk_i=1$ or relative fitness $fk_i=2$.
-\end{itemize}
-\subsection*{Module Prototypes}
-\begin{itemize}
-\item getUniqueGenotype[genotypeElem]\\
-\item testGenoType[genotype, fitnesscriteria, indx]\\
-\item SolveDiffEqs[$r_0$, $\mu_0$, s, c, K, genotypeabundances, genotypes, timeuntilnextenvchange]\\
-\item PruneUnfitClasses[genotypes, genotypeabundances]\\
-\item getNextMutants[genotypes]
-\end{itemize}
-\newpage
-\section*{Notes From Meeting No. 1}
-Program modules in this order
-\begin{enumerate}
-\item{Modify NDSolve to accommodate the 2 dimensional fitness space.  Define data structures for holding genotype information and modify parameters to include separate selection coefficients.}
-\item{Modify existing pruning modules to prune two dimensional sets of genotypes.}
-\item{Test new modules with following scenarios.}
-\begin{itemize}
-\item{Run system until for lengthy period to check intuition that all fitness class disappear except fittest one.}
-\item{Run code up to certain time, then add new fitness class and check that new fitness class is appropriately incorporated into system, then run again until fittest class remains.}
-\item{Run code along individual dimensions to test that they follow expected behavior.}
-\item{Run code that simulate environmental shift and test that it follows expected behavior.}
-\end{itemize}
-\end{enumerate}
-
-\section{Substitional Load}
-Here we define the substitutional load for the 2D model.  The total selective deaths for each class $i=1,...,k$ are given by the quantity,
-\[
-
-\]
-\begin{equation}
-
-\end{equation}
-
 \bibliographystyle{plain}
-\bibliography{eebprojs2016bib}
+\bibliography{biblio}
 

### 2016-12-09 manuscript.tex
small updates to bib and manuscript.

@@
 
+\begin{abstract}
+This paper will discuss something.
+\end{abstract}
+
 \section*{Introduction}
@@
 \section*{Methods}
-\subsection*{Model Description}
+\subsection*{A Two Trait Model with Selection and Mutations}
 The construction of our model employs much of the theoretical framework developed by Desai and Fisher \citep{DesFish07}, but with the key distinction of including two traits that evolve through selective and mutation forces.
-We consider asexual populations with no recombination, and assume
-
-\subsection*{Parameters, Model Basics, and Definitions}
-The population size $N$ is assumed to large but fixed.
-Beneficial mutations occur at a rate $U_k$, where $k=1,2$ designates the trait.
+We consider asexual populations with no recombination.
+We assume that the population size $N$ large but fixed.
+Beneficial mutations occur at a rate $U_k$, where $k=1,2$ and designates the trait in question.
 A beneficial mutation modifying trait $k$ provides a fixed incremental gain $\Delta c_k$, in the relative growth rate $r_{i,j}$ of a mutant, where $r_{i,j}=1+\Delta c_1 i + \Delta c_2 j$.
-We adopt the convention of using an index $i$ and $j$ to indicate the number of beneficial mutations in the first and second trait, respectively.
-The fitness of an individual or a fitness class can be represented as the point $(i,j) \in \mathbb{N}^2$.
-
+We adopt the convention of using an indices $i$ and $j$ to indicate the number of beneficial mutations in the first and second trait, respectively.
+This allows us to represent the fitness of an individual, or its fitness class, by a the point $(i,j) \in \mathbb{N}^2$.
+The discrete fitness distribution of the population in the two dimensional trait space is then given by the set of frequencies $\{p_{i,j}\}_{i,j}$.
 Relative fitness is defined as its growth rate in excess of the population's mean growth rate $r_{i,j}-\bar{r}$, .
@@
 From this equation it follows that the relative fitness of an individual from the fitness class $(i,j)$ equals $d(\ln(p_{i,j}))/dt$.
-
-\subsection*{Adaptation in Two Dimensions}
-%
-The population is divided into subpopulations, called fitness classes, which
 

### 2017-07-13 manuscript.tex
updated plot of fitness landscape and moved write up to original latex file.

@@
 \usepackage{natbib}
+\usepackage{graphicx}
+\graphicspath{{figures/}}
+\usepackage{subfigure}
 
 \begin{document}
-\title{The Speed of Evolution in a Two Dimensional Trait Space}
+\title{Modeling How Evolution in One Trait is Affected by Adaptation in Another}
 \author{Kevin Gomez, Joanna Masel, Jason Bertram}
@@
 \section*{Introduction}
-
-%-----------------------------------------------------------------------------------------
-\section*{Methods}
-\subsection*{A Two Trait Model with Selection and Mutations}
-The construction of our model employs much of the theoretical framework developed by Desai and Fisher \citep{DesFish07}, but with the key distinction of including two traits that evolve through selective and mutation forces.
-We consider asexual populations with no recombination.
-We assume that the population size $N$ large but fixed.
-Beneficial mutations occur at a rate $U_k$, where $k=1,2$ and designates the trait in question.
-A beneficial mutation modifying trait $k$ provides a fixed incremental gain $\Delta c_k$, in the relative growth rate $r_{i,j}$ of a mutant, where $r_{i,j}=1+\Delta c_1 i + \Delta c_2 j$.
-We adopt the convention of using an indices $i$ and $j$ to indicate the number of beneficial mutations in the first and second trait, respectively.
-This allows us to represent the fitness of an individual, or its fitness class, by a the point $(i,j) \in \mathbb{N}^2$.
-The discrete fitness distribution of the population in the two dimensional trait space is then given by the set of frequencies $\{p_{i,j}\}_{i,j}$.
-Relative fitness is defined as its growth rate in excess of the population's mean growth rate $r_{i,j}-\bar{r}$, .
-The mean fitness of a population can also be written as $\bar{r}=1+\Delta c_1 \bar{i} + \Delta c_2 \bar{j}$, where the means are marginal with respect to the joint distribution $p_{i,j}$.
-
-The frequency of a fitness classes which grows deterministically is denoted as $p_{i,j}$, and its abundance is governed by the coupled equation defined in (1).
+Natural selection frequently acts on multiple traits simultaneously.
+A single trait’s response to selection leads to a shift towards the fittest phenotype in a population’s mean.
+With multiple traits, the response to selection involves both direct and indirect effects that determine the evolution of traits \cite{Scarcelli23102007,Lovell2013,Wagner2011}.
+Indirect effects arise from genetic mechanisms that underlie trait expression which may be associated or shared in a manner that enforces specific patterns of co-expression and give rise to trait correlations.
+Plant and animal breeders preceding Darwon trait associations in domesticated stocks influence their response to artificial selection, and early breeders preceding Darwin often had to account for this in their selection programs (cite).
+Darwin’s own considerations of this fact naturally led him to posit an important role for them in adaptive evolution (Darwin p.
+144-150).
+However, the lack of quantitative descriptions hindered the exploration of have been two major approaches to address questions in this area with the common aim of elucidating further on the complex dynamics of adaptation with multiple traits.
+One of the earliest theoretical treatments of adaptation in a multivariate trait space was put forth by R.A.
+Fisher in The Genetical Theory of Natural Selection (1930).
+The model provides a simple geometric argument demonstrating how adaptive evolution would be constrained in higher dimensional trait spaces.
+Kimura (1983) gave further extensions to this model by incorporating probability of fixation, and Orr (1998, 2000) provided additional insights on the the optimal dimensions that ensured maximum adaptation.
+Some of the earliest works that attempted to address multi-trait adaptation were developed by Fisher (1930)However, early evolutionary biology could provide little development in this area, in part due to standing questions concerning the relationships between traits and their genetic foundation, such as how to reconcile Mendelian genetics and with the inheritance of continuous characters.
+Thus questions concerning whole organisms, which are of main interest to plant and animal breeders, and the level of whole organisms.
+The extension of quantitative genetics to multivariate phenotypes, formulated by Lande (Lande 1979; 1980), provides the first characterization of selection on multiple traits, formally expressed in the breeder’s equation ($\Delta z ̅=G \beta$).
+The phenotypes of quantitative traits change ($\Delta z$) in response to a selection gradient ($\beta$) acting on them, given a specific set of trait variances and covariance between them ($G$).
+Lande’s formulation is largely independent of the principles of heredity.
+However, genetic correlations are known to be the produce of either pleiotropy, linkage disequilibria, and trans acting elements.
+In practice, geneticists can estimate $\beta$ and $G$ to determine the short-term response of quantitative traits to selection \cite{lynch1998genetics}, without knowledge of the underlying causes of that create correlations between traits.
+However, most quantitative geneticist adhere to two schools of thought on this matter; these are known as the Birmingham and Edinburgh schools.
+The differences in their views stem from the types of organism they studied and the types of questions they aimed to answer.
+Quantitative geneticists in the Edinburgh school have traditionally worked on questions concerning animal breeding, and much of their statistical models aim to clarifying the process by which phenotypic quantitative traits change in response to artificial selection in animals.
+Autho
+In such populations, the genetic foundation of phenotypes is hidden, and random breeding are inevitably core elements that must be accounted for in their genetic analysis.  Consequently, 
+ arise from associations between alleles that form through either pleiotropy, linkage disequilibrium, or trans acting effects. 
+There are two schools of thought among quantitative geneticists that differ in the principle cause of associations between alleles.
+One of them holds pleiotropy as the largely responsible for  pleiotropic effects each of them The views held by each are largely the result of distinct questions and applications Geneticist in the Edinburgh school hold that genetic correlations are the product of pleiotropic effects, while those in the Birmingham school believe that linkage
+The majority of quantitative genetics follow  
+Geneticist pleiotropy.
+We focus on asexual micro-organisms that lack recombination and examine how the evolution of their traits is affected by linkage.
+A variety of organisms fall under this category, which includes many single celled Eukaryotes and a variety of bacteria (cite work).
+In this setting, linkage disequilibria is entirely determined by the combined action of mutations, selection and drift.
+Standard results in population genetics demonstrate that the amount of linkage disequilibria between various alleles will depend on the beneficial mutation rate ($U_b$), the selection coefficient ($s$), and strength of drift measured by population’s size ($N$).
+Logically, one expects variances and covariances arising from linkage to also depend on the population’s parameters, but a general expression that details this explicit relationship has not been uncovered.
+The majority of results concerning linkage comes from experiments with inbred lines in plants, stemming from the Birmingham school, and that have limited application to populations of interest to us Jinks and Mather (1981).
+In the case of adaptive traits, with fitness conferred taken as their quantitative measure, the relationship between variance and the population parameters is known explicitly.
+The work of Desai and Fisher (2007) provide the variance in fitness explicitly shown to work of gives the variance of adaptive traits in terms of the population’s parameters $N$, $U_b$, and $s$.
+The variance of traits and the covariance between them are estimated, and assume that the underlying traits satisfy basic assumptions, such as being polygenic.
+The variance in fitness was not something that could be determine from populations parameters, until only recently.
+Desai and Fisher’s work (2007) provides the rate of adaptation of the population in terms of its parameters,
 \begin{equation} 
-\frac{dp_{i,j}}{dt}=(r_{i,j}-\bar{r})p_{i,j}
+v =s^2  \frac{2 \ln(Ns)-\ln(s/U_b)}{\ln^2(s/U_b)}
 \end{equation}  
-From this equation it follows that the relative fitness of an individual from the fitness class $(i,j)$ equals $d(\ln(p_{i,j}))/dt$.
-
-\subsection*{Mutation-Selection Balance in a Two Dimensional Trait Space}
+and since the rate of adaptation is the variance of fitness ($v=\sigma^2$).
+If we consider a second adaptive trait evolving concurrently, with identical mutation rate and selection coefficient, then the rate of adaptation in each trait should be half of the total rate of adaptation $v(2U_b)$ since evolution in each contributes equally the overall adaptation of the population.
+A simple application of Desai and Fisher’s formula quickly shows that the rate of adaptation for a single trait is reduced when a second trait is considered since
+\[ 1/2  v(2U)<v(U) \]
+Naturally, we should expect such a result since we have doubled the mutation rate, and consequently, increased the amount of clonal interference that occurs within the population.
+However, the mere fact that we are also considering a second trait indicates that there to this picture than this.
+The rate of adaptation in one trait is determined by its variance and covariance with other traits, and this relation is expressed in the breeder’s equation.
+In our situation, both quantitative traits are adaptive and we measure them by the fitness they endow the organism.
+Consequently, the rate of fitness increase of the focal trait $v_1$ must be related
+\begin{equation}
+v_1=\sigma_1^2+\sigma_{1,2}
+\end{equation}
 
@@
 
+\begin{figure}
+\includegraphics[scale=1.0]{fig1_N-10p11_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
+\end{figure}
 \section*{Discussion}
 
-\section*{Appendices}
-The following notes provide some basic calculations and formulas needed for constructing and running the two dimensional model.
+\section*{Methods}
+We consider a population with fixed size N, in which individuals have two traits that contribute to its fitness, its growth rate, and which evolve through beneficial mutations; deleterious mutations are not considered.
+The beneficial mutation rate for each trait is denoted U_k (trait k=1,2), and each increases trait k’s contribution to relative fitness by s_k.
+We can express an individual fitness as
+
+\[ r_{ij}=r+i s_1+j s_2 \]
+
+where i and j represent the number of beneficial mutations in trait one and two, while $r$ is the growth rate a common ancestor to the population. 
+An individual’s relative fitness is given by
+
+\[ (r_{ij}-r ̅)=(i-\bar{i} ) s_1+(j-\bar{j} ) s_2 \]
+respectively, and i ̅ and j ̅ are the population mean.
+The two traits of individuals are treated as independently evolving loci in which beneficial mutations occur at a fixed rate $U_k$ ($k=1,2$).
+Each beneficial mutation is either lost to drift, or the subpopulation carrying it will grow sufficiently large to ensure that the beneficial mutation fixes in the population through selection.
+The probability of the latter event, denoted $\pi_{fix}$, is proportional to the selective advantage of the mutation.
+In these cases a beneficial mutation is said to have established in the population.
+The selective advantage of an individual is measured by its relative fitness, which in our model, is its expected growth rate minus the growth rate of an average individual.
+Each new beneficial mutation in trait $k$ increases this fitness by $s_k$; we do not consider deleterious and neutral mutations in our model.
+If an individual has i mutations in the first trait and $j$ in the second, while the average number carried in the two traits is $\bar{i}$ and $\bar{j}$, its relative fitness equals the difference $(r_{ij}-\bar{r})=(i-\bar{i}) s_1+(j-\bar{j}) s_2$.
+We group members of the population that carry the same number of beneficial mutations in each trait, $i$ mutations in the first and $j$ in the second, into fitness classes that will often be denoted by subscripts $ij$.
+The set of fitness class abundances $\{n_{ij} \}$ induce a distribution over the trait space, represented by $N^2$, containing all fitness classes.
+This two-dimensional distribution, which we will often denote by the set of frequencies $\{p_{ij}\}$, travels in the direction of increasing fitness as the population adapts (Fig 1) in an analogous fashion to the motion of Desai and Fisher’s (2007) one dimensional traveling wave.
+We also examine the properties of the marginal distributions $\{p_{i\cdot}\}$ and $\{p_{\cdot j}\}$ for each trait, whose behaviors over time are significantly affected by genetic correlations that arise from the confluences of mutations, selection and drift.
+\begin{equation}
+p_{i \cdot }= \sum_j p_{ij} \hspace{0.5in} p_{\cdot j}= \sum_i p_{ij}
+\end{equation}
+In order to study how genetic correlations arise and influence the evolution of traits in adapting populations, we focus on regimes of adaptation characterized by significant levels of genetic variation in each trait.
+The results for one-dimensional traveling waves provides the relationship between variance in fitness from one trait evolving and population parameters $N$, $U_k$, and $s_k$.
+When mutations appear infrequently relative to the average time required for beneficial mutation to fix, we obtain the simplest behavior possible in which beneficial mutations establish in succession and spread to the population long before the next one appears ($N U_k \ll 1/\ln(N s_k)$).
+These populations rarely have any variance in fitness stemming from trait diversity, and are not the focus of our investigations.
+We focus instead on the case where many beneficial mutations establish prior to any one fixing ($N U_k \geq 1/\ln(N s_k)$), where the effects of clonal interference cause variation in fitness which persists over time in the type of populations we consider.
+For large populations with strong selection in each trait ($N^{-1}\ll s_k \ll 1$), changes in the abundances of fitness class are either dominated by stochastic fluctuations due to drift ($n_{ij}<\pi_{fix}^{-1}$), or growth deterministically otherwise due to selection.
+We will refer to fitness classes in the first group as the front of the wave, and their growth can be modeled as branching processes that are fed new mutants from adjacent less fit classes.
+The fitness classes in the stochastic front will consist of those which are fitter and adjacent to ones that have established whenever $s_k/U_k \gg 1$.
+During the stochastic phase of growth, fitness classes are much too small to influence dynamics that are driven by selection, so that they may be ignored until the moment they do establish.
+As for the one-dimensional traveling wave, the key variable of interest is the time-between establishments τ of new beneficial mutations.
+This random variable summarizes all of the stochastic effects that accumulate over time during the growth of a fitness class, and determines the rate at which fitness classes in the stochastic front transition into the deterministic regime where selection is the prominent force.
+We will be interested in the rate of these transitions, because they will determine two very important aspects of the two-dimensional distribution’s evolution that influence trait evolution.
+The abundances of fitness classes that have established, whose behavior is dominated by the action of selection, deterministically grow and decline in size.
+We refer to the fitness classes in this group as the bulk of the population.
+The quantities i ̅ and j ̅ are marginal means computed from the marginal distributions (Eq.
+1).
+\begin{equation}
+\bar{i}=∑_i i p_{i \cdot} \hspace{0.5in} \bar{j} = ∑_j p_{⋅j} j
+\end{equation}
+Relative fitness, written as the difference $(r_{ij}-\bar{r})=(i-\bar{i}) s_1+(j-\bar{j}) s_2$, determines the growth and decline in abundances n_ij that follow the differential equation
+\begin{equation}
+\dot{n}_{ij}=n_{ij} (r_{ij}-\bar{r})-n_{ij} \left(\frac{N(t)-N}{N}\right) r 
+\end{equation}
+where $N(t)=∑_{ij} n_{ij}(t)$ above, and the right-most term is the density-regulation component of growth.
+Stochastic fluctuations in the total population size are negligible ($|N(t)-N| \ll N$), and so we may also consider the frequencies of fitness classes in the bulk to obey Eq.
+3 below.
+\begin{equation}
+\dot{p}_{ij}=p_{ij} (r_{ij}-\bar{r})
+\end{equation}
+From Eq. 3, the rate of adaptation can be expressed as the sum of the individual rates of adaptation in each trait, indicated by subscripts, as $\dot{\bar{r}}=\dot{\bar{r}}_1 +\dot{\bar{r}}_2$.
+Each of the two individual rates are related to the variance in fitness from the respective trait and its covariance in fitness with other (Eq. 4).
+\begin{equation}
+\dot{\bar{r}}_1=\sigma_1^2+\sigma_{1,2}\\
+\dot{\bar{r}}_2=\sigma_2^2+\sigma_{1,2}
+\end{equation}
+Mutation-selection balance induces steady state values for the speed of evolution in the two, from which it possible to determine the two variances, the single covariance, and furthermore, the resulting G-matrix of the two traits (Eq.
+5).
+\[
+G= \left(
+\begin{array}{cc}
+\sigma_1^2& \sigma_{1,2}\\
+\sigma_{1,2} & \sigma_2^2
+\end{array}
+\right) \]
+Linkage disequilibria continually arises throughout the adaptation process and limits the response of traits to selection.
+In the case of two alleles, linkage disequilibria is a measure of the statistical association between them.
+For simplicity, we can regard all ith beneficial mutations occurring in trait one as the same allele, and likewise with the jth mutation in trait two.
+This allows us to treat the difference $D_{ij}=p_{ij}-p_{i\cdot} p_{\cdot j}$ as a measure of linkage disequilibrium between the two alleles corresponding to the $i^{th}$ mutation in trait one, and the $j^{th}$ in trait two.
+In turn, the covariance in fitness can be written out as Eq.
+6.
+\begin{equation}
+\sigma_{1,2}=\sum_{ij}D_{ij}(i-\bar{i})s_1 (j-\bar{j}) s_2
+\end{equation}
+and it becomes evident that linkage disequilibrium directly affects the evolution of the two traits.
+We can also note from this expression that making a distinction between alleles that yields the same fitness advantage does not change the covariance.
+
+Discuss the details of the simulations.
+
+\section*{Appendix A:}
 

### 2017-07-13 plots.py

@@ -299,9 +299,12 @@ ax5 = fig5.add_subplot(111, projection='3d')
 ax5.plot_wireframe(X,Y,Z,rstride=1, cstride=1,linewidth=0.5,color='k')
 ax5.scatter(genotypes_fitness[kindx][:,0],genotypes_fitness[kindx][:,1],genotypes_fitness[kindx][:,2],c='blue',s=65,marker="s")
 ax5.view_init(elev=90,azim=-90)
-ax5.set_xlabel("Trait 1 Benificial Mutations")
-ax5.set_ylabel("Trait 2 Benificial Mutations")
-ax5.set_zlabel("Relative Fitness")
+ax5.set_xticks([])                               
+ax5.set_yticks([])                               
+ax5.set_zticks([])
+ax5.set_xlabel("Trait 1 Mutations")
+ax5.set_ylabel("Trait 2 Mutations")
+ax5.set_zlabel("Fitness")
 
 
 # second animation

### 2017-07-17 manuscript.tex
more updates

@@
-\documentclass{article}
+\documentclass[11pt,one column]{article}
 \usepackage[margin=0.5in]{geometry}
@@
 \usepackage{subfigure}
+\usepackage{multicol}
 
 \begin{document}
+\begin{multicols}{2}
+
 \title{Modeling How Evolution in One Trait is Affected by Adaptation in Another}
@@
 \date{2017}
-
 \maketitle
-\newpage
 
@@
 Natural selection frequently acts on multiple traits simultaneously.
-A single trait’s response to selection leads to a shift towards the fittest phenotype in a population’s mean.
-With multiple traits, the response to selection involves both direct and indirect effects that determine the evolution of traits \cite{Scarcelli23102007,Lovell2013,Wagner2011}.
-Indirect effects arise from genetic mechanisms that underlie trait expression which may be associated or shared in a manner that enforces specific patterns of co-expression and give rise to trait correlations.
-Plant and animal breeders preceding Darwon trait associations in domesticated stocks influence their response to artificial selection, and early breeders preceding Darwin often had to account for this in their selection programs (cite).
+A single trait’s response to selection leads to a shift in the population mean towards the fittest phenotype.
+The response to selection with multiple traits involves both direct and indirect effects that together determine their evolution \cite{Scarcelli23102007,Lovell2013,Wagner2011}.
+Indirect effects arise from genetic mechanisms that underlie trait expression which may be associated or shared in a manner that enforces specific patterns of co-expression and give rise to trait correlations.\par
+
+Evolutionary theory for the in our understanding of trait correlations and their role in evolution surfaced as the conAdvances in our understanding of trait correlatiosn and their role in evolution have been limited, Plant and animal breeders preceding Darwin trait associations in domesticated stocks influence their response to artificial selection, and early breeders preceding Darwin often had to account for this in their selection programs (cite).
 Darwin’s own considerations of this fact naturally led him to posit an important role for them in adaptive evolution (Darwin p.
@@
 Some of the earliest works that attempted to address multi-trait adaptation were developed by Fisher (1930)However, early evolutionary biology could provide little development in this area, in part due to standing questions concerning the relationships between traits and their genetic foundation, such as how to reconcile Mendelian genetics and with the inheritance of continuous characters.
-Thus questions concerning whole organisms, which are of main interest to plant and animal breeders, and the level of whole organisms.
+Thus questions concerning whole organisms, which are of main interest to plant and animal breeders, and the level of whole organisms.\par
+
 The extension of quantitative genetics to multivariate phenotypes, formulated by Lande (Lande 1979; 1980), provides the first characterization of selection on multiple traits, formally expressed in the breeder’s equation ($\Delta z ̅=G \beta$).
@@
 Lande’s formulation is largely independent of the principles of heredity.
-However, genetic correlations are known to be the produce of either pleiotropy, linkage disequilibria, and trans acting elements.
+However, genetic correlations are known to be the produce of either pleiotropy, linkage disequilibria, and trans acting elements.\par
 In practice, geneticists can estimate $\beta$ and $G$ to determine the short-term response of quantitative traits to selection \cite{lynch1998genetics}, without knowledge of the underlying causes of that create correlations between traits.
 However, most quantitative geneticist adhere to two schools of thought on this matter; these are known as the Birmingham and Edinburgh schools.
-The differences in their views stem from the types of organism they studied and the types of questions they aimed to answer.
+The differences in their views stem from the types of organism they studied and the types of questions they aimed to answer.\par
+
 Quantitative geneticists in the Edinburgh school have traditionally worked on questions concerning animal breeding, and much of their statistical models aim to clarifying the process by which phenotypic quantitative traits change in response to artificial selection in animals.
 Autho
-In such populations, the genetic foundation of phenotypes is hidden, and random breeding are inevitably core elements that must be accounted for in their genetic analysis.  Consequently, 
- arise from associations between alleles that form through either pleiotropy, linkage disequilibrium, or trans acting effects. 
+In such populations, the genetic foundation of phenotypes is hidden, and random breeding are inevitably core elements that must be accounted for in their genetic analysis.
+Consequently, arise from associations between alleles that form through either pleiotropy, linkage disequilibrium, or trans acting effects.\par
+
 There are two schools of thought among quantitative geneticists that differ in the principle cause of associations between alleles.
-One of them holds pleiotropy as the largely responsible for  pleiotropic effects each of them The views held by each are largely the result of distinct questions and applications Geneticist in the Edinburgh school hold that genetic correlations are the product of pleiotropic effects, while those in the Birmingham school believe that linkage
-The majority of quantitative genetics follow  
-Geneticist pleiotropy.
+One of them holds pleiotropy as the largely responsible for  pleiotropic effects each of them The views held by each are largely the result of distinct questions and applications Geneticist in the Edinburgh school hold that genetic correlations are the product of pleiotropic effects, while those in the Birmingham school believe that linkage The majority of quantitative genetics follow Geneticist pleiotropy.\par
+
 We focus on asexual micro-organisms that lack recombination and examine how the evolution of their traits is affected by linkage.
@@
 Logically, one expects variances and covariances arising from linkage to also depend on the population’s parameters, but a general expression that details this explicit relationship has not been uncovered.
-The majority of results concerning linkage comes from experiments with inbred lines in plants, stemming from the Birmingham school, and that have limited application to populations of interest to us Jinks and Mather (1981).
+The majority of results concerning linkage comes from experiments with inbred lines in plants, stemming from the Birmingham school, and that have limited application to populations of interest to us Jinks and Mather (1981).\par
+
 In the case of adaptive traits, with fitness conferred taken as their quantitative measure, the relationship between variance and the population parameters is known explicitly.
@@
 Naturally, we should expect such a result since we have doubled the mutation rate, and consequently, increased the amount of clonal interference that occurs within the population.
-However, the mere fact that we are also considering a second trait indicates that there to this picture than this.
+However, the mere fact that we are also considering a second trait indicates that there to this picture than this.\par
+
 The rate of adaptation in one trait is determined by its variance and covariance with other traits, and this relation is expressed in the breeder’s equation.
@@
 \section*{Results}
-
+Discuss results here
 \begin{figure}
-\includegraphics[scale=1.0]{fig1_N-10p11_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
+\includegraphics[scale=1.0]{fig2_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
 \end{figure}
 \section*{Discussion}
-
 \section*{Methods}
 We consider a population with fixed size N, in which individuals have two traits that contribute to its fitness, its growth rate, and which evolve through beneficial mutations; deleterious mutations are not considered.
-The beneficial mutation rate for each trait is denoted U_k (trait k=1,2), and each increases trait k’s contribution to relative fitness by s_k.
+The beneficial mutation rate for each trait is denoted $U_k$ (trait $k=1,2$), and each increases trait $k$’s contribution to relative fitness by $s_k$.
 We can express an individual fitness as
-
 \[ r_{ij}=r+i s_1+j s_2 \]
-
 where i and j represent the number of beneficial mutations in trait one and two, while $r$ is the growth rate a common ancestor to the population. 
 An individual’s relative fitness is given by
-
 \[ (r_{ij}-r ̅)=(i-\bar{i} ) s_1+(j-\bar{j} ) s_2 \]
@@
 \end{equation}
-Relative fitness, written as the difference $(r_{ij}-\bar{r})=(i-\bar{i}) s_1+(j-\bar{j}) s_2$, determines the growth and decline in abundances n_ij that follow the differential equation
+Relative fitness, written as the difference $(r_{ij}-\bar{r})=(i-\bar{i}) s_1+(j-\bar{j}) s_2$, determines the growth and decline in abundances $n_{ij}$ that follow the differential equation
 \begin{equation}
@@
 
-\section*{Appendix A:}
-
 \bibliographystyle{plain}
 \bibliography{biblio}
+\end{multicols}
 
+\newpage
+
+\section*{Appendix A:}
+\section*{Appendix B:}
+\section*{Appendix C:}
 \end{document}

### 2017-07-17 manuscript.tex
more updates

@@
 
-Evolutionary theory for the in our understanding of trait correlations and their role in evolution surfaced as the conAdvances in our understanding of trait correlatiosn and their role in evolution have been limited, Plant and animal breeders preceding Darwin trait associations in domesticated stocks influence their response to artificial selection, and early breeders preceding Darwin often had to account for this in their selection programs (cite).
-Darwin’s own considerations of this fact naturally led him to posit an important role for them in adaptive evolution (Darwin p.
+Darwin himself realized that trait correlations could significantly influence adaptive evolution (Darwin p.
 144-150).
-However, the lack of quantitative descriptions hindered the exploration of have been two major approaches to address questions in this area with the common aim of elucidating further on the complex dynamics of adaptation with multiple traits.
+However, the first theoretical insights  but it would take many more years and the insights of the modern synthesis before evolutionary biologists could begin addressing questions in this area.
+The lack of quantitative descriptions hindered the exploration of have been two major approaches to address questions in this area with the common aim of elucidating further on the complex dynamics of adaptation with multiple traits.
 One of the earliest theoretical treatments of adaptation in a multivariate trait space was put forth by R.A.
@@
 \section*{Appendix A:}
-\section*{Appendix B:}
-\section*{Appendix C:}
+
 \end{document}

### 2017-07-20 manuscript.tex
more updates

@@
 \usepackage{graphicx}
+\usepackage{capt-of}
+\usepackage{multicol}
+
 \graphicspath{{figures/}}
-\usepackage{subfigure}
-\usepackage{multicol}
 
@@
 
-Darwin himself realized that trait correlations could significantly influence adaptive evolution (Darwin p.
+Trait correlations are statistical measures of associated co-expression in traits primarily used in quantitative genetics.  Within the
+
+Evolution with multiple traits can produce much more complex responses in populations as they adapt; a fact that was clear to Darwin (Darwin p.
 144-150).
-However, the first theoretical insights  but it would take many more years and the insights of the modern synthesis before evolutionary biologists could begin addressing questions in this area.
-The lack of quantitative descriptions hindered the exploration of have been two major approaches to address questions in this area with the common aim of elucidating further on the complex dynamics of adaptation with multiple traits.
-One of the earliest theoretical treatments of adaptation in a multivariate trait space was put forth by R.A.
-Fisher in The Genetical Theory of Natural Selection (1930).
-The model provides a simple geometric argument demonstrating how adaptive evolution would be constrained in higher dimensional trait spaces.
-Kimura (1983) gave further extensions to this model by incorporating probability of fixation, and Orr (1998, 2000) provided additional insights on the the optimal dimensions that ensured maximum adaptation.
-Some of the earliest works that attempted to address multi-trait adaptation were developed by Fisher (1930)However, early evolutionary biology could provide little development in this area, in part due to standing questions concerning the relationships between traits and their genetic foundation, such as how to reconcile Mendelian genetics and with the inheritance of continuous characters.
-Thus questions concerning whole organisms, which are of main interest to plant and animal breeders, and the level of whole organisms.\par
-
+Today, evolutionary biologists are in agreement that these indirect influences stemming from simultaneous trait evolution have important consequences that affect populations on a micro- and macro-evolutionary scale.
+Plant and animal breeders have been keen on this fact for much longer, and not surprisingly, our most comprehensive theory of adaptation in complex organisms with multiple traits was developed from the quantitative tools used in by them.
 The extension of quantitative genetics to multivariate phenotypes, formulated by Lande (Lande 1979; 1980), provides the first characterization of selection on multiple traits, formally expressed in the breeder’s equation ($\Delta z ̅=G \beta$).
@@
 However, genetic correlations are known to be the produce of either pleiotropy, linkage disequilibria, and trans acting elements.\par
+
 In practice, geneticists can estimate $\beta$ and $G$ to determine the short-term response of quantitative traits to selection \cite{lynch1998genetics}, without knowledge of the underlying causes of that create correlations between traits.
@@
 Discuss results here
-\begin{figure}
-\includegraphics[scale=1.0]{fig2_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
-\end{figure}
+{\centering
+\includegraphics[width=1.0\linewidth]{fig2_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1}
+\captionof{figure}{Mean rates of adaptation, variance and covariance.} \label{fig.1}}
+
+{\centering
+\includegraphics[width=1.0\linewidth]{fig3_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1}
+\captionof{figure}{Variance and covariance fluctuations over time.}
+\label{fig.2}}
+
 \section*{Discussion}
@@
 
+\section*{scrap text}
+ for example,  the rate of adaptation, and developed his now famous geometric model to infer that complexity would lead to slower adaptation in living organisms.
+Other majors works addressing evolutionary biology would follow, such as (cite), but quantitative genetics to provide some of the most The model provides a simple geometric argument demonstrating how adaptive evolution would be constrained in higher dimensional trait spaces.
+Kimura (1983) gave further extensions to this model by incorporating probability of fixation, and Orr (1998, 2000) provided additional insights on the the optimal dimensions that ensured maximum adaptation.
+Some of the earliest works that attempted to address multi-trait adaptation were developed by Fisher (1930)However, early evolutionary biology could provide little development in this area, in part due to standing questions concerning the relationships between traits and their genetic foundation, such as how to reconcile Mendelian genetics and with the inheritance of continuous characters.
+Thus questions concerning whole organisms, which are of main interest to plant and animal breeders, and the level of whole organisms.\par
+
 \end{document}

### 2017-07-20 manuscriptoutline.tex
more updates

\documentclass[12pt,one column]{article}
\usepackage[margin=.5in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}

\begin{document}
\section*{Introduction}
\begin{itemize}
\item What are we talking about? 
	\begin{itemize}
	\item Trait correlations (TC) and their role in adaptive evolution \cite{Scarcelli23102007,Lovell2013,Wagner2011}
	\end{itemize}
\item What do we already know about trait correlations?
	\begin{itemize}
	\item TC in the minds of evolutionary biologist.
	\item Influential works on TC and their insights. 
	\item Advances in genetics and the Landau's breeder equation.
	\item The causes of TC and how they shape them over time over time.
	\item The Edinburgh view of trait correlations and adaptation: F.A.E. Crew, C.H. Waddington, D.S. Falconer, F.W. Robertson  
	\begin{itemize}
		\item Edinburgh school on animal genetics and evolution allowed for higher heterozygosity and number of alleles
	\end{itemize}
	\item The Birmingham view of trait correlations and adaptation: Mathers, Jinks, Kearsey, Pooni
	\begin{itemize}
		\item Birmingham school dealt with almost pure inbred plant lines, with only two alleles at each segregating locus
	\end{itemize}
	\end{itemize}
\item Why are we talking about this subject?
	\begin{itemize}
	\item Pleiotropy based models for G-matrix evolution already exist.
	\item LD's role in shaping TC has not been explored despite being necessary.
	\end{itemize}
\item What are we going to say about the 
	\begin{itemize}
	\item Trait correlations and their role in adaptive evolution
	\end{itemize}
\end{itemize}

\section*{Results}
\begin{itemize}
	\item one
\end{itemize}

\section*{Discussion}
\begin{itemize}
	\item
\end{itemize}

\section*{Methods}
\begin{itemize}
	\item
\end{itemize}

\bibliographystyle{plain}
\bibliography{biblio}

\end{document}

### 2017-07-20 manuscript.tex
more updates for paper

@@
 The response to selection with multiple traits involves both direct and indirect effects that together determine their evolution \cite{Scarcelli23102007,Lovell2013,Wagner2011}.
-Indirect effects arise from genetic mechanisms that underlie trait expression which may be associated or shared in a manner that enforces specific patterns of co-expression and give rise to trait correlations.\par
+Indirect effects arise from genetic mechanisms that underlie trait expression which may be associated or shared in a manner that enforces specific patterns of co-expression.
+These effects can significantly influence the evolution of populations on a micro- and macro-evolutionary scale.
+This was partly evident to Darwin (Darwin p.
+144-150) through his observations of animal and plant breeding.
+Evolutionary biologists since then have explore related questions experimentally and theoretically, but the most comprehensive framework for understanding  among geneticists and breeders .\par
 
-Trait correlations are statistical measures of associated co-expression in traits primarily used in quantitative genetics.  Within the
+In quantitative genetics, indirect effects stemming from the evolution in genetically associated traits is measured by genetic correlations, or alternatively, genetic covariances, and they form part of the statistical description of phenotypic responses to selection
 
-Evolution with multiple traits can produce much more complex responses in populations as they adapt; a fact that was clear to Darwin (Darwin p.
-144-150).
-Today, evolutionary biologists are in agreement that these indirect influences stemming from simultaneous trait evolution have important consequences that affect populations on a micro- and macro-evolutionary scale.
-Plant and animal breeders have been keen on this fact for much longer, and not surprisingly, our most comprehensive theory of adaptation in complex organisms with multiple traits was developed from the quantitative tools used in by them.
-The extension of quantitative genetics to multivariate phenotypes, formulated by Lande (Lande 1979; 1980), provides the first characterization of selection on multiple traits, formally expressed in the breeder’s equation ($\Delta z ̅=G \beta$).
+The extension of quantitative genetics to multivariate phenotypes, formulated by Lande (Lande 1979; 1980), provides the first characterization of selection on multiple traits, formally expressed in the breeder’s equation ($\Delta z = \beta$).
 The phenotypes of quantitative traits change ($\Delta z$) in response to a selection gradient ($\beta$) acting on them, given a specific set of trait variances and covariance between them ($G$).

### 2017-07-21 manuscript.tex
more updates

@@
 Indirect effects arise from genetic mechanisms that underlie trait expression which may be associated or shared in a manner that enforces specific patterns of co-expression.
-These effects can significantly influence the evolution of populations on a micro- and macro-evolutionary scale.
-This was partly evident to Darwin (Darwin p.
-144-150) through his observations of animal and plant breeding.
+These effects can significantly influence the evolution of populations on a micro- and macro-evolutionary scale; a fact that was evident to Darwin through his influences from animal and plant breeding (Darwin p.
+144-150).
 Evolutionary biologists since then have explore related questions experimentally and theoretically, but the most comprehensive framework for understanding  among geneticists and breeders .\par

### 2017-07-21 manuscriptoutline.tex
more updates

@@
 \usepackage{amssymb}
+\usepackage{graphicx}
+\usepackage{capt-of}
+
+\graphicspath{{figures/}}
 
@@
 \begin{itemize}
-\item What are we talking about? 
+\item Natural selection frequently acts on multiple traits simultaneously. 
 	\begin{itemize}
-	\item Trait correlations (TC) and their role in adaptive evolution \cite{Scarcelli23102007,Lovell2013,Wagner2011}
+	\item A single trait’s response to selection leads to a shift in the population mean towards the fittest phenotype.
+	\item The response to selection with multiple traits involves both direct and indirect effects that together determine their evolution \cite{Scarcelli23102007,Lovell2013,Wagner2011}.  
+	\item Indirect effects arise from genetic mechanisms that underlie trait expression which may be associated or shared in a manner that enforces specific patterns of co-expression.
 	\end{itemize}
-\item What do we already know about trait correlations?
+\item Associations between traits influence macro- and micro-evolutionary outcomes by constraining or facilitating adaptation.
 	\begin{itemize}	
-	\item TC in the minds of evolutionary biologist.
-	\item Influential works on TC and their insights. 
-	\item Advances in genetics and the Landau's breeder equation.
-	\item The causes of TC and how they shape them over time over time.
-	\item The Edinburgh view of trait correlations and adaptation: F.A.E. Crew, C.H. Waddington, D.S. Falconer, F.W. Robertson  
+	\item Darwin himself suggested as much in his work (Darwin p.144-150), while evolutionary biologist since then have explored a range of related questions.
+	\item Today we know, among other things, that they can adversely affect the rate of adaptation, and determine fundamental features of the evolutionary trajectories that populations follow\cite{Felsenstein1979, Arnold2001, Arnold2008}.
+	\item On short time scales, populations will exhibit much more complex responses to natural selection, some of which may produce seemingly pardoxical features in individual trait evolution and its standing variation \cite{walsh2009abundant}
+	\item Only recently have we begun to understand the nuances of multi-trait adaptation that perhaps where to some extent apparent animal and plant breeders in the past. 
+	\item In the case of quantitative traits, we have a much more sophisticated mathematical framework that grew out the work by quantitative geneticists motivated by questions most relevant to breeders.
+	\end{itemize}
+\item Genetic correlations, or alternatively genetic covariances, provide one way of measuring the degree to which quantitative traits are genetically associated. 
 	\begin{itemize}
-		\item Edinburgh school on animal genetics and evolution allowed for higher heterozygosity and number of alleles
+	\item They form integral part of machinery that describes the phenotypic evolution of multiple traits under selection, summarized in Lande's equation ($\Delta \bar{z} = \textbf{G} \hspace{.05in}\beta$), also known as the breeder's equation.
+	\item Genetic covariances form part of the components of the \textbf{G} matrix, whose geometric features determine what the change in population mean of traits ($\Delta \bar{z}$) will be given a selection gradient ($\beta$) on those traits \cite{Arnold2008}.
+	\item Lande’s formulation depends on assumptions that prescribe important features of the underlying genetics controlling trait expression, but say nothing about the causes of genetic correlations.
+	\item Trait correlations can arise for a number of reasons, and they often have a genetic basis \cite{Saltz2017}, which are either due to pleiotropy, linkage disequilibria, and trans acting elements.
+	\item Most quantitative geneticist adhere to two schools of thought on this matter; known as the Birmingham and Edinburgh schools.
 	\end{itemize}
-	\item The Birmingham view of trait correlations and adaptation: Mathers, Jinks, Kearsey, Pooni
+\item The Edinburgh school have traditionally been traditionally motivated by questions most closely related to animal breeding, and much of there framework has developed around its difficulties. 
 	\begin{itemize} 
-		\item Birmingham school dealt with almost pure inbred plant lines, with only two alleles at each segregating locus
+	\item In such populations, the genetic foundation of phenotypes is hidden, and random breeding are inevitably core elements that must be accounted for in their genetic analysis.  
+	\item Consequently, arise from associations between alleles that form through either pleiotropy, linkage disequilibrium, or trans acting effects.
 	\end{itemize}
+\item There are two schools of thought among quantitative geneticists that differ in the principle cause of associations between alleles.  
+	\begin{itemize}
+	\item One of them holds pleiotropy as the largely responsible for  pleiotropic effects each of them. 
+	\item The views held by each are largely the result of distinct questions and applications Geneticist in the Edinburgh school hold that genetic correlations are the product of pleiotropic effects, while those in the Birmingham school believe that linkage disequilibrium is the primary cause.
+	\item The majority of quantitative genetics follow Geneticist pleiotropy.
 	\end{itemize}
-\item Why are we talking about this subject?
+\item We focus on asexual micro-organisms that lack recombination and examine how the evolution of their traits is affected by linkage.
 	\begin{itemize}
-	\item Pleiotropy based models for G-matrix evolution already exist.
-	\item LD's role in shaping TC has not been explored despite being necessary.
+	\item A variety of organisms fall under this category, which includes many single celled Eukaryotes and a variety of bacteria (cite work).
+	\item In this setting, linkage disequilibria is entirely determined by the combined action of mutations, selection and drift. 
+	\item Standard results in population genetics demonstrate that the amount of linkage disequilibria between various alleles will depend on the beneficial mutation rate ($U_b$), the selection coefficient ($s$), and strength of drift measured by population’s size ($N$).
+	\item Logically, one expects variances and covariances arising from linkage to also depend on the population’s parameters, but a general expression that details this explicit relationship has not been uncovered.
+	\item The majority of results concerning linkage comes from experiments with inbred lines in plants, stemming from the Birmingham school, and that have limited application to populations of interest to us Jinks and Mather (1981).
 	\end{itemize}
-\item What are we going to say about the 
+\item In the case of adaptive traits, with fitness conferred taken as their quantitative measure, the relationship between variance and the population parameters is known explicitly. 
 	\begin{itemize}
-	\item Trait correlations and their role in adaptive evolution
+	\item The work of Desai and Fisher (2007) provide the variance in fitness explicitly shown to work of gives the variance of adaptive traits in terms of the population’s parameters $N$, $U_b$, and $s$.
+	\item The variance of traits and the covariance between them are estimated, and assume that the underlying traits satisfy basic assumptions, such as being polygenic. 
+	\item The variance in fitness was not something that could be determine from populations parameters, until only recently. 
+	\item Desai and Fisher’s work (2007) provides the rate of adaptation of the population in terms of its parameters, \[v =s^2  \frac{2 \ln(Ns)-\ln(s/U_b)}{\ln^2(s/U_b)}\] and since the rate of adaptation is the variance of fitness ($v=\sigma^2$).
+	\item If we consider a second adaptive trait evolving concurrently, with identical mutation rate and selection coefficient, then the rate of adaptation in each trait should be half of the total rate of adaptation $v(2U_b)$ since evolution in each contributes equally the overall adaptation of the population.
+	\item A simple application of Desai and Fisher’s formula quickly shows that the rate of adaptation for a single trait is reduced when a second trait is considered since \[ 1/2  v(2U)<v(U). \]
+	\item Naturally, we should expect such a result since we have doubled the mutation rate, and consequently, increased the amount of clonal interference that occurs within the population. 
+	\item However, the mere fact that we are also considering a second trait indicates that there to this picture than this.
+	\item The rate of adaptation in one trait is determined by its variance and covariance with other traits, and this relation is expressed in the breeder’s equation.
+In our situation, both quantitative traits are adaptive and we measure them by the fitness they endow the organism.
+Consequently, the rate of fitness increase of the focal trait $v_1$ must be related \[v_1=\sigma_1^2+\sigma_{1,2}.\]
 	\end{itemize}
@@
 \begin{itemize}
-	\item one
+	\item 
 \end{itemize}
+
+{\centering
+\includegraphics[width=1.0\linewidth]{fig2_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1}
+\captionof{figure}{Average rate of adaptation, variance and covariance. Parameters $N=10^9$, $s_1=s_2=0.01$, $U_1=U_2=10^{-5}$}
+\label{fig.1}}
+
+{\centering
+\includegraphics[width=1.0\linewidth]{fig1_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1}
+\captionof{figure}{Variance and covariance fluctuations over time. Parameters $N=10^9$, $s_1=s_2=0.01$, $U_1=U_2=10^{-5}$}
+\label{fig.2}}
 
@@
 \begin{itemize}
-	\item
+	\item two
 \end{itemize}
@@
 \begin{itemize}
-	\item
+	\item three
 \end{itemize}
@@
 
+\section*{Appedices}
+\begin{itemize}
+	\item QED
+\end{itemize}
+
 \end{document}

### 2017-07-22 manuscriptoutline.tex
more updates

@@
 	\end{itemize}
+
+\end{itemize}
+
+\section*{Methods}
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.5\linewidth]{fig2_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1} 
+	\caption{Average rate of adaptation, variance and covariance. Parameters $N=10^9$, $s_1=s_2=0.01$, $U_1=U_2=10^{-5}$} 
+	\label{fig.1}
+\end{figure}
+
+\begin{table}[h]
+	\centering
+ 	\begin{tabular}{ | c | c | c | c |}
+    \hline
+    Organism 												& Population Size 	& Selection Coefficient & Beneficial Mutation Rate \\ \hline
+    \textit{E}. coli \cite{Perfeito2007}					& $10^9 - 10^12$ 	& $0.5 - 2 \%$  & $10^{-5} - 10^{-4}$ \\ \hline
+    \textit{S}. cerevisiae \cite{desai2007speed,Levy2015}	& $10^9 - 10^12$ 	& $0.5 - 2 \%$  & $10^{-5} - 10^{-4}$ \\
+    \hline
+	\end{tabular}
+	\caption{Population parameters measured in evolution experiments for key microorganisms.} 
+	\label{Table 1}
+\end{table}
+
+\begin{itemize}
+	\item We consider a population of asexuals whose size ($N$) is constant evolving in a static environment.
+	\begin{itemize}
+		\item Individuals have two traits that contribute to its relative fitness, and which evolve through beneficial mutations; deleterious mutations are not considered.
+		\item The beneficial mutation rate for each trait is denoted $U_k$ (trait $k=1,2$), and each increases trait $k$’s contribution to relative fitness by $s_k$.  
+		\item We can express an individual fitness as \[ r_{ij}=r+i s_1+j s_2.
+\] where $i$ and $j$ represent the number of beneficial mutations in trait one and two, while $r$ is the growth rate a common ancestor to the population.
+		\item An individual’s relative fitness is given by \[ (r_{ij}-r ̅)=(i-\bar{i} ) s_1+(j-\bar{j} ) s_2 \] respectively, and $\bar{i}$ and $\bar{j}$ are the population mean.  
+		\item The two traits of individuals are treated as independently evolving loci in which beneficial mutations occur at a fixed rate $U_k$ ($k=1,2$).  
+		\item Each beneficial mutation is either lost to drift, or the subpopulation carrying it will grow sufficiently large to ensure that the beneficial mutation fixes in the population through selection.
+		\item The probability of the latter event, denoted $\pi_{fix}$, is proportional to the selective advantage of the mutation.
+In these cases a beneficial mutation is said to have established in the population.
+		\item The selective advantage of an individual is measured by its relative fitness, which in our model, is its expected growth rate minus the growth rate of an average individual.
+Each new beneficial mutation in trait $k$ increases this fitness by $s_k$; we do not consider deleterious and neutral mutations in our model.
+		\item If an individual has i mutations in the first trait and $j$ in the second, while the average number carried in the two traits is $\bar{i}$ and $\bar{j}$, its relative fitness equals the difference $(r_{ij}-\bar{r})=(i-\bar{i}) s_1+(j-\bar{j}) s_2$.
+		\item We group members of the population that carry the same number of beneficial mutations in each trait, $i$ mutations in the first and $j$ in the second, into fitness classes that will often be denoted by subscripts $ij$.
+		\item The set of fitness class abundances $\{n_{ij} \}$ induce a distribution over the trait space, represented by $N^2$, containing all fitness classes. 
+		\item This two-dimensional distribution, which we will often denote by the set of frequencies $\{p_{ij}\}$, travels in the direction of increasing fitness as the population adapts (Fig 1) in an analogous fashion to the motion of Desai and Fisher’s (2007) one dimensional traveling wave.
+		\item We also examine the properties of the marginal distributions $\{p_{i\cdot}\}$ and $\{p_{\cdot j}\}$ for each trait, whose behaviors over time are significantly affected by genetic correlations that arise from the confluences of mutations, selection and drift.
+\[ p_{i \cdot }= \sum_j p_{ij} \hspace{0.5in} p_{\cdot j}= \sum_i p_{ij} \]
+		\item In order to study how genetic correlations arise and influence the evolution of traits in adapting populations, we focus on regimes of adaptation characterized by significant levels of genetic variation in each trait.
+		\item The results for one-dimensional traveling waves provides the relationship between variance in fitness from one trait evolving and population parameters $N$, $U_k$, and $s_k$. 
+		\item When mutations appear infrequently relative to the average time required for beneficial mutation to fix, we obtain the simplest behavior possible in which beneficial mutations establish in succession and spread to the population long before the next one appears ($N U_k \ll 1/\ln(N s_k)$).
+		\item These populations rarely have any variance in fitness stemming from trait diversity, and are not the focus of our investigations. 
+		\item We focus instead on the case where many beneficial mutations establish prior to any one fixing ($N U_k \geq 1/\ln(N s_k)$), where the effects of clonal interference cause variation in fitness which persists over time in the type of populations we consider.
+		\item For large populations with strong selection in each trait ($N^{-1}\ll s_k \ll 1$), changes in the abundances of fitness class are either dominated by stochastic fluctuations due to drift ($n_{ij}<\pi_{fix}^{-1}$), or growth deterministically otherwise due to selection.
+		\item We will refer to fitness classes in the first group as the front of the wave, and their growth can be modeled as branching processes that are fed new mutants from adjacent less fit classes. 
+		\item The fitness classes in the stochastic front will consist of those which are fitter and adjacent to ones that have established whenever $s_k/U_k \gg 1$.
+During the stochastic phase of growth, fitness classes are much too small to influence dynamics that are driven by selection, so that they may be ignored until the moment they do establish.
+		\item As for the one-dimensional traveling wave, the key variable of interest is the time-between establishments $\tau$ of new beneficial mutations. 
+		\item This random variable summarizes all of the stochastic effects that accumulate over time during the growth of a fitness class, and determines the rate at which fitness classes in the stochastic front transition into the deterministic regime where selection is the prominent force.
+		\item We will be interested in the rate of these transitions, because they will determine two very important aspects of the two-dimensional distribution’s evolution that influence trait evolution. 
+		\item The abundances of fitness classes that have established, whose behavior is dominated by the action of selection, deterministically grow and decline in size. 
+		\item We refer to the fitness classes in this group as the bulk of the population.
+The quantities i ̅ and j ̅ are marginal means computed from the marginal distributions (Eq.
+1).
+\[ \bar{i}=∑_i i p_{i \cdot} \hspace{0.5in} \bar{j} = ∑_j p_{⋅j} j \]
+		\item Relative fitness, written as the difference $(r_{ij}-\bar{r})=(i-\bar{i}) s_1+(j-\bar{j}) s_2$, determines the growth and decline in abundances $n_{ij}$ that follow the differential equation \[ \dot{n}_{ij}=n_{ij} (r_{ij}-\bar{r})-n_{ij} \left(\frac{N(t)-N}{N}\right) r \] where $N(t)=∑_{ij} n_{ij}(t)$ above, and the right-most term is the density-regulation component of growth.
+		\item Stochastic fluctuations in the total population size are negligible ($|N(t)-N| \ll N$), and so we may also consider the frequencies of fitness classes in the bulk to obey Eq.
+3 below.
+\[ \dot{p}_{ij}=p_{ij} (r_{ij}-\bar{r}) \]
+		\item From Eq.
+3, the rate of adaptation can be expressed as the sum of the individual rates of adaptation in each trait, indicated by subscripts, as $\dot{\bar{r}}=\dot{\bar{r}}_1 +\dot{\bar{r}}_2$.
+		\item Each of the two individual rates are related to the variance in fitness from the respective trait and its covariance in fitness with other (Eq.
+4).
+\[ \dot{\bar{r}}_1=\sigma_1^2+\sigma_{1,2}\\ \dot{\bar{r}}_2=\sigma_2^2+\sigma_{1,2} \]
+		\item Mutation-selection balance induces steady state values for the speed of evolution in the two, from which it possible to determine the two variances, the single covariance, and furthermore, the resulting G-matrix of the two traits (Eq.
+5).
+\[ G= \left( \begin{array}{cc} \sigma_1^2& \sigma_{1,2}\\ \sigma_{1,2} & \sigma_2^2 \end{array} \right) \]
+		\item Linkage disequilibria continually arises throughout the adaptation process and limits the response of traits to selection.
+In the case of two alleles, linkage disequilibria is a measure of the statistical association between them.
+		\item For simplicity, we can regard all ith beneficial mutations occurring in trait one as the same allele, and likewise with the jth mutation in trait two.
+This allows us to treat the difference $D_{ij}=p_{ij}-p_{i\cdot} p_{\cdot j}$ as a measure of linkage disequilibrium between the two alleles corresponding to the $i^{th}$ mutation in trait one, and the $j^{th}$ in trait two.
+		\item In turn, the covariance in fitness can be written out as Eq.
+6.
+\[ \sigma_{1,2}=\sum_{ij}D_{ij}(i-\bar{i})s_1 (j-\bar{j}) s_2 \] and it becomes evident that linkage disequilibrium directly affects the evolution of the two traits.
+		\item We can also note from this expression that making a distinction between alleles that yields the same fitness advantage does not change the covariance.
 		\item In the case of adaptive traits, with fitness conferred taken as their quantitative measure, the relationship between variance and the population parameters is known explicitly. 
-	\begin{itemize}
 		\item The work of Desai and Fisher (2007) provide the variance in fitness explicitly shown to work of gives the variance of adaptive traits in terms of the population’s parameters $N$, $U_b$, and $s$.
@@
 	\end{itemize}
+	\item We examined the resulting dynamics of trait evolution using numerical simulations of our model carried out in the software package Mathematica.
+	\begin{itemize}
+		\item We focused on modeling the symmetric case in which the two traits had identical mutation rate and selection coefficients.
+		\item In our numerical implementation our choice of parameters ensured that the treatment of abundance growth in classes could be separated into deterministic or stochastic groups, identified above as the population bulk and the stochastic front.
+		\item Consequently, simulations consisted of two computational modules running iteratively each  abundances in these two groups
+	\end{itemize}
+	\item Behavior of the bulk
+	\begin{itemize}
+		\item Our simulations took advantage of the separation between deterministic and stochastic phases of abundant growth, which is possible whenever the parameters satisfy $1/N \ll s \ll 1$ 
+		\item Simulations were initialized as a monoclonal population, and allowed to run through a standard burn in phase equivalent to  to overcome transient effect 
+		\item 
+	\end{itemize}
+	\item Evolution of Bulk follows deterministic ODE's (eqtn.?), plus trimming
+	\item Stochastic front modeled as minimum of establishing mutant class from stochastic front.
+	\item 
 \end{itemize}
@@
 {\centering
-\includegraphics[width=1.0\linewidth]{fig2_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1}
+\includegraphics[width=0.5\linewidth]{fig2_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1}
 \captionof{figure}{Average rate of adaptation, variance and covariance. Parameters $N=10^9$, $s_1=s_2=0.01$, $U_1=U_2=10^{-5}$}
@@
 {\centering
-\includegraphics[width=1.0\linewidth]{fig1_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1}
+\includegraphics[width=0.5\linewidth]{fig1_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1}
 \captionof{figure}{Variance and covariance fluctuations over time. Parameters $N=10^9$, $s_1=s_2=0.01$, $U_1=U_2=10^{-5}$}
@@
 	\item two
-\end{itemize}
-
-\section*{Methods}
-\begin{itemize}
-	\item three
 \end{itemize}

### 2017-07-22 plotdata.py

@@ -1,0 +1,82 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Mon Jul 17 13:27:18 2017
+
+@author: kgomez81
+"""
+
+# import packages for script
+import pickle
+import scipy as sp
+import numpy as np
+import matplotlib.pyplot as plt
+
+# set parameters of simulation and create required variables
+N=1e9; s1=1e-2; s2=1e-2; U1=1e-5; U2=1e-5;
+#vU_thry = s1*s1*(2*np.log(N*s1)-np.log(s1/(1*U1)))/((np.log(s1/(1*U1)))**2)
+#v2U_thry = 0.5*s1*s1*(2*np.log(N*s1)-np.log(s1/(2*U1)))/((np.log(s1/(2*U1)))**2)
+#tau_est = 0.5*s1/v2U_thry
+data_name = '_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp2'
+[times,genotypes,abundances] = [[],[],[]]
+
+def get_sample_window(times,start_time,end_time):
+    [num_pts,start_indx,end_indx] = [len(times),0,0]    
+    for i in range(len(times)):
+        if times[start_indx] <= start_time:
+            start_indx = start_indx + 1
+        if times[end_indx] <= end_time:
+            end_indx = end_indx + 1
+    return [start_indx,end_indx]
+
+#------------------------------------------------------------------------------
+#------------------------------------------------------------------------------
+#------------------------------------------------------------------------------
+#------------------------------------------------------------------------------
+# get simulation data and store genotypes as lists since they vary in dimensions over time
+data_file=open('./Documents/kgrel2d/data/pythondata/times'+data_name+'.dat')
+times = data_file.read().splitlines()
+times = np.array(map(float,times))
+data_file.close()
+data_file=open('./Documents/kgrel2d/data/pythondata/genotypes'+data_name+'.dat')
+genotypes = data_file.read().splitlines()
+data_file.close()
+data_file=open('./Documents/kgrel2d/data/pythondata/abundances'+data_name+'.dat')
+abundances = data_file.read().splitlines()
+data_file.close()
+del data_file
+num_pts = len(times)
+
+# clean up mathematica data format
+for i in range(num_pts):
+    genotypes[i]='genotypes[i]=np.array(['+genotypes[i].replace('\t',',')+'])'
[…]

### 2017-07-26 manuscriptoutline.tex
updates

--- 2017-07-20 manuscriptoutline.tex
@@
-\documentclass[12pt,one column]{article}
+\documentclass[11pt,one column]{article}
 \usepackage[margin=.5in]{geometry}
@@
 \begin{itemize}
-\item What are we talking about? 
+\item Natural selection frequently acts on multiple traits simultaneously. 
 	\begin{itemize}
-	\item Trait correlations (TC) and their role in adaptive evolution \cite{Scarcelli23102007,Lovell2013,Wagner2011}
+	\item A single trait’s response to selection leads to a shift in the population mean towards the fittest phenotype.
+	\item The response to selection with multiple traits involves both direct and indirect effects that together determine their evolution \cite{Scarcelli23102007,Lovell2013,Wagner2011}.  
+	\item Indirect effects originate from genetic mechanisms that underlie trait expression, which may be associated or shared in a manner that enforces specific patterns of co-expression.
+	\item Darwin, while this fact, recognized that such associations likely played an important role in directing the evolution of organisms under natural selection (Darwin p. 130-133).
+	\item Presently, we know that these genetic associations can fundamentally alter how populations respond to selection and direct their adaptation towards certain evolutionary outcomes, such as speciation or extinction \cite{Felsenstein1979, Arnold2001, Arnold2008}.
+	\item This makes them important to many micro- and macro-evolutionary questions central to evolutionary biology.
 	\end{itemize}
-\item What do we already know about trait correlations?
+\item On short time scales, selection on phenotypes composed of multiple quantitative traits is best understood using the framework of quantitative genetics. 
 	\begin{itemize}	
-	\item TC in the minds of evolutionary biologist.
-	\item Influential works on TC and their insights. 
-	\item Advances in genetics and the Landau's breeder equation.
-	\item The causes of TC and how they shape them over time over time.
-	\item The Edinburgh view of trait correlations and adaptation: F.A.E. Crew, C.H. Waddington, D.S. Falconer, F.W. Robertson  
+	\item The strength of a genetic associations between traits are estimated and summarized by genetic correlations, or genetic covariances.
+	\item Genetic covariances form part of the components of the \textbf{G} matrix.
+	\item The response of phenotypes to selection is mathematically described by Lande's equation ($\Delta \bar{z} = \textbf{G} \hspace{.05in}\beta$).
+	\item Here, $\Delta \bar{z}$ is the vector whose components are the change in population means of each trait, and $\beta$ is the selection gradient on the set of traits. 
+	\item Lande’s formulation provides a means of estimating responses to selection correlated traits, but it does not indicate gives rise to them.
+	\item Trait correlations can generally come about for a number of reasons, but they often have a genetic basis \cite{Saltz2017} arising from either pleiotropy, linkage disequilibria, or trans acting elements, but most agree that the first two are much more significant factors.
+	\end{itemize}
+\item Historically, there are two schools of thought on which of the first two is responsible for most of the genetic correlations.
 	\begin{itemize} 
-		\item Edinburgh school on animal genetics and evolution allowed for higher heterozygosity and number of alleles
+	\item The first of these was known as the Edinburgh view which held that pleiotropy was primary cause of most genetic correlations.
+	\item This belief was largely motivated by their interests in the relationship between parent-offspring phenotypes in randomly mating populations.
+	\item A large portion of their work revolved around animal genetics and its applications in evolution, and linkage disequilibrium was expected to rapidly decay by recombination (\cite{fox2006evolutionary}, chp 20).
+	\item Geneticists adhering to the view of the Birmingham school held that linkage disequilibria could not be so easily dismissed.  
+	\item Their experiments often involved inbred lines with plants where genetic correlations due to linkage disequilibria could persist despite recombination.
 	\end{itemize}
-	\item The Birmingham view of trait correlations and adaptation: Mathers, Jinks, Kearsey, Pooni
-	\begin{itemize}
-		\item Birmingham school dealt with almost pure inbred plant lines, with only two alleles at each segregating locus
-	\end{itemize}
-	\end{itemize}
-\item Why are we talking about this subject?
-	\begin{itemize}
-	\item Pleiotropy based models for G-matrix evolution already exist.
-	\item LD's role in shaping TC has not been explored despite being necessary.
-	\end{itemize}
-\item What are we going to say about the 
-	\begin{itemize}
-	\item Trait correlations and their role in adaptive evolution
-	\end{itemize}
-\end{itemize}
-
-\section*{Results}
-\begin{itemize}
-	\item one
-\end{itemize}
-
-\section*{Discussion}
+\item Today most geneticists assume that pleiotropy is responsible for the genetic correlations found in organisms.
 	\begin{itemize}
 	\item 
+	\item Their is an ample amount of evidence supporting widespread pleiotropy, as well as a number of experiments that have uncovered pleiotropic effects causing genetic correlations \cite{Saltz2017}.
+	\item However, as geneticists adopting the Birmingham view have pointed out, the effects of linkage disequilibria in many settings are significant and can endure. 
+	\item Many asexauls micro-organisms, which include some Eukaryotes and all Prokaryotes, experience very little recombination of their DNA or lack it entirely.
+	\item In these cases, genetic correlations from linkage disequilibria are just as substantial to evolution as those arising from pleiotropy.
+	\item The fact the majority of life on the planet falls into this category underscores the Birmingham view to some degree.
+	\end{itemize}
+\item The evolution of the 
+	\begin{itemize}
+	\item We focus on asexual micro-organisms that lack recombination and examine how the evolution of their traits is affected by linkage.	
+	\item In the case of adaptive traits, with fitness conferred taken as their quantitative measure, the relationship between variance and the population parameters is known explicitly. 
+	\item The work of Desai and Fisher (2007) provide the variance in fitness explicitly shown to work of gives the variance of adaptive traits in terms of the population’s parameters $N$, $U_b$, and $s$.
+	\item The variance of traits and the covariance between them are estimated, and assume that the underlying traits satisfy basic assumptions, such as being polygenic. 
+	\item The variance in fitness was not something that could be determine from populations parameters, until only recently. 
+	\item Desai and Fisher’s work (2007) provides the rate of adaptation of the population in terms of its parameters, \[v =s^2  \frac{2 \ln(Ns)-\ln(s/U_b)}{\ln^2(s/U_b)}\] and since the rate of adaptation is the variance of fitness ($v=\sigma^2$).
+	\item If we consider a second adaptive trait evolving concurrently, with identical mutation rate and selection coefficient, then the rate of adaptation in each trait should be half of the total rate of adaptation $v(2U_b)$ since evolution in each contributes equally the overall adaptation of the population.
+	\item A simple application of Desai and Fisher’s formula quickly shows that the rate of adaptation for a single trait is reduced when a second trait is considered since \[ 1/2  v(2U)<v(U). \]
+	\item Naturally, we should expect such a result since we have doubled the mutation rate, and consequently, increased the amount of clonal interference that occurs within the population. 
+	\item However, the mere fact that we are also considering a second trait indicates that there to this picture than this.
+	\item The rate of adaptation in one trait is determined by its variance and covariance with other traits, and this relation is expressed in the breeder’s equation.
+In our situation, both quantitative traits are adaptive and we measure them by the fitness they endow the organism.
+Consequently, the rate of fitness increase of the focal trait $v_1$ must be related \[v_1=\sigma_1^2+\sigma_{12}.\]
+	\end{itemize}
+\item (\textbf{Approach}) To examine the effects on the evolution of one trait by adaptation in another, we develop a model of a traveling wave over a two dimensional trait space.
+	\begin{itemize}
+	\item 
+	\end{itemize}
 \end{itemize}
@@
 \begin{itemize}
+\item (\textbf{Model parameters}) In our model we consider a population of asexuals at carrying capacity $N$ evolving in a static environment.
+	\begin{itemize}
+	\item Individuals have two traits contributing to their relative fitness which are controlled by separately loci that independently evolve.
+	\item Beneficial mutations contribute fitter alleles to the genetic diversity of each and occur at rate $U_k$ (trait $k=1,2$); deleterious mutations are not considered.
+	\item Mutations increase trait \textit{k}'s contribution to relative fitness by $s_k$, and we may express an individual's relative fitness as $r_{ij} = (i-\bar{i} ) s_1+(j-\bar{j} ) s_2$.
+	\item Here $i$ and $j$ represent the number of beneficial mutations that have accumulated in traits one and two, respectively, while $\bar{i}$ and $\bar{j}$ are the mean number of them in the average individual.
+	\item We divide the population into classes defined as groups of individuals with an identical number of beneficial mutations in each trait, and denote their abundances with $n_{ij}$ and frequencies with $p_{ij}$.
+	\end{itemize}	
+\item (\textbf{Pieces of the model}) We focus on the evolutionary regime characterized by large populations with strong selection, for which $1/N \ll s \ll 1$.
+	\begin{itemize}
+	\item The set of abundances $n_{ij}$ can then be categorized into two groups based on whether the stochastic fluctuations negligibly influence changes in abundances or don't. 
+	\item Following the convention of Desai and Fisher (2007), we refer to the former group as the bulk, and we will call the latter set of classes the stochastic front.
+	\item Stochastic fluctuations caused by drift can be ignored once abundances have grown sufficiently large and selection begins to dominate their dynamics, such a class is said to have established.
+	\end{itemize}
+\item (\textbf{Dynamics of the Bulk}) The dynamics of the bulk are driven by selection, and have deterministically over time.
+	\begin{itemize}
+	\item Since the selective advantage of these classes is $r_{ij}$, their abundances in a manner that can be approximated by the ordinary differential equation \[ \dot{n}_{ij}(t) = S_{ij} n_{ij}(t).
+\]
+	\item From Eq. 3, the rate of adaptation can be expressed as the sum of the individual rates of adaptation in each trait, indicated by subscripts, as $\dot{\bar{r}}=\dot{\bar{r}}_1 +\dot{\bar{r}}_2$.
+	\end{itemize}
+\item (\textbf{Stochastic Front})
+	\begin{itemize}
+	\item As for the one-dimensional traveling wave, the key variable of interest is the time-between establishments $\tau$ of new beneficial mutations. 
+	\item This random variable summarizes all of the stochastic effects that accumulate over time during the growth of a fitness class, and determines the rate at which fitness classes in the stochastic front transition into the deterministic regime where selection is the prominent force.
+	\item We will be interested in the rate of these transitions, because they will determine two very important aspects of the two-dimensional distribution’s evolution that influence trait evolution. 
+	\item The probability of the latter event, denoted $\pi_{fix}$, is proportional to the selective advantage of the mutation.
+In these cases a beneficial mutation is said to have established in the population.
+	\item The threshold separating these two characteristically different behaviors in growth/decline occurs at size equal to the inverse of the class' probability of fixation.
+	\end{itemize}
+\item (\textbf{Rate of Adaptation})
+	\begin{itemize}
+	\item In the case of adaptive traits, with fitness conferred taken as their quantitative measure, the relationship between variance and the population parameters is known explicitly. 
+	\item The work of Desai and Fisher (2007) provide the variance in fitness explicitly shown to work of gives the variance of adaptive traits in terms of the population’s parameters $N$, $U_b$, and $s$.
+	\item The variance of traits and the covariance between them are estimated, and assume that the underlying traits satisfy basic assumptions, such as being polygenic. 
+	\item The variance in fitness was not something that could be determine from populations parameters, until only recently. 
+	\item Desai and Fisher’s work (2007) provides the rate of adaptation of the population in terms of its parameters, \[v =s^2  \frac{2 \ln(Ns)-\ln(s/U_b)}{\ln^2(s/U_b)}\] and since the rate of adaptation is the variance of fitness ($v=\sigma^2$).
+	\item If we consider a second adaptive trait evolving concurrently, with identical mutation rate and selection coefficient, then the rate of adaptation in each trait should be half of the total rate of adaptation $v(2U_b)$ since evolution in each contributes equally the overall adaptation of the population.
+	\item A simple application of Desai and Fisher’s formula quickly shows that the rate of adaptation for a single trait is reduced when a second trait is considered since \[ 1/2  v(2U)<v(U). \]
+	\item Naturally, we should expect such a result since we have doubled the mutation rate, and consequently, increased the amount of clonal interference that occurs within the population. 
+	\end{itemize}
+\item (\textbf{Covariance and Trait Evolution})
+	\begin{itemize}
+	\item The results for one-dimensional traveling waves provides the relationship between variance in fitness from one trait evolving and population parameters $N$, $U_k$, and $s_k$. 
+	\item Mutation-selection balance induces steady state values for the speed of evolution in the two, from which it possible to determine the two variances, the single covariance, and furthermore, the resulting G-matrix of the two traits (Eq.
+5).
+\[ G= \left( \begin{array}{cc} \sigma_1^2& \sigma_{1,2}\\ \sigma_{12} & \sigma_2^2 \end{array} \right) \]
+	\item However, the mere fact that we are also considering a second trait indicates that there to this picture than this.
+	\item The rate of adaptation in one trait is determined by its variance and covariance with other traits, and this relation is expressed in the breeder’s equation.
+In our situation, both quantitative traits are adaptive and we measure them by the fitness they endow the organism.
+Consequently, the rate of fitness increase of the focal trait $v_1$ must be related \[v_1=\sigma_1^2+\sigma_{12}.\]
+	\item Each of the two individual rates are related to the variance in fitness from the respective trait and its covariance in fitness with other (Eq.
+4).
+\[ \dot{\bar{r}}_1=\sigma_1^2+\sigma_{12}\] \[ \dot{\bar{r}}_2=\sigma_2^2+\sigma_{12} \]
+	\item We examine how genetic correlations arise and influence the evolution of traits in adapting populations, we focus on regimes of adaptation characterized by significant levels of genetic variation in each trait.
+	\item Linkage disequilibria continually arises throughout the adaptation process and limits the response of traits to selection.
+In the case of two alleles, linkage disequilibria is a measure of the statistical association between them.
+	\item In turn, the covariance in fitness can be written out as Eq.
+6.
+\[ \sigma_{12}=\sum_{ij}D_{ij}(i-\bar{i})s_1 (j-\bar{j}) s_2 \] and it becomes evident that linkage disequilibrium directly affects the evolution of the two traits.
+	\item We can also note from this expression that making a distinction between alleles that yields the same fitness advantage does not change the covariance.
+
+	\end{itemize}
+\item (\textbf{Simulation description}) We examined the resulting dynamics of trait evolution using numerical simulations of our model carried out in Mathematica.
+	\begin{itemize}
+	\item We focused on modeling the symmetric case, in which the two traits had identical mutation rate and selection coefficients.
+	\item These, along with the population size, were chosen to match values reported in evolutionary experiments involving relevant micro-organisms (see Table 1).  
+	\item We also focused parameter sets in which $1/N \ll s \ll 1$ to ensure that changes in abundances for the population bulk could be treated separately from those in the stochastic front.
+	\end{itemize}
+\item Our numerical implementation consisted of two routines which ran iteratively for the duration of the simulation.
+	\begin{itemize}
+	\item Each iteration began with solving for the time-dependent abundances in the bulk governed by equation [??] over a time interval of one-thousand generations.
+	\item We used Mathematica's standard numerical ODE solver to obtain the solutions and used them to determine which class established next and the time at which it did so.
+	\item Abundances in the bulk were then examined at the time of establishment.
+	\item Any classes in the bulk with less than one individual were removed, while abundances were updated to their size at the time of establishment for those that weren't.
+	\item The establishing class was then added to the bulk with initial size randomly determined using equation [??].
+	\item The next iteration then began with solving for the abundances of the new bulk, followed by a repetition of the steps just mentioned.  
+	\end{itemize}	
+\item We simulated the time of the next establishment by taking the first mutant whose lineage manages to fix. 
+	\begin{itemize}
+	\item The times that mutants appear follow a time-inhomogeneous Poisson process with rate functions given by the abundances of the adjacent classes from which they originated. 
+	\item Consequently, if $\tau_l$ denotes the arrival time of the $l$-th mutant generated from the class $n_{ij}(t)$ with an additional mutation in trait $k$, then \[ \int_0^{\tau_l} U_k n_{ij}(t) dt \] has the distribution of $l$ i.i.d exponentially distributed summed random variables.
+	\item We began with identifying the first mutant to appear by solving for the set of $\tau_1$ 's of each class in the stochastic front.  
+	\item The probability of fixation for this mutant was calculated and used to check if the the mutant's lineage would establish.  
+	\item If mutation did establish, then the class was incorporated into the bulk as described above.  
+	\end{itemize}
+\item Simulations were initialized with monoclonal populations at carrying capacity consisting of individuals having no beneficial mutations in either trait and relative fitness set zero.
+	\begin{itemize}
+	\item We allowed the simulation to run for 5,000 generations to allow the population to achieve beneficial mutation-selection balance before collecting data.
+	\item This ensured that the transient effects from the initializing would not distort the statistics collected for the distribution.
+	\item Following the burn period of the simulation, we recorded the existing classes in the bulk and their respective abundances every one 100 generations and intermittent times at which new classes established.
+	\item The data was used to compute the marginal mean fitness and variances of each trait, as well as their covariance.
+	\item From these quantities we also computed the instantaneous rate of adaptation (eqtn. ??), the mean rate of fitness increase in each trait and the overall rate of adaptation.
+	\end{itemize}
+\end{itemize}
+
+% figure to include in methods
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.35\linewidth]{2dfitnessdistributionexample.png} 
+	\caption{Two-dimensional fitness distribution.} 
+	\label{fig.1}
+\end{figure}
+
+\begin{table}[h]
+	\centering
+ 	\begin{tabular}{ | c | c | c | c |}
+    \hline
+    Organism 												& Population Size 	& Selection Coefficient & Beneficial Mutation Rate \\ \hline
+    \textit{E}. coli \cite{Perfeito2007}					& $10^9 - 10^{12}$ 	& $0.5 - 2 \%$  & $10^{-5} - 10^{-4}$ \\ \hline
+    \textit{S}. cerevisiae \cite{desai2007speed,Levy2015}	& $10^9 - 10^{12}$ 	& $0.5 - 2 \%$  & $10^{-5} - 10^{-4}$ \\
+    \hline
+	\end{tabular}
+	\caption{Population parameters measured in evolution experiments for key microorganisms.} 
+	\label{Table 1}
+\end{table}
+
+\section*{Results}
+\begin{itemize}
+\item Mean trait variances and covariances over the period of the simulation
+\item Fluctuations in the components of the \textbf{G}-matrix.
 \item 
+\end{itemize}
+
+%figures to include in results
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.35\linewidth]{fig2_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1}
+\caption{Average rate of adaptation, variance and covariance. Parameters $N=10^9$, $s_1=s_2=0.01$, $U_1=U_2=10^{-5}$}
+\label{fig.1}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.35\linewidth]{fig1_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1}
+\captionof{figure}{Variance and covariance fluctuations over time. Parameters $N=10^9$, $s_1=s_2=0.01$, $U_1=U_2=10^{-5}$}
+\label{fig.2}
+\end{figure}
+
+\section*{Discussion}
+\begin{itemize}
+\item Negative covariance driven by LD
+\item \textbf{G}-matrix instability and its implication for the long term evolution of populations
 \end{itemize}
@@
 
+\section*{Appedices}
+\begin{itemize}
+\item relationship between LD and covariance
+\end{itemize}
+
 \end{document}

### 2017-07-26 manuscriptoutline.tex
updates

@@
 \end{itemize}
+
 \item On short time scales, selection on phenotypes composed of multiple quantitative traits is best understood using the framework of quantitative genetics. 
@@
 \item The response of phenotypes to selection is mathematically described by Lande's equation ($\Delta \bar{z} = \textbf{G} \hspace{.05in}\beta$).
-	\item Here, $\Delta \bar{z}$ is the vector whose components are the change in population means of each trait, and $\beta$ is the selection gradient on the set of traits. 
+\item Here, $\Delta \bar{z}$ is the response of the mean of each trait, and $\beta$ is the selection gradient on the set of traits. 
 \item Lande’s formulation provides a means of estimating responses to selection correlated traits, but it does not indicate gives rise to them.
-	\item Trait correlations can generally come about for a number of reasons, but they often have a genetic basis \cite{Saltz2017} arising from either pleiotropy, linkage disequilibria, or trans acting elements, but most agree that the first two are much more significant factors.
+\item Trait correlations can generally come about for a number of reasons, but they often have a genetic basis \cite{Saltz2017} arising from either pleiotropy, linkage disequilibria, or trans acting elements.
+\item Most evolutionary biologist agree that the first two are much more important.
 \end{itemize}
-\item Historically, there are two schools of thought on which of the first two is responsible for most of the genetic correlations.
+
+\item Historically, there have been two schools of thought on which of the first two is responsible for most of the genetic correlations.
 \begin{itemize} 
-	\item The first of these was known as the Edinburgh view which held that pleiotropy was primary cause of most genetic correlations.
-	\item This belief was largely motivated by their interests in the relationship between parent-offspring phenotypes in randomly mating populations.
-	\item A large portion of their work revolved around animal genetics and its applications in evolution, and linkage disequilibrium was expected to rapidly decay by recombination (\cite{fox2006evolutionary}, chp 20).
-	\item Geneticists adhering to the view of the Birmingham school held that linkage disequilibria could not be so easily dismissed.  
-	\item Their experiments often involved inbred lines with plants where genetic correlations due to linkage disequilibria could persist despite recombination.
+\item The first of these was put forth by the Edinburgh school and held that pleiotropy was primary cause of most genetic correlations.
+\item This belief was largely motivated by their interests in relationships between parent-offspring phenotypes within randomly mating populations and their implication for their evolution.
+\item In these sorts of populations, linkage disequilibrium can rapidly decay by recombination on evolutionary time scales (\cite{fox2006evolutionary}, chp 20), while pleiotropic effects tend to persist.
+\item The alternative view which came out of the Birmingham school held that linkage disequilibria was much more significant.  
+\item Their interest led them to experiments involving inbred lines with plants where the genetic correlations from linkage disequilibria were just as substantial.
+\item Inarguably, the question of which view is perhaps correct largely depends the details of the population.
+\item 
 \end{itemize}
-\item Today most geneticists assume that pleiotropy is responsible for the genetic correlations found in organisms.
+
+\item As populations evolve, the components of the \textbf{G}-matrix change as well, which may constrain or fascilitate adaptation.
 \begin{itemize}
-	\item 
-	\item Their is an ample amount of evidence supporting widespread pleiotropy, as well as a number of experiments that have uncovered pleiotropic effects causing genetic correlations \cite{Saltz2017}.
-	\item However, as geneticists adopting the Birmingham view have pointed out, the effects of linkage disequilibria in many settings are significant and can endure. 
-	\item Many asexauls micro-organisms, which include some Eukaryotes and all Prokaryotes, experience very little recombination of their DNA or lack it entirely.
-	\item In these cases, genetic correlations from linkage disequilibria are just as substantial to evolution as those arising from pleiotropy.
+\item The question of how stable the \textbf{G}-matrix is over time, is directly related to the question of how genetic correlations evolve, and that of course, depends on their causes.
+\item Today most geneticists assume that pleiotropy underlies many of the genetic correlations we observe directing adaptation, and as a result, have developed many sophisticated models to examine how genetic correlations arising from .
+\item We focus asexauls micro-organisms, which include some Eukaryotes and all Prokaryotes, experience very little recombination of their DNA or lack it entirely.	
 \item The fact the majority of life on the planet falls into this category underscores the Birmingham view to some degree.
 \end{itemize}
+
 \item The evolution of the 
@@
 \end{itemize}
+
 \item (\textbf{Approach}) To examine the effects on the evolution of one trait by adaptation in another, we develop a model of a traveling wave over a two dimensional trait space.
@@
 \end{itemize}
+
 \end{itemize}
@@
 \begin{itemize}
-\item (\textbf{Model parameters}) In our model we consider a population of asexuals at carrying capacity $N$ evolving in a static environment.
+\item In our model we consider a population of asexuals at carrying capacity $N$ evolving in a static environment.
 \begin{itemize}
@@
 \item Beneficial mutations contribute fitter alleles to the genetic diversity of each and occur at rate $U_k$ (trait $k=1,2$); deleterious mutations are not considered.
-	\item Mutations increase trait \textit{k}'s contribution to relative fitness by $s_k$, and we may express an individual's relative fitness as $r_{ij} = (i-\bar{i} ) s_1+(j-\bar{j} ) s_2$.
-	\item Here $i$ and $j$ represent the number of beneficial mutations that have accumulated in traits one and two, respectively, while $\bar{i}$ and $\bar{j}$ are the mean number of them in the average individual.
-	\item We divide the population into classes defined as groups of individuals with an identical number of beneficial mutations in each trait, and denote their abundances with $n_{ij}$ and frequencies with $p_{ij}$.
+\item Each new mutation in trait \textit{k} additively increases the trait's contribution to an individual's fitness by an amount $s_k$.
+\item An individual's overall fitness is $r_{ij} = i s_1+j s_2$, if it has accumulated $i$ mutations in the first trait and $j$ in the second.
+\item Its relative fitness is $(r_{ij}-\bar{r})$, where $\bar{r} = \bar{i} s_1+\bar{j} s_2$ is the population mean fitness, and $\bar{i}$ and $\bar{j}$ are the number of mutations in the average individual.
+\item We divide the population into classes composed of individuals with an identical number of beneficial mutations in each trait, and denote their abundances with $n_{ij}$ and frequencies with $p_{ij}$.
 \end{itemize}
-\item (\textbf{Pieces of the model}) We focus on the evolutionary regime characterized by large populations with strong selection, for which $1/N \ll s \ll 1$.
+
+\item We focus on the evolutionary regime characterized by large populations with strong selection for which $1/N \ll s \ll 1$.
 \begin{itemize}
-	\item The set of abundances $n_{ij}$ can then be categorized into two groups based on whether the stochastic fluctuations negligibly influence changes in abundances or don't. 
-	\item Following the convention of Desai and Fisher (2007), we refer to the former group as the bulk, and we will call the latter set of classes the stochastic front.
-	\item Stochastic fluctuations caused by drift can be ignored once abundances have grown sufficiently large and selection begins to dominate their dynamics, such a class is said to have established.
+\item This allows us to categorize the set of abundances $n_{ij}$ into two groups.
+\item Classes in the first group are dominated by selection, and we refer to this group as the bulk following the convention of Desai and Fisher (2007).
+\item Abundances of the remaining set of classes are largely affected by drift and flucuate randomly as a result.
+\item We refer to this group as the stochastic front.
 \end{itemize}
-\item (\textbf{Dynamics of the Bulk}) The dynamics of the bulk are driven by selection, and have deterministically over time.
+
+\item The bulk behaves nearly deterministic over time and we can approximate its behavior with a system of ordinary differential equations.
 \begin{itemize}
-	\item Since the selective advantage of these classes is $r_{ij}$, their abundances in a manner that can be approximated by the ordinary differential equation \[ \dot{n}_{ij}(t) = S_{ij} n_{ij}(t).
-\]
-	\item From Eq. 3, the rate of adaptation can be expressed as the sum of the individual rates of adaptation in each trait, indicated by subscripts, as $\dot{\bar{r}}=\dot{\bar{r}}_1 +\dot{\bar{r}}_2$.
+\item Abundances in the bulk change in proportion to their selective advantage and accordingly their dynamics follow the ODE \[ \dot{n}_{ij}(t) = r_{ij} n_{ij}(t). \]
 \end{itemize}
+
 \item (\textbf{Stochastic Front})
@@
 \end{itemize}
+
 \item (\textbf{Rate of Adaptation})
@@
 \item The work of Desai and Fisher (2007) provide the variance in fitness explicitly shown to work of gives the variance of adaptive traits in terms of the population’s parameters $N$, $U_b$, and $s$.
+\item From Eq. 3, the rate of adaptation can be expressed as the sum of the individual rates of adaptation in each trait, indicated by subscripts, as $\dot{\bar{r}}=\dot{\bar{r}}_1 +\dot{\bar{r}}_2$.
 \item The variance of traits and the covariance between them are estimated, and assume that the underlying traits satisfy basic assumptions, such as being polygenic. 
@@
 \end{itemize}
+
 \item (\textbf{Covariance and Trait Evolution})
@@
 \item We can also note from this expression that making a distinction between alleles that yields the same fitness advantage does not change the covariance.
+\end{itemize}
 
-	\end{itemize}
 \item (\textbf{Simulation description}) We examined the resulting dynamics of trait evolution using numerical simulations of our model carried out in Mathematica.
@@
 \end{itemize}
+
 \item Our numerical implementation consisted of two routines which ran iteratively for the duration of the simulation.
@@
 \end{itemize}
+
 \item We simulated the time of the next establishment by taking the first mutant whose lineage manages to fix. 
@@
 \end{itemize}
+
 \item Simulations were initialized with monoclonal populations at carrying capacity consisting of individuals having no beneficial mutations in either trait and relative fitness set zero.

### 2017-07-27 manuscriptoutline.tex
udpates

@@
 \begin{itemize}
-\item A single trait’s response to selection leads to a shift in the population mean towards the fittest phenotype.
-\item The response to selection with multiple traits involves both direct and indirect effects that together determine their evolution \cite{Scarcelli23102007,Lovell2013,Wagner2011}.  
+\item A single trait's response to selection leads to a shift in the population mean towards the fittest phenotype.
+\item With multiple traits, the response involves both direct and indirect effects that together determine the evolution each \cite{Scarcelli23102007,Lovell2013,Wagner2011}.  
 \item Indirect effects originate from genetic mechanisms that underlie trait expression, which may be associated or shared in a manner that enforces specific patterns of co-expression.
-\item Darwin, while this fact, recognized that such associations likely played an important role in directing the evolution of organisms under natural selection (Darwin p. 130-133).
-\item Presently, we know that these genetic associations can fundamentally alter how populations respond to selection and direct their adaptation towards certain evolutionary outcomes, such as speciation or extinction \cite{Felsenstein1979, Arnold2001, Arnold2008}.
-\item This makes them important to many micro- and macro-evolutionary questions central to evolutionary biology.
+\item These associations can alter fundamental features of adaptation and have significant evolutionary consequences \cite{Felsenstein1979, Arnold2001, Arnold2008}.
 \end{itemize}
 
-\item On short time scales, selection on phenotypes composed of multiple quantitative traits is best understood using the framework of quantitative genetics. 
+\item For quantitative traits, genetic associations and their effects on adaptation are mathematically described by Landau's equation (($\Delta \bar{z} = \textbf{G} \hspace{.05in}\beta$))
 \begin{itemize}	
-\item The strength of a genetic associations between traits are estimated and summarized by genetic correlations, or genetic covariances.
-\item Genetic covariances form part of the components of the \textbf{G} matrix.
-\item The response of phenotypes to selection is mathematically described by Lande's equation ($\Delta \bar{z} = \textbf{G} \hspace{.05in}\beta$).
-\item Here, $\Delta \bar{z}$ is the response of the mean of each trait, and $\beta$ is the selection gradient on the set of traits. 
-\item Lande’s formulation provides a means of estimating responses to selection correlated traits, but it does not indicate gives rise to them.
-\item Trait correlations can generally come about for a number of reasons, but they often have a genetic basis \cite{Saltz2017} arising from either pleiotropy, linkage disequilibria, or trans acting elements.
-\item Most evolutionary biologist agree that the first two are much more important.
+\item It states that the change in the population mean $\Delta \bar{z}$ equals the product of the selection gradient $\beta$ and what is known as the \textbf{G}-matrix, whose components are the additive genetic covariances between traits.
+\item Lande’s result has been used to extend much of our understanding of how phenotypes evolve under selection. 
+\item When combined with the notions of adaptive landscapes, a rich set of ideas emerge that emphasize the role of the \textbf{G}-matrix in directing adaptive evolution at both micro- and macro-evolutionary scales.
+\item This is one reason why the evolution of the \textbf{G}-matrix has been of key interest to evolutionary biologist. 
+\item Furthermore, these ideas make it apparent that the evolution of traits cannot understood in isolation, and in fact must be considered alongside what we know about the dynamics of the \textbf{G}-matrix.
 \end{itemize}
 
-\item Historically, there have been two schools of thought on which of the first two is responsible for most of the genetic correlations.
+\item The behavior of the \textbf{G}-matrix over time largely depends on what are the causes of the genetic correlations.
 \begin{itemize}
-\item The first of these was put forth by the Edinburgh school and held that pleiotropy was primary cause of most genetic correlations.
-\item This belief was largely motivated by their interests in relationships between parent-offspring phenotypes within randomly mating populations and their implication for their evolution.
-\item In these sorts of populations, linkage disequilibrium can rapidly decay by recombination on evolutionary time scales (\cite{fox2006evolutionary}, chp 20), while pleiotropic effects tend to persist.
-\item The alternative view which came out of the Birmingham school held that linkage disequilibria was much more significant.  
-\item Their interest led them to experiments involving inbred lines with plants where the genetic correlations from linkage disequilibria were just as substantial.
-\item Inarguably, the question of which view is perhaps correct largely depends the details of the population.
-\item 
+\item Correlations between quantitative traits can arise for a number of reasons, but they often have a genetic basis \cite{Saltz2017}.
+\item They arise from either pleiotropy, linkage disequilibria, or are the product of trans acting elements; the first two that are thought to be the most significant factors.
+\item Historically, there have been two schools of thought what the principle causes of genetic correlations are.
+\item The first was put forth by the Edinburgh school, which held that pleiotropy was principle source, since linkage disequilibrium could rapidly decay by recombination (\cite{fox2006evolutionary}, chp 20).
+\item This belief was largely motivated by interests the relationships between parent-offspring phenotypes within randomly mating populations, and its implications for evolution.
+\item The alternative view came out of the Birmingham school, which largely experimented with inbred lines of plants.
+\item Geneticists in this group believed that linkage disequilibria was much more important; a fact that often reflected in their experiments.
+\item Presently, the Edinburgh view is more widely accepted, and as a result, most models for the evolution of the \textbf{G}-matrix assume genetic correlations are pleiotropic proceed with examining how evolutionary processes influence them.
+\item We know far less about the about the behavior genetic correlations from linkage disequilibria over time, but this is not much of an issue for most complex organisms.
 \end{itemize}
 
-\item As populations evolve, the components of the \textbf{G}-matrix change as well, which may constrain or fascilitate adaptation.
+\item In the case of asexuals, the Birmingham view is more appropriate, and determining how the genetic correlations due to LD change over time is critical to our understanding of adaptation.
 \begin{itemize}
-\item The question of how stable the \textbf{G}-matrix is over time, is directly related to the question of how genetic correlations evolve, and that of course, depends on their causes.
-\item Today most geneticists assume that pleiotropy underlies many of the genetic correlations we observe directing adaptation, and as a result, have developed many sophisticated models to examine how genetic correlations arising from .
-\item We focus asexauls micro-organisms, which include some Eukaryotes and all Prokaryotes, experience very little recombination of their DNA or lack it entirely.	
-\item The fact the majority of life on the planet falls into this category underscores the Birmingham view to some degree.
+\item Linkage disequilibria persists for much longer periods of time in a asexuals, which can adversely affect adaptation.
+\item This fact has been used to argue for the evolutionary advantages of sex and recombination \cite{Barton2005}.
+\item Consequently, we can learn a substantial amount about the evolution of their traits by examining how genetic correlations from LD evolve over time and influence adaptation.
+\item Do so would not only reveal a great deal about the oldest and most abundant forms of life on the planet, such as bacteria, but could prove useful to our use of them in evolutionary experiments.
 \end{itemize}
 
-\item The evolution of the 
+\item Our best models for adaptation in asexuals make use of what are known as traveling waves, and are based on population genetics.
 \begin{itemize}
-\item We focus on asexual micro-organisms that lack recombination and examine how the evolution of their traits is affected by linkage.	
-\item In the case of adaptive traits, with fitness conferred taken as their quantitative measure, the relationship between variance and the population parameters is known explicitly. 
-\item The work of Desai and Fisher (2007) provide the variance in fitness explicitly shown to work of gives the variance of adaptive traits in terms of the population’s parameters $N$, $U_b$, and $s$.
-\item The variance of traits and the covariance between them are estimated, and assume that the underlying traits satisfy basic assumptions, such as being polygenic. 
-\item The variance in fitness was not something that could be determine from populations parameters, until only recently. 
-\item Desai and Fisher’s work (2007) provides the rate of adaptation of the population in terms of its parameters, \[v =s^2  \frac{2 \ln(Ns)-\ln(s/U_b)}{\ln^2(s/U_b)}\] and since the rate of adaptation is the variance of fitness ($v=\sigma^2$).
-\item If we consider a second adaptive trait evolving concurrently, with identical mutation rate and selection coefficient, then the rate of adaptation in each trait should be half of the total rate of adaptation $v(2U_b)$ since evolution in each contributes equally the overall adaptation of the population.
-\item A simple application of Desai and Fisher’s formula quickly shows that the rate of adaptation for a single trait is reduced when a second trait is considered since \[ 1/2  v(2U)<v(U). \]
-\item Naturally, we should expect such a result since we have doubled the mutation rate, and consequently, increased the amount of clonal interference that occurs within the population. 
-\item However, the mere fact that we are also considering a second trait indicates that there to this picture than this.
-\item The rate of adaptation in one trait is determined by its variance and covariance with other traits, and this relation is expressed in the breeder’s equation.
-In our situation, both quantitative traits are adaptive and we measure them by the fitness they endow the organism.
-Consequently, the rate of fitness increase of the focal trait $v_1$ must be related \[v_1=\sigma_1^2+\sigma_{12}.\]
+\item They have been used to establish important relationships between the population's parameter and key features of evolution in asexuals.
+\item Among these a model of Desai and Fisher \cite{desai2007beneficial}, which they used to derive an expression for the rate of adaptation, given by $v(N,s,U) \approx 2s^2 \ln(N s)/\ln^2(s/U)$.
+\item Here, $N$ is the population size, $s$ is the effective selective advantage of a beneficial mutation, and $U$ is the beneficial mutation rate.
+\item This is an important result, because it directly provides the additive variance in fitness by applying Fisher's fundamental theorem ($v=\sigma^2$).
+\item If we regard the population as having one adaptive trait, then this also equals the trait's additive variance and realize that we have said something more about the quantitative genetics involved.
 \end{itemize}
 
-\item (\textbf{Approach}) To examine the effects on the evolution of one trait by adaptation in another, we develop a model of a traveling wave over a two dimensional trait space.
+\item Unfortunately, if we apply Desai and Fisher's result to the case of two adaptive traits, we still can still solve for the overall fitness variance $\sigma^2$ of the population, but nothing can be said about the variances and covariance of the two traits.
 \begin{itemize}
-\item 
+\item This highlights the shortcomings of the traveling wave model in the case of multiple traits, for which we should expect Landau's equation to hold.
+\item As an example, consider a single adaptive trait evolving with parameters $N$, $s$ and $U$, then its variance and rate of adaption are $v(U;N,s)$ according to the formula given above.
+\item If the population acquires a second adaptive trait with identical parameters $s$ and $U$, then the new rate of adaptation is clearly $v(2U;N,s)$.
+\item By symmetry we know that rate of adaptation in each trait alone must be $0.5 \hspace{.02 in} v(2U;N,s)$, and furthermore, the rate of adaptation in trait one is now less with the second trait.
+\item Clearly, we lack any information about how trait two is affecting trait one's evolution, but Landau's equation tells us that these effects must somehow translate into changes in the additive variance of trait one and its covariance with the second.
+\end{itemize}
+
+\item To examine how adaptation in one trait is affected by evolution in another, we extend the work of Desai and Fisher's traveling wave model and consider a two dimensional trait space.
+\begin{itemize}
+\item We focus on the simplest case in which individuals have two independently evolving traits contributing to fitness.
+\item Our aim is to explore the effects of genetic correlations arising from linkage disequilibrium contribute to the interactions that direct trait evolution.
+\item So, we assume that the two traits are controlled separate non-pleiotropic loci.
+\item Our exploration of the question will done using simulations, whose results we present after discussing key details of the model in the next section.
+\item We will conclude with a discussion of the our findings and their implications.
 \end{itemize}
@@
 \begin{itemize}
-\item In our model we consider a population of asexuals at carrying capacity $N$ evolving in a static environment.
+\item We consider an asexually reproducing population at carrying capacity $N$ evolving in a static environment.
 \begin{itemize}
-\item Individuals have two traits contributing to their relative fitness which are controlled by separately loci that independently evolve.
-\item Beneficial mutations contribute fitter alleles to the genetic diversity of each and occur at rate $U_k$ (trait $k=1,2$); deleterious mutations are not considered.
-\item Each new mutation in trait \textit{k} additively increases the trait's contribution to an individual's fitness by an amount $s_k$.
-\item An individual's overall fitness is $r_{ij} = i s_1+j s_2$, if it has accumulated $i$ mutations in the first trait and $j$ in the second.
-\item Its relative fitness is $(r_{ij}-\bar{r})$, where $\bar{r} = \bar{i} s_1+\bar{j} s_2$ is the population mean fitness, and $\bar{i}$ and $\bar{j}$ are the number of mutations in the average individual.
-\item We divide the population into classes composed of individuals with an identical number of beneficial mutations in each trait, and denote their abundances with $n_{ij}$ and frequencies with $p_{ij}$.
+\item Individuals have two traits that contribute to their relative fitness, each controlled by separate loci that independently evolve.
+\item Beneficial mutations occur at a rate of $U_k$ (trait $k=1,2$); deleterious mutations are not considered.
+\item Each mutation additively increases the trait's contribution to an individual's fitness by an amount $s_k$.
+\item If the individual has accumulated $i$ mutations in the first trait and $j$ in the second, then its overall fitness is $r_{ij} = i s_1+j s_2$.
+\item Its relative fitness is then $(r_{ij}-\bar{r})$, where $\bar{r} = \bar{i} s_1+\bar{j} s_2$ is the population mean fitness, with $\bar{i}$ and $\bar{j}$ being the number of mutations found in the average individual.
+\item The population is divided into groups composed of individuals with an identical number of beneficial mutations in each trait; we will refer to them as classes.
+\item We denote their abundances by $n_{ij}$ and their frequencies by $p_{ij}$.
 \end{itemize}
 
-\item We focus on the evolutionary regime characterized by large populations with strong selection for which $1/N \ll s \ll 1$.
+\item We assumed that our population was large and evolving in a strong-selection strong-mutation regime, which required that the population parameters satisfied $1/N \ll s_k \ll 1$ and $N\hspace{.02in}U_k \gg 1$.
 \begin{itemize}
-\item This allows us to categorize the set of abundances $n_{ij}$ into two groups.
-\item Classes in the first group are dominated by selection, and we refer to this group as the bulk following the convention of Desai and Fisher (2007).
-\item Abundances of the remaining set of classes are largely affected by drift and flucuate randomly as a result.
-\item We refer to this group as the stochastic front.
+\item The first relation provides a separation of deterministic and stochastic behavior in abundances.
+\item This allowed use to consider the dynamics of the population as two interacting pieces, which greatly simplified formulation and implementation of the model.
+\item The second was necessary to ensure variation in each trait, which is a critical aspect of the question we aimed to answer.
+\item Following the work of Desai and Fisher (2007), we refer to set of classes whose abundances behave deterministically as the bulk.
+\item The set of remaining classes whose abundances behave stochastically due to drift are referred to as the stochastic front.
 \end{itemize}
 
-\item The bulk behaves nearly deterministic over time and we can approximate its behavior with a system of ordinary differential equations.
+\item The dynamics of the bulk are described by a system of ordinary differential equations.
 \begin{itemize}
-\item Abundances in the bulk change in proportion to their selective advantage and accordingly their dynamics follow the ODE \[ \dot{n}_{ij}(t) = r_{ij} n_{ij}(t). \]
+\item Abundances change in proportion to their selective advantage, and consequently, their behavior can be approximated by ODE 
+\begin{equation} 
+\dot{n}_{ij}(t) = (r_{ij}-\bar{r}) n_{ij}(t). 
+\end{equation}
+\item The population's size remains roughly constant over time as a result.
+\item Moreover, our assumptions concerning $N$ and $s$ imply that the sum of abundances in the stochastic front are negligible in comparison to the population's size.
+\item This allows us to model changes in the $\{p_{ij} \}$ for the bulk with an analogous ODE to equation 1.
+\item Additive variances and covariance, which tell us about how traits influence the evolution of one another, are entirely determined by these frequencies.
 \end{itemize}
 
-\item (\textbf{Stochastic Front})
+\item Classes in the stochastic front intermittently transition from stochastic to deterministic behavior and become part of the bulk.
 \begin{itemize}
-\item As for the one-dimensional traveling wave, the key variable of interest is the time-between establishments $\tau$ of new beneficial mutations. 
-\item This random variable summarizes all of the stochastic effects that accumulate over time during the growth of a fitness class, and determines the rate at which fitness classes in the stochastic front transition into the deterministic regime where selection is the prominent force.
-\item We will be interested in the rate of these transitions, because they will determine two very important aspects of the two-dimensional distribution’s evolution that influence trait evolution. 
-\item The probability of the latter event, denoted $\pi_{fix}$, is proportional to the selective advantage of the mutation.
-In these cases a beneficial mutation is said to have established in the population.
-\item The threshold separating these two characteristically different behaviors in growth/decline occurs at size equal to the inverse of the class' probability of fixation.
+\item This occurs when a class has grown sufficiently large so as to escape the effects of drift.
+\item The key aspects of this process are the rate at which it occurs and initial size of classes once they transition; both are entirely determined by the parameters $N$, $U_k$ and $s_k$. 
+\item The stochastic growth of classes in the front follow a branching process fed by incoming mutants generated by adjacent less fit classes in the bulk.
+\item Each of these continue to exhibit stochastic fluctuations until it has reached a size that 
 \end{itemize}
@@
 \begin{itemize}
+
 \item The results for one-dimensional traveling waves provides the relationship between variance in fitness from one trait evolving and population parameters $N$, $U_k$, and $s_k$. 

### 2017-08-05 manuscript.tex
more updates
updates

--- 2017-07-20 manuscript.tex
@@
-\documentclass[11pt,one column]{article}
-\usepackage[margin=0.5in]{geometry}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{natbib}
-\usepackage{graphicx}
-\usepackage{capt-of}
-\usepackage{multicol}
+\documentclass[9pt,twocolumn,twoside]{gsajnl}
+% Use the documentclass option 'lineno' to view line numbers
 
-\graphicspath{{figures/}}
+\articletype{inv} % article type
+% {inv} Investigation 
+% {gs} Genomic Selection
+% {goi} Genetics of Immunity 
+% {gos} Genetics of Sex 
+% {mp} Multiparental Populations
+
+\title{Modeling How Evolution in One Trait is Affected by Adaptation in Another}
+
+\author[$\ast$,1]{Kevin Gomez}
+\author[$\dagger$]{Jason Bertram}
+\author[$\dagger$]{Joanna Masel}
+
+\affil[$\ast$]{Department of Applied Mathematics, University of Arizona, and}
+\affil[$\dagger$]{Department of Evolution and Ecology, Univeristy of Arizona}
+
+\keywords{Keyword; Keyword2; Keyword3; ...}
+
+\runningtitle{GENETICS Journal Template on Overleaf} % For use in the footer
+
+%% For the footnote.
+%% Give the last name of the first author if only one author;
+% \runningauthor{FirstAuthorLastname}
+%% last names of both authors if there are two authors;
+% \runningauthor{FirstAuthorLastname and SecondAuthorLastname}
+%% last name of the first author followed by et al, if more than two authors.
+\runningauthor{Gomez \textit{et al.}}
+
+\begin{abstract}
+Abstract.
+\end{abstract}
+
+\setboolean{displaycopyright}{true}
 
 \begin{document}
-\begin{multicols}{2}
 
-\title{Modeling How Evolution in One Trait is Affected by Adaptation in Another}
-\author{Kevin Gomez, Joanna Masel, Jason Bertram}
-\date{2017}
 \maketitle
+\thispagestyle{firststyle}
+\marginmark
+\firstpagefootnote
+\correspondingauthoraffiliation{Please insert the affiliation correspondence address and email for the corresponding author.
+The corresponding author should be marked with a `1' in the author list, as shown in the example.}
+\vspace{-11pt}%
 
-\begin{abstract}
-This paper will discuss something.
-\end{abstract}
+\section{Introduction}
 
-\section*{Introduction}
 Natural selection frequently acts on multiple traits simultaneously.
-A single trait’s response to selection leads to a shift in the population mean towards the fittest phenotype.
-The response to selection with multiple traits involves both direct and indirect effects that together determine their evolution \cite{Scarcelli23102007,Lovell2013,Wagner2011}.
-Indirect effects arise from genetic mechanisms that underlie trait expression which may be associated or shared in a manner that enforces specific patterns of co-expression.
-These effects can significantly influence the evolution of populations on a micro- and macro-evolutionary scale.
-This was partly evident to Darwin (Darwin p.
-144-150) through his observations of animal and plant breeding.
-Evolutionary biologists since then have explore related questions experimentally and theoretically, but the most comprehensive framework for understanding  among geneticists and breeders .\par
+A single trait's response to selection leads to a shift in the population mean towards the fittest phenotype.
+With multiple traits, the response involves both direct and indirect effects that together determine the evolution each \citep{lande1983measurement,Scarcelli23102007,Lovell2013,Wagner2011}.
+Indirect effects originate from genetic mechanisms that underlie the phenotypic values of traits, which may be associated or shared in a manner that enforces specific relationships between them.
+These associations can alter fundamental features of adaptation and have significant evolutionary consequences \citep{Felsenstein1979, Arnold2001, Arnold2008}.\par
+% * <masel@email.arizona.edu> 2017-07-28T22:53:26.215Z:
+% 
+% > trait expression,
+% "Expression" is a funny word to use.
+For most biologists, it means gene expression.
+By calling it "trait expression", you are signalling that you mean something else, but I think readers might still find it confusing.
+The term "expression" pre-supposes that the genetic basis for something is there, and focuses on whether it is somehow realized or not; I don't think that is what you mean.
+% 
+% ^.
+% * <masel@email.arizona.edu> 2017-07-28T22:52:47.005Z:
+% 
+% > the response involves both direct and indirect effects
+% You need to define what you mean by direct and indirect effects: the reader doesn't know.
+% 
+% ^.
 
-In quantitative genetics, indirect effects stemming from the evolution in genetically associated traits is measured by genetic correlations, or alternatively, genetic covariances, and they form part of the statistical description of phenotypic responses to selection
+For quantitative traits, genetic associations and their effects on adaptation are mathematically described by Lande's equation ($\Delta \bar{z} = \textbf{G} \hspace{.05in}\beta$).
+It states that the change in the population mean $\Delta \bar{z}$ equals the product of the selection gradient $\beta$ and what is known as the \textbf{G}-matrix, whose components are the additive genetic covariances between traits.
+Lande’s result has been used to extend much of our understanding of how phenotypes evolve under selection, yielding a rich set of ideas that emphasize the role of the \textbf{G}-matrix in directing adaptive evolution at both micro- and macro-evolutionary scales.
+This is one reason why the evolution of the \textbf{G}-matrix has been of key interest to evolutionary biologist.
+Furthermore, these ideas make it apparent that the evolution of traits cannot understood in isolation, and in fact must be considered alongside what we know about the dynamics of the \textbf{G}-matrix.\par
+% * <masel@email.arizona.edu> 2017-07-29T12:43:16.916Z:
+% 
+% > When combined with the notions of adaptive landscapes
+% The link to adaptive landscapes is not obvious, including to me.
+G-matrix theory is all continuous, and the only possible landscape is a phenotype-fitness map.
+The adaptive landscape you are going to deal with is a genotype-fitness map, ie a completely different object.
+% 
+% ^.
 
-The extension of quantitative genetics to multivariate phenotypes, formulated by Lande (Lande 1979; 1980), provides the first characterization of selection on multiple traits, formally expressed in the breeder’s equation ($\Delta z = \beta$).
-The phenotypes of quantitative traits change ($\Delta z$) in response to a selection gradient ($\beta$) acting on them, given a specific set of trait variances and covariance between them ($G$).
-Lande’s formulation is largely independent of the principles of heredity.
-However, genetic correlations are known to be the produce of either pleiotropy, linkage disequilibria, and trans acting elements.\par
+The behavior of the \textbf{G}-matrix over time largely depends on what are the causes of the genetic correlations.
+Correlations between quantitative traits can arise for a number of reasons, but they often have a genetic basis \citep{Saltz2017}.
+They arise from either pleiotropy, linkage disequilibria, or are the product of trans acting elements; the first two that are thought to be the most significant factors.
+Historically, there have been two schools of thought what the principle causes of genetic correlations are.
+The first was put forth by the Edinburgh school, which held that pleiotropy was principle source, since linkage disequilibrium could rapidly decay by recombination (\citep{fox2006evolutionary}, chp 20).
+This belief was largely motivated by interests the relationships between parent-offspring phenotypes within randomly mating populations, and its implications for evolution.
+The alternative view came out of the Birmingham school, which largely experimented with inbred lines of plants.
+Geneticists in this group believed that linkage disequilibria was much more important; a fact that often reflected in their experiments.
+Presently, the Edinburgh view is more widely accepted, and as a result, most models for the evolution of the \textbf{G}-matrix assume genetic correlations are pleiotropic and proceed with examining how evolutionary processes influence them.
+We know far less about the about the behavior genetic correlations from linkage disequilibria over time, but this is not much of an issue for most complex organisms.\par
 
-In practice, geneticists can estimate $\beta$ and $G$ to determine the short-term response of quantitative traits to selection \cite{lynch1998genetics}, without knowledge of the underlying causes of that create correlations between traits.
-However, most quantitative geneticist adhere to two schools of thought on this matter; these are known as the Birmingham and Edinburgh schools.
-The differences in their views stem from the types of organism they studied and the types of questions they aimed to answer.\par
+In the case of asexuals, the Birmingham view is more appropriate, and determining how the genetic correlations due to LD change over time is critical to our understanding of adaptation.
+Linkage disequilibria persists for much longer periods of time in a asexuals, which can adversely affect adaptation.
+This fact has been used to argue for the evolutionary advantages of sex and recombination \citep{Barton2005,Otto2009}.
+Consequently, we can learn a substantial amount about the evolution of their traits by examining how genetic correlations from LD evolve over time and influence adaptation.
+Do so would not only reveal a great deal about the oldest and most abundant forms of life on the planet, such as bacteria, but could prove useful to our use of them in evolutionary experiments.\par
 
-Quantitative geneticists in the Edinburgh school have traditionally worked on questions concerning animal breeding, and much of their statistical models aim to clarifying the process by which phenotypic quantitative traits change in response to artificial selection in animals.
-Autho
-In such populations, the genetic foundation of phenotypes is hidden, and random breeding are inevitably core elements that must be accounted for in their genetic analysis.
-Consequently, arise from associations between alleles that form through either pleiotropy, linkage disequilibrium, or trans acting effects.\par
+Our best models for adaptation in asexuals make use of what are known as traveling waves, and are based on population genetics.
+They have been used to establish important relationships between the population's parameter and key features of evolution in asexuals.
+Among these is the model developed by \cite{desai2007beneficial}, which they used to derive an expression for the rate of adaptation, given by $v(N,s,U) = 2s^2 \ln(N s)/\ln^2(s/U)$.
+Here, $N$ is the population size, $s$ is the effective selective advantage of a beneficial mutation, and $U$ is the beneficial mutation rate.
+This is an important result, because it directly provides the additive variance in fitness by applying Fisher's fundamental theorem ($v=\sigma^2$).
+If we regard the population as having one adaptive trait, then this also equals the trait's additive variance and realize that we have said something more about the quantitative genetics involved.\par
 
-There are two schools of thought among quantitative geneticists that differ in the principle cause of associations between alleles.
-One of them holds pleiotropy as the largely responsible for  pleiotropic effects each of them The views held by each are largely the result of distinct questions and applications Geneticist in the Edinburgh school hold that genetic correlations are the product of pleiotropic effects, while those in the Birmingham school believe that linkage The majority of quantitative genetics follow Geneticist pleiotropy.\par
+Unfortunately, if we apply Desai and Fisher's result to the case of two adaptive traits, we still can still solve for the overall fitness variance $\sigma^2$ of the population, but nothing can be said about the variances and covariance of the two traits.
+This highlights the shortcomings of the traveling wave model in the case of multiple traits, for which we should expect Lande's equation to hold.
+As an example, consider a single adaptive trait evolving with parameters $N$, $s$ and $U$, then its variance and rate of adaption are $v(U;N,s)$ according to the formula given above.
+If the population acquires a second adaptive trait with identical parameters $s$ and $U$, then the new rate of adaptation is clearly $v(2U;N,s)$.
+By symmetry we know that rate of adaptation in each trait alone must be $0.5 \hspace{.02 in} v(2U;N,s)$, and furthermore, the rate of adaptation in trait one is now less with the second trait.
+Clearly, we lack any information about how trait two is affecting trait one's evolution, but Lande's equation tells us that these effects must somehow translate into changes in the additive variance of trait one and its covariance with the second.\par
+% * <masel@email.arizona.edu> 2017-07-29T13:00:14.234Z:
+% 
+% > Clearly, we lack any information about how trait two is affecting trait one's evolution,
+% Spell out what exactly it is that we don't know, ie the degree to which the reduction in adaptation rate of trait 1 is due to a change in variance or a change in covariance.
+% 
+% 
+% ^.
 
-We focus on asexual micro-organisms that lack recombination and examine how the evolution of their traits is affected by linkage.
-A variety of organisms fall under this category, which includes many single celled Eukaryotes and a variety of bacteria (cite work).
-In this setting, linkage disequilibria is entirely determined by the combined action of mutations, selection and drift.
-Standard results in population genetics demonstrate that the amount of linkage disequilibria between various alleles will depend on the beneficial mutation rate ($U_b$), the selection coefficient ($s$), and strength of drift measured by population’s size ($N$).
-Logically, one expects variances and covariances arising from linkage to also depend on the population’s parameters, but a general expression that details this explicit relationship has not been uncovered.
-The majority of results concerning linkage comes from experiments with inbred lines in plants, stemming from the Birmingham school, and that have limited application to populations of interest to us Jinks and Mather (1981).\par
+To examine how adaptation in one trait is affected by evolution in another, we extend the work of Desai and Fisher's traveling wave model and consider a two dimensional trait space.
+We examine simplest case in which individuals have two independently evolving traits contributing to fitness.
+Our aim is to explore the effects of genetic correlations arising from linkage disequilibrium contribute to the interactions that direct trait evolution.
+So, we assume that the two traits are controlled separate non-pleiotropic loci.
+Our exploration of the question will done using simulations, whose results we present after discussing key details of the model in the next section.
+We will conclude with a discussion of the our findings and their implications.
+% * <masel@email.arizona.edu> 2017-07-29T13:01:43.541Z:
+% 
+% > two independently evolving traits
+% They are not evolving independently. What you mean instead is that there is no pleiotropy. You might also summarize other key simplifications here, eg constant N, no epistasis etc.
+% 
+% ^.
 
-In the case of adaptive traits, with fitness conferred taken as their quantitative measure, the relationship between variance and the population parameters is known explicitly.
-The work of Desai and Fisher (2007) provide the variance in fitness explicitly shown to work of gives the variance of adaptive traits in terms of the population’s parameters $N$, $U_b$, and $s$.
-The variance of traits and the covariance between them are estimated, and assume that the underlying traits satisfy basic assumptions, such as being polygenic.
-The variance in fitness was not something that could be determine from populations parameters, until only recently.
-Desai and Fisher’s work (2007) provides the rate of adaptation of the population in terms of its parameters,
+\section{Materials and Methods}
+\label{sec:materials:methods}
+
+We considered an asexual population at carrying capacity $N$ in which haploid individuals had two traits contributing to their relative fitness.
+Each trait was assumed to be under the control of a single locus.
+We assumed that beneficial mutations improving each trait appeared at a rate of $U_k$ ($k=1,2$) and increase th fitness by $s_k$; deleterious mutations were excluded.
+Individuals that had accumulated $i$ mutations in the first and $j$ in the second were taken to have an overall fitness of $r_{ij} = i s_1+j s_2$.
+Their relative fitness was the difference $(r_{ij}-\bar{r})$, where  $\bar{r} $  is the population mean fitness.
+
+In analogous approach to that of \cite{desai2007beneficial}, we divided our population into classes bases on the number of beneficial mutations in each trait.
+Their abundances are accordingly denoted $n_{ij}$ and their frequencies by $p_{ij}$.
+We assumed that $1/N \ll s \ll 1$ to ensure that the set of classes could be neatly separated into two groups based on whether their behavior was deterministic and stochastic.
+Following the convention of Desai and Fisher, we refer to the set of deterministically behaving classes as the bulk, while those that grow stochastically and have a positive selective advantage ($r_{ij}>\bar{r}$) will be referred to as the stochastic front (see \textbf{fig.1}).\par
+
+The dynamics of the bulk was modeled by a system of ordinary differential equations, in which abundances changed in proportion to their selective advantage (\textbf{eqt. 1}). 
 \begin{equation} 
-v =s^2  \frac{2 \ln(Ns)-\ln(s/U_b)}{\ln^2(s/U_b)}
+\dot{n}_{ij}(t) = (r_{ij}-\bar{r}) n_{ij}(t). 
 \end{equation}
-and since the rate of adaptation is the variance of fitness ($v=\sigma^2$).
-If we consider a second adaptive trait evolving concurrently, with identical mutation rate and selection coefficient, then the rate of adaptation in each trait should be half of the total rate of adaptation $v(2U_b)$ since evolution in each contributes equally the overall adaptation of the population.
-A simple application of Desai and Fisher’s formula quickly shows that the rate of adaptation for a single trait is reduced when a second trait is considered since
-\[ 1/2  v(2U)<v(U) \]
-Naturally, we should expect such a result since we have doubled the mutation rate, and consequently, increased the amount of clonal interference that occurs within the population.
-However, the mere fact that we are also considering a second trait indicates that there to this picture than this.\par
+Abundances in the stochastic front were negligible in size relative to $N$ due to our parameter assumptions stated earlier.
+Since the bulk represents the majority of population, we assumed that the frequencies of the bulk followed an analogous ODE.
+\begin{equation} 
+\dot{p}_{ij}(t) = (r_{ij}-\bar{r}) p_{ij}(t) 
+\end{equation}
+These frequencies determine all of the quantitative genetic measures that are of interest to us.\par
 
-The rate of adaptation in one trait is determined by its variance and covariance with other traits, and this relation is expressed in the breeder’s equation.
-In our situation, both quantitative traits are adaptive and we measure them by the fitness they endow the organism.
-Consequently, the rate of fitness increase of the focal trait $v_1$ must be related
+Classes in the stochastic front intermittently transition from stochastic to deterministic behavior and become part of the bulk.
+Such classes are said to have established.
+Their growth can be modeled as a branching processes fed by incoming mutants generated in adjacent, less fit, classes within the bulk.
+The chances of a mutant lineage establishing is given by the probability of fixation.
+This was obtained using the expression below derived by \citep{Uecker2011} [eqt.
+16b].
+% * <masel@email.arizona.edu> 2017-07-29T16:52:06.887Z:
+% 
+% > This was obtained using the expression below derived by \citep{Uecker2011} [eqt. 16b].
+% Do you use the full generality of Uecker & Hermisson, or are earlier expressions, eg as used in Desai & Fisher, sufficient for your work?
+% 
+% 
+% ^.
 \begin{equation}
-v_1=\sigma_1^2+\sigma_{1,2}
+p_{fix} = 2^{-1}  \left(1 + \int_0^\infty (b_{ij} +d_{ij}) e^{-\int_0^{t'}(b_{ij}-d_{ij}) d t'} \right)
 \end{equation}
+The terms $b_{ij}=1+(r_{ij}-\bar{r})$ and $d_{ij}=1$ are the respective birth and death rates of the stochastically growing class.
+A mutant lineage that is destined to establish transitions into its deterministic regime of growth once it has roughly reached the size $n_{ij} = 1/p_{fix}$.
+Rather than attempting to model the timing of this transition, we instead focus on equivalently choosing the initial size $\nu$ that produces the same deterministic trajectory.
+The cumulative distribution of the random variable $\nu$, expressed below, is provided by \citep{Uecker2011} [eqt.
+40].
+\begin{equation}
+P(\nu \leq \nu_0)= 1-e^{-\nu_0/p_{fix}}
+\end{equation}
+We assumed that the next class to transition into the bulk would be the first in which  a mutant lineage appear that was destined to establish.\par
 
-\section*{Results}
-Discuss results here
-{\centering
-\includegraphics[width=1.0\linewidth]{fig2_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1}
-\captionof{figure}{Mean rates of adaptation, variance and covariance.} \label{fig.1}}
+In order to investigate how of traits affect one another, we identified the contributions of each trait to the evolution of the population.
+The rate of adaptation can obtained by differentiating the population's mean relative fitness $\bar{r}=\sum p_{ij} r_{ij} $.
+We combined these results with \textbf{equation 2} and arrived at the expression $\dot{\bar{r}}=\dot{\bar{r_1}} +\dot{\bar{r_2}}$.
+In our simple model, the population's rate of adaptation could therefore be expressed as the sum of the individual rates.
+We found that each of these rates could then be rewritten in terms of trait variances and covariance and obtained the system below.
+% * <masel@email.arizona.edu> 2017-07-29T17:21:52.605Z:
+% 
+% >  We found that each of these rates could then be rewritten in terms of trait variances and covariance
+% This is not something that "we found" or "obtained".
+It is simply a statement of Lande's equation for an appropriate (normalized) definition of what a trait is.
+Also, remember to define the various sigma terms.
+% 
+% ^.
+% * <masel@email.arizona.edu> 2017-07-29T17:18:22.431Z:
+% 
+% > arrived at the expression $\dot{\bar{r}}=\dot{\bar{r_1}} +\dot{\bar{r_2}}$
+% I'm OK going straight to the expression. Your derivation of it is hard to follow, but I'm not convinced it is necessary.
+% 
+% ^.
+\begin{equation}
+\left(
+\begin{array}{c}
+\dot{\bar{r_1}} \\
+\dot{\bar{r_2}} 
+\end{array}
+\right)
+=
+\left(
+\begin{array}{cc}
+\sigma_1^2 & \sigma_{12} \\
+\sigma_{12} & \sigma_2^2 
+\end{array}
+\right)
+\left(
+\begin{array}{c}
+1 \\
+1 
+\end{array}
+\right)
+\end{equation}
+This is exactly Lande's equation with a selection gradient and trait response whose components have been normalized.
 
-{\centering
-\includegraphics[width=1.0\linewidth]{fig3_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1}
-\captionof{figure}{Variance and covariance fluctuations over time.}
-\label{fig.2}}
+In our model, covariance between traits is arises entirely from linkage disequilibria.
+We can assume that specific alleles at either locus correspond to the number of beneficial mutations that have accumulated in the corresponding trait.
+It follows that the $p_{ij}$ are the gametic frequencies for the genome consisting of allele $i$ paired with allele $j$, and $D_{ij} = p_{ij}-p_{i\cdot} p_{\cdot j}$ is their coefficient of linkage disequilibrium ( $p_{i\cdot}$ and $p_{\cdot j}$ are marginal distributions).
+By applying \textbf{equation 2}, we can derive a mathematical relationship between covariance and linkage disequilibria.
+% * <masel@email.arizona.edu> 2017-07-29T17:34:47.794Z:
+% 
+% > specific alleles at either locus correspond to the number of beneficial mutations that have accumulated in the corresponding trait
+% Multiple loci can contribute to the same trait.
+An allele cannot be a "number of beneficial mutations", because an allele must pertain to one locus.
+Stick to the word "genotype" not "allele", eg "genotype ij has i beneficial mutations in trait 1 and j beneficial mutations in trait 2 compared to the ancestral population".
+This means that you cannot calculate linkage disequilibrium as you do: linkage disequilibrium is generally defined as being between two particular beneficial mutations.
+What you calculated was simply a covariance.
+Because of this confusion, I got a bit lost here and so skipped Equation 6.
+% 
+% ^.
+\begin{equation}
+\sigma_{12}=\sum_{ij}D_{ij}[(i-\bar{i})s_1][(j-\bar{j})s_2]. 
+\end{equation} 
+As the population adapts, mutations and selection  increase and decrease the amounts LD which ultimately determine how traits interact in their evolution.\par
 
-\section*{Discussion}
-\section*{Methods}
-We consider a population with fixed size N, in which individuals have two traits that contribute to its fitness, its growth rate, and which evolve through beneficial mutations; deleterious mutations are not considered.
-The beneficial mutation rate for each trait is denoted $U_k$ (trait $k=1,2$), and each increases trait $k$’s contribution to relative fitness by $s_k$.
-We can express an individual fitness as
-\[ r_{ij}=r+i s_1+j s_2 \]
-where i and j represent the number of beneficial mutations in trait one and two, while $r$ is the growth rate a common ancestor to the population. 
-An individual’s relative fitness is given by
-\[ (r_{ij}-r ̅)=(i-\bar{i} ) s_1+(j-\bar{j} ) s_2 \]
-respectively, and i ̅ and j ̅ are the population mean.
-The two traits of individuals are treated as independently evolving loci in which beneficial mutations occur at a fixed rate $U_k$ ($k=1,2$).
-Each beneficial mutation is either lost to drift, or the subpopulation carrying it will grow sufficiently large to ensure that the beneficial mutation fixes in the population through selection.
-The probability of the latter event, denoted $\pi_{fix}$, is proportional to the selective advantage of the mutation.
-In these cases a beneficial mutation is said to have established in the population.
-The selective advantage of an individual is measured by its relative fitness, which in our model, is its expected growth rate minus the growth rate of an average individual.
-Each new beneficial mutation in trait $k$ increases this fitness by $s_k$; we do not consider deleterious and neutral mutations in our model.
-If an individual has i mutations in the first trait and $j$ in the second, while the average number carried in the two traits is $\bar{i}$ and $\bar{j}$, its relative fitness equals the difference $(r_{ij}-\bar{r})=(i-\bar{i}) s_1+(j-\bar{j}) s_2$.
-We group members of the population that carry the same number of beneficial mutations in each trait, $i$ mutations in the first and $j$ in the second, into fitness classes that will often be denoted by subscripts $ij$.
-The set of fitness class abundances $\{n_{ij} \}$ induce a distribution over the trait space, represented by $N^2$, containing all fitness classes.
-This two-dimensional distribution, which we will often denote by the set of frequencies $\{p_{ij}\}$, travels in the direction of increasing fitness as the population adapts (Fig 1) in an analogous fashion to the motion of Desai and Fisher’s (2007) one dimensional traveling wave.
-We also examine the properties of the marginal distributions $\{p_{i\cdot}\}$ and $\{p_{\cdot j}\}$ for each trait, whose behaviors over time are significantly affected by genetic correlations that arise from the confluences of mutations, selection and drift.
-\begin{equation}
-p_{i \cdot }= \sum_j p_{ij} \hspace{0.5in} p_{\cdot j}= \sum_i p_{ij}
-\end{equation}
-In order to study how genetic correlations arise and influence the evolution of traits in adapting populations, we focus on regimes of adaptation characterized by significant levels of genetic variation in each trait.
-The results for one-dimensional traveling waves provides the relationship between variance in fitness from one trait evolving and population parameters $N$, $U_k$, and $s_k$.
-When mutations appear infrequently relative to the average time required for beneficial mutation to fix, we obtain the simplest behavior possible in which beneficial mutations establish in succession and spread to the population long before the next one appears ($N U_k \ll 1/\ln(N s_k)$).
-These populations rarely have any variance in fitness stemming from trait diversity, and are not the focus of our investigations.
-We focus instead on the case where many beneficial mutations establish prior to any one fixing ($N U_k \geq 1/\ln(N s_k)$), where the effects of clonal interference cause variation in fitness which persists over time in the type of populations we consider.
-For large populations with strong selection in each trait ($N^{-1}\ll s_k \ll 1$), changes in the abundances of fitness class are either dominated by stochastic fluctuations due to drift ($n_{ij}<\pi_{fix}^{-1}$), or growth deterministically otherwise due to selection.
-We will refer to fitness classes in the first group as the front of the wave, and their growth can be modeled as branching processes that are fed new mutants from adjacent less fit classes.
-The fitness classes in the stochastic front will consist of those which are fitter and adjacent to ones that have established whenever $s_k/U_k \gg 1$.
-During the stochastic phase of growth, fitness classes are much too small to influence dynamics that are driven by selection, so that they may be ignored until the moment they do establish.
-As for the one-dimensional traveling wave, the key variable of interest is the time-between establishments τ of new beneficial mutations.
-This random variable summarizes all of the stochastic effects that accumulate over time during the growth of a fitness class, and determines the rate at which fitness classes in the stochastic front transition into the deterministic regime where selection is the prominent force.
-We will be interested in the rate of these transitions, because they will determine two very important aspects of the two-dimensional distribution’s evolution that influence trait evolution.
-The abundances of fitness classes that have established, whose behavior is dominated by the action of selection, deterministically grow and decline in size.
-We refer to the fitness classes in this group as the bulk of the population.
-The quantities i ̅ and j ̅ are marginal means computed from the marginal distributions (Eq.
-1).
-\begin{equation}
-\bar{i}=∑_i i p_{i \cdot} \hspace{0.5in} \bar{j} = ∑_j p_{⋅j} j
-\end{equation}
-Relative fitness, written as the difference $(r_{ij}-\bar{r})=(i-\bar{i}) s_1+(j-\bar{j}) s_2$, determines the growth and decline in abundances $n_{ij}$ that follow the differential equation
-\begin{equation}
-\dot{n}_{ij}=n_{ij} (r_{ij}-\bar{r})-n_{ij} \left(\frac{N(t)-N}{N}\right) r 
-\end{equation}
-where $N(t)=∑_{ij} n_{ij}(t)$ above, and the right-most term is the density-regulation component of growth.
-Stochastic fluctuations in the total population size are negligible ($|N(t)-N| \ll N$), and so we may also consider the frequencies of fitness classes in the bulk to obey Eq.
-3 below.
-\begin{equation}
-\dot{p}_{ij}=p_{ij} (r_{ij}-\bar{r})
-\end{equation}
-From Eq. 3, the rate of adaptation can be expressed as the sum of the individual rates of adaptation in each trait, indicated by subscripts, as $\dot{\bar{r}}=\dot{\bar{r}}_1 +\dot{\bar{r}}_2$.
-Each of the two individual rates are related to the variance in fitness from the respective trait and its covariance in fitness with other (Eq. 4).
-\begin{equation}
-\dot{\bar{r}}_1=\sigma_1^2+\sigma_{1,2}\\
-\dot{\bar{r}}_2=\sigma_2^2+\sigma_{1,2}
-\end{equation}
-Mutation-selection balance induces steady state values for the speed of evolution in the two, from which it possible to determine the two variances, the single covariance, and furthermore, the resulting G-matrix of the two traits (Eq.
-5).
+We examined the resulting dynamics of trait evolution using numerical simulations of our model using Mathematica.
+We focused on modeling the symmetric case, in which the two traits had identical mutation rate and selection coefficients.
+These, along with the population size, were chosen to match values reported in evolutionary experiments involving relevant micro-organisms (see Table 1).
+
+Our numerical implementation consisted of two routines which ran iteratively for the duration of the simulation.
+Each iteration began with solving for the time-dependent abundances in the bulk governed by \textbf{equation 1} over a time interval of one-thousand generations.
+We used Mathematica's standard numerical ODE solver to obtain the solutions and used them to determine the appearance time of the next mutant lineage destined to establish.
+Having computed the timing of this event, we evaluated the solutions and accordingly updated the abundances of the bulk.
+Any classes with less than one individual were removed.
+The establishing class was then added to the bulk with an appropriate initial size.
+Following this step, the next iteration began with solving for the abundances of the new bulk, followed by the subsequent steps outlined above.\par
+
+The contributions of the stochastic front were incorporated into the adaptive evolution of the population by simulating each establishment event.
+We did this by identifying the first mutant lineage destined to establish among all mutants appearing in the stochastic front.
+Mutants appear according to a time-inhomogeneous Poisson process with rate functions given by the abundances of the adjacent classes from which they originated.
+Consequently, the arrival time $\tau_l$ of the $l^{th}$ mutant generated from the class $n_{ij}(t)$ having a mutation in trait $k$ can be used to form the random variable
 \[ 
-G= \left(
-\begin{array}{cc}
-\sigma_1^2& \sigma_{1,2}\\
-\sigma_{1,2} & \sigma_2^2
-\end{array}
-\right) \]
-Linkage disequilibria continually arises throughout the adaptation process and limits the response of traits to selection.
-In the case of two alleles, linkage disequilibria is a measure of the statistical association between them.
-For simplicity, we can regard all ith beneficial mutations occurring in trait one as the same allele, and likewise with the jth mutation in trait two.
-This allows us to treat the difference $D_{ij}=p_{ij}-p_{i\cdot} p_{\cdot j}$ as a measure of linkage disequilibrium between the two alleles corresponding to the $i^{th}$ mutation in trait one, and the $j^{th}$ in trait two.
-In turn, the covariance in fitness can be written out as Eq.
-6.
-\begin{equation}
-\sigma_{1,2}=\sum_{ij}D_{ij}(i-\bar{i})s_1 (j-\bar{j}) s_2
-\end{equation}
-and it becomes evident that linkage disequilibrium directly affects the evolution of the two traits.
-We can also note from this expression that making a distinction between alleles that yields the same fitness advantage does not change the covariance.
+\int_0^{\tau_l} U_k n_{ij}(t) dt ,
+\] 
+which has the distribution of $l$ i.i.d exponentially distributed summed random variables.
+By randomly sampling from the this distribution, we could then solve for each arrival time $\tau_l$.
+We then computed the probability of fixation using \textbf{equation 5}.
+The integral was numerically approximated up to an appropriate cutoff.
+To determine if a mutation would establish, we sampled a uniform random variable and checked it against the calculated $p_{fix}$.
+The next class to establish was chosen accordingly and incorporated into the bulk with a random initial size generated with from \textbf{equation 4}.\par
 
-Discuss the details of the simulations.
+Simulations were initialized with monoclonal populations at carrying capacity consisting of individuals having no beneficial mutations in either trait and relative fitness set zero.
+We allowed the simulation to run for 5,000 generations to allow the population to achieve beneficial mutation-selection balance before collecting data.
+This ensured that the transient effects from the initializing would not distort the statistics collected for the distribution.
+Following the burn period of the simulation, we recorded the existing classes in the bulk and their respective abundances every one 100 generations and intermittent times at which new classes established.
+The data was used to compute the marginal mean fitness and variances of each trait, as well as their covariance.
+From these quantities we also computed the instantaneous rate of adaptation (\textbf{eqtn.
+5}), the mean rate of fitness increase in each trait and the overall rate of adaptation.
 
-\bibliographystyle{plain}
+\section{Results and Discussion}
+
+Results and Discussion.
+
+% \section{In-text Citations}
+
+% Add citations using the \verb|\citep{}| command, for example \citep{neher2013genealogies} or for multiple citations, \citep{neher2013genealogies, rodelsperger2014characterization}
+
+% \section{Examples of Article Components}
+% \label{sec:examples}
+
+% The sections below show examples of different header levels, which you can use in the primary sections of the manuscript (Results, Discussion, etc.) to organize your content.
+
+% \section{First level section header}
+
+% Use this level to group two or more closely related headings in a long article.
+
+% \subsection{Second level section header}
+
+% Second level section text.
+
+% \subsubsection{Third level section header:}
+
+% Third level section text. These headings may be numbered, but only when the numbers must be cited in the text.
+
+% \section{Figures and Tables}
+
+% Figures and Tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.
+
+% \subsection{Sample Figure}
+
+% Figure \ref{fig:spectrum} shows an example figure.
+
+% \begin{figure}[htbp]
+% \centering
+% \includegraphics[width=\linewidth]{example-figure}
+% \caption{Example figure from \url{10.1534/genetics.114.173807}.
+Please include your figures in the manuscript for the review process.
+You can upload figures to Overleaf via the Project menu.
+Upon acceptance, we'll ask for your figure files to be uploaded in any of the following formats: TIFF (.tiff), JPEG (.jpg), Microsoft PowerPoint (.ppt), EPS (.eps), or Adobe Illustrator (.ai).
+Images should be a minimum of 300 dpi in resolution and 500 dpi minimum if line art images.
+RGB, CMYK, and Grayscale are all acceptable.
+Halftones should be high contrast with sharp detail, because some loss of detail and contrast is inevitable in the production process.
+Figures should be 10-20 cm in width and 1-25 cm in height.
+Graph axes must be exactly perpendicular and all lines of equal density.
+% Label multiple figure parts with A, B, etc.
+in bolded type, and use Arrows and numbers to draw attention to areas you want to highlight.
+Legends should start with a brief title and should be a self-contained description of the content of the figure that provides enough detail to fully understand the data presented.
+All conventional symbols used to indicate figure data points are available for typesetting; unconventional symbols should not be used.
+Italicize all mathematical variables (both in the figure legend and figure) , genotypes, and additional symbols that are normally italicized.
+% }%
+% \label{fig:spectrum}
+% \end{figure}
+
+% \subsection{Sample Table}
+
+% Table \ref{tab:shape-functions} shows an example table.
+Avoid shading, color type, line drawings, graphics, or other illustrations within tables.
+Use tables for data only; present drawings, graphics, and illustrations as separate figures.
+Histograms should not be used to present data that can be captured easily in text or small tables, as they take up much more space.
+
+% Tables numbers are given in Arabic numerals. Tables should not be numbered 1A, 1B, etc., but if necessary, interior parts of the table can be labeled A, B, etc. for easy reference in the text.
+
+% \begin{table*}[htbp]
+% \centering
+% \caption{\bf Students and their grades}
+% \begin{tableminipage}{\textwidth}
+% \begin{tabularx}{\textwidth}{XXXX}
+% \hline
+% Student & Grade\footnote{This is an example of a footnote in a table.
+Lowercase, superscript italic letters (a, b, c, etc.) are used by default.
+You can also use *, **, and *** to indicate conventional levels of statistical significance, explained below the table.} & Rank & Notes \\
+% \hline
+% Alice & 82\% & 1 & Performed very well.\\
+% Bob & 65\% & 3 & Not up to his usual standard.\\
+% Charlie & 73\% & 2 & A good attempt.\\
+% \hline
+% \end{tabularx}
+%   \label{tab:shape-functions}
+% \end{tableminipage}
+% \end{table*}
+
+% \section{Sample Equation}
+
+% Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
+% \begin{equation}
+% S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
+%       = \frac{1}{n}\sum_{i}^{n} X_i
+% \label{eq:refname1}
+% \end{equation}
+% denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.
+
 \bibliography{biblio}
-\end{multicols}
-
-\newpage
-
-\section*{Appendix A:}
-
-\section*{scrap text}
- for example,  the rate of adaptation, and developed his now famous geometric model to infer that complexity would lead to slower adaptation in living organisms.
-Other majors works addressing evolutionary biology would follow, such as (cite), but quantitative genetics to provide some of the most The model provides a simple geometric argument demonstrating how adaptive evolution would be constrained in higher dimensional trait spaces.
-Kimura (1983) gave further extensions to this model by incorporating probability of fixation, and Orr (1998, 2000) provided additional insights on the the optimal dimensions that ensured maximum adaptation.
-Some of the earliest works that attempted to address multi-trait adaptation were developed by Fisher (1930)However, early evolutionary biology could provide little development in this area, in part due to standing questions concerning the relationships between traits and their genetic foundation, such as how to reconcile Mendelian genetics and with the inheritance of continuous characters.
-Thus questions concerning whole organisms, which are of main interest to plant and animal breeders, and the level of whole organisms.\par
 

### 2017-08-05 plotdata.py

@@ -10,6 +10,7 @@ import pickle
 import scipy as sp
 import numpy as np
 import matplotlib.pyplot as plt
+from scipy.stats import multivariate_normal
 
 # set parameters of simulation and create required variables
 N=1e9; s1=1e-2; s2=1e-2; U1=1e-5; U2=1e-5;
@@ -27,6 +28,48 @@ def get_sample_window(times,start_time,end_time):
         if times[end_indx] <= end_time:
             end_indx = end_indx + 1
     return [start_indx,end_indx]
+
+def orient_my_2D_distr():
+    
+    return flipped_arry
+
+def generate_2D_discrete_gauss(distr_grid, min_fit_classes,
+                                     mean_fit_class, cov_matrix, tot_pop_size):
+    # distr_grid should be a square array with 1's where the subpop is nonzero
+    # mean indx set (i,j) represents the mean of the gaussian
+
+    [genotypes,abundances] = [[],[]]    
+    
+    for i in range(len(distr_grid[:,0])):
+        for j in range(len(distr_grid[0,:])):
+            if(distr_grid[i,j] != 0):
+                genotypes = genotypes+[[i+min_fit_classes[0],
+                                                j+min_fit_classes[1]]]
+                abundances = abundances+[multivariate_normal.pdf([i,j],
+                                                mean_fit_class,cov_matrix)]
+    
+    abundances = [(tot_pop_size/sum(abundances))*abundances[i] 
+                                            for i in range(len(abundances))]
+    return [genotypes,abundances]
+
+def get_2D_distr(genotypes,abundances,box_dim):
+    #box_dim = [[boxwidth1 (> max_geno1-min_geno1),min indx1 = min_fit_class1],
+    #               ...,[boxwidth2 (> max_geno2-min_geno2),min indx2 = min_fit_class2,...]]
+    tot_pop_size = sum(abundances)
+    
+    dim1_data = [np.min([genotypes[i][0] for i in range(len(abundances))]),
+                 np.max([genotypes[i][0] for i in range(len(abundances))])]
+    dim2_data = [np.min([genotypes[i][1] for i in range(len(abundances))]),
+                 np.max([genotypes[i][1] for i in range(len(abundances))])]
+                 
+    my_distr = np.zeros([box_dim[0][0],box_dim[1][0]])
+        
+    for i in range(len(abundances)):
+        indx1 = genotypes[i][0] - dim1_data[1] + box_dim[0][1]
[…]

### 2017-08-05 plots.py

@@ -307,4 +307,43 @@ ax5.set_ylabel("Trait 2 Mutations")
 ax5.set_zlabel("Fitness")
 
 
-# second animation
+# Grid Plot for 2D distribution
+
+min_fit_classes = [0,0]
+mean_fit_class = [7,7]
+cov_matrix = [[1,0],[0,1]]
+cut_off = 0.0001
+distr_grid = np.zeros([16,16])
+box_dim = [[20,7],[20,7]]
+[arry_dim1,arry_dim2]=[np.shape(distr_grid[:,0])[0],np.shape(distr_grid[0,:])[0]]
+
+for i in range(arry_dim1):
+    for j in range(arry_dim2):
+        if(multivariate_normal.pdf([i,j],mean_fit_class,cov_matrix) >= cut_off):
+            distr_grid[j,i] = 1
+            
+[genotypes,abundances] = generate_2D_discrete_gauss(distr_grid, 
+                                min_fit_classes,mean_fit_class, cov_matrix, N)
+
+[distr_grid,dim1,dim2] = get_2D_distr(genotypes,abundances,box_dim)
+
+fig, ax = plt.subplots()
+
+image = np.ones(np.shape(distr_grid)) - distr_grid
+#image = np.asarray([[.1,0,1],[.1,0,1],[.1,0,1]])
+#image = np.asarray([[.1,.1,.1],[0,0,0],[1,1,1]])
+
+ax.imshow(image, cmap=plt.cm.gray, interpolation='nearest')
+ax.set_title('dropped spines')
+
+# Move left and bottom spines outward by 10 points
+ax.spines['left'].set_position(('outward', 10))
+ax.spines['bottom'].set_position(('outward', 10))
+# Hide the right and top spines
+ax.spines['right'].set_visible(False)
+ax.spines['top'].set_visible(False)
+# Only show ticks on the left and bottom spines
+ax.yaxis.set_ticks_position('left')
+ax.xaxis.set_ticks_position('bottom')
+
+plt.show()

### 2017-08-16 manuscript.tex
reorganized the pythong modules, and added updates for the manuscript.

@@
 A single trait's response to selection leads to a shift in the population mean towards the fittest phenotype.
-With multiple traits, the response involves both direct and indirect effects that together determine the evolution each \citep{lande1983measurement,Scarcelli23102007,Lovell2013,Wagner2011}.
-Indirect effects originate from genetic mechanisms that underlie the phenotypic values of traits, which may be associated or shared in a manner that enforces specific relationships between them.
-These associations can alter fundamental features of adaptation and have significant evolutionary consequences \citep{Felsenstein1979, Arnold2001, Arnold2008}.\par
+With multiple traits, the response involves both direct and indirect effects from selection that together determine the evolution each \citep{lande1983measurement,Scarcelli23102007,Lovell2013,Wagner2011}.
+Direct effects are those induced by selection favoring fitter phenotypic values in each trait.
+However, the traits themselves may be associated at the genetic level to an extent that only certain combinations of their phenotypic values are possible, and selection on any one of them will necessarily lead to changes in the others.
+For example, we can consider two traits that are controlled by a single pleiotropic locus, but with only the first under selection.
+Alleles frequencies at the locus change in response to selection against the first trait, and they also induce a shift in the mean of the second.
+These are indirect effects experienced by the second trait, which are caused by selection on the first.
+Associations such as these can alter fundamental features of adaptation and have significant evolutionary consequences \citep{Felsenstein1979, Arnold2001, Arnold2008}.\par
 % * <masel@email.arizona.edu> 2017-07-28T22:53:26.215Z:
@@
 % 
-% ^.
+% ^ <kgomez81@math.arizona.edu> 2017-08-07T14:10:25.416Z.
 
@@
 It states that the change in the population mean $\Delta \bar{z}$ equals the product of the selection gradient $\beta$ and what is known as the \textbf{G}-matrix, whose components are the additive genetic covariances between traits.
-Lande’s result has been used to extend much of our understanding of how phenotypes evolve under selection, yielding a rich set of ideas that emphasize the role of the \textbf{G}-matrix in directing adaptive evolution at both micro- and macro-evolutionary scales.
-This is one reason why the evolution of the \textbf{G}-matrix has been of key interest to evolutionary biologist.
-Furthermore, these ideas make it apparent that the evolution of traits cannot understood in isolation, and in fact must be considered alongside what we know about the dynamics of the \textbf{G}-matrix.\par
+Lande’s result has been used to extend much of our understanding of how phenotypes evolve under selection, yielding a rich set of ideas that emphasize the role of the \textbf{G}-matrix in directing adaptive evolution at both micro- and macro-evolutionary scales.\par
 % * <masel@email.arizona.edu> 2017-07-29T12:43:16.916Z:
@@
 % 
-% ^.
+% ^ <kgomez81@math.arizona.edu> 2017-08-07T14:10:30.176Z.
 
@@
 By symmetry we know that rate of adaptation in each trait alone must be $0.5 \hspace{.02 in} v(2U;N,s)$, and furthermore, the rate of adaptation in trait one is now less with the second trait.
-Clearly, we lack any information about how trait two is affecting trait one's evolution, but Lande's equation tells us that these effects must somehow translate into changes in the additive variance of trait one and its covariance with the second.\par
+Lande's result implies that trait one's rate of adaptation, when evolving with the second, must equal the sum of its variances and covariance ($v_1 =\sigma_1^2 +\sigma_{12}$).
+Clearly, adding the second trait produces a new covariance term that affects $v_1$, but it also induces a change in the variance of the first trait.
+What we do not know is the amount that each of these contribute to the observed reduction in trait one's rate of adaptation.\par
 % * <masel@email.arizona.edu> 2017-07-29T13:00:14.234Z:
@@
 To examine how adaptation in one trait is affected by evolution in another, we extend the work of Desai and Fisher's traveling wave model and consider a two dimensional trait space.
-We examine simplest case in which individuals have two independently evolving traits contributing to fitness.
+We examine the simplest case where individuals have two non-pleiotropic traits that contribute equally to fitness.
+We assume that beneficial mutations in each trait occur at identical rates, affect each trait independently, and exhibit no epistatic interactions between them.
+Lastly, we will focus on a population whose size remains fixed.
 Our aim is to explore the effects of genetic correlations arising from linkage disequilibrium contribute to the interactions that direct trait evolution.
 So, we assume that the two traits are controlled separate non-pleiotropic loci.
-Our exploration of the question will done using simulations, whose results we present after discussing key details of the model in the next section.
+Our exploration of the question will be done using simulations, whose results we present after discussing key details of the model in the next section.
 We will conclude with a discussion of the our findings and their implications.
@@
 Individuals that had accumulated $i$ mutations in the first and $j$ in the second were taken to have an overall fitness of $r_{ij} = i s_1+j s_2$.
-Their relative fitness was the difference $(r_{ij}-\bar{r})$, where  $\bar{r} $  is the population mean fitness.
+Their relative fitness is given by the difference $(r_{ij}-\bar{r})$, where  $\bar{r} $  is the population mean fitness.
 
@@
 Following the convention of Desai and Fisher, we refer to the set of deterministically behaving classes as the bulk, while those that grow stochastically and have a positive selective advantage ($r_{ij}>\bar{r}$) will be referred to as the stochastic front (see \textbf{fig.1}).\par
+
+\begin{figure}
+\includegraphics[width=1\linewidth]{fig1.pdf}
+\label{Figure 1.}
+\caption{Two dimensional fitness distribution.}
+\end{figure}
 
@@
 
-Classes in the stochastic front intermittently transition from stochastic to deterministic behavior and become part of the bulk.
+Classes in the stochastic front intermittently transition from stochastic to deterministic behavior and become part of the bulk once they have grown sufficiently large.
 Such classes are said to have established.
-Their growth can be modeled as a branching processes fed by incoming mutants generated in adjacent, less fit, classes within the bulk.
-The chances of a mutant lineage establishing is given by the probability of fixation.
-This was obtained using the expression below derived by \citep{Uecker2011} [eqt.
-16b].
+Prior to establishment, random fluctuations due to drift drive the dynamics of abundances.
+The behavior of these classes can be modeled as a branching process fed by incoming mutants generated in adjacent, less fit, classes within the bulk.
+Each mutant lineage that appears in the stochastic front either goes establishes or goes extinct.
+The probability of fixation measures the chances of the former outcome, and it depends on the selective advantage of the mutant.
+If its genotype is $(i,j)$, then its selective advantage is $s_{ij}=(r_{ij}-\bar{r})$.
+Standard mathematical results for branching processes can be used to solve for the lineage's probability of fixation which is given by equation 3.
+\begin{equation}
+\pi_{fix} = \frac{s_{ij}}{ 1+s_{ij}}
+\end{equation}
+The lineage establishes when it has reached a size if roughly $1/s_{ij}$, and its subsequent growth is exponential.
+\cite{desai2007beneficial} expressed the abundances of these lineages as $n_{ij}(t) = \frac{1}{s_{ij}} \exp[{s_{ij}(t-\tau)}]$.
+The random variable $\tau$ is the time of establishment and it captures the cumulative effects of the stochastic fluctuations.
+In their initial analysis, Desai and Fisher solve for the probability density function of $\tau$ for lineages destined to establish.
+In our model, we find it more useful to consider the related random variable $\nu = \frac{1}{s_{ij}}\exp[-\Delta r_{ij}\tau]$ representing an initial size that produces the deterministic trajectory observed after establishment.
+Using the expression given by Desai and Fisher for the PDF of $\tau$, we can arrive at the cumulative distribution of  $\nu$ (equation 5).
+\begin{equation}
+P(\nu \leq \nu_0) = 1- e^{-\pi_{fix} \nu_0}
+\end{equation}
+Each class in the stochastic front may consist of many mutant lineages.
+However, it is the the arrival time of the first mutant lineage destined to establish that determines which class will join the bulk next.
+Once the genotype of this mutant lineage is obtained, its initial size $\nu$ can be sampled.
+The class can then be regarded as belonging to the bulk at the determined arrival time with initial size $\nu$.
+\par
 % * <masel@email.arizona.edu> 2017-07-29T16:52:06.887Z:
@@
 % 
-% ^.
-\begin{equation}
-p_{fix} = 2^{-1}  \left(1 + \int_0^\infty (b_{ij} +d_{ij}) e^{-\int_0^{t'}(b_{ij}-d_{ij}) d t'} \right)
-\end{equation}
-The terms $b_{ij}=1+(r_{ij}-\bar{r})$ and $d_{ij}=1$ are the respective birth and death rates of the stochastically growing class.
-A mutant lineage that is destined to establish transitions into its deterministic regime of growth once it has roughly reached the size $n_{ij} = 1/p_{fix}$.
-Rather than attempting to model the timing of this transition, we instead focus on equivalently choosing the initial size $\nu$ that produces the same deterministic trajectory.
-The cumulative distribution of the random variable $\nu$, expressed below, is provided by \citep{Uecker2011} [eqt.
-40].
-\begin{equation}
-P(\nu \leq \nu_0)= 1-e^{-\nu_0/p_{fix}}
-\end{equation}
-We assumed that the next class to transition into the bulk would be the first in which  a mutant lineage appear that was destined to establish.\par
-
+% ^ <kgomez81@math.arizona.edu> 2017-08-04T23:43:18.217Z.
 In order to investigate how of traits affect one another, we identified the contributions of each trait to the evolution of the population.
-The rate of adaptation can obtained by differentiating the population's mean relative fitness $\bar{r}=\sum p_{ij} r_{ij} $.
-We combined these results with \textbf{equation 2} and arrived at the expression $\dot{\bar{r}}=\dot{\bar{r_1}} +\dot{\bar{r_2}}$.
-In our simple model, the population's rate of adaptation could therefore be expressed as the sum of the individual rates.
-We found that each of these rates could then be rewritten in terms of trait variances and covariance and obtained the system below.
+The rate of adaptation can obtained by differentiating the population's mean relative fitness $\bar{r}=\sum_{ij} p_{ij} r_{ij}$, which is itself the sum of the two marginal means, $\bar{r_1}=\sum_{ij} p_{ij} is_1$ and $\bar{r_2}=\sum_{ij} p_{ij} j s_2$.
+It follows that the total rate of adaptation is sum of rates of adaptation corresponding to each trait ($\dot{\bar{r}}=\dot{\bar{r_1}} +\dot{\bar{r_2}}$).
+We can apply Lande's result to the two components $\dot{\bar{r_1}} $ and $\dot{\bar{r_2}} $ and rewrite them in terms of the trait variances $\sigma_k^2$ (trait $k=1,2$) and their covariance $\sigma_{12}$.
+In doing so, we obtain the system expressed in equation 6.
 % * <masel@email.arizona.edu> 2017-07-29T17:21:52.605Z:
@@
 \end{equation}
-This is exactly Lande's equation with a selection gradient and trait response whose components have been normalized.
-
-In our model, covariance between traits is arises entirely from linkage disequilibria.
-We can assume that specific alleles at either locus correspond to the number of beneficial mutations that have accumulated in the corresponding trait.
-It follows that the $p_{ij}$ are the gametic frequencies for the genome consisting of allele $i$ paired with allele $j$, and $D_{ij} = p_{ij}-p_{i\cdot} p_{\cdot j}$ is their coefficient of linkage disequilibrium ( $p_{i\cdot}$ and $p_{\cdot j}$ are marginal distributions).
-By applying \textbf{equation 2}, we can derive a mathematical relationship between covariance and linkage disequilibria.
-% * <masel@email.arizona.edu> 2017-07-29T17:34:47.794Z:
-% 
-% > specific alleles at either locus correspond to the number of beneficial mutations that have accumulated in the corresponding trait
-% Multiple loci can contribute to the same trait.
-An allele cannot be a "number of beneficial mutations", because an allele must pertain to one locus.
-Stick to the word "genotype" not "allele", eg "genotype ij has i beneficial mutations in trait 1 and j beneficial mutations in trait 2 compared to the ancestral population".
-This means that you cannot calculate linkage disequilibrium as you do: linkage disequilibrium is generally defined as being between two particular beneficial mutations.
-What you calculated was simply a covariance.
-Because of this confusion, I got a bit lost here and so skipped Equation 6.
-% 
-% ^.
-\begin{equation}
-\sigma_{12}=\sum_{ij}D_{ij}[(i-\bar{i})s_1][(j-\bar{j})s_2]. 
-\end{equation} 
-As the population adapts, mutations and selection  increase and decrease the amounts LD which ultimately determine how traits interact in their evolution.\par
-
-We examined the resulting dynamics of trait evolution using numerical simulations of our model using Mathematica.
-We focused on modeling the symmetric case, in which the two traits had identical mutation rate and selection coefficients.
-These, along with the population size, were chosen to match values reported in evolutionary experiments involving relevant micro-organisms (see Table 1).
+The components of the selection gradient and trait response have been normalized in the expression above.
 
@@
 The integral was numerically approximated up to an appropriate cutoff.
-To determine if a mutation would establish, we sampled a uniform random variable and checked it against the calculated $p_{fix}$.
+To determine if a mutation would establish, we sampled a uniform random variable and checked it against the calculated $\pi_{fix}$.
 The next class to establish was chosen accordingly and incorporated into the bulk with a random initial size generated with from \textbf{equation 4}.\par
@@
 
-Results and Discussion.
+\subsection{Results}
+Our simulations were carried out for the symmetric case with parameters $s$ and $U$ being identical for both traits, but in collecting our data, we regarded trait one as the focal trait.
+The population parameters $N$, $s$, and $U$ were chosen using the results \citep{desai2007beneficial} for fitness variance.
+Specifically, we ensured that the two traits would have a sufficient amount of genetic diversity to study how their evolution is affected by interactions with the other trait ($\sigma_k^2(N,s_k,U_k) > s_k^2$ ).
+We also focused on parameter sets that were in agreement with reported values of $N$, $s$, and $U$ in evolutionary experiments involving model asexual micro-organisms, or .
+We measured the fitness variance of trait one over time, as well as its covariance, and then calculated the contribution of each to the reduction in trait one's rate of adaptation.
 
+In the one dimensional traveling wave model, $N$, $s$, and $U$  prescribe a specific
+
+We also selected parameters that coincided with in evolutionary experiments for micro-organisms that typically are regarded to be asexual, or have very limited amounts of sex.
+In table 1 we summarize these ranges for various micro-organisms of interest, these
+
+\begin{table*}[ht]
+\centering
+\begin{tabular}{ | c | c | c | c |}
+\hline Micro-organism& Population Size ($N$)	& Selection Coefficient ($s$) & Beneficial Mutation Rate ($U$) \\
+\hline \begin{tabular}{c} \textit{E}. coli \\ \citep{Perfeito2007} \end{tabular} & $10^6 - 10^{14}$ & $0.5 \% - 2 \%$  & $10^{-9} - 10^{-5}$ \\
+\hline \begin{tabular}{c} \textit{S}. cerevisiae \\ \citep{desai2007speed,Levy2015} \end{tabular} & $10^7 - 10^{12}$ & $0.5 \%- 3 \%$  & $10^{-5} - 10^{-4}$ \\
+\hline
+\end{tabular}
+\caption{Population parameters measured in evolution experiments for key microorganisms. Beneficial mutations rates are given in units of mutations per genome per generation.} 
+\label{Table 1}
+\end{table*}
+
+\begin{figure}
+\includegraphics[width=1\linewidth]{fig2a_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
+\label{Figure 2.}
+\caption{Trait 1 variance and covariance.}
+\end{figure}
+Figure 1 also demonstrates that the covariance over time fluctuates
+
+\subsection{Discussion}
+
+In our model, covariance between traits is arises entirely from linkage disequilibrium.
+We can assume that specific alleles at either locus correspond to the number of beneficial mutations that have accumulated in the corresponding trait.
+It follows that the $p_{ij}$ are the gametic frequencies for the genome consisting of allele $i$ paired with allele $j$, and $D_{ij} = p_{ij}-p_{i\cdot} p_{\cdot j}$ is their coefficient of linkage disequilibrium ( $p_{i\cdot}$ and $p_{\cdot j}$ are marginal distributions).
+By applying \textbf{equation 2}, we can derive a mathematical relationship between covariance and linkage disequilibrium.
+% * <masel@email.arizona.edu> 2017-07-29T17:34:47.794Z:
+% 
+% > specific alleles at either locus correspond to the number of beneficial mutations that have accumulated in the corresponding trait
+% Multiple loci can contribute to the same trait.
+An allele cannot be a "number of beneficial mutations", because an allele must pertain to one locus.
+Stick to the word "genotype" not "allele", eg "genotype ij has i beneficial mutations in trait 1 and j beneficial mutations in trait 2 compared to the ancestral population".
+This means that you cannot calculate linkage disequilibrium as you do: linkage disequilibrium is generally defined as being between two particular beneficial mutations.
+What you calculated was simply a covariance.
+Because of this confusion, I got a bit lost here and so skipped Equation 6.
+% 
+% ^.
+\begin{equation}
+\sigma_{12}=\sum_{ij}D_{ij}[(i-\bar{i})s_1][(j-\bar{j})s_2]. 
+\end{equation} 
+As the population adapts, mutations and selection  increase and decrease the amounts LD which ultimately determine how traits interact in their evolution.\par
+
+We examined the resulting dynamics of trait evolution using numerical simulations of our model using Mathematica.
+We focused on modeling the symmetric case, in which the two traits had identical mutation rate and selection coefficients.
+These, along with the population size, were chosen to match values reported in evolutionary experiments involving relevant micro-organisms (see Table 1).
 % \section{In-text Citations}

### 2017-08-16 manuscriptoutline.tex
reorganized the pythong modules, and added updates for the manuscript.

@@
 
-\begin{table}[h]
+\begin{table*}[b]
 	\centering
@@
 	\end{tabular}
-	\caption{Population parameters measured in evolution experiments for key microorganisms.} 
+	\caption{Population parameters measured in evolution experiments for key microorganisms. Beneficial mutation rates are given.} 
 	\label{Table 1}
@@
 \section*{Results}
+What are my questions?
+\begin{enumerate}
+\item What are the changes to variance trait one and its covariance with trait two? 
+\item Which of them is responsible for the reduction of adaptation and to what degree?
+\item How does variance of the trait and its covariance behave over time?
+\end{enumerate}
+
 \begin{itemize}

### 2017-08-16 plotdata.py

@@ -5,29 +5,37 @@ Created on Mon Jul 17 13:27:18 2017
 Script to process data generated by simulations in Mathematica code.
 """
 
-# import packages for script
+# import packages needed for script
 import pickle
 import scipy as sp
 import numpy as np
+import copy as cpy
 
 # set parameters of simulation and create required variables
 [N,s1,s2,U1,U2] = [1e9,1e-2,1e-2,1e-5,1e-5];
 
+# calculate desai and fisher (2007) theoretical "v" and "tau_est"
+vU_thry = s1*s1*(2*np.log(N*s1)-np.log(s1/(1*U1)))/((np.log(s1/(1*U1)))**2)
+v2U_thry = 0.5*s1*s1*(2*np.log(N*s1)-np.log(s1/(2*U1)))/((np.log(s1/(2*U1)))**2)
+tau_est = 0.5*s1/v2U_thry
+
 # section of code for processing new data from Mathematica simulations
 data_name = '_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp2'
+#folder_location = 'Documents/kgrel2d/'  # use this location in linux
+folder_location = ''     # use this location if windows
 [times,genotypes,abundances] = [[],[],[]]
 
 # get simulation data and store genotypes as lists since they vary in dimensions over time
-data_file=open('./Documents/kgrel2d/data/pythondata/times'+data_name+'.dat')
+data_file=open('./'+folder_location+'data/pythondata/times'+data_name+'.dat')
 times = data_file.read().splitlines()
 times = np.array(map(float,times))
 data_file.close()
 
-data_file=open('./Documents/kgrel2d/data/pythondata/genotypes'+data_name+'.dat')
+data_file=open('./'+folder_location+'data/pythondata/genotypes'+data_name+'.dat')
 genotypes = data_file.read().splitlines()
 data_file.close()
 
-data_file=open('./Documents/kgrel2d/data/pythondata/abundances'+data_name+'.dat')
+data_file=open('./'+folder_location+'data/pythondata/abundances'+data_name+'.dat')
 abundances = data_file.read().splitlines()
 data_file.close()
 
@@ -44,20 +52,24 @@ for i in range(num_pts):
     exec(abundances[i])
 
[…]

### 2017-08-16 plotfunctions.py

@@ -5,10 +5,15 @@ Created on Tue Aug 08 18:08:03 2017
 Library of functions used in plots.py and plotdata.py
 """
 
-# import packages for script
+
+#--------FUNCTIONS REQUIRE PACKAGES LISTED:------------------------------------
+from mpl_toolkits.mplot3d import Axes3D
+from scipy.stats import multivariate_normal
+import pickle
+import matplotlib.pyplot as plt
 import scipy as sp
 import numpy as np
-from scipy.stats import multivariate_normal
+# -----------------------------------------------------------------------------
 
 # -----------------------------------------------------------------------------
 def get_sample_window(times,start_time,end_time):
@@ -29,7 +34,7 @@ def generate_2D_discrete_gauss(distr_grid, min_fit,mean_fit, cov_matrix, N):
 # distr_grid =  square array with 1's where the subpop is nonzero
 # mean_fit_class = (i,j) in the array that will be the mean of the gaussian
 # returns: two lists genotypes & abundances corresponding to desired distribution
-
+    
     [genotypes,abundances] = [[],[]]    
     
     for i in range(len(distr_grid[:,0])):
@@ -60,5 +65,110 @@ def get_2D_distr(genotypes,abundances,box_dim):
         indx1 = genotypes[i][0] - dim1_data[1] + box_dim[0][1]
         indx2 = genotypes[i][1] - dim2_data[1] + box_dim[1][1]
         my_distr[indx1,indx2] = abundances[i]/tot_pop_size
-
+        
     return [np.flipud(my_distr.transpose()),dim1_data,dim2_data]
+
+# -----------------------------------------------------------------------------
+def generate_figure(figNum,data_name,folder_location):
+# figure 1: representation of two-dimensional distribution  (no data required)
+        
+    if(figNum == 1):
+        
+        # set up distribution data for the figure
+        [min_fit_clss,mean_fit_clss, cov_mtrx, box_dim] = [[0,0],[7,7],[[1,0],[0,1]],[[20,7],[20,7]]]
+        [cut_off, distr_grid] = [1e-4,np.zeros([box_dim[0][0],box_dim[1][0]])]
+        [arry_dim1,arry_dim2]=[np.shape(distr_grid[:,0])[0],np.shape(distr_grid[0,:])[0]]
+        
+        # set up non-zero classes on array
[…]

### 2017-08-16 plots.py

@@ -5,124 +5,16 @@ Created on Mon Jan 12 13:27:18 2017
 Script for plots used in "Modeling How Evolution in One Trait is Affected by Adapatation in Another"
 """
 
-# import packages needed for script
-from mpl_toolkits.mplot3d import Axes3D
-from scipy.stats import multivariate_normal
-import pickle
+#---------------CHANGE WORKING DIRECTORY FIRST---------------------------------
+#cd C://Users/dirge/Documents/kgrel2d/
 
-import matplotlib.pyplot as plt
-import scipy as sp
-import numpy as np
-
-# set parameters of simulation and create required variables
-[N,s1,s2,U1,U2] = [1e9,1e-2,1e-2,1e-5,1e-5];
-
-# calculate desai and fisher (2007) theoretical "v" and "tau_est"
-vU_thry = s1*s1*(2*np.log(N*s1)-np.log(s1/(1*U1)))/((np.log(s1/(1*U1)))**2)
-v2U_thry = 0.5*s1*s1*(2*np.log(N*s1)-np.log(s1/(2*U1)))/((np.log(s1/(2*U1)))**2)
-tau_est = 0.5*s1/v2U_thry
+# -----------------------------------------------------------------------------
+import plotfunctions as pltfun
 
 # load data stored in file generated by plotdata
-data_name = '_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1'
+data_name = '_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp2'
+#folder_location = 'Documents/kgrel2d/'
+folder_location = ''
 
-#------------------------------------------------------------------------------
-#------------------------------------------------------------------------------
-#------------------------------------------------------------------------------
-# figure 1: representation of two-dimensional distribution  (no data required)
-
-# set up distribution data for the figure
-[min_fit_clss,mean_fit_clss, cov_mtrx, box_dim] = [[0,0],[7,7],[[1,0],[0,1]],[[20,7],[20,7]]]
-[cut_off, distr_grid] = [1e-4,np.zeros([box_dim[0][0],box_dim[1][0]])]
-[arry_dim1,arry_dim2]=[np.shape(distr_grid[:,0])[0],np.shape(distr_grid[0,:])[0]]
-
-# set up non-zero classes on array
[…]

### 2017-08-17 plotdata.py
"more updates"

@@ -20,7 +20,7 @@ v2U_thry = 0.5*s1*s1*(2*np.log(N*s1)-np.log(s1/(2*U1)))/((np.log(s1/(2*U1)))**2)
 tau_est = 0.5*s1/v2U_thry
 
 # section of code for processing new data from Mathematica simulations
-data_name = '_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp2'
+data_name = '_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1'
 #folder_location = 'Documents/kgrel2d/'  # use this location in linux
 folder_location = ''     # use this location if windows
 [times,genotypes,abundances] = [[],[],[]]

### 2017-08-17 plotfunctions.py

@@ -50,29 +50,28 @@ def generate_2D_discrete_gauss(distr_grid, min_fit,mean_fit, cov_matrix, N):
 # -----------------------------------------------------------------------------
 def get_2D_distr(genotypes,abundances,box_dim):
 # box_dim = gives array data to bound distr correponding to genotypes & abund.
+#           [[width1,margin1],[width2,margin2]]
 # returns: an array whose elements are the abundances of the fit classes
 
     tot_pop_size = sum(abundances)
     
-    dim1_data = [np.min([genotypes[i][0] for i in range(len(abundances))]),
-                 np.max([genotypes[i][0] for i in range(len(abundances))])]
-    dim2_data = [np.min([genotypes[i][1] for i in range(len(abundances))]),
-                 np.max([genotypes[i][1] for i in range(len(abundances))])]
+    dim1_data = [np.min(genotypes[:,0]), np.max(genotypes[:,0])]
                  
-    my_distr = np.zeros([box_dim[0][0],box_dim[1][0]])
+    dim2_data = [np.min(genotypes[:,1]), np.max(genotypes[:,1])]
+                 
+    my_distr = np.zeros([np.max([box_dim[0][0],box_dim[1][0]]),np.max([box_dim[0][0],box_dim[1][0]])])
         
-    for i in range(len(abundances)):
-        indx1 = genotypes[i][0] - dim1_data[1] + box_dim[0][1]
-        indx2 = genotypes[i][1] - dim2_data[1] + box_dim[1][1]
-        my_distr[indx1,indx2] = abundances[i]/tot_pop_size
+    for i in range(len(genotypes)):
+        indx1 = genotypes[i][0] - dim1_data[0] + box_dim[0][1]
+        indx2 = genotypes[i][1] - dim2_data[0] + box_dim[1][1]
+        my_distr[indx1,indx2] = abundances[0][i]/tot_pop_size
         
-    return [np.flipud(my_distr.transpose()),dim1_data,dim2_data]
+    return [my_distr,dim1_data,dim2_data]
 
 # -----------------------------------------------------------------------------
 def generate_figure(figNum,data_name,folder_location):
 # figure 1: representation of two-dimensional distribution  (no data required)
-        
-    if(figNum == 1):
+        if(figNum == 1):
         
[…]

### 2017-08-17 plotdata.py
updates to figures python code.

@@ -12,7 +12,7 @@ import numpy as np
 import copy as cpy
 
 # set parameters of simulation and create required variables
-[N,s1,s2,U1,U2] = [1e9,1e-2,1e-2,1e-5,1e-5];
+[N,s1,s2,U1,U2] = [1e9,1e-2,1e-2,1e-5,1e-5]
 
 # calculate desai and fisher (2007) theoretical "v" and "tau_est"
 vU_thry = s1*s1*(2*np.log(N*s1)-np.log(s1/(1*U1)))/((np.log(s1/(1*U1)))**2)

### 2017-08-17 plotfunctions.py

@@ -9,6 +9,7 @@ Library of functions used in plots.py and plotdata.py
 #--------FUNCTIONS REQUIRE PACKAGES LISTED:------------------------------------
 from mpl_toolkits.mplot3d import Axes3D
 from scipy.stats import multivariate_normal
+from numpy import inf
 import pickle
 import matplotlib.pyplot as plt
 import scipy as sp
@@ -53,13 +54,14 @@ def get_2D_distr(genotypes,abundances,box_dim):
 #           [[width1,margin1],[width2,margin2]]
 # returns: an array whose elements are the abundances of the fit classes
 
-    tot_pop_size = sum(abundances)
+    tot_pop_size = sum(abundances[0])  # be careful with your sums of arrays!!!
     
     dim1_data = [np.min(genotypes[:,0]), np.max(genotypes[:,0])]
                  
     dim2_data = [np.min(genotypes[:,1]), np.max(genotypes[:,1])]
                  
-    my_distr = np.zeros([np.max([box_dim[0][0],box_dim[1][0]]),np.max([box_dim[0][0],box_dim[1][0]])])
+    my_distr = np.zeros([np.max([box_dim[0][0],box_dim[1][0]]),
+                         np.max([box_dim[0][0],box_dim[1][0]])])
         
     for i in range(len(genotypes)):
         indx1 = genotypes[i][0] - dim1_data[0] + box_dim[0][1]
@@ -69,9 +71,11 @@ def get_2D_distr(genotypes,abundances,box_dim):
     return [my_distr,dim1_data,dim2_data]
 
 # -----------------------------------------------------------------------------
-def generate_figure(figNum,data_name,folder_location):
+def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param,fname):
 # figure 1: representation of two-dimensional distribution  (no data required)
-        if(figNum == 1):
+    [N,s1,s2,U1,U2] = pop_param
+    
+    if(figNum == 1):
         
         # set up distribution data for the figure
         [min_fit_clss,mean_fit_clss, cov_mtrx, box_dim] = [[0,0],[7,7],[[1,0],[0,1]],[[20,7],[20,7]]]
@@ -104,11 +108,10 @@ def generate_figure(figNum,data_name,folder_location):
         ax1.yaxis.set_ticks_position('left')
         ax1.xaxis.set_ticks_position('bottom')
         
-        plt.show()
[…]

### 2017-08-17 plots.py

@@ -12,9 +12,15 @@ Script for plots used in "Modeling How Evolution in One Trait is Affected by Ada
 import plotfunctions as pltfun
 
 # load data stored in file generated by plotdata
-data_name = '_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp2'
+data_name = '_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1'
 #folder_location = 'Documents/kgrel2d/'
 folder_location = ''
 
-# generate figure 1. (parameters N=1e9, s1=s2=1e-2, U1=U2=1e-5)
-pltfun.generate_figure(2,data_name,folder_location)
+pop_param = [1e9,1e-2,1e-2,1e-5,1e-5]
+[sim_start,sim_end,snapshot] = [1e4,2e4,1.313e4]
+
+# generate figure 2. (parameters N=1e9, s1=s2=1e-2, U1=U2=1e-5)
+pltfun.generate_figure(2,data_name,folder_location,snapshot,snapshot,pop_param,'fig2')
+
+# generate figure 3. (parameters N=1e9, s1=s2=1e-2, U1=U2=1e-5)
+#pltfun.generate_figure(3,data_name,folder_location,sim_start,sim_end,pop_param,'fig3')

### 2017-08-17 plotfunctions.py
need to add mean fitness line

@@ -143,24 +143,21 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
                     
         # plot figure 1 with general with constructed discrete gaussian
         fig2, ax2 = plt.subplots(1,1,figsize=[10,8])
-#        fig2.subplots_adjust(bottom=0.25,left=0.25)
 
-        fit_distr_2d = ax2.pcolor(distr_grid,cmap=plt.cm.gray_r)
-
+        fit_distr_2d = ax2.pcolormesh(distr_grid,cmap=plt.cm.gray_r)
         cbar = plt.colorbar(fit_distr_2d)
-        ax2.axis('tight')
-        
+        ax2.axis('tight')        
         ax2.set_xticks(np.arange(distr_grid.shape[1])+0.5)
-        ax2.set_yticks(np.arange(distr_grid.shape[0])+0.5)
-        
+        ax2.set_yticks(np.arange(distr_grid.shape[0])+0.5)        
         ax2.set_xticklabels(class_xlabels[1:],rotation=90)
-        ax2.set_yticklabels(class_ylabels[1:])
-        
+        ax2.set_yticklabels(class_ylabels[1:])        
         ax2.set_xlabel('Beneficial Mutaitons Trait 1',fontsize=18,labelpad=20)
         ax2.set_ylabel('Beneficial Mutaitons Trait 2',fontsize=18,labelpad=10)
-        ax2.tick_params(axis='both',labelsize=14)
-        
+        ax2.tick_params(axis='both',labelsize=14)        
         cbar.ax.text(2.5,0.65,'Log$_{10}$ of Abundances',rotation=90,fontsize=18)
+
+#        ax2.scatter([45],[33])
+#        fit_line = plt.plot([i+40 for i in range(13)],[-i+38 for i in range(13)],ls="--", c=".3")
 
         fig2.savefig('./'+folder_location+'figures/'+fname+data_name+'.pdf')
         

### 2017-09-13 manuscript.tex
Added section for dynamics of G

@@
-\documentclass[9pt,twocolumn,twoside]{gsajnl}
-% Use the documentclass option 'lineno' to view line numbers
+\documentclass[11pt,twocolumn]{article} 
+\usepackage{natbib}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{graphicx}
+\usepackage{caption}
+\usepackage{subcaption}
+\usepackage[margin=.5in]{geometry}
 
-\articletype{inv} % article type
-% {inv} Investigation 
-% {gs} Genomic Selection
-% {goi} Genetics of Immunity 
-% {gos} Genetics of Sex 
-% {mp} Multiparental Populations
+\graphicspath{{figures/}}
 
 \title{Modeling How Evolution in One Trait is Affected by Adaptation in Another}
+\date{Fall 2016}
+\author{Kevin Gomez,Jason Bertram,Joanna Masel}
 
-\author[$\ast$,1]{Kevin Gomez}
-\author[$\dagger$]{Jason Bertram}
-\author[$\dagger$]{Joanna Masel}
-
-\affil[$\ast$]{Department of Applied Mathematics, University of Arizona, and}
-\affil[$\dagger$]{Department of Evolution and Ecology, Univeristy of Arizona}
-
-\keywords{Keyword; Keyword2; Keyword3; ...}
-
-\runningtitle{GENETICS Journal Template on Overleaf} % For use in the footer
-
-%% For the footnote.
-%% Give the last name of the first author if only one author;
-% \runningauthor{FirstAuthorLastname}
-%% last names of both authors if there are two authors;
-% \runningauthor{FirstAuthorLastname and SecondAuthorLastname}
-%% last name of the first author followed by et al, if more than two authors.
-\runningauthor{Gomez \textit{et al.}}
+\begin{document}
+\maketitle
+\newpage
 
@@
 
-\setboolean{displaycopyright}{true}
-
-\begin{document}
-
-\maketitle
-\thispagestyle{firststyle}
-\marginmark
-\firstpagefootnote
-\correspondingauthoraffiliation{Please insert the affiliation correspondence address and email for the corresponding author.
-The corresponding author should be marked with a `1' in the author list, as shown in the example.}
-\vspace{-11pt}%
-
 \section{Introduction}
@@
 A single trait's response to selection leads to a shift in the population mean towards the fittest phenotype.
-With multiple traits, the response involves both direct and indirect effects from selection that together determine the evolution each \citep{lande1983measurement,Scarcelli23102007,Lovell2013,Wagner2011}.
-Direct effects are those induced by selection favoring fitter phenotypic values in each trait.
-However, the traits themselves may be associated at the genetic level to an extent that only certain combinations of their phenotypic values are possible, and selection on any one of them will necessarily lead to changes in the others.
+With multiple traits the response involves both direct and indirect effects that together determine the evolution each trait \citep{lande1983measurement,Lovell2013,Wagner2011}.
+Direct effects are those induced by selection favoring fitter phenotypic values in each of the traits.
+However, traits may be associated at the genetic level in a manner that permits only certain combinations of their phenotypic values.
+As a result, selection on any one trait produces to changes in the others due to the constraints on their possible combinations.
 For example, we can consider two traits that are controlled by a single pleiotropic locus, but with only the first under selection.
-Alleles frequencies at the locus change in response to selection against the first trait, and they also induce a shift in the mean of the second.
+Allele frequencies at the locus change in response to selection against the first trait, and they also induce a shift in the mean of the second.
 These are indirect effects experienced by the second trait, which are caused by selection on the first.
 Associations such as these can alter fundamental features of adaptation and have significant evolutionary consequences \citep{Felsenstein1979, Arnold2001, Arnold2008}.\par
+% * <masel@email.arizona.edu> 2017-08-22T16:40:07.527Z:
+% 
+% > permits only certain combinations of their phenotypic values
+% The rest of this paragraph seems to be about correlated traits, but here you talk about hard constraints.
+The two are not equivalent, indeed you can have either one without the other.
+I'm not totally sure what you want to achieve in this paragraph, other than giving a non-mathematical version of the next paragraph on Lande's equation.
+If so, I wonder if it is better to merge the two paragraphs, while removing the language about absolute constraints.
+% 
+% ^ <kgomez81@math.arizona.edu> 2017-08-22T20:17:12.979Z:
+% 
+% 
+% 
+% ^.
 % * <masel@email.arizona.edu> 2017-07-28T22:53:26.215Z:
@@
 % 
-% ^.
+% ^ <kgomez81@math.arizona.edu> 2017-08-22T14:05:36.066Z.
 % * <masel@email.arizona.edu> 2017-07-28T22:52:47.005Z:
@@
 We know far less about the about the behavior genetic correlations from linkage disequilibria over time, but this is not much of an issue for most complex organisms.\par
+% * <masel@email.arizona.edu> 2017-08-22T16:50:23.214Z:
+% 
+% >  most models for the evolution of the \textbf{G}-matrix assume genetic correlations are pleiotropic and proceed with examining how evolutionary processes influence them
+% An important exception to note is the "Bulmer effect", which I believe describes how selection can cause LD even in the absence of linkage, which then of course appears in G.
+This is distinct from our work where linkage to the background on which a mutation first appear is the cause of LD, albeit later shaped by selection.
+I believe there is some material on the Bulmer effect in Bruce Walsh's book.
+It is important to review it just enough to distinguish what you are doing from what Bulmer already did.
+% 
+% ^.
+% * <masel@email.arizona.edu> 2017-08-22T16:47:43.713Z:
+% 
+% > the product of trans acting elements
+% This is not self-explanatory, and insofar as I can guess what it means, it seems to me to be a special case of pleiotropy rather than a third distinct possibility.
+% 
+% ^.
 
@@
 Consequently, we can learn a substantial amount about the evolution of their traits by examining how genetic correlations from LD evolve over time and influence adaptation.
-Do so would not only reveal a great deal about the oldest and most abundant forms of life on the planet, such as bacteria, but could prove useful to our use of them in evolutionary experiments.\par
+Doing so would not only reveal a great deal about some of the oldest and most abundant forms of life on the planet, such as bacteria, but could prove useful to future evolutionary experiments.\par
+
+Lande’s equation describes the phenotypic evolution of populations with multiple characters from generation to generation.
+Whether one can use this framework to uncover the history of natural selection in populations over large evolutionary timescales largely depends on the forces that influence the evolution of the G-matrix.
+Over the past few decades there has been a significant amount of effort to shed light on the matter of G-matrix stability.
+The analysis of genetic models for the adaptive evolution of multiple phenotypic traits indicate some of the conditions needed for stability\citep{Turelli1988,Jones2003}.
+These works exclude the potential influences of linkage disequilibrium on the dynamics of variances and covariances over time, and as a result, our understanding of the role of linkage disequilibria in G-matrix evolution remains unclear.
+Nonetheless, the question is significant there is empirical evidence suggesting that in the G-matrix is often conserved in natural populations \citep{Arnold1999,Roff2000,Steppan2002}.
+However, much of what we know about G-matrix stability corresponds the examining evolutionary forces on
 
 Our best models for adaptation in asexuals make use of what are known as traveling waves, and are based on population genetics.
-They have been used to establish important relationships between the population's parameter and key features of evolution in asexuals.
-Among these is the model developed by \cite{desai2007beneficial}, which they used to derive an expression for the rate of adaptation, given by $v(N,s,U) = 2s^2 \ln(N s)/\ln^2(s/U)$.
-Here, $N$ is the population size, $s$ is the effective selective advantage of a beneficial mutation, and $U$ is the beneficial mutation rate.
-This is an important result, because it directly provides the additive variance in fitness by applying Fisher's fundamental theorem ($v=\sigma^2$).
-If we regard the population as having one adaptive trait, then this also equals the trait's additive variance and realize that we have said something more about the quantitative genetics involved.\par
+These models have established important relationships between the population's parameter and key features of evolution in asexuals.
+Among these is the traveling wave model developed by \cite{desai2007beneficial} for a population whose size $N$ is fixed in which only beneficial mutations occur at the rate $U$.
+They also make the key assumption that all mutations all provide the same selective advantage $s$, which allows them to show that the population's rate of adaptation is given by  $v(N,s,U) = 2s^2 \ln(N s)/\ln^2(s/U)$.
+This is an important result, because it directly provides the additive variance in fitness by simply applying Fisher's fundamental theorem ($v=\sigma^2$).
+If we regard the population as having one adaptive trait, then $\sigma^2$ is in fact the trait's additive variance and we obtain a quantitative genetics result as well.\par
+% * <masel@email.arizona.edu> 2017-08-22T17:28:18.736Z:
+% 
+% > the model developed by \cite{desai2007beneficial}
+% This is the place to briefly summarize the core assumptions of the model.
+You have already given the assumption of strict asexuality, the others are constant N, absence of deleterious mutations, and lack of variation in s among beneficial mutants.
+The latter (combined with asexuality) is the key innovation that makes the model work, because it relieves the need to track the identity of individual mutations.
+% 
+% ^ <kgomez81@math.arizona.edu> 2017-08-27T02:56:42.167Z:
+% 
+% I've added a statement about Desai and Fisher's assumptions following the citation, and I've also slightly changed the sentences after.
+%
+% ^.
 
@@
 Clearly, adding the second trait produces a new covariance term that affects $v_1$, but it also induces a change in the variance of the first trait.
-What we do not know is the amount that each of these contribute to the observed reduction in trait one's rate of adaptation.\par
+What we do not know is the amount that each of these contributes to the observed reduction in trait one's rate of adaptation.
+This is exactly what we aim to determine.
+\par
 % * <masel@email.arizona.edu> 2017-07-29T13:00:14.234Z:
@@
 % 
-% ^.
+% ^ <kgomez81@math.arizona.edu> 2017-08-22T14:11:45.747Z.
 
-To examine how adaptation in one trait is affected by evolution in another, we extend the work of Desai and Fisher's traveling wave model and consider a two dimensional trait space.
-We examine the simplest case where individuals have two non-pleiotropic traits that contribute equally to fitness.
-We assume that beneficial mutations in each trait occur at identical rates, affect each trait independently, and exhibit no epistatic interactions between them.
+To explore how adaptation in one trait is affected by evolution in another, we extend the work of Desai and Fisher's traveling wave model and consider a two dimensional trait space.
+We examine the simplest case where individuals have two non-pleiotropic traits that contribute equally to fitness, each controlled by separate loci.
+We assume that beneficial mutations occur with identical rates in each trait, affect each independently, and exhibit no epistatic interactions.
 Lastly, we will focus on a population whose size remains fixed.
-Our aim is to explore the effects of genetic correlations arising from linkage disequilibrium contribute to the interactions that direct trait evolution.
-So, we assume that the two traits are controlled separate non-pleiotropic loci.
-Our exploration of the question will be done using simulations, whose results we present after discussing key details of the model in the next section.
-We will conclude with a discussion of the our findings and their implications.
-% * <masel@email.arizona.edu> 2017-07-29T13:01:43.541Z:
-% 
-% > two independently evolving traits
-% They are not evolving independently. What you mean instead is that there is no pleiotropy. You might also summarize other key simplifications here, eg constant N, no epistasis etc.
-% 
-% ^.
+Our aim is to examine the effects of genetic correlations arising solely from linkage disequilibrium and determine how they influence trait evolution.
 
@@
 Each trait was assumed to be under the control of a single locus.
-We assumed that beneficial mutations improving each trait appeared at a rate of $U_k$ ($k=1,2$) and increase th fitness by $s_k$; deleterious mutations were excluded.
-Individuals that had accumulated $i$ mutations in the first and $j$ in the second were taken to have an overall fitness of $r_{ij} = i s_1+j s_2$.
+We assumed that beneficial mutations improving each trait appeared at a rate of $U$ and increase th fitness by $s$; deleterious mutations were excluded.
+Individuals that had accumulated $i$ mutations in the first and $j$ in the second were taken to have an overall fitness of $r_{ij} = i s+j s$.
 Their relative fitness is given by the difference $(r_{ij}-\bar{r})$, where  $\bar{r} $  is the population mean fitness.
+We will often right $r_i$ in place of $i s$ and $r_j$ in place of $j s$, so that $r_{ij} = r_i + r_j$.
+% * <masel@email.arizona.edu> 2017-08-22T19:12:14.027Z:
+% 
+% > appeared at a rate of $U_k$ ($k=1,2$) and increase th fitness by $s_k$
+% Since you are going to make U and s independent of k, it's easier on the reader if you do that from the outset, rather than set up the general case and then  collapse it.
+% 
+% ^ <kgomez81@math.arizona.edu> 2017-08-23T16:45:52.343Z:
+% 
+% I will change the notation the instances of U_k and s_k to just U and s here and in the rest of the document. 
+%
+% ^.
 
-In analogous approach to that of \cite{desai2007beneficial}, we divided our population into classes bases on the number of beneficial mutations in each trait.
+In an analogous approach to that of \cite{desai2007beneficial}, we divided our population into classes based on the number of beneficial mutations in each trait.
 Their abundances are accordingly denoted $n_{ij}$ and their frequencies by $p_{ij}$.
 We assumed that $1/N \ll s \ll 1$ to ensure that the set of classes could be neatly separated into two groups based on whether their behavior was deterministic and stochastic.
-Following the convention of Desai and Fisher, we refer to the set of deterministically behaving classes as the bulk, while those that grow stochastically and have a positive selective advantage ($r_{ij}>\bar{r}$) will be referred to as the stochastic front (see \textbf{fig.1}).\par
+Following the convention of Desai and Fisher, we refer to the set of deterministically behaving classes as the bulk, while those that grow stochastically and have a positive selective advantage ($r_{ij}>\bar{r}$) will be referred to as the stochastic front.
+Unlike in the case of Desai and Fisher's  traveling wave whose stochastic front is a single point, our stochastic front is the boundary of the bulk's distribution above the line crossing the population's mean fitness.
+This is shown in figure 1, where classes in the bulk are shown in shades of blue, while those belonging to the stochastic front are shown in shades of red.
+One reason why the growth of these two groups is treated differently has to do with the fact that only beneficial mutations in the stochastic front are consequential to adaptation.
+In asexuals, a large percentage of beneficial mutations are wasted since most occur poor genetic backgrounds, and their lineages remain relatively insignificant size and as a result, they do not contribute to the population's adaptive evolution.
+These are the beneficial mutations that occur within the bulk.
+On the other hand, mutations that do manage to occur on good genetic backgrounds will eventually grow large enough in size to influence the population's mean relative fitness over time.
+These are exactly the beneficial mutations produced at the edge of the bulk that give rise to classes in the stochastic front, marked by arrows in figure 1.
+Thus, the bulk's distribution determines the position and behavior of the stochastic front, but in time, the latter also shapes how the former shifts forward (fix this!).\par
+% * <masel@email.arizona.edu> 2017-08-22T19:17:50.152Z:
+% 
+% > those that grow stochastically and have a positive selective advantage ($r_{ij}>\bar{r}$) will be referred to as the stochastic front
+% This is a good place to point a key difference between you and prior work, namely that your front is a curve not a point.
+% 
+% ^ <kgomez81@math.arizona.edu> 2017-08-23T17:03:48.195Z:
+% 
+% I added a bit more after this sentence, and changed the subsequent one.
+% "Unlike in the case of Desai and Fisher's  traveling wave whose stochastic front is a single point, our stochastic front is the boundary of the bulk's distribution above the line crossing the population's mean fitness.
+This is shown in figure 1, where classes in the bulk are shown in shades of blue, while those belonging to the stochastic front are shown in shades of red.
+"
+%
+% ^.
 
@@
 \label{Figure 1.}
-\caption{Two dimensional fitness distribution.}
+% * <masel@email.arizona.edu> 2017-08-22T22:26:35.280Z:
+% 
+% > Figure 1.
+% Do you think it is possible to merge Figures 1 and 2 into a single figure? Ie, to take the features you illustrate in Figure 1, put them into Figure 2, and then cut Figure 1 out?
+% 
+% ^.
+\caption{\footnotesize (reproduce this for the actual data and cut out figure 2) Reproduction of two dimensional distribution over the trait space from simulation.
+Each box is a class consisting of individuals with an identical number of beneficial mutations in traits one and two, and their respective colors indicates their abundance.
+The red dot within the distribution marks the point whose coordinates are the mean number of beneficial mutations in each trait, while the red line passing through it represents the line of constant relative fitness associated with the mean.
+Classes that lie on this line have the same relative fitness, but vary in their number of beneficial mutations that have accumulated in each trait.
+In the perpendicular direction, the relative fitness of individuals maximally increases by improving in both traits.
+The set of classes shown in shades of blue are considered the bulk, while those in shades of red belong to the stochastic front.
+For classes within the bulk, beneficial mutations and random sampling produce negligible effects in the changes of their abundances.
+Moreover, beneficial mutations in this region do not contribute to the adaptation of the populations.
+The dynamics of classes in the bulk are completely driven by selection and can be modeled deterministically (equation 2).
+In asexuals, only beneficial mutations that occur on the fittest backgrounds produce lineages that may eventually grow large enough to affect the rate of increasing mean relative fitness.
+These are exactly the beneficial mutations at the edge of the bulk that give rise to classes in the stochastic front, marked by arrows.
+All other lineages are out-competed and their abundances remain insignificant with respect to the population's fixed size.
+As new classes establish along the stochastic front, the bulk expands towards the direction of increasing relative fitness, while classes below the red line dwindle and eventually disappear.
+This leads to the two dimensional distribution shifting in the direction of increasing relative fitness over time.}
 \end{figure}
@@
 The behavior of these classes can be modeled as a branching process fed by incoming mutants generated in adjacent, less fit, classes within the bulk.
+
 Each mutant lineage that appears in the stochastic front either goes establishes or goes extinct.
@@
 Standard mathematical results for branching processes can be used to solve for the lineage's probability of fixation which is given by equation 3.
+% * <masel@email.arizona.edu> 2017-08-22T21:57:02.810Z:
+% 
+% > Standard mathematical results for branching processes can be used to solve for the lineage's probability of fixation which is given by equation 3
+% Eq.
+3 is not in a familiar form to me, and you don't give a citation.
+The result we discussed at your oral exams was pi = s/sigma^2 where sigma is the standard deviation of offspring number given a particular fitness/expectation.
+D&F for-convenience choice was sigma=1, so pi=s.
+% 
+% ^ <kgomez81@math.arizona.edu> 2017-08-23T17:43:47.591Z:
+% 
+% Their initial discussion of the branching process uses "s" to state the general results of mutant lineages establishing, but they later use "qs" for their analysis of establishments at the nose.
+Equation 3 is taken from equation 16 in Desai and Fisher, but is also a general result found in Allen's book "intro to stoch proc" page 262.
+%
+% ^.
 \begin{equation}
@@
 In our model, we find it more useful to consider the related random variable $\nu = \frac{1}{s_{ij}}\exp[-\Delta r_{ij}\tau]$ representing an initial size that produces the deterministic trajectory observed after establishment.
-Using the expression given by Desai and Fisher for the PDF of $\tau$, we can arrive at the cumulative distribution of  $\nu$ (equation 5).
+Using the expression given by Desai and Fisher for the PDF of $\tau$, we can arrive at the cumulative distribution of  $\nu$ (equation 5) by following the analysis of \cite{Uecker2011}.
+% * <masel@email.arizona.edu> 2017-08-22T22:14:21.200Z:
+% 
+% > useful to consider the related random variable $\nu = \frac{1}{s_{ij}}\exp[-\Delta r_{ij}\tau]$ representing an initial size that produces the deterministic trajectory observed after establishment
+% If doing this was Uecker & Hermisson's idea, then this is the place to give them credit/citation.
+% 
+% ^ <kgomez81@math.arizona.edu> 2017-08-23T17:49:29.395Z:
+% 
+% I modified the last sentence, but this is probably not clear.  I'll have to go back to Uecker and Hermisson to come up with a more detailed explanation concerning their analysis.
+% 
+% Using the expression given by Desai and Fisher for the PDF of $\tau$, we can arrive at the cumulative distribution of  $\nu$ (equation 5) by following the analysis of \cite{Uecker2011}. 
+% 
+% ^.
 \begin{equation}
@@
 Each class in the stochastic front may consist of many mutant lineages.
-However, it is the the arrival time of the first mutant lineage destined to establish that determines which class will join the bulk next.
+However, it is the arrival time of the first mutant lineage destined to establish that determines which class will join the bulk next.
 Once the genotype of this mutant lineage is obtained, its initial size $\nu$ can be sampled.
@@
 \par
+% * <masel@email.arizona.edu> 2017-08-22T22:18:30.247Z:
+% 
+% >  its initial size $\nu$ can be sampled
+% I think this is the first time in the Methods that you talk about sampling.
+The rest could be read as a general mathematical description, rather than a methodology for simulation, so this could cause confusion.
+Right now you begin talking about numerical implementation two paragraphs down from here, and this reference to sampling doesn't have that context.
+% 
+% ^.
 % * <masel@email.arizona.edu> 2017-07-29T16:52:06.887Z:
@@
 It follows that the total rate of adaptation is sum of rates of adaptation corresponding to each trait ($\dot{\bar{r}}=\dot{\bar{r_1}} +\dot{\bar{r_2}}$).
-We can apply Lande's result to the two components $\dot{\bar{r_1}} $ and $\dot{\bar{r_2}} $ and rewrite them in terms of the trait variances $\sigma_k^2$ (trait $k=1,2$) and their covariance $\sigma_{12}$.
+We can apply Lande's result to the two components $\dot{\bar{r_1}} $ and $\dot{\bar{r_2}} $ and rewrite them in terms of the trait variances $\sigma_1^2 = \sum_{ij} p_{ij} (r_i-\bar{r_1})^2$ and $\sigma_2^2 = \sum_{ij} p_{ij} (r_j-\bar{r_2})^2$, and their covariance $\sigma_{12} =\sum_{ij} p_{ij} (r_i-\bar{r_1})(r_j-\bar{r_2})$.
 In doing so, we obtain the system expressed in equation 6.
+% * <masel@email.arizona.edu> 2017-08-22T22:41:29.374Z:
+% 
+% >  equation 6
+% Instead of hard-coding equation numbers, you can create auto-numbered links - ask Jason how.
+Right now the number doesn't match.
+You should also link this more clearly to the way it was presented in the introduction, ie you should explain that this is simply $\Delta \bar{z} = \textbf{G} \hspace{.05in}\beta$ in the case of two traits both contributing directly to fitness
+% 
+% ^.
 % * <masel@email.arizona.edu> 2017-07-29T17:21:52.605Z:
@@
 We did this by identifying the first mutant lineage destined to establish among all mutants appearing in the stochastic front.
-Mutants appear according to a time-inhomogeneous Poisson process with rate functions given by the abundances of the adjacent classes from which they originated.
+Mutants awhich are parallel to the line shownwhich are parallel to the line shownppear according to a time-inhomogeneous Poisson process with rate functions given by the abundances of the adjacent classes from which they originated.
 Consequently, the arrival time $\tau_l$ of the $l^{th}$ mutant generated from the class $n_{ij}(t)$ having a mutation in trait $k$ can be used to form the random variable
+% * <masel@email.arizona.edu> 2017-08-22T22:55:54.046Z:
+% 
+% > can be used to form the random variable
+% This is confusing. The equation is for the cumulant for the inhomogeneous rate, the equation does not describe a random variable.
+% 
+% ^.
 \[ 
@@
 The next class to establish was chosen accordingly and incorporated into the bulk with a random initial size generated with from \textbf{equation 4}.\par
+% * <masel@email.arizona.edu> 2017-08-22T22:57:52.494Z:
+% 
+% >  \textbf{equation 5}
+% Again, replace all these with auto-numbered links to specific equations, not to hard-numbered text.
+% 
+% ^.
 
@@
 
-\section{Results and Discussion}
+\section{Results}
+\subsection{The Effect of a Second Trait on the Evolution of the First}
+Desai and Fisher's results indicate that the rate of adaptation of one trait will inevitably fall by adding an additional trait, but our simulations indicate that this is the result of an increase in the trait's variance, and the formation of negative covariance with the additional trait.
+For our simulations, we focused on examining sets of $N$, $s$ and $U$ within ranges that were reported in various papers evolutionary experiments involving either E.
+coli or S.
+cerevisiae \citep{desai2007speed,Levy2015,Perfeito2007}.
+We ran three sets of simulations, each set constructed with one varying parameter among the three.
+The mean values for trait one's variance and covariance were recorded in proportion to the variance of trait one when evolving independently.
+We found that in every case the addition of a second trait always lead to an increase in the variance of the original trait (fig 2a-2c).
+The increases ranged from 40\% to 100\%, depending the parameter being varied and its value.
+Normally this would lead to a rise in trait one's rate of adaptation, but the addition of a trait also resulted in covariance term that was always both negative and substantially larger in magnitude (fig .
 
-\subsection{Results}
-Our simulations were carried out for the symmetric case with parameters $s$ and $U$ being identical for both traits, but in collecting our data, we regarded trait one as the focal trait.
-The population parameters $N$, $s$, and $U$ were chosen using the results \citep{desai2007beneficial} for fitness variance.
-Specifically, we ensured that the two traits would have a sufficient amount of genetic diversity to study how their evolution is affected by interactions with the other trait ($\sigma_k^2(N,s_k,U_k) > s_k^2$ ).
-We also focused on parameter sets that were in agreement with reported values of $N$, $s$, and $U$ in evolutionary experiments involving model asexual micro-organisms, or .
-We measured the fitness variance of trait one over time, as well as its covariance, and then calculated the contribution of each to the reduction in trait one's rate of adaptation.
+In all cases we found that the reduction in trait one's rate of adaptation equaled 40\% its original variances.
 
-In the one dimensional traveling wave model, $N$, $s$, and $U$  prescribe a specific
+In all cases for the selected parameter sets, The addition of a second trait increases the variance of the first as shown
 
-We also selected parameters that coincided with in evolutionary experiments for micro-organisms that typically are regarded to be asexual, or have very limited amounts of sex.
-In table 1 we summarize these ranges for various micro-organisms of interest, these
+\begin{figure}[h]
+\begin{subfigure}[b]{1\linewidth}
+\includegraphics[width=.45\linewidth]{fig5d.pdf}
+\includegraphics[width=.45\linewidth]{fig5e.pdf}
+\end{subfigure}
 
-\begin{table*}[ht]
-\centering
-\begin{tabular}{ | c | c | c | c |}
-\hline Micro-organism& Population Size ($N$)	& Selection Coefficient ($s$) & Beneficial Mutation Rate ($U$) \\
-\hline \begin{tabular}{c} \textit{E}. coli \\ \citep{Perfeito2007} \end{tabular} & $10^6 - 10^{14}$ & $0.5 \% - 2 \%$  & $10^{-9} - 10^{-5}$ \\
-\hline \begin{tabular}{c} \textit{S}. cerevisiae \\ \citep{desai2007speed,Levy2015} \end{tabular} & $10^7 - 10^{12}$ & $0.5 \%- 3 \%$  & $10^{-5} - 10^{-4}$ \\
-\hline
-\end{tabular}
-\caption{Population parameters measured in evolution experiments for key microorganisms. Beneficial mutations rates are given in units of mutations per genome per generation.} 
-\label{Table 1}
-\end{table*}
+\begin{subfigure}[b]{1\linewidth}
+\includegraphics[width=.45\linewidth]{fig5f.pdf}
+\end{subfigure}
+
+\label{fig}
+\caption{\footnotesize Contributions to rate of adaptation from changes in trait variance and covariance.
+(a) $s$ varies with $N$ and $U$ being kept fixed ($N= 10^7$, $U=10^-5$).
+(b) $U$ varies with $N$ and $s$ being kept fixed ($N= 10^7$,$s=10^-2$).
+(c) $N$ varies with $s$ and $U$ being kept fixed ($s=10^-2$,$U=10^-5$).}
+\end{figure}
+
+\subsection{Independence of the Rate of Adaptation's Reduction}
+In every simulation, the overall calculated reduction rate of adaptation scaled by the initial variance of the trait was found to be independent of the parameters of $N$, $U$ and $s$.
+
+ with the second trait.
+In simulations of our model we observed two important changes Simulations of our model indicate that the expected reduction caused by the adding an additional trait is due  covariance term.
+variance of the trait.
+Figure 2 shows an example of the two dimensional distribution taken from simulations; the classes of the stochastic front are not shown.
+One of the distribution's key features is its oblateness along the lines of constant relative fitness which are parallel to the line shown that crosses the population's mean relative fitness.
+As new classes establish in the stochastic front, genetic variation accumulates on these lines of constant relative fitness and can increase the distribution's width substantially along these directions.
+In contrast, the width of the distribution in the perpendicular direction is kept in check by beneficial mutation-selection balance, and will tend to be shorter.
+This is caused by the fact that the shape of the distribution determines its covariance, and subsequently, influences the rate of adaptation in each trait.
+When the two dimensional distribution is oblate in the manner shown in figure 2, the covariance between two traits will be negative.
+In particular, there will be genotypes that tend to be either superior in one of the traits with respect to its population mean, but inferior in the other relative to its mean.
+The shape of the distribution fluctuates over time but the covariance remains largely negative as shown in its sampled trajectory figure 3 drawn from simulations.
+Positive covariance is relatively brief in comparison due its effects on the population's rate of adaptation in each trait.
+Specifically, positive covariance increases their rate of adaptation (Eqt 5).
+This shortens the width of the distribution along the direction of maximally increasing relative fitness, while the width along the perpendicular direction remains unaffected.
+This restores distribution's oblate shape typical as pictured in figure 2 and once again leads to negative covariance.
+Period of negative covariance endure much longer because of the opposite effects it has on the rate of adaptation.
+Specifically, a slower rate of adaptation in each trait provide additional chances for classes in the stochastic front to establish and increase the distribution's width further.
+These lead to large spikes in negative covariance such as those shown in figure 3.
+% * <masel@email.arizona.edu> 2017-08-22T23:04:41.253Z:
+% 
+% > which are parallel to the line shown
+% The line is not shown.
+% 
+% ^.
+% * <masel@email.arizona.edu> 2017-08-22T23:02:02.846Z:
+% 
+% > Quantitative genetic interactions
+% I don't know what a "quantitative genetic interaction" is. An interaction must be between two specified objects A and B.
+% 
+% ^.
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{fig2a_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
-\label{Figure 2.}
-\caption{Trait 1 variance and covariance.}
+\includegraphics[width=1\linewidth]{fig2-13130_0_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
+\label{Figure}
+\caption{\footnotesize Two dimensional distribution of bulk with negative covariance.
+The plot shown represents a snapshot of the two dimensional distribution of the bulk taken from an actual simulation.
+The classes in the stochastic front are not shown, and the line crossing the distribution represents the line of constant relative fitness passing the population's mean relative fitness.
+In this instance the distribution is oblate along the direction parallel to the lines of constant relative fitness, indicating negative covariance between the two traits.
+This shape is maintained for extended periods of time as a result of the effects that negative covariance has on the rates of adaptation for each trait.
+By slowing down the progression of the mean, this ensures that classes at the edge of the bulk and in the stochastic front maintain their selective advantage for greater periods of time.
+As a result, classes in the edge of the bulk grow faster and produce more mutations, while classes in the stochastic front have an increased influx of mutants and an larger probability of establishing.
+(Simulation parameters: $N=10^9$, $s=10^{-2}$, and $U=10^{-5}$ )}
 \end{figure}
-Figure 1 also demonstrates that the covariance over time fluctuates
 
-\subsection{Discussion}
+\begin{figure}
+\includegraphics[width=1\linewidth]{fig3_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
+\label{Figure}
+\caption{\footnotesize An average negative covariance between traits one and two causes a reduction in trait one's mean rate of adaptation.
+The figure includes sampled trajectories of the the fitness variance in trait one and its covariance over a period of 10,000 generations.
+The average values of these trajectories are shown as horizontal lines, and their calculated values are included in the legend.
+The plot also include a horizontal line representing trait one's variance when evolving without a second trait, and the legend includes its calculated value ($\sigma^2$).
+For this particular simulation, the rate of adaptation of trait one is drops by 38\% after adding the second trait.
+This is in agreement with the theoretical calculations  discussed in the introduction.
+We observed that the mean rate of adaptation for trait one without the second trait was $v_1 = 5.31 \times 10^{-5}$ ($= \sigma^2 $).
+Upon adding the second trait, trait one's mean rate of adaptation fell by 37\% to the value $v_1  = 3.35 \times 10^{-5}$ ($= \sigma_1^2+\sigma_{12}$).
+We found that adding the second trait causes a 49\% increase in the fitness variance of the first trait, which would translate into a 49\% increases in trait one's rate of adaptation.
+However, the covariance term whose effect is to reduce the rate of adaptation due to its negative sign, decreases trait one's mean rate of adaptation by 86\%.
+Summing the two contributions provides a net change of -37\% in the rate of adaptation of trait one.
+(Simulation parameters: $N=10^9$, $s=10^{-2}$, and $U=10^{-5}$ for our simulations)}
+\end{figure}
 
+In our simulations, we observed highly correlated behavior between the variance trait one and its covariance with the second.
+Their fluctuations shown in figure 3 are nearly in sync throughout duration of the simulations.
+This behavior can be explained by the effects that covariance has on the distribution of individual traits.
+As previously mentioned, negative covariance measure associations between fitter values of one trait with poorer one in the other.
+Consequently, the larger the magnitude of the negative covariance, the more likely it is that we will observe extreme values in each trait, with each these associated with the opposite opposite versions in the other trait.
+(This section still needs more work)\par
+
+In selecting parameters, we restricted our attention to values of $N$, $s$, and $U$ which ensured that genetic variation would be present in each trait.
+This was necessary to study the quantitative genetic interactions described  by equation 5, since each trait must first have some amount of variance to ensure the possibility of a covariance term.
+To do so, we applied Desai and Fisher's results and fitness variance formula, and considered parameter sets that satisfied the relationship $\sigma^2(N,s,U) \geq s^2 / \log(s/U)$.
+For our simulation parameters, we also restricted our attention to values of $N$, $s$, and $U$ that were typical of widely studied asexaul micro-organisms, or micro-organisms with very limited amounts of sex.
+We examined various literature sources that reported these values as measured in evolutionary experiments \citep{Perfeito2007,desai2007speed,Levy2015}.
+
+\subsection{The Dynamics of \textbf{G} Over time}
+Under beneficial mutation-selection balance, the time scale governing fluctuations in the G-matrix is proportional to the inverse of the expected rate of adaptation $v(N,s,2U)$ as given by Desai and Fisher.
+In our model, the components of the G-matrix are highly correlated in their dynamics due to their adherence to equation 5.
+Since the left hand side is approximately constant in beneficial mutation-selection balance, the components of G on average satisfy a linear relationship.
+Focusing on the covariance terms, we can see that its dynamics must approximately coincide with those of the one dimensional fitness distribution associated with the two dimensional trait distribution.
+As shown in figure 4b, the peak of the one dimensional Gaussian represents the largest portion of the population and it occurs when the population's mean number of beneficial mutations is an integer.
+This is the case when the mean of the two dimensional passes through a line of constant fitness, shown in fig 4a.
+The distribution of abundances for classes along this line of constant fitness determine the covariance to a large degree (see Appendix).
+As peak shifts, as shown in fig 4b, the covariance transitions from one value to another, each determine by the distribution of abundances for classes along the line of constant fitness which coincides with the mean of the two dimensional trait distribution.
+These major transition are governed by the rate of adaptation, while in between them the covariance will be convex combination of the two.
+Minor fluctuations in these dynamics are contributed by the residual portion of abundances not included in these two groups.
+
+\begin{figure}
+\begin{subfigure}[b]{1\linewidth}
+\label{a}
+\caption{Two Dimensional Traveling Wave}
+\includegraphics[width=1\linewidth]{fig9a.png}\\
+\end{subfigure}
+
+\begin{subfigure}[b]{1\linewidth}
+\label{b}
+\caption{Abundances Along Axis of Total Relative Fitness}
+\includegraphics[width=1\linewidth]{fig9b.png}
+\end{subfigure}
+
+\label{Fig 5}
+\caption{\footnotesize Fluctuations in the G-matrix, whose components are highly correlated, occur on time scales proportional to the rate of adaptation $v(N,s,2U)$.
+The relationship between the two dimensional trait distribution and its associated one dimensional fitness distribution governs the the dynamics of covariance over time, and changes in the latter are determined by $v$.
+(a) Two sets of classes are distinguished in the two dimensional distribution shown.
+The first set, whose boxes are shaded in with patterned gray, represent the set of classes with 38 total beneficial mutations.
+At some time $t$, the population's mean fitness lies exactly on the line of constant fitness associated with this group indicated by the black dot.
+The second set of classes are shaded in patterned black and they represent the classes with 39 total beneficial mutations.
+At some time $t+tau_q$, the black dot shift and reside on the line of constant fitness associated with this second group.
+(b) The two distributions are the one dimensional traveling waves decribed by Desai and Fisher, which are associated with the two dimensional trait distribution in (a), but for the two distinct time points $t$ and $t+\tau_q$.
+Both are Gaussian, and consequently highly peaked at the population's mean fitness.
+At time $t$ the peak lies exactly on the marker for 38 total beneficial mutations.
+This indicates that the abundances of classes in the first group discussed in (a) largely determine the covariance of the two dimensional trait distribution.
+At time $t+\tau_q$, the covariance shifts to a new value that then determined by the distribution of abundances in second group discussed in (a).
+In between these times the covariance is nearly a convex combination of its values at these two times.
+However, all major transitions in covariance are in fact occurring on the same time scale as advances in the mean, i.e.
+the rate of adaptation.}
+\end{figure}
+
+\section{Discussion}
+
+\subsection{Comparison with Barton and Otto's Negative Linkage Disequilibria Results}
 In our model, covariance between traits is arises entirely from linkage disequilibrium.
@@
 It follows that the $p_{ij}$ are the gametic frequencies for the genome consisting of allele $i$ paired with allele $j$, and $D_{ij} = p_{ij}-p_{i\cdot} p_{\cdot j}$ is their coefficient of linkage disequilibrium ( $p_{i\cdot}$ and $p_{\cdot j}$ are marginal distributions).
-By applying \textbf{equation 2}, we can derive a mathematical relationship between covariance and linkage disequilibrium.
+By applying \textbf{equation 2}, we can derive a mathematical relationship between covariance and linkage disequilibrium (need to cite chapter 5, Walsh and Lynch).
 % * <masel@email.arizona.edu> 2017-07-29T17:34:47.794Z:
@@
 These, along with the population size, were chosen to match values reported in evolutionary experiments involving relevant micro-organisms (see Table 1).
-% \section{In-text Citations}
-
-% Add citations using the \verb|\citep{}| command, for example \citep{neher2013genealogies} or for multiple citations, \citep{neher2013genealogies, rodelsperger2014characterization}
-
-% \section{Examples of Article Components}
-% \label{sec:examples}
-
-% The sections below show examples of different header levels, which you can use in the primary sections of the manuscript (Results, Discussion, etc.) to organize your content.
-
-% \section{First level section header}
-
-% Use this level to group two or more closely related headings in a long article.
-
-% \subsection{Second level section header}
-
-% Second level section text.
-
-% \subsubsection{Third level section header:}
-
-% Third level section text. These headings may be numbered, but only when the numbers must be cited in the text.
-
-% \section{Figures and Tables}
-
-% Figures and Tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.
-
-% \subsection{Sample Figure}
-
-% Figure \ref{fig:spectrum} shows an example figure.
-
-% \begin{figure}[htbp]
-% \centering
-% \includegraphics[width=\linewidth]{example-figure}
-% \caption{Example figure from \url{10.1534/genetics.114.173807}.
-Please include your figures in the manuscript for the review process.
-You can upload figures to Overleaf via the Project menu.
-Upon acceptance, we'll ask for your figure files to be uploaded in any of the following formats: TIFF (.tiff), JPEG (.jpg), Microsoft PowerPoint (.ppt), EPS (.eps), or Adobe Illustrator (.ai).
-Images should be a minimum of 300 dpi in resolution and 500 dpi minimum if line art images.
-RGB, CMYK, and Grayscale are all acceptable.
-Halftones should be high contrast with sharp detail, because some loss of detail and contrast is inevitable in the production process.
-Figures should be 10-20 cm in width and 1-25 cm in height.
-Graph axes must be exactly perpendicular and all lines of equal density.
-% Label multiple figure parts with A, B, etc.
-in bolded type, and use Arrows and numbers to draw attention to areas you want to highlight.
-Legends should start with a brief title and should be a self-contained description of the content of the figure that provides enough detail to fully understand the data presented.
-All conventional symbols used to indicate figure data points are available for typesetting; unconventional symbols should not be used.
-Italicize all mathematical variables (both in the figure legend and figure) , genotypes, and additional symbols that are normally italicized.
-% }%
-% \label{fig:spectrum}
-% \end{figure}
-
-% \subsection{Sample Table}
-
-% Table \ref{tab:shape-functions} shows an example table.
-Avoid shading, color type, line drawings, graphics, or other illustrations within tables.
-Use tables for data only; present drawings, graphics, and illustrations as separate figures.
-Histograms should not be used to present data that can be captured easily in text or small tables, as they take up much more space.
-
-% Tables numbers are given in Arabic numerals. Tables should not be numbered 1A, 1B, etc., but if necessary, interior parts of the table can be labeled A, B, etc. for easy reference in the text.
-
-% \begin{table*}[htbp]
-% \centering
-% \caption{\bf Students and their grades}
-% \begin{tableminipage}{\textwidth}
-% \begin{tabularx}{\textwidth}{XXXX}
-% \hline
-% Student & Grade\footnote{This is an example of a footnote in a table.
-Lowercase, superscript italic letters (a, b, c, etc.) are used by default.
-You can also use *, **, and *** to indicate conventional levels of statistical significance, explained below the table.} & Rank & Notes \\
-% \hline
-% Alice & 82\% & 1 & Performed very well.\\
-% Bob & 65\% & 3 & Not up to his usual standard.\\
-% Charlie & 73\% & 2 & A good attempt.\\
-% \hline
-% \end{tabularx}
-%   \label{tab:shape-functions}
-% \end{tableminipage}
-% \end{table*}
-
-% \section{Sample Equation}
-
-% Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
-% \begin{equation}
-% S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
-%       = \frac{1}{n}\sum_{i}^{n} X_i
-% \label{eq:refname1}
-% \end{equation}
-% denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.
 
 \bibliography{biblio}
+\bibliographystyle{plain}
 

### 2017-09-14 plotdata.py
adding code for new plot

@@ -5,6 +5,10 @@ Created on Mon Jul 17 13:27:18 2017
 Script to process data generated by simulations in Mathematica code.
 """
 
+# --------------------------------------------------------------------------------
+# Create data for figure showing time plots of variance and covariance over time
+# --------------------------------------------------------------------------------
+
 # import packages needed for script
 import pickle
 import scipy as sp
@@ -123,7 +127,6 @@ vUthryp = np.ones([num_exp,1])
 v2Uthryp = np.ones([num_exp,1])
 NsUparam = [[] for l in range(num_exp)]
 
-
 def get_sample_window(times,start_time,end_time):
 # returns: indeces of times that correspond to start_time and end_time
  
@@ -236,7 +239,10 @@ pickle_file = open(pickle_file_name,'wb')
 pickle.dump([var, cov, vUthry, v2Uthry, varp, covp, vUthryp, v2Uthryp,NsUparam],pickle_file,pickle.HIGHEST_PROTOCOL)
 pickle_file.close()
 
-# ------------------------------------------------------------------------------
+# --------------------------------------------------------------------------------
+# Create summarized data for Joanna and Jason var-cov plots
+# --------------------------------------------------------------------------------
+
 # import packages needed for script
 import pickle
 import scipy as sp
@@ -276,3 +282,89 @@ for k in range(num_exp):
     pickle_file = open(pickle_file_name,'wb') 
     pickle.dump([tau_q,median_cov_time,mean_cov,dcov_dt,cov_times,parameters],pickle_file,pickle.HIGHEST_PROTOCOL)
     pickle_file.close()
+
+# --------------------------------------------------------------------------------
+# Create summarized data for Joanna and Jason var-cov plots
+# --------------------------------------------------------------------------------
+
+# import packages needed for script
+import pickle
+import scipy as sp
+import numpy as np
+import copy as cpy
+
+def get_trait_mean_var(genotypes,abundances,traitno):
+    
+    if ((traitno == 1) | (traitno == 2)):
[…]

### 2017-09-14 plotfunctions.py

@@ -118,6 +118,32 @@ def get_2D_distr(genotypes,abundances,box_dim):
     return [my_distr,xlabels,ylabels]
 
 # -----------------------------------------------------------------------------
+def get_cov_by_fitness_line(genotypes,abundances):
+    
+    mean_fit1 = get_trait_mean_var(genotypes,abundances,1)[0]
+    mean_fit2 = get_trait_mean_var(genotypes,abundances,2)[0]
+    num_genotypes = len(abundances[0])
+    
+    fit1D = [genotypes[i,0]+genotypes[i,1] for i in range(num_genotypes)]
+    fit1Dshrt = list(set(fit1D))
+    fit1Dcovs = []    
+    tempcov = 0
+    tempfreq = 0
+    popsize = sum(abundances[0])
+    
+    for i in range(len(fit1Dshrt)):
+        tempcov = 0
+        tempfreq = 0
+        for j in range(num_genotypes):
+            if(fit1D[j]==fit1Dshrt[i]):
+                tempfreq += abundances[0][j]/popsize
+                tempcov += (genotypes[j,0]-mean_fit1)*(genotypes[j,1]-mean_fit2)*(abundances[0][j]/popsize)
+        fit1Dcovs = fit1Dcovs +[[fit1Dshrt[i],tempfreq,tempcov/tempfreq]]
+    
+    # should return [[fit_i,p_i,cov_i,] for i = min_fit,...,max_fit]
+    return fit1Dcovs
+
+# -----------------------------------------------------------------------------
 def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param,fname,traitno,num_exp):
 # figure 1: representation of two-dimensional distribution  (no data required)
     [N,s1,s2,U1,U2] = pop_param
@@ -302,152 +328,8 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
         ax4.set_title('1D distribution trait '+str(traitno))
         fig4.savefig('./'+folder_location+'figures/'+fname+data_name+'.png')
 
-# figure 5: plot of covariances and trait 1 varainces
+# figure 5: plots of var-cov means wrt 
     if (figNum == 5):
-        var = np.ones([num_exp,1])
-        cov = np.ones([num_exp,1])
[…]

### 2017-09-14 plots.py

@@ -45,6 +45,4 @@ num_exp = 33
 #   
    
 # generate set for figure 5
-#pltfun.generate_figure(5,data_name,folder_location,sim_start,sim_end,pop_param,'fig',traitno,num_exp)
-#pltfun.generate_figure(6,data_name,folder_location,sim_start,sim_end,pop_param,'fig',traitno,num_exp)
-pltfun.generate_figure(7,'7',folder_location,sim_start,sim_end,pop_param,'fig',traitno,num_exp)
+pltfun.generate_figure(5,data_name,folder_location,sim_start,sim_end,pop_param,'fig',traitno,num_exp)

### 2017-09-29 manuscript.tex
added code to predict covariance

@@
 Simulations were initialized with monoclonal populations at carrying capacity consisting of individuals having no beneficial mutations in either trait and relative fitness set zero.
+We constructed three parameter sets, each one varying either $N$, $s$ and $U$ while fixing the others.
+Each was varied across a range constructed from values reported by evolutionary experiments involving either E.
+coli or S.
+cerevisiae \citep{desai2007speed,Levy2015,Perfeito2007}.
+We ran three sets of simulations, each set constructed with one varying parameter among the three.
+The mean values for trait one's variance and covariance were recorded in proportion to the variance of trait one when evolving independently.
+
 We allowed the simulation to run for 5,000 generations to allow the population to achieve beneficial mutation-selection balance before collecting data.
 This ensured that the transient effects from the initializing would not distort the statistics collected for the distribution.
-Following the burn period of the simulation, we recorded the existing classes in the bulk and their respective abundances every one 100 generations and intermittent times at which new classes established.
+Following the burn period of the simulation, we recorded the existing classes in the bulk and their respective abundances.
 The data was used to compute the marginal mean fitness and variances of each trait, as well as their covariance.
@@
 \section{Results}
-\subsection{The Effect of a Second Trait on the Evolution of the First}
-Desai and Fisher's results indicate that the rate of adaptation of one trait will inevitably fall by adding an additional trait, but our simulations indicate that this is the result of an increase in the trait's variance, and the formation of negative covariance with the additional trait.
-For our simulations, we focused on examining sets of $N$, $s$ and $U$ within ranges that were reported in various papers evolutionary experiments involving either E.
-coli or S.
-cerevisiae \citep{desai2007speed,Levy2015,Perfeito2007}.
-We ran three sets of simulations, each set constructed with one varying parameter among the three.
-The mean values for trait one's variance and covariance were recorded in proportion to the variance of trait one when evolving independently.
-We found that in every case the addition of a second trait always lead to an increase in the variance of the original trait (fig 2a-2c).
-The increases ranged from 40\% to 100\%, depending the parameter being varied and its value.
-Normally this would lead to a rise in trait one's rate of adaptation, but the addition of a trait also resulted in covariance term that was always both negative and substantially larger in magnitude (fig .
+In each of our simulations we observed a 40\% decrease in the rate of adaptation of a trait as a result adding a second.
+The decrease was consistent across three parameter sets which varied either $N$, $s$ and $U$.
+These results also agree with predictions that can be derived from Desai and Fisher's theoretical work ($v(N,s,U)$).
+However, our simulations show that upon adding the second trait, the net change in $v_1$ arises from an increase in the first trait's variance and the formation of negative covariance.
+The first results in an increase in $v_1$, as a direct consequence of equation 5.
+This gain is more than offset by the negative covariance created between the two traits, whose magnitude is always substantially larger than the variance increase.
+Figure 2 presents the calculated scaled mean variance of trait one, and its scaled covariance with the second trait, as a function of each varying population parameter.
+The graph demonstrates that within these parameter ranges, the new rate of adaptation $v_1$ as a percentage of the prior in every case is nearly 60\% less due solely to the large amounts of negative covariance that builds up between the two traits.\par
 
-In all cases we found that the reduction in trait one's rate of adaptation equaled 40\% its original variances.
-
-In all cases for the selected parameter sets, The addition of a second trait increases the variance of the first as shown
-
+%------------------------------------------------------------------------------------------------------------
+%------------------------------------------------------------------------------------------------------------
 \begin{figure}[h]
@@
 \end{subfigure}
-
 \begin{subfigure}[b]{1\linewidth}
+\centering
 \includegraphics[width=.45\linewidth]{fig5f.pdf}
 \end{subfigure}
-
 \label{fig}
@@
 \end{figure}
+%------------------------------------------------------------------------------------------------------------
+%------------------------------------------------------------------------------------------------------------
 
-\subsection{Independence of the Rate of Adaptation's Reduction}
+The reduction in the scaled mean value of $v_1$ plotted in figure 2 is independent of the choice of parameters $N$, $s$ and $U$ within ranges characterizing asexual evolution.
+The insensitivity of the decrease in $v_1$ to changes in the population parameters is clear from Desai and Fisher's expression for the rate of adaptation, which exhibits a logarithmic dependence on $N$, $s$ and $U$.
+In fact, one can generally express the decrease $v_1$ resulting from the addition of $(n-1)$ traits and arrive at the expression
+\[
+\frac{(1/n) v(N,s,nU)}{v(N,s,U)} 
+= \frac{\ln^2(s/U)\ln(nN^2sU)}{2\ln^2(s/nU)\ln(N^2sU)}. 
+\]
+
+In all cases we found that the reduction in trait one's rate of adaptation equaled 40\% its original variances.
+
+In all cases for the selected parameter sets, The addition of a second trait increases the variance of the first as shown
+
 In every simulation, the overall calculated reduction rate of adaptation scaled by the initial variance of the trait was found to be independent of the parameters of $N$, $U$ and $s$.
@@
 
+%------------------------------------------------------------------------------------------------------------
+%------------------------------------------------------------------------------------------------------------
 \begin{figure}
@@
 \end{figure}
+%------------------------------------------------------------------------------------------------------------
+%------------------------------------------------------------------------------------------------------------
 
+%------------------------------------------------------------------------------------------------------------
+%------------------------------------------------------------------------------------------------------------
 \begin{figure}
@@
 \end{figure}
+%------------------------------------------------------------------------------------------------------------
+%------------------------------------------------------------------------------------------------------------
 
@@
 
-\subsection{The Dynamics of \textbf{G} Over time}
-Under beneficial mutation-selection balance, the time scale governing fluctuations in the G-matrix is proportional to the inverse of the expected rate of adaptation $v(N,s,2U)$ as given by Desai and Fisher.
+Under beneficial mutation-selection balance, the time scale governing fluctuations in the G-matrix is proportional to the inverse of the expected rate of adaptation $v(N,s,2U)$ given by Desai and Fisher's 2007 results.
 In our model, the components of the G-matrix are highly correlated in their dynamics due to their adherence to equation 5.
-Since the left hand side is approximately constant in beneficial mutation-selection balance, the components of G on average satisfy a linear relationship.
-Focusing on the covariance terms, we can see that its dynamics must approximately coincide with those of the one dimensional fitness distribution associated with the two dimensional trait distribution.
-As shown in figure 4b, the peak of the one dimensional Gaussian represents the largest portion of the population and it occurs when the population's mean number of beneficial mutations is an integer.
-This is the case when the mean of the two dimensional passes through a line of constant fitness, shown in fig 4a.
-The distribution of abundances for classes along this line of constant fitness determine the covariance to a large degree (see Appendix).
-As peak shifts, as shown in fig 4b, the covariance transitions from one value to another, each determine by the distribution of abundances for classes along the line of constant fitness which coincides with the mean of the two dimensional trait distribution.
-These major transition are governed by the rate of adaptation, while in between them the covariance will be convex combination of the two.
+The left hand side of this expression is approximately constant in beneficial mutation-selection balance, and as a result, the components of G will on average satisfy a linear relationship.
+The dynamics of the covariance terms coincide with those of the associated one dimensional fitness distribution the two dimensional trait distribution.
+Figure 4b, illustrates the Gaussian shape of the associated one dimensional fitness distribution.
+Its peak represents the largest portion of the population and is composed of classes that lie exactly on the line of constant fitness passing through the population's mean fitness.
+The proportions of the classes within the peak determine the covariance of the two dimensional trait distribution.
+These proportions are approximately fixed in relative size to one another at the time the classes establish when forming part of the stochastic front in a manner describe by \cite{Desai2013}.
+Thus changes in the value of covariance in fact follow the transitions of the peaks the one dimensional traveling wave shift.
+These major transitions are consequently governed by the rate of adaptation and determined at the stochastic front.
+In between the times of these transitions, the covariance will simply be a convex combination of the two adjacent sets of classes that lie on the lines of constant fitness below and above the mean.
 Minor fluctuations in these dynamics are contributed by the residual portion of abundances not included in these two groups.

### 2017-09-29 plotdata.py

@@ -292,6 +292,7 @@ import pickle
 import scipy as sp
 import numpy as np
 import copy as cpy
+import matplotlib.pyplot as plt
 
 def get_trait_mean_var(genotypes,abundances,traitno):
     
@@ -309,28 +310,63 @@ def get_trait_mean_var(genotypes,abundances,traitno):
 def get_cov_by_fitness_line(genotypes,abundances,s):
     
     mean_fit = get_trait_mean_var(genotypes,abundances,0)[0]
-    mean_fit1 = get_trait_mean_var(genotypes,abundances,1)[0]
-    mean_fit2 = get_trait_mean_var(genotypes,abundances,2)[0]
     num_genotypes = len(abundances[0])
     
     fit1D = [genotypes[i,0]+genotypes[i,1] for i in range(num_genotypes)]
     fit1Dshrt = list(set(fit1D))
+    fit1Dshrt.sort()
     fit1Dcovs = []    
     tempcov = 0
     tempfreq = 0
+    tempmean1 = 0
+    tempmean2 = 0
     popsize = sum(abundances[0])
     
     for i in range(len(fit1Dshrt)):
         tempcov = 0
         tempfreq = 0
+        tempmean1 = 0
+        tempmean2 = 0
         for j in range(num_genotypes):
             if(fit1D[j]==fit1Dshrt[i]):
+                tempmean1 += s*genotypes[j,0]*abundances[0][j]/popsize
+                tempmean2 += s*genotypes[j,1]*abundances[0][j]/popsize                
                 tempfreq += abundances[0][j]/popsize
-                tempcov += s**2*(genotypes[j,0]-mean_fit1-0.5*(fit1D[j]-mean_fit))*(genotypes[j,1]-mean_fit2-0.5*(fit1D[j]-mean_fit))*(abundances[0][j]/popsize)
-        fit1Dcovs = fit1Dcovs +[[fit1Dshrt[i]-mean_fit,tempfreq,tempcov/tempfreq]]
+                tempcov += s**2*genotypes[j,0]*genotypes[j,1]*(abundances[0][j]/popsize)
+        tempcov = tempcov/tempfreq - (tempmean1/tempfreq)*(tempmean2/tempfreq)
+        fit1Dcovs = fit1Dcovs+[[fit1Dshrt[i]-mean_fit,tempfreq,tempcov]]
     
[…]

### 2017-09-29 plotfunctions.py

@@ -341,46 +341,46 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
         [end1,end2,end3] = [num_exp/3,2*num_exp/3,num_exp]
         NsUparam = np.asarray(NsUparam)        
         
-        fig6d, ax5d = plt.subplots(1,1,figsize=[8,8])
-        ax5d.plot(NsUparam[start1:end1,1],-np.asarray(covp[start1:end1])-np.asarray(varp[start1:end1])+np.asarray(vUthryp[start1:end1]),c="black",label='$v_1$',linewidth=2.0,linestyle = '-')                
-        ax5d.plot(NsUparam[start1:end1,1],np.asarray(varp[start1:end1]),c="black",label='$\sigma_{1}^2$',linewidth=2.0,linestyle = ':')
-        ax5d.plot(NsUparam[start1:end1,1],np.asarray(covp[start1:end1]),c="black",label='$\sigma_{12}$',linewidth=2.0,linestyle = '-.')        
-        ax5d.axhline(linewidth=0.5, color = 'k')
-        ax5d.set_ylabel('Scaled Variance-Covariance',fontsize=18)
-        ax5d.set_xlabel('selection coeff ($10^{-3} - 2 x 10^{-2}$)',fontsize=18)
-        ax5d.tick_params(axis='both',labelsize=14)
-        ax5d.ticklabel_format(style='sci',axis='x',scilimits=(0,0))
-        ax5d.set_ylim((-2,2))
-        ax5d.set_xlim((1e-3,2e-2))  
-        ax5d.legend()
-        fig5d.savefig('./'+folder_location+'figures/'+fname+data_name+'d.pdf')
+        fig5a, ax5a = plt.subplots(1,1,figsize=[16,8])
+        ax5a.plot(NsUparam[start1:end1,1],-np.asarray(covp[start1:end1])-np.asarray(varp[start1:end1])+np.asarray(vUthryp[start1:end1]),c="black",label='$v_1$',linewidth=2.0,linestyle = '-')                
+        ax5a.plot(NsUparam[start1:end1,1],np.asarray(varp[start1:end1]),c="black",label='$\sigma_{1}^2$',linewidth=2.0,linestyle = ':')
+        ax5a.plot(NsUparam[start1:end1,1],np.asarray(covp[start1:end1]),c="black",label='$\sigma_{12}$',linewidth=2.0,linestyle = '-.')        
[…]

### 2017-09-29 plotfunctions.py
fixed plots

@@ -351,7 +351,7 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
         ax5a.tick_params(axis='both',labelsize=14)
         ax5a.ticklabel_format(style='sci',axis='x',scilimits=(0,0))
         ax5a.set_ylim((-2,2))
-        ax5a.set_xlim((1e-3,2e-2))  
+        ax5a.set_xlim((min(NsUparam[start1:end1,1]),max(NsUparam[start1:end1,1])))  
         ax5a.legend()
         fig5a.savefig('./'+folder_location+'figures/'+fname+data_name+'d.pdf')
         
@@ -365,7 +365,7 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
         ax5b.tick_params(axis='both',labelsize=14)
         ax5b.ticklabel_format(style='sci',axis='x',scilimits=(0,0))
         ax5b.set_ylim((-2,2))
-        ax5b.set_xlim((1e-6,2e-5))          
+        ax5b.set_xlim((min(NsUparam[start2:end2,2]),max(NsUparam[start2:end2,2])))          
         ax5b.legend()
         fig5b.savefig('./'+folder_location+'figures/'+fname+data_name+'e.pdf')
         
@@ -379,7 +379,7 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
         ax5c.tick_params(axis='both',labelsize=14)
         ax5c.ticklabel_format(style='sci',axis='x',scilimits=(0,0))
         ax5c.set_ylim((-2,2))
-        ax5c.set_xlim((1e6,1e9)) 
+        ax5c.set_xlim((min(NsUparam[start3:end3,0]),max(NsUparam[start3:end3,0]))) 
         ax5c.legend()        
         fig5c.savefig('./'+folder_location+'figures/'+fname+data_name+'f.pdf')
 

### 2017-09-29 manuscript.tex
more updates to results section.

@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{fig1.pdf}
+\includegraphics[width=1\linewidth]{fig1a.pdf}
 \label{Figure 1.}
@@
 \includegraphics[width=.45\linewidth]{fig5f.pdf}
+\includegraphics[width=.45\linewidth]{fig5g.pdf}
 \end{subfigure}
@@
 
-The reduction in the scaled mean value of $v_1$ plotted in figure 2 is independent of the choice of parameters $N$, $s$ and $U$ within ranges characterizing asexual evolution.
-The insensitivity of the decrease in $v_1$ to changes in the population parameters is clear from Desai and Fisher's expression for the rate of adaptation, which exhibits a logarithmic dependence on $N$, $s$ and $U$.
-In fact, one can generally express the decrease $v_1$ resulting from the addition of $(n-1)$ traits and arrive at the expression
+(i) the pop param don't affect the deacrease
+
+The decreases shown in the scaled mean values of $v_1$ that are plotted in figure 2 are nearly independent of the population parameters picked within ranges characterizing asexual evolution.
+In all three parameter sets, the decrease nearly is always approximately 40\% of the trait's prior variance.
+The insensitivity follows from Desai and Fisher's expression for the rate of adaptation, which exhibits a logarithmic dependence on $N$, $s$ and $U$.
+This leads to a general expression that can be obtained for the expected reduction in a trait's rate of adaptation as a result of adding  $(n-1)$ traits.
 \[
-\frac{(1/n) v(N,s,nU)}{v(N,s,U)} 
-= \frac{\ln^2(s/U)\ln(nN^2sU)}{2\ln^2(s/nU)\ln(N^2sU)}. 
+\% \Delta v_1 
+= 1-\frac{1}{n}\frac{v(N,s,nU)}{v(N,s,U)}
 \]
-
-In all cases we found that the reduction in trait one's rate of adaptation equaled 40\% its original variances.
-
-In all cases for the selected parameter sets, The addition of a second trait increases the variance of the first as shown
-
-In every simulation, the overall calculated reduction rate of adaptation scaled by the initial variance of the trait was found to be independent of the parameters of $N$, $U$ and $s$.
-
- with the second trait.
-In simulations of our model we observed two important changes Simulations of our model indicate that the expected reduction caused by the adding an additional trait is due  covariance term.
-variance of the trait.
-Figure 2 shows an example of the two dimensional distribution taken from simulations; the classes of the stochastic front are not shown.
-One of the distribution's key features is its oblateness along the lines of constant relative fitness which are parallel to the line shown that crosses the population's mean relative fitness.
-As new classes establish in the stochastic front, genetic variation accumulates on these lines of constant relative fitness and can increase the distribution's width substantially along these directions.
-In contrast, the width of the distribution in the perpendicular direction is kept in check by beneficial mutation-selection balance, and will tend to be shorter.
-This is caused by the fact that the shape of the distribution determines its covariance, and subsequently, influences the rate of adaptation in each trait.
-When the two dimensional distribution is oblate in the manner shown in figure 2, the covariance between two traits will be negative.
-In particular, there will be genotypes that tend to be either superior in one of the traits with respect to its population mean, but inferior in the other relative to its mean.
-The shape of the distribution fluctuates over time but the covariance remains largely negative as shown in its sampled trajectory figure 3 drawn from simulations.
-Positive covariance is relatively brief in comparison due its effects on the population's rate of adaptation in each trait.
-Specifically, positive covariance increases their rate of adaptation (Eqt 5).
-This shortens the width of the distribution along the direction of maximally increasing relative fitness, while the width along the perpendicular direction remains unaffected.
-This restores distribution's oblate shape typical as pictured in figure 2 and once again leads to negative covariance.
-Period of negative covariance endure much longer because of the opposite effects it has on the rate of adaptation.
-Specifically, a slower rate of adaptation in each trait provide additional chances for classes in the stochastic front to establish and increase the distribution's width further.
-These lead to large spikes in negative covariance such as those shown in figure 3.
-% * <masel@email.arizona.edu> 2017-08-22T23:04:41.253Z:
-% 
-% > which are parallel to the line shown
-% The line is not shown.
-% 
-% ^.
-% * <masel@email.arizona.edu> 2017-08-22T23:02:02.846Z:
-% 
-% > Quantitative genetic interactions
-% I don't know what a "quantitative genetic interaction" is. An interaction must be between two specified objects A and B.
-% 
-% ^.
+Only changes $s$ appear to affect the expected decrease in scaled $v_1$ (fig 2d), while changes in the number of traits seem to be most adverse.
+The extent of this effect is marginally less with each additional trait, but quite substantive in the first few cases.
 

### 2017-09-30 manuscript.tex
more updates to manuscript

@@
 \usepackage{subcaption}
+\usepackage{hyperref}
 \usepackage[margin=.5in]{geometry}
@@
 Historically, there have been two schools of thought what the principle causes of genetic correlations are.
-The first was put forth by the Edinburgh school, which held that pleiotropy was principle source, since linkage disequilibrium could rapidly decay by recombination (\citep{fox2006evolutionary}, chp 20).
+The first was put forth by the Edinburgh school, which held that pleiotropy was principle source, since linkage disequilibrium could rapidly decay by recombination \citep[Chapter~20]{fox2006evolutionary}.
 This belief was largely motivated by interests the relationships between parent-offspring phenotypes within randomly mating populations, and its implications for evolution.
@@
 These models have established important relationships between the population's parameter and key features of evolution in asexuals.
-Among these is the traveling wave model developed by \cite{desai2007beneficial} for a population whose size $N$ is fixed in which only beneficial mutations occur at the rate $U$.
+Among these is the traveling wave model developed by \citet{desai2007beneficial} for a population whose size $N$ is fixed in which only beneficial mutations occur at the rate $U$.
 They also make the key assumption that all mutations all provide the same selective advantage $s$, which allows them to show that the population's rate of adaptation is given by  $v(N,s,U) = 2s^2 \ln(N s)/\ln^2(s/U)$.
@@
 
-In an analogous approach to that of \cite{desai2007beneficial}, we divided our population into classes based on the number of beneficial mutations in each trait.
-Their abundances are accordingly denoted $n_{ij}$ and their frequencies by $p_{ij}$.
+In an analogous approach to that of \citet{desai2007beneficial}, we divided our population into classes based on the number of beneficial mutations in each trait.
+Abundances are accordingly denoted $n_{ij}$ and their frequencies by $p_{ij}$.
 We assumed that $1/N \ll s \ll 1$ to ensure that the set of classes could be neatly separated into two groups based on whether their behavior was deterministic and stochastic.
@@
 
+% *****************************************************************************************************************
+% ***********************************FIGURE 1*2D*Travelling*Wave***************************************************
+% *****************************************************************************************************************
 \begin{figure}
@@
 \label{Figure 1.}
-% * <masel@email.arizona.edu> 2017-08-22T22:26:35.280Z:
-% 
-% > Figure 1.
-% Do you think it is possible to merge Figures 1 and 2 into a single figure? Ie, to take the features you illustrate in Figure 1, put them into Figure 2, and then cut Figure 1 out?
-% 
-% ^.
-\caption{\footnotesize (reproduce this for the actual data and cut out figure 2) Reproduction of two dimensional distribution over the trait space from simulation.
-Each box is a class consisting of individuals with an identical number of beneficial mutations in traits one and two, and their respective colors indicates their abundance.
-The red dot within the distribution marks the point whose coordinates are the mean number of beneficial mutations in each trait, while the red line passing through it represents the line of constant relative fitness associated with the mean.
-Classes that lie on this line have the same relative fitness, but vary in their number of beneficial mutations that have accumulated in each trait.
-In the perpendicular direction, the relative fitness of individuals maximally increases by improving in both traits.
-The set of classes shown in shades of blue are considered the bulk, while those in shades of red belong to the stochastic front.
-For classes within the bulk, beneficial mutations and random sampling produce negligible effects in the changes of their abundances.
-Moreover, beneficial mutations in this region do not contribute to the adaptation of the populations.
-The dynamics of classes in the bulk are completely driven by selection and can be modeled deterministically (equation 2).
-In asexuals, only beneficial mutations that occur on the fittest backgrounds produce lineages that may eventually grow large enough to affect the rate of increasing mean relative fitness.
-These are exactly the beneficial mutations at the edge of the bulk that give rise to classes in the stochastic front, marked by arrows.
-All other lineages are out-competed and their abundances remain insignificant with respect to the population's fixed size.
-As new classes establish along the stochastic front, the bulk expands towards the direction of increasing relative fitness, while classes below the red line dwindle and eventually disappear.
-This leads to the two dimensional distribution shifting in the direction of increasing relative fitness over time.}
+\caption{\footnotesize Evolutuion of a two dimensional traveling wave.
+The distribution depicted was sampled from simulations.
+Each box represents a class composed of individuals with an identical number of beneficial mutations in each trait.
+Classes with solid coloring form part of the bulk, while those with patterned interiors represent the stochastic front.
+The red dot marks the mean number of beneficial mutations in each trait.
+The dashed red line represents a lines of constant fitness.
+Classes whose centers lie along this line have relative fitness equal to the population mean, but vary in the number of beneficial mutations that have accumulated in each trait.
+Relative fitness increases along the perpendicular direction.
+Classes .
+Beneficial mutations and random sampling produce negligible effects in the changes of their abundances.
+In asexuals, only beneficial mutations that occur on the fittest backgrounds produce lineages that may eventually fix.
+These are marked with arrows into the stochastic front.
+Other lineages are out-competed and remain insignificant in size with respect to bulk.
+As new classes, the bulk expands in the direction of increasing relative fitness, and classes below the red line shrink and eventually disappear.
+As a result, the two dimensional trait distribution continually shifts in the direction of increasing relative fitness over time.
+(Simulation parameters: $N=10^9$, $s=10^{-2}$, and $U=10^{-5}$)}
 \end{figure}
+% *****************************************************************************************************************
+% *******************************END*FIGURE 1**********************************************************************
+% *****************************************************************************************************************
 
 The dynamics of the bulk was modeled by a system of ordinary differential equations, in which abundances changed in proportion to their selective advantage (\textbf{eqt. 1}). 
-\begin{equation} 
+\begin{equation} \label{eq:1}
 \dot{n}_{ij}(t) = (r_{ij}-\bar{r}) n_{ij}(t). 
@@
 Since the bulk represents the majority of population, we assumed that the frequencies of the bulk followed an analogous ODE.
-\begin{equation} 
+\begin{equation} \label{eq:2}
 \dot{p}_{ij}(t) = (r_{ij}-\bar{r}) p_{ij}(t) 
@@
 % ^.
-\begin{equation}
+\begin{equation} \label{eq:3}
 \pi_{fix} = \frac{s_{ij}}{ 1+s_{ij}}
@@
 % ^.
-\begin{equation}
+\begin{equation} \label{eq:4}
 P(\nu \leq \nu_0) = 1- e^{-\pi_{fix} \nu_0}
@@
 % ^.
-\begin{equation}
-\left(
+\begin{equation}\label{eq:5}
+\left[
 \begin{array}{c}
@@
 \end{array}
-\right)
+\right]
 =
-\left(
+\left[
 \begin{array}{cc}
@@
 \end{array}
-\right)
-\left(
+\right]
+\left[
 \begin{array}{c}
@@
 \end{array}
-\right)
+\right]
 \end{equation}
-The components of the selection gradient and trait response have been normalized in the expression above.
+The components of the selection gradient and trait response have been normalized in the expression above. \par
 
 Our numerical implementation consisted of two routines which ran iteratively for the duration of the simulation.
-Each iteration began with solving for the time-dependent abundances in the bulk governed by \textbf{equation 1} over a time interval of one-thousand generations.
+Each iteration began with solving for the time-dependent abundances in the bulk governed by \eqref{eq:1} over a time interval of one-thousand generations.
 We used Mathematica's standard numerical ODE solver to obtain the solutions and used them to determine the appearance time of the next mutant lineage destined to establish.
@@
 \section{Results}
+Adaptation in a second trait
+
 In each of our simulations we observed a 40\% decrease in the rate of adaptation of a trait as a result adding a second.
@@
 
-%------------------------------------------------------------------------------------------------------------
-%------------------------------------------------------------------------------------------------------------
+% *****************************************************************************************************************
+% ***********************************FIGURE 2*Scaled Var/Cov Changes & Trait Reduction*****************************
+% *****************************************************************************************************************
 \begin{figure}[h]
 \begin{subfigure}[b]{1\linewidth}
-\includegraphics[width=.45\linewidth]{fig5d.pdf}
-\includegraphics[width=.45\linewidth]{fig5e.pdf}
+\centering
+\includegraphics[width=.48\linewidth]{fig5d.pdf}
+\includegraphics[width=.48\linewidth]{fig5e.pdf}
 \end{subfigure}
@@
 \centering
-\includegraphics[width=.45\linewidth]{fig5f.pdf}
-\includegraphics[width=.45\linewidth]{fig5g.pdf}
+\includegraphics[width=.48\linewidth]{fig5f.pdf}
+\includegraphics[width=.48\linewidth]{fig5g.pdf}
 \end{subfigure}
@@
 \end{figure}
-%------------------------------------------------------------------------------------------------------------
-%------------------------------------------------------------------------------------------------------------
-
-(i) the pop param don't affect the deacrease
+% *****************************************************************************************************************
+% ***********************************END*FIGURE*2******************************************************************
+% *****************************************************************************************************************
 
@@
 \]
-Only changes $s$ appear to affect the expected decrease in scaled $v_1$ (fig 2d), while changes in the number of traits seem to be most adverse.
-The extent of this effect is marginally less with each additional trait, but quite substantive in the first few cases.
-
-%------------------------------------------------------------------------------------------------------------
-%------------------------------------------------------------------------------------------------------------
-\begin{figure}
-\includegraphics[width=1\linewidth]{fig2-13130_0_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
-\label{Figure}
-\caption{\footnotesize Two dimensional distribution of bulk with negative covariance.
-The plot shown represents a snapshot of the two dimensional distribution of the bulk taken from an actual simulation.
-The classes in the stochastic front are not shown, and the line crossing the distribution represents the line of constant relative fitness passing the population's mean relative fitness.
-In this instance the distribution is oblate along the direction parallel to the lines of constant relative fitness, indicating negative covariance between the two traits.
-This shape is maintained for extended periods of time as a result of the effects that negative covariance has on the rates of adaptation for each trait.
-By slowing down the progression of the mean, this ensures that classes at the edge of the bulk and in the stochastic front maintain their selective advantage for greater periods of time.
-As a result, classes in the edge of the bulk grow faster and produce more mutations, while classes in the stochastic front have an increased influx of mutants and an larger probability of establishing.
-(Simulation parameters: $N=10^9$, $s=10^{-2}$, and $U=10^{-5}$ )}
-\end{figure}
-%------------------------------------------------------------------------------------------------------------
-%------------------------------------------------------------------------------------------------------------
-
-%------------------------------------------------------------------------------------------------------------
-%------------------------------------------------------------------------------------------------------------
-\begin{figure}
-\includegraphics[width=1\linewidth]{fig3_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
-\label{Figure}
-\caption{\footnotesize An average negative covariance between traits one and two causes a reduction in trait one's mean rate of adaptation.
-The figure includes sampled trajectories of the the fitness variance in trait one and its covariance over a period of 10,000 generations.
-The average values of these trajectories are shown as horizontal lines, and their calculated values are included in the legend.
-The plot also include a horizontal line representing trait one's variance when evolving without a second trait, and the legend includes its calculated value ($\sigma^2$).
-For this particular simulation, the rate of adaptation of trait one is drops by 38\% after adding the second trait.
-This is in agreement with the theoretical calculations  discussed in the introduction.
-We observed that the mean rate of adaptation for trait one without the second trait was $v_1 = 5.31 \times 10^{-5}$ ($= \sigma^2 $).
-Upon adding the second trait, trait one's mean rate of adaptation fell by 37\% to the value $v_1  = 3.35 \times 10^{-5}$ ($= \sigma_1^2+\sigma_{12}$).
-We found that adding the second trait causes a 49\% increase in the fitness variance of the first trait, which would translate into a 49\% increases in trait one's rate of adaptation.
-However, the covariance term whose effect is to reduce the rate of adaptation due to its negative sign, decreases trait one's mean rate of adaptation by 86\%.
-Summing the two contributions provides a net change of -37\% in the rate of adaptation of trait one.
-(Simulation parameters: $N=10^9$, $s=10^{-2}$, and $U=10^{-5}$ for our simulations)}
-\end{figure}
-%------------------------------------------------------------------------------------------------------------
-%------------------------------------------------------------------------------------------------------------
+Of the three parameters, changes in $s$ affect the expected decrease in scaled $v_1$ most (fig 2d).
+The total number of additional traits also lead to larger and larger reductions in $v_1$.
+However, the extent of this effect is marginally less with each additional trait, but quite substantive with the first few.
+\par
 
@@
 
-In selecting parameters, we restricted our attention to values of $N$, $s$, and $U$ which ensured that genetic variation would be present in each trait.
-This was necessary to study the quantitative genetic interactions described  by equation 5, since each trait must first have some amount of variance to ensure the possibility of a covariance term.
-To do so, we applied Desai and Fisher's results and fitness variance formula, and considered parameter sets that satisfied the relationship $\sigma^2(N,s,U) \geq s^2 / \log(s/U)$.
-For our simulation parameters, we also restricted our attention to values of $N$, $s$, and $U$ that were typical of widely studied asexaul micro-organisms, or micro-organisms with very limited amounts of sex.
-We examined various literature sources that reported these values as measured in evolutionary experiments \citep{Perfeito2007,desai2007speed,Levy2015}.
-
-Under beneficial mutation-selection balance, the time scale governing fluctuations in the G-matrix is proportional to the inverse of the expected rate of adaptation $v(N,s,2U)$ given by Desai and Fisher's 2007 results.
-In our model, the components of the G-matrix are highly correlated in their dynamics due to their adherence to equation 5.
-The left hand side of this expression is approximately constant in beneficial mutation-selection balance, and as a result, the components of G will on average satisfy a linear relationship.
-The dynamics of the covariance terms coincide with those of the associated one dimensional fitness distribution the two dimensional trait distribution.
-Figure 4b, illustrates the Gaussian shape of the associated one dimensional fitness distribution.
-Its peak represents the largest portion of the population and is composed of classes that lie exactly on the line of constant fitness passing through the population's mean fitness.
-The proportions of the classes within the peak determine the covariance of the two dimensional trait distribution.
-These proportions are approximately fixed in relative size to one another at the time the classes establish when forming part of the stochastic front in a manner describe by \cite{Desai2013}.
-Thus changes in the value of covariance in fact follow the transitions of the peaks the one dimensional traveling wave shift.
-These major transitions are consequently governed by the rate of adaptation and determined at the stochastic front.
-In between the times of these transitions, the covariance will simply be a convex combination of the two adjacent sets of classes that lie on the lines of constant fitness below and above the mean.
-Minor fluctuations in these dynamics are contributed by the residual portion of abundances not included in these two groups.
-
-\begin{figure}
+% *****************************************************************************************************************
+% ***********************************FIGURE*3*COVARIANCE*TIME*SCALES*ARGUMENT*??***********************************
+% *****************************************************************************************************************
+\begin{figure}[h!]
 \begin{subfigure}[b]{1\linewidth}
 \label{a}
-\caption{Two Dimensional Traveling Wave}
-\includegraphics[width=1\linewidth]{fig9a.png}\\
+\caption{Two Dimensional Distribution \& Covariance Fluctations}
+\includegraphics[width=0.48\linewidth]{fig3_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
+\includegraphics[width=0.48\linewidth]{fig9a.png}
 \end{subfigure}
-
 \begin{subfigure}[b]{1\linewidth}
@@
 \end{subfigure}
+\label{Fig 5}
+\caption{\footnotesize Evolution of the G-matrix over time.
+(a) Time dependent trajectories of trait one's variance and covariance are plotted with their mean values.
+Both are highly correlated in their behavior and fluctuate substantially over time.
+(b) Two dimensional trait distribution with two sets of classes identified.
+The first set, whose boxes are shaded in with patterned gray, represent the set of classes with 38 total beneficial mutations.
+At time $t$, the population mean fitness lies on the line of constant fitness given by 38 total beneficial mutations; mean fitness is marked by the black dot.
+At time $t+tau_q$ the mean shifts to the next line of constant fitness associated with 39 total beneficial mutations.
+Here, $\tau_q \sim v^(-1)$ is the mean time between establishements described in \cite{desai2007beneficial}.
+(b) Depictions of the associated one dimensional traveling wave at times $t$ and $t+\tau_q$.
+Due to the Gaussian shape of the one dimensional traveling wave, the covariance between traits is set by proportions of classes within the peak.
+At time $t$ the group with 38 total beneficial mutations determine covariance, while at time $t+\tau_q$ it will be the group with 39 total beneficial mutations that sets covariance.
+In between these times the covariance is approximately a convex combination of its values at these two times.}
+\end{figure}
+% *****************************************************************************************************************
+% ***********************************END*FIGURE*3******************************************************************
+% *****************************************************************************************************************
 
-\label{Fig 5}
-\caption{\footnotesize Fluctuations in the G-matrix, whose components are highly correlated, occur on time scales proportional to the rate of adaptation $v(N,s,2U)$.
-The relationship between the two dimensional trait distribution and its associated one dimensional fitness distribution governs the the dynamics of covariance over time, and changes in the latter are determined by $v$.
-(a) Two sets of classes are distinguished in the two dimensional distribution shown.
-The first set, whose boxes are shaded in with patterned gray, represent the set of classes with 38 total beneficial mutations.
-At some time $t$, the population's mean fitness lies exactly on the line of constant fitness associated with this group indicated by the black dot.
-The second set of classes are shaded in patterned black and they represent the classes with 39 total beneficial mutations.
-At some time $t+tau_q$, the black dot shift and reside on the line of constant fitness associated with this second group.
-(b) The two distributions are the one dimensional traveling waves decribed by Desai and Fisher, which are associated with the two dimensional trait distribution in (a), but for the two distinct time points $t$ and $t+\tau_q$.
-Both are Gaussian, and consequently highly peaked at the population's mean fitness.
-At time $t$ the peak lies exactly on the marker for 38 total beneficial mutations.
-This indicates that the abundances of classes in the first group discussed in (a) largely determine the covariance of the two dimensional trait distribution.
-At time $t+\tau_q$, the covariance shifts to a new value that then determined by the distribution of abundances in second group discussed in (a).
-In between these times the covariance is nearly a convex combination of its values at these two times.
-However, all major transitions in covariance are in fact occurring on the same time scale as advances in the mean, i.e.
-the rate of adaptation.}
-\end{figure}
+Under beneficial mutation-selection balance, the time scale governing the dynamics of the G-matrix components is given by the mean time between establishements $\tau_q \sim 1/v(N,s,2U)$ described by \cite{desai2007beneficial}.
+Figure 3 shows the fluctuations in the variance of trait one, and its covariance with trait two.
+The two are are highly correlated in their dynamics as a result of their adherence to equation 5.
+When the population is beneficial mutation-selection balance, the left hand side of this expression is approximately constant, and as a result, variance and covariance satisfy a linear relationship on average.
+This allows us to describe the dynamics of G by understanding the behavior of covariance over time, which is entirely driven by changes in the associated one dimensional fitness distribution depicted in figure 3c.
+The peak of the traveling wave represents the largest portion of the population.
+The frequencies of classes in this group, which are those whose fitness is equal to the population's mean fitness, to a large extent determine the covariance of the two dimensional distribution.
+After $\tau_q$ generations the mean fitness shift forward, and the peak of the one dimensional traveling wave will then coincide with next group of classes that lie on the next line of constant fitness, whose fitness subsequently equals the mean fitness, as shown in figure 3b.
+These major transitions in covariance are governed by the rate of adaptation.
+In between the times of these transitions, the covariance will simply be a convex combination of the two adjacent sets of classes that lie on the lines of constant fitness below and above the mean.
+Minor fluctuations in these dynamics are contributed by the residual portion of abundances not included in these two groups.\par
 
 \section{Discussion}
+
+An interesting consequence of the dynamics of covariance invovles the relationship between stochastic front and its role in setting the covariance of the two dimensional trait distribution.
+In between establishement events for the one dimensional traveling wave, classes that lie along the line of constant fitness passing through the fittest part of the stochastic front their relative size to one another.
+OneThis process is exactly  manner describe by \cite{Desai2013}.
+Thus changes in the value of covariance in fact follow the transitions of the peaks the one dimensional traveling wave shift.
+These major transitions are consequently governed by the rate of adaptation and determined at the stochastic front.
 
@@
 \bibliography{biblio}
-\bibliographystyle{plain}
+\bibliographystyle{plainnat}
 

### 2017-09-30 plotfunctions.py

@@ -9,6 +9,7 @@ Library of functions used in plots.py and plotdata.py
 #--------FUNCTIONS REQUIRE PACKAGES LISTED:------------------------------------
 from mpl_toolkits.mplot3d import Axes3D
 from scipy.stats import multivariate_normal
+from matplotlib.ticker import FuncFormatter
 from numpy import inf
 import pickle
 import matplotlib.pyplot as plt
@@ -142,6 +143,9 @@ def get_cov_by_fitness_line(genotypes,abundances):
     
     # should return [[fit_i,p_i,cov_i,] for i = min_fit,...,max_fit]
     return fit1Dcovs
+    
+def get_vNSU(N,s,U,n):
+    return 1-(np.log(s/U)**2*np.log(n*s*U*N**2))/(n*np.log(s/n/U)**2*np.log(s*U*N**2))
 
 # -----------------------------------------------------------------------------
 def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param,fname,traitno,num_exp):
@@ -271,11 +275,12 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
         rate_adpt2_avg = var2_avg + cov_avg
         
         # plot data for figure 2a and 2b 
-        fig3a, ax3a = plt.subplots(1,1,figsize=[8,8])
+        fig3a, ax3a = plt.subplots(1,1,figsize=[8,10])
+        fig3a.subplots_adjust(bottom=0.3)
         ax3b = plt.twinx(ax3a)
         ax3a.plot(times,var1_avg,c="black",label='var($r_1|r_2$)=' + str(round(var1_avg[0],7)),linewidth=2.0,linestyle = '--')
         ax3a.plot(times,cov_avg,c="black",label='cov($r_1$,$r_2$)=' + str(round(cov_avg[0],7)),linewidth=2.0,linestyle = '-.')
-        ax3a.plot(times,vU_thry*np.ones(np.shape(var1_avg)),c="black",label='var($r_1$)=' + str(round(vU_thry,7)),linewidth=2.0,linestyle = '-')        
+#        ax3a.plot(times,vU_thry*np.ones(np.shape(var1_avg)),c="black",label='var($r_1$)=' + str(round(vU_thry,7)),linewidth=2.0,linestyle = '-')        
[…]

### 2017-09-30 plots.py

@@ -12,8 +12,8 @@ Script for plots used in "Modeling How Evolution in One Trait is Affected by Ada
 import plotfunctions as pltfun
 
 # load data stored in file generated by plotdata
-#data_name = '_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1'
-data_name = '5'
+data_name = '_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1'
+#data_name = '5'
 #folder_location = 'Documents/kgrel2d/'
 folder_location = ''
 

### 2017-09-30 manuscript.tex
more updates to manuscript and graphs

@@
 \graphicspath{{figures/}}
+\newcommand{\G}{\textbf{G}}
 
@@
 
-For quantitative traits, genetic associations and their effects on adaptation are mathematically described by Lande's equation ($\Delta \bar{z} = \textbf{G} \hspace{.05in}\beta$).
-It states that the change in the population mean $\Delta \bar{z}$ equals the product of the selection gradient $\beta$ and what is known as the \textbf{G}-matrix, whose components are the additive genetic covariances between traits.
-Lande’s result has been used to extend much of our understanding of how phenotypes evolve under selection, yielding a rich set of ideas that emphasize the role of the \textbf{G}-matrix in directing adaptive evolution at both micro- and macro-evolutionary scales.\par
+For quantitative traits, genetic associations and their effects on adaptation are mathematically described by Lande's equation ($\Delta \bar{z} = \G \hspace{.05in}\beta$).
+The expression provides the expected change in the population mean of a trait $\Delta \bar{z}$ as the product of the selection gradient $\beta$ and what is known as the \G-matrix, whose components are the additive genetic covariances between traits.
+Lande’s result has been used to extend much of our understanding of how phenotypes evolve under selection, yielding a rich set of ideas that emphasize the role of the \G-matrix in directing adaptive evolution at both micro- and macro-evolutionary scales.\par
 % * <masel@email.arizona.edu> 2017-07-29T12:43:16.916Z:
@@
 
-The behavior of the \textbf{G}-matrix over time largely depends on what are the causes of the genetic correlations.
+The behavior of the \G-matrix over time largely depends on what are the causes of the genetic correlations.
 Correlations between quantitative traits can arise for a number of reasons, but they often have a genetic basis \citep{Saltz2017}.
@@
 Geneticists in this group believed that linkage disequilibria was much more important; a fact that often reflected in their experiments.
-Presently, the Edinburgh view is more widely accepted, and as a result, most models for the evolution of the \textbf{G}-matrix assume genetic correlations are pleiotropic and proceed with examining how evolutionary processes influence them.
+Presently, the Edinburgh view is more widely accepted, and as a result, most models for the evolution of the \G-matrix assume genetic correlations are pleiotropic and proceed with examining how evolutionary processes influence them.
 We know far less about the about the behavior genetic correlations from linkage disequilibria over time, but this is not much of an issue for most complex organisms.\par
@@
 % 
-% >  most models for the evolution of the \textbf{G}-matrix assume genetic correlations are pleiotropic and proceed with examining how evolutionary processes influence them
+% >  most models for the evolution of the \G-matrix assume genetic correlations are pleiotropic and proceed with examining how evolutionary processes influence them
 % An important exception to note is the "Bulmer effect", which I believe describes how selection can cause LD even in the absence of linkage, which then of course appears in G.
@@
 Lande’s equation describes the phenotypic evolution of populations with multiple characters from generation to generation.
-Whether one can use this framework to uncover the history of natural selection in populations over large evolutionary timescales largely depends on the forces that influence the evolution of the G-matrix.
-Over the past few decades there has been a significant amount of effort to shed light on the matter of G-matrix stability.
+Whether one can use this framework to uncover the history of natural selection in populations over large evolutionary timescales largely depends on the forces that influence the evolution of the \G-matrix.
+Over the past few decades there has been a significant amount of effort to shed light on the matter of \G-matrix stability.
 The analysis of genetic models for the adaptive evolution of multiple phenotypic traits indicate some of the conditions needed for stability\citep{Turelli1988,Jones2003}.
-These works exclude the potential influences of linkage disequilibrium on the dynamics of variances and covariances over time, and as a result, our understanding of the role of linkage disequilibria in G-matrix evolution remains unclear.
-Nonetheless, the question is significant there is empirical evidence suggesting that in the G-matrix is often conserved in natural populations \citep{Arnold1999,Roff2000,Steppan2002}.
-However, much of what we know about G-matrix stability corresponds the examining evolutionary forces on
+These works exclude the potential influences of linkage disequilibrium on the dynamics of variances and covariances over time, and as a result, our understanding of the role of linkage disequilibria in \G-matrix evolution remains unclear.
+Nonetheless, the question is significant there is empirical evidence suggesting that in the \G-matrix is often conserved in natural populations \citep{Arnold1999,Roff2000,Steppan2002}.
+However, much of what we know about \G-matrix stability corresponds the examining evolutionary forces on
 
@@
 % *****************************************************************************************************************
-\begin{figure}
+\begin{figure}[h!]
 \includegraphics[width=1\linewidth]{fig1a.pdf}
@@
 
-The dynamics of the bulk was modeled by a system of ordinary differential equations, in which abundances changed in proportion to their selective advantage (\textbf{eqt. 1}). 
+The dynamics of the bulk was modeled by a system of ordinary differential equations, in which abundances changed in proportion to their selective advantage \eqref{eq:1}. 
 \begin{equation} \label{eq:1}
@@
 We can apply Lande's result to the two components $\dot{\bar{r_1}} $ and $\dot{\bar{r_2}} $ and rewrite them in terms of the trait variances $\sigma_1^2 = \sum_{ij} p_{ij} (r_i-\bar{r_1})^2$ and $\sigma_2^2 = \sum_{ij} p_{ij} (r_j-\bar{r_2})^2$, and their covariance $\sigma_{12} =\sum_{ij} p_{ij} (r_i-\bar{r_1})(r_j-\bar{r_2})$.
-In doing so, we obtain the system expressed in equation 6.
+In doing so, we obtain the system expressed in \eqref{eq:5}.
 % * <masel@email.arizona.edu> 2017-08-22T22:41:29.374Z:
@@
 Right now the number doesn't match.
-You should also link this more clearly to the way it was presented in the introduction, ie you should explain that this is simply $\Delta \bar{z} = \textbf{G} \hspace{.05in}\beta$ in the case of two traits both contributing directly to fitness
+You should also link this more clearly to the way it was presented in the introduction, ie you should explain that this is simply $\Delta \bar{z} = \G \hspace{.05in}\beta$ in the case of two traits both contributing directly to fitness
 % 
@@
 We did this by identifying the first mutant lineage destined to establish among all mutants appearing in the stochastic front.
-Mutants awhich are parallel to the line shownwhich are parallel to the line shownppear according to a time-inhomogeneous Poisson process with rate functions given by the abundances of the adjacent classes from which they originated.
+Mutants awhich are parallel to the line shownwhich are parallel to the line shown appear according to a time-inhomogeneous Poisson process with rate functions given by the abundances of the adjacent classes from which they originated.
 Consequently, the arrival time $\tau_l$ of the $l^{th}$ mutant generated from the class $n_{ij}(t)$ having a mutation in trait $k$ can be used to form the random variable
@@
 By randomly sampling from the this distribution, we could then solve for each arrival time $\tau_l$.
-We then computed the probability of fixation using \textbf{equation 5}.
+We then computed the probability of fixation using \eqref{eq:5}.
 The integral was numerically approximated up to an appropriate cutoff.
 To determine if a mutation would establish, we sampled a uniform random variable and checked it against the calculated $\pi_{fix}$.
-The next class to establish was chosen accordingly and incorporated into the bulk with a random initial size generated with from \textbf{equation 4}.\par
+The next class to establish was chosen accordingly and incorporated into the bulk with a random initial size generated with from \eqref{eq:4}.\par
 % * <masel@email.arizona.edu> 2017-08-22T22:57:52.494Z:
@@
 The data was used to compute the marginal mean fitness and variances of each trait, as well as their covariance.
-From these quantities we also computed the instantaneous rate of adaptation (\textbf{eqtn.
-5}), the mean rate of fitness increase in each trait and the overall rate of adaptation.
+From these quantities we also computed the instantaneous rate of adaptation \eqref{eq:5}, the mean rate of fitness increase in each trait and the overall rate of adaptation.
 
@@
 However, our simulations show that upon adding the second trait, the net change in $v_1$ arises from an increase in the first trait's variance and the formation of negative covariance.
-The first results in an increase in $v_1$, as a direct consequence of equation 5.
+The first results in an increase in $v_1$, as a direct consequence of \eqref{eq:5}.
 This gain is more than offset by the negative covariance created between the two traits, whose magnitude is always substantially larger than the variance increase.
@@
 % *****************************************************************************************************************
-\begin{figure}[h]
+\begin{figure}[h!]
 \begin{subfigure}[b]{1\linewidth}
@@
 \label{Fig 5}
-\caption{\footnotesize Evolution of the G-matrix over time.
+\caption{\footnotesize Evolution of the \G-matrix over time.
 (a) Time dependent trajectories of trait one's variance and covariance are plotted with their mean values.
@@
 
-Under beneficial mutation-selection balance, the time scale governing the dynamics of the G-matrix components is given by the mean time between establishements $\tau_q \sim 1/v(N,s,2U)$ described by \cite{desai2007beneficial}.
+Under beneficial mutation-selection balance, the time scale governing the dynamics of the \G-matrix components is given by the mean time between establishements $\tau_q \sim 1/v(N,s,2U)$ described by \cite{desai2007beneficial}.
 Figure 3 shows the fluctuations in the variance of trait one, and its covariance with trait two.
-The two are are highly correlated in their dynamics as a result of their adherence to equation 5.
+The two are are highly correlated in their dynamics as a result of their adherence to \eqref{eq:5}.
 When the population is beneficial mutation-selection balance, the left hand side of this expression is approximately constant, and as a result, variance and covariance satisfy a linear relationship on average.
-This allows us to describe the dynamics of G by understanding the behavior of covariance over time, which is entirely driven by changes in the associated one dimensional fitness distribution depicted in figure 3c.
+This allows us to describe the dynamics of \G-matrix by understanding the behavior of covariance over time, which is entirely driven by changes in the associated one dimensional fitness distribution depicted in figure 3c.
 The peak of the traveling wave represents the largest portion of the population.

### 2017-09-30 plotfunctions.py

@@ -347,17 +347,19 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
         [end1,end2,end3] = [num_exp/3,2*num_exp/3,num_exp]
         NsUparam = np.asarray(NsUparam)      
         
+        formatter = FuncFormatter(to_percent)
+        
         fig5d, ax5d = plt.subplots(1,1,figsize=[8,8])
         ax5d.plot(NsUparam[start1:end1,1],-np.asarray(covp[start1:end1])-np.asarray(varp[start1:end1])+np.asarray(vUthryp[start1:end1]),c="black",label='$v_1$',linewidth=2.0,linestyle = '-')                
-        ax5d.plot(NsUparam[start1:end1,1],np.asarray(varp[start1:end1]),c="red",label='$\sigma_{1}^2$',linewidth=2.0,linestyle = ':')
-        ax5d.plot(NsUparam[start1:end1,1],np.asarray(covp[start1:end1]),c="blue",label='$\sigma_{12}$',linewidth=2.0,linestyle = '-.')        
+        ax5d.plot(NsUparam[start1:end1,1],np.asarray(varp[start1:end1]),c="red",label='$\sigma_{1}^2$',linewidth=2.0,linestyle = '-')
+        ax5d.plot(NsUparam[start1:end1,1],np.asarray(covp[start1:end1]),c="blue",label='$\sigma_{12}$',linewidth=2.0,linestyle = '-')        
         ax5d.axhline(linewidth=0.5, color = 'k')
-        ax5d.yaxis.set_major_formatter(FuncFormatter(lambda y, _: '{:.0%}'.format(y)))         
+        ax5d.yaxis.set_major_formatter(formatter)         
         ax5d.set_ylabel('Scaled Variance-Covariance',fontsize=18)
         ax5d.set_xlabel('selection coeff ($10^{-3} - 2 x 10^{-2}$)',fontsize=18)
         ax5d.tick_params(axis='both',labelsize=14)
         ax5d.ticklabel_format(style='sci',axis='x',scilimits=(0,0))
-        ax5d.set_ylim((-2,2))
+        ax5d.set_ylim((-2.5,2.5))
         ax5d.set_xlim((min(NsUparam[start1:end1,1]),max(NsUparam[start1:end1,1])))  
         ax5d.legend(loc=7)
         fig5d.savefig('./'+folder_location+'figures/'+fname+data_name+'d.pdf')
[…]

### 2017-10-01 manuscript.tex
more updates to plots for figure 3, and completed more of its description.

@@
 
-\section{Introduction}
+\section*{Introduction}
+\label{sec:introduction}
 
@@
 
-\section{Materials and Methods}
+\section*{Materials and Methods}
 \label{sec:materials:methods}
@@
 \label{Figure 1.}
-\caption{\footnotesize Evolutuion of a two dimensional traveling wave.
-The distribution depicted was sampled from simulations.
-Each box represents a class composed of individuals with an identical number of beneficial mutations in each trait.
-Classes with solid coloring form part of the bulk, while those with patterned interiors represent the stochastic front.
-The red dot marks the mean number of beneficial mutations in each trait.
-The dashed red line represents a lines of constant fitness.
-Classes whose centers lie along this line have relative fitness equal to the population mean, but vary in the number of beneficial mutations that have accumulated in each trait.
-Relative fitness increases along the perpendicular direction.
-Classes .
-Beneficial mutations and random sampling produce negligible effects in the changes of their abundances.
-In asexuals, only beneficial mutations that occur on the fittest backgrounds produce lineages that may eventually fix.
-These are marked with arrows into the stochastic front.
-Other lineages are out-competed and remain insignificant in size with respect to bulk.
-As new classes, the bulk expands in the direction of increasing relative fitness, and classes below the red line shrink and eventually disappear.
-As a result, the two dimensional trait distribution continually shifts in the direction of increasing relative fitness over time.
+\caption{\footnotesize Illustration of a two dimensional trait distribution taken from simulation.
+Boxes correspond to classes whose members have identical numbers of beneficial mutations in each trait.
+The bulk is composed of classes with solid interiors, while those with patterned interiors form part of the stochastic front.
+The red dot marks the number of beneficial mutations in the average individual.
+Classes are grouped with respect to their total relative fitness, but vary in the number of beneficial mutations that have accumulated in each trait.
+These groups are represented by the lines passing through their centers that are parallel to the dashed line.
+These are refered to lines of constant fitness.
+The red line marks the set of classes whose fitness equals the population mean fitness.
+Fitness increases along the perpendicular direction to this line.
+Only beneficial mutations occuring on fit backgrounds produce lineages that grow sufficiently large to affect adaptation.
+These are represented by arrows from the bulk into the stochastic front.
+Other mutant lineages are out-competed and do not significantly contribute to the population's adapatation.
+The stochastict front expands in the direction of increasing relative fitness, while classes in the bulk below the red line disappear.
+These two processes continually shift the two dimensional trait distribution in the direction of increasing fitness over time.
 (Simulation parameters: $N=10^9$, $s=10^{-2}$, and $U=10^{-5}$)}
@@
 
-\section{Results}
-Adaptation in a second trait
+\section*{Results}
+\label{sec:results}
+Much Additional traits
 
-In each of our simulations we observed a 40\% decrease in the rate of adaptation of a trait as a result adding a second.
+In each of our simulations we observed a 60\% decrease in the rate of adaptation of a trait as a result adding a second.
 The decrease was consistent across three parameter sets which varied either $N$, $s$ and $U$.
@@
 \begin{figure}[h!]
-\begin{subfigure}[b]{1\linewidth}
+\begin{subfigure}[b!]{0.48\linewidth}
 \centering
-\includegraphics[width=.48\linewidth]{fig5d.pdf}
-\includegraphics[width=.48\linewidth]{fig5e.pdf}
+\includegraphics[width=1\linewidth]{fig5d.pdf}
+\caption{}
 \end{subfigure}
-\begin{subfigure}[b]{1\linewidth}
+\begin{subfigure}[b!]{0.48\linewidth}
 \centering
-\includegraphics[width=.48\linewidth]{fig5f.pdf}
-\includegraphics[width=.48\linewidth]{fig5g.pdf}
+\includegraphics[width=1\linewidth]{fig5e.pdf}
+\caption{}
 \end{subfigure}
+
+\begin{subfigure}[b]{0.48\linewidth}
+\centering
+\includegraphics[width=1\linewidth]{fig5f.pdf}
+\caption{}
+\end{subfigure}
+\begin{subfigure}[b]{0.48\linewidth}
+\centering
+\includegraphics[width=1\linewidth]{fig5g.pdf}
+\caption{}
+\end{subfigure}
+
 \label{fig}
-\caption{\footnotesize Contributions to rate of adaptation from changes in trait variance and covariance.
-(a) $s$ varies with $N$ and $U$ being kept fixed ($N= 10^7$, $U=10^-5$).
-(b) $U$ varies with $N$ and $s$ being kept fixed ($N= 10^7$,$s=10^-2$).
-(c) $N$ varies with $s$ and $U$ being kept fixed ($s=10^-2$,$U=10^-5$).}
+\caption{\footnotesize 
+(a-c) Magnitudes of the mean variance, covariance, and rate of adaptation of a trait divided by its expected rate of adaptation when evolving without a second.
+Each plot includes the measured values produced by simulations with either varying $s$, $U$, or $N$.
+(d) Expected decrease in the rate of adapatation of trait, when additional traits are added for three choices of $s$.
+Changes in the scaled mean variances and covarainces vary only slightly with changes in each of three parameters.
+The decrease in a trait's rate of adaptation is nearly 60\% across changes in each of the three parameters.
+Fixed simulation parameters in (a-c) are: $s=10^-2$, $U=10^-5$ and $N=10^9$.}
 \end{figure}
@@
 
-In our simulations, we observed highly correlated behavior between the variance trait one and its covariance with the second.
-Their fluctuations shown in figure 3 are nearly in sync throughout duration of the simulations.
-This behavior can be explained by the effects that covariance has on the distribution of individual traits.
-As previously mentioned, negative covariance measure associations between fitter values of one trait with poorer one in the other.
-Consequently, the larger the magnitude of the negative covariance, the more likely it is that we will observe extreme values in each trait, with each these associated with the opposite opposite versions in the other trait.
-(This section still needs more work)\par
-
 % *****************************************************************************************************************
@@
 \begin{figure}[h!]
+\begin{subfigure}[b!]{0.48\linewidth}
+\label{fig:3a}
+\includegraphics[width=1\linewidth, height=6cm]{fig3_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
+\caption{}
+\end{subfigure}
+\begin{subfigure}[b!]{0.48\linewidth}
+\label{fig:3b}
+\includegraphics[width=1\linewidth, height=6cm]{fig9a.png}
+\caption{}
+\end{subfigure}
+
 \begin{subfigure}[b]{1\linewidth}
-\label{a}
-\caption{Two Dimensional Distribution \& Covariance Fluctations}
-\includegraphics[width=0.48\linewidth]{fig3_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
-\includegraphics[width=0.48\linewidth]{fig9a.png}
+\label{fig:3c}
+\includegraphics[width=1\linewidth]{fig9b.png}
+\caption{}
 \end{subfigure}
-\begin{subfigure}[b]{1\linewidth}
-\label{b}
-\caption{Abundances Along Axis of Total Relative Fitness}
-\includegraphics[width=1\linewidth]{fig9b.png}
-\end{subfigure}
-\label{Fig 5}
+\label{fig:3}
 \caption{\footnotesize Evolution of the \G-matrix over time.
-(a) Time dependent trajectories of trait one's variance and covariance are plotted with their mean values.
-Both are highly correlated in their behavior and fluctuate substantially over time.
-(b) Two dimensional trait distribution with two sets of classes identified.
-The first set, whose boxes are shaded in with patterned gray, represent the set of classes with 38 total beneficial mutations.
-At time $t$, the population mean fitness lies on the line of constant fitness given by 38 total beneficial mutations; mean fitness is marked by the black dot.
-At time $t+tau_q$ the mean shifts to the next line of constant fitness associated with 39 total beneficial mutations.
-Here, $\tau_q \sim v^(-1)$ is the mean time between establishements described in \cite{desai2007beneficial}.
-(b) Depictions of the associated one dimensional traveling wave at times $t$ and $t+\tau_q$.
-Due to the Gaussian shape of the one dimensional traveling wave, the covariance between traits is set by proportions of classes within the peak.
-At time $t$ the group with 38 total beneficial mutations determine covariance, while at time $t+\tau_q$ it will be the group with 39 total beneficial mutations that sets covariance.
-In between these times the covariance is approximately a convex combination of its values at these two times.}
+(a) Time dependent trajectories of variance and covariance with their mean values (constant lines).
+(b) Two dimensional trait distribution with classes having 38 and 39 total beneficial mutations identified.
+(c) Depictions of the associated one dimensional traveling wave at times $t$ and $t+\tau_q$.
+Major transitions in their values occur over a period of $\tau_q \sim v^{-1}$ described by \citet{desai2007beneficial}.
+At time $t$ the mean fitness, marked by the black dot, resides on the line associted with classes with 38 total beneficial mutations (a).
+These classes compose the peak of the one dimensional Gaussian fitness distribution which is exponentially larger than all others bins.
+As a result, their frequencies determine variance and covariance.
+At time $t+tau_q$ the mean shifts to the next line of constant fitness for 39 total beneficial mutations, and their frequencies will then determine variance and covariance.
+In between these times, their values are approximately a convex combinations of those at the time points.}
 \end{figure}
@@
 
-Under beneficial mutation-selection balance, the time scale governing the dynamics of the \G-matrix components is given by the mean time between establishements $\tau_q \sim 1/v(N,s,2U)$ described by \cite{desai2007beneficial}.
+Under beneficial mutation-selection balance the time scale governing the dynamics of the \G-matrix components is given by the mean time between establishements $\tau_q \sim 1/v(N,s,2U)$ described by \cite{desai2007beneficial}.
 Figure 3 shows the fluctuations in the variance of trait one, and its covariance with trait two.
@@
 
-\section{Discussion}
+\section*{Discussion}
+\label{sec:discussion}
 
@@
 
-\subsection{Comparison with Barton and Otto's Negative Linkage Disequilibria Results}
+\subsection*{Comparison with Barton and Otto's Negative Linkage Disequilibria Results}
 In our model, covariance between traits is arises entirely from linkage disequilibrium.
@@
 
+\section*{Acknowledgments}
+\label{sec:acknowledgments}
+
 \bibliography{biblio}

### 2017-10-01 plotfunctions.py

@@ -9,8 +9,8 @@ Library of functions used in plots.py and plotdata.py
 #--------FUNCTIONS REQUIRE PACKAGES LISTED:------------------------------------
 from mpl_toolkits.mplot3d import Axes3D
 from scipy.stats import multivariate_normal
-from matplotlib.ticker import FuncFormatter
 from numpy import inf
+import matplotlib.ticker as mtick
 import pickle
 import matplotlib.pyplot as plt
 import scipy as sp
@@ -276,7 +276,7 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
         
         # plot data for figure 2a and 2b 
         fig3a, ax3a = plt.subplots(1,1,figsize=[8,10])
-        fig3a.subplots_adjust(bottom=0.3)
+        fig3a.subplots_adjust(bottom=0.35)
         ax3b = plt.twinx(ax3a)
         ax3a.plot(times,var1_avg,c="black",label='var($r_1|r_2$)=' + str(round(var1_avg[0],7)),linewidth=2.0,linestyle = '--')
         ax3a.plot(times,cov_avg,c="black",label='cov($r_1$,$r_2$)=' + str(round(cov_avg[0],7)),linewidth=2.0,linestyle = '-.')
@@ -286,9 +286,9 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
         ax3a.set_xlabel('Time (Generations)',fontsize=18)
         ax3a.set_ylim((-3e-4,4e-4))
         ax3a.set_xlim((1e4,2e4))
-        ax3a.tick_params(axis='both',labelsize=14)
+        ax3a.tick_params(axis='both',labelsize=18)
         ax3a.ticklabel_format(style='sci',axis='both',scilimits=(0,0))
-        ax3a.legend(loc='upper center', bbox_to_anchor=(0.5, -0.2),fancybox=True, shadow=True, ncol=2)
+        ax3a.legend(loc='upper center', bbox_to_anchor=(0.5, -0.2),fancybox=True, shadow=True, ncol=1,fontsize=20)
         
         ax3b.plot(times,fit_var[:,0],c="black",linestyle="--",linewidth=1.0)
         ax3b.plot(times,fit_cov[:],c="black",linestyle="-.",linewidth=1.0)
[…]

### 2017-10-01 plots.py

@@ -12,8 +12,8 @@ Script for plots used in "Modeling How Evolution in One Trait is Affected by Ada
 import plotfunctions as pltfun
 
 # load data stored in file generated by plotdata
-data_name = '_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1'
-#data_name = '5'
+#data_name = '_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1'
+data_name = '5'
 #folder_location = 'Documents/kgrel2d/'
 folder_location = ''
 

### 2017-10-01 manuscript.tex
updates to manuscript

@@
 \title{Modeling How Evolution in One Trait is Affected by Adaptation in Another}
-\date{Fall 2016}
+\date{Fall 2017}
 \author{Kevin Gomez,Jason Bertram,Joanna Masel}
@@
 \includegraphics[width=1\linewidth]{fig1a.pdf}
-\label{Figure 1.}
 \caption{\footnotesize Illustration of a two dimensional trait distribution taken from simulation.
@@
 These two processes continually shift the two dimensional trait distribution in the direction of increasing fitness over time.
-(Simulation parameters: $N=10^9$, $s=10^{-2}$, and $U=10^{-5}$)}
+(Simulation parameters: $N=10^9$, $s=10^{-2}$, and $U=10^{-5}$)}\label{fig:1}
 \end{figure}
@@
 In our model, we find it more useful to consider the related random variable $\nu = \frac{1}{s_{ij}}\exp[-\Delta r_{ij}\tau]$ representing an initial size that produces the deterministic trajectory observed after establishment.
-Using the expression given by Desai and Fisher for the PDF of $\tau$, we can arrive at the cumulative distribution of  $\nu$ (equation 5) by following the analysis of \cite{Uecker2011}.
+Using the expression given by Desai and Fisher for the PDF of $\tau$, we can arrive at the cumulative distribution of  $\nu$ (equation 5) by following the analysis of \citet{Uecker2011}.
 % * <masel@email.arizona.edu> 2017-08-22T22:14:21.200Z:
@@
 \label{sec:results}
-Much Additional traits
+As expected, we found that an additional trait led to a decrease in the rate of adaptation in the first.
+We found that the reduction was primarily caused by a build of negative covariance between the two traits.
+When the second trait is allowed to involve with the first, the variance of the focal trait increases.
+The magnitude of this increase is shown in Figure \ref{fig:2} in excess of the due to the appearance of genotypes buildup of linkage disequibria which inhibits the purg
 
-In each of our simulations we observed a 60\% decrease in the rate of adaptation of a trait as a result adding a second.
+addition of a trait leads to 60\% decrease in the rate of adaptation of a trait as a result adding a second.
 The decrease was consistent across three parameter sets which varied either $N$, $s$ and $U$.
@@
 \includegraphics[width=1\linewidth]{fig5d.pdf}
-\caption{}
+\caption{}\label{fig:2a}
 \end{subfigure}
@@
 \includegraphics[width=1\linewidth]{fig5e.pdf}
-\caption{}
+\caption{}\label{fig:2b}
 \end{subfigure}
@@
 \includegraphics[width=1\linewidth]{fig5f.pdf}
-\caption{}
+\caption{}\label{fig:2c}
 \end{subfigure}
@@
 \includegraphics[width=1\linewidth]{fig5g.pdf}
-\caption{}
+\caption{}\label{fig:2d}
 \end{subfigure}
-
-\label{fig}
 \caption{\footnotesize 
-(a-c) Magnitudes of the mean variance, covariance, and rate of adaptation of a trait divided by its expected rate of adaptation when evolving without a second.
+(a-c) Magnitudes of the mean variance, covariance, and rate of adaptation for trait one divided by its expected rate of adaptation when evolving without a second.
 Each plot includes the measured values produced by simulations with either varying $s$, $U$, or $N$.
@@
 The decrease in a trait's rate of adaptation is nearly 60\% across changes in each of the three parameters.
-Fixed simulation parameters in (a-c) are: $s=10^-2$, $U=10^-5$ and $N=10^9$.}
+Fixed simulation parameters in (a-c) are: $s=10^{-2}$, $U=10^{-5}$ and $N=10^9$.}\label{fig:2}
 \end{figure}
@@
 \begin{subfigure}[b!]{0.48\linewidth}
-\label{fig:3a}
 \includegraphics[width=1\linewidth, height=6cm]{fig3_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
-\caption{}
+\caption{}\label{fig:3a}
 \end{subfigure}
 \begin{subfigure}[b!]{0.48\linewidth}
-\label{fig:3b}
 \includegraphics[width=1\linewidth, height=6cm]{fig9a.png}
-\caption{}
+\caption{}\label{fig:3b}
 \end{subfigure}
@@
 \begin{subfigure}[b]{1\linewidth}
-\label{fig:3c}
 \includegraphics[width=1\linewidth]{fig9b.png}
-\caption{}
+\caption{}\label{fig:3c}
 \end{subfigure}
-\label{fig:3}
 \caption{\footnotesize Evolution of the \G-matrix over time.
@@
 At time $t+tau_q$ the mean shifts to the next line of constant fitness for 39 total beneficial mutations, and their frequencies will then determine variance and covariance.
-In between these times, their values are approximately a convex combinations of those at the time points.}
+In between these times, their values are approximately a convex combinations of those at the time points.}\label{fig:3}
 \end{figure}

### 2017-10-01 plotfunctions.py

@@ -287,11 +287,12 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
         ax3a.set_ylim((-3e-4,4e-4))
         ax3a.set_xlim((1e4,2e4))
         ax3a.tick_params(axis='both',labelsize=18)
+        ax3a.grid(b='on', which='both', axis='both')
         ax3a.ticklabel_format(style='sci',axis='both',scilimits=(0,0))
         ax3a.legend(loc='upper center', bbox_to_anchor=(0.5, -0.2),fancybox=True, shadow=True, ncol=1,fontsize=20)
         
-        ax3b.plot(times,fit_var[:,0],c="black",linestyle="--",linewidth=1.0)
-        ax3b.plot(times,fit_cov[:],c="black",linestyle="-.",linewidth=1.0)
+        ax3b.plot(times,fit_var[:,0],c="black",linestyle="--",linewidth=2.0)
+        ax3b.plot(times,fit_cov[:],c="black",linestyle="-.",linewidth=2.0)
         ax3b.axhline(linewidth=0.5, color = 'k')        
         ax3b.set_ylim((-3e-4,4e-4))
         ax3b.set_xlim((1e4,2e4))        
@@ -353,16 +354,17 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
         fig5d, ax5d = plt.subplots(1,1,figsize=[8,9])
         fig5d.subplots_adjust(bottom=0.3)
         fig5d.subplots_adjust(left=0.2)        
-        ax5d.plot(NsUparam[start1:end1,1],100*(-np.asarray(covp[start1:end1])-np.asarray(varp[start1:end1])+np.asarray(vUthryp[start1:end1])),c="black",label='$v_{1|2}$',linewidth=2.0,linestyle = '-')                
-        ax5d.plot(NsUparam[start1:end1,1],100*np.asarray(varp[start1:end1]),c="red",label='var($r_1|r_2$)',linewidth=2.0,linestyle = '-')
-        ax5d.plot(NsUparam[start1:end1,1],-100*np.asarray(covp[start1:end1]),c="blue",label='|cov($r_1,r_2$)|',linewidth=2.0,linestyle = '-')        
[…]

### 2017-10-01 manuscript.tex
more simulations started and fixed data collection to speed up simulations.

@@
 \caption{\footnotesize 
-(a-c) Magnitudes of the mean variance, covariance, and rate of adaptation for trait one divided by its expected rate of adaptation when evolving without a second.
+(a-c) Magnitudes of changes in mean variance, covariance, and rate of adaptation of trait one divided by its expected rate of adaptation when evolving without the second.
 Each plot includes the measured values produced by simulations with either varying $s$, $U$, or $N$.
-(d) Expected decrease in the rate of adapatation of trait, when additional traits are added for three choices of $s$.
-Changes in the scaled mean variances and covarainces vary only slightly with changes in each of three parameters.
-The decrease in a trait's rate of adaptation is nearly 60\% across changes in each of the three parameters.
+(d) The expected decrease in trait one's rate of adapatation for various numbers of traits added.
+Three choices of $s$ are shown to depict its effect.
+Adaptation in a single trait is affected in two ways when evolving with a second trait.
+First, its variance increases due to classes with poorer backgrounds in the first trait persisting longer when they are fit in the second.
+In (a-c), the increase (dashed line) ranges from 60\% to 100\% depending on the parameter.
+The second effect, involves the formation of negative covariance between the two traits.
+Its magnitude (dotted line) is substantially larger, ranging from 80\% to 140\%, and always 40\% larger than the increase in variance.
+This results in a net decrease of 40\% in the rate of adaptation (solid line).
 Fixed simulation parameters in (a-c) are: $s=10^{-2}$, $U=10^{-5}$ and $N=10^9$.}\label{fig:2}

### 2017-10-01 plotfunctions.py

@@ -145,7 +145,8 @@ def get_cov_by_fitness_line(genotypes,abundances):
     return fit1Dcovs
     
 def get_vNSU(N,s,U,n):
-    return 1-(np.log(s/U)**2*np.log(n*s*U*N**2))/(n*np.log(s/n/U)**2*np.log(s*U*N**2))
+    vrate = 1-(np.log(s/U)**2*(2*np.log(N*s)-np.log(s/n/U)))/(n*np.log(s/n/U)**2*(2*np.log(N*s)-np.log(s/U))) 
+    return vrate
 
 # -----------------------------------------------------------------------------
 def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param,fname,traitno,num_exp):
@@ -278,8 +279,8 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
         fig3a, ax3a = plt.subplots(1,1,figsize=[8,10])
         fig3a.subplots_adjust(bottom=0.35)
         ax3b = plt.twinx(ax3a)
-        ax3a.plot(times,var1_avg,c="black",label='var($r_1|r_2$)=' + str(round(var1_avg[0],7)),linewidth=2.0,linestyle = '--')
-        ax3a.plot(times,cov_avg,c="black",label='cov($r_1$,$r_2$)=' + str(round(cov_avg[0],7)),linewidth=2.0,linestyle = '-.')
+        ax3a.plot(times,var1_avg,c="black",label='var($r_1|r_2$)=' + str(round(var1_avg[0],7)),linewidth=3.0,linestyle = '--')
+        ax3a.plot(times,cov_avg,c="black",label='cov($r_1$,$r_2$)=' + str(round(cov_avg[0],7)),linewidth=3.0,linestyle = ':')
 #        ax3a.plot(times,vU_thry*np.ones(np.shape(var1_avg)),c="black",label='var($r_1$)=' + str(round(vU_thry,7)),linewidth=2.0,linestyle = '-')        
         ax3a.axhline(linewidth=0.5, color = 'k')
         ax3a.set_ylabel('Fitness Variances & Covariance',fontsize=18)
@@ -291,8 +292,8 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
         ax3a.ticklabel_format(style='sci',axis='both',scilimits=(0,0))
[…]

### 2017-10-04 manuscript.tex
more updates

@@
 \graphicspath{{figures/}}
-\newcommand{\G}{\textbf{G}}
+\newcommand{\G}{\textbf{G }}
 
@@
 For quantitative traits, genetic associations and their effects on adaptation are mathematically described by Lande's equation ($\Delta \bar{z} = \G \hspace{.05in}\beta$).
-The expression provides the expected change in the population mean of a trait $\Delta \bar{z}$ as the product of the selection gradient $\beta$ and what is known as the \G-matrix, whose components are the additive genetic covariances between traits.
-Lande’s result has been used to extend much of our understanding of how phenotypes evolve under selection, yielding a rich set of ideas that emphasize the role of the \G-matrix in directing adaptive evolution at both micro- and macro-evolutionary scales.\par
+The expression provides the expected change in the population mean of a trait $\Delta \bar{z}$ as the product of the selection gradient $\beta$ and what is known as the \G  matrix, whose components are the additive genetic covariances between traits.
+Lande’s result has been used to extend much of our understanding of how phenotypes evolve under selection, yielding a rich set of ideas that emphasize the role of the \G  matrix in directing adaptive evolution at both micro- and macro-evolutionary scales.\par
 % * <masel@email.arizona.edu> 2017-07-29T12:43:16.916Z:
@@
 
-The behavior of the \G-matrix over time largely depends on what are the causes of the genetic correlations.
-Correlations between quantitative traits can arise for a number of reasons, but they often have a genetic basis \citep{Saltz2017}.
-They arise from either pleiotropy, linkage disequilibria, or are the product of trans acting elements; the first two that are thought to be the most significant factors.
-Historically, there have been two schools of thought what the principle causes of genetic correlations are.
-The first was put forth by the Edinburgh school, which held that pleiotropy was principle source, since linkage disequilibrium could rapidly decay by recombination \citep[Chapter~20]{fox2006evolutionary}.
-This belief was largely motivated by interests the relationships between parent-offspring phenotypes within randomly mating populations, and its implications for evolution.
-The alternative view came out of the Birmingham school, which largely experimented with inbred lines of plants.
-Geneticists in this group believed that linkage disequilibria was much more important; a fact that often reflected in their experiments.
-Presently, the Edinburgh view is more widely accepted, and as a result, most models for the evolution of the \G-matrix assume genetic correlations are pleiotropic and proceed with examining how evolutionary processes influence them.
-We know far less about the about the behavior genetic correlations from linkage disequilibria over time, but this is not much of an issue for most complex organisms.\par
+The behavior of \G over time largely depends on what are the causes of the genetic correlations between traits.
+These can arise for different reasons, but often have a genetic basis \citep{Saltz2017}.
+Pleiotropy and linkage disequilibria are by far the most common causes.
+Historically, there have been two schools of thought on which of these underlie most of the genetic correlations detected.
+Pleiotropy was favored by the Edinburgh school, which adopted the view that recombination could quickly eliminate linkage disequilibrium and remove its effects \citep[Chapter~20]{fox2006evolutionary}.
+This belief was largely motivated by their interests in relationships between parent-offspring phenotypes within randomly mating animal populations.
+In such groups pleiotropy is indeed the most prominent cause of genetic correlations that can persist over evolutionary time scales.
+An alternative view was put forth by the Birmingham school.
+Quantitative geneticists in this group largely worked on inbred lines of plants, and as result, where often confronted with the effects of linkage disequilibria in the results of their experiments.
+Presently, the Edinburgh view is more widely accepted, and as a result, most models for the evolution of the \G  matrix assume genetic correlations are pleiotropic and proceed with examining how evolutionary processes influence them.
+We know far less about the about the dynamics of genetic correlations arising from linkage disequilibria.\par
 % * <masel@email.arizona.edu> 2017-08-22T16:50:23.214Z:
 % 
-% >  most models for the evolution of the \G-matrix assume genetic correlations are pleiotropic and proceed with examining how evolutionary processes influence them
+% >  most models for the evolution of the \G  matrix assume genetic correlations are pleiotropic and proceed with examining how evolutionary processes influence them
 % An important exception to note is the "Bulmer effect", which I believe describes how selection can cause LD even in the absence of linkage, which then of course appears in G.
@@
 
-In the case of asexuals, the Birmingham view is more appropriate, and determining how the genetic correlations due to LD change over time is critical to our understanding of adaptation.
-Linkage disequilibria persists for much longer periods of time in a asexuals, which can adversely affect adaptation.
-This fact has been used to argue for the evolutionary advantages of sex and recombination \citep{Barton2005,Otto2009}.
-Consequently, we can learn a substantial amount about the evolution of their traits by examining how genetic correlations from LD evolve over time and influence adaptation.
-Doing so would not only reveal a great deal about some of the oldest and most abundant forms of life on the planet, such as bacteria, but could prove useful to future evolutionary experiments.\par
+For most asexuals, recombination is limited or absent, which makes the Birmingham view appropriate when considering the source of genetic correlations.
+Linkage disequilibrium certainly persists long enough to affect adaptation in these organisms.
+This fact has been the basis of arguments favoring the evolutionary advantages of sex and recombination \citep{Barton2005,Otto2009}.
+Consequently, accounting for the effects of genetic correlations from linkage disequilibrium becomes crucial for studying how traits in asexuals evolve.
+Doing so could potentially lead to insights about evolution of the oldest and most abundant life forms on the planet, such as bacteria, and prove useful to future evolutionary experiments.\par
 
-Lande’s equation describes the phenotypic evolution of populations with multiple characters from generation to generation.
-Whether one can use this framework to uncover the history of natural selection in populations over large evolutionary timescales largely depends on the forces that influence the evolution of the \G-matrix.
-Over the past few decades there has been a significant amount of effort to shed light on the matter of \G-matrix stability.
-The analysis of genetic models for the adaptive evolution of multiple phenotypic traits indicate some of the conditions needed for stability\citep{Turelli1988,Jones2003}.
-These works exclude the potential influences of linkage disequilibrium on the dynamics of variances and covariances over time, and as a result, our understanding of the role of linkage disequilibria in \G-matrix evolution remains unclear.
-Nonetheless, the question is significant there is empirical evidence suggesting that in the \G-matrix is often conserved in natural populations \citep{Arnold1999,Roff2000,Steppan2002}.
-However, much of what we know about \G-matrix stability corresponds the examining evolutionary forces on
+Adaptation in asexuals and the resulting evolution of their traits through selection will be governed by Lande’s equation.
+However, by considering genetic correlations arising from linkage disequilibria, key questions arise about the dynamics of variances and covariances which have important implications for the use of Lande's framework to answer questions about long term features of adaptation.
+In particular, the stability of the \G matrix is chief among these.
+When the \G matrix is stable, it can be used to extrapolate historical patterns of selection.
+This assumption is critical to the use of \G for selection studies and any macro-evolutionary conclusions that can be drawn from them.
+Empirical evidence suggests that at least portions of the \G matrix may be conserved among certain phylogenetic groups \citep{Arnold1999,Roff2000,Steppan2002}, but other studies also indicate that \G is unstable.
+Models for the evolution of the \G matrix have been used to provide insights on this matter, but these assume genetic correlations arise from pleiotropy \citep{Turelli1988}.
+The stability of \G when additive genetic variances and covariance arise from linkage disequilibria have not been explored, but deserves exploration.
 
-Our best models for adaptation in asexuals make use of what are known as traveling waves, and are based on population genetics.
-These models have established important relationships between the population's parameter and key features of evolution in asexuals.
-Among these is the traveling wave model developed by \citet{desai2007beneficial} for a population whose size $N$ is fixed in which only beneficial mutations occur at the rate $U$.
-They also make the key assumption that all mutations all provide the same selective advantage $s$, which allows them to show that the population's rate of adaptation is given by  $v(N,s,U) = 2s^2 \ln(N s)/\ln^2(s/U)$.
-This is an important result, because it directly provides the additive variance in fitness by simply applying Fisher's fundamental theorem ($v=\sigma^2$).
+Population genetic models based on traveling waves have been used to study the dynamics of adaptation in asexuals \citep{rouzine2003solitary,desai2007beneficial}.
+One of their chief accomplishments is their ability to provide details  about the relationship between the rate of adaptation while accounting for the effects of clonal interference.
+The traveling wave model by Desai and Fisher provides useful estimates of the rate of adaptation using a relatively simple framework.
+Their model treats asexual populations of fixed size $N$, in which only beneficial mutation occur at the rate $U$, with each having selection coefficient $s$.
+The expression they obtain for the rate of adaptation in terms of these parameters is given by \eqref{eq:1}.
+\begin{equation}\label{eq:1}
+v(N,s,U) = s^2\frac{2\log(N s)-\log(s/U)}{\log^2(s/U)}
+\end{equation} 
+This is an important result not only because it elucidates on the effects that the population parameters have on the rate of adaptation, but also because Equation \eqref{eq:1} is also the additive variance in fitness $\sigma^2$ of the population by Fisher's fundamental theorem.
 If we regard the population as having one adaptive trait, then $\sigma^2$ is in fact the trait's additive variance and we obtain a quantitative genetics result as well.\par
@@
 
-The dynamics of the bulk was modeled by a system of ordinary differential equations, in which abundances changed in proportion to their selective advantage \eqref{eq:1}. 
-\begin{equation} \label{eq:1}
+The dynamics of the bulk was modeled by a system of ordinary differential equations, in which abundances changed in proportion to their selective advantage \eqref{eq:2}. 
+\begin{equation} \label{eq:2}
 \dot{n}_{ij}(t) = (r_{ij}-\bar{r}) n_{ij}(t). 
@@
 Since the bulk represents the majority of population, we assumed that the frequencies of the bulk followed an analogous ODE.
-\begin{equation} \label{eq:2}
+\begin{equation} \label{eq:3}
 \dot{p}_{ij}(t) = (r_{ij}-\bar{r}) p_{ij}(t) 
@@
 % ^.
-\begin{equation} \label{eq:3}
+\begin{equation} \label{eq:4}
 \pi_{fix} = \frac{s_{ij}}{ 1+s_{ij}}
@@
 % ^.
-\begin{equation} \label{eq:4}
+\begin{equation} \label{eq:5}
 P(\nu \leq \nu_0) = 1- e^{-\pi_{fix} \nu_0}
@@
 We can apply Lande's result to the two components $\dot{\bar{r_1}} $ and $\dot{\bar{r_2}} $ and rewrite them in terms of the trait variances $\sigma_1^2 = \sum_{ij} p_{ij} (r_i-\bar{r_1})^2$ and $\sigma_2^2 = \sum_{ij} p_{ij} (r_j-\bar{r_2})^2$, and their covariance $\sigma_{12} =\sum_{ij} p_{ij} (r_i-\bar{r_1})(r_j-\bar{r_2})$.
-In doing so, we obtain the system expressed in \eqref{eq:5}.
+In doing so, we obtain the system expressed in \eqref{eq:6}.
 % * <masel@email.arizona.edu> 2017-08-22T22:41:29.374Z:
@@
 % ^.
-\begin{equation}\label{eq:5}
+\begin{equation}\label{eq:6}
 \left[
@@
 Our numerical implementation consisted of two routines which ran iteratively for the duration of the simulation.
-Each iteration began with solving for the time-dependent abundances in the bulk governed by \eqref{eq:1} over a time interval of one-thousand generations.
+Each iteration began with solving for the time-dependent abundances in the bulk governed by \eqref{eq:2} over a time interval of one-thousand generations.
 We used Mathematica's standard numerical ODE solver to obtain the solutions and used them to determine the appearance time of the next mutant lineage destined to establish.
@@
 By randomly sampling from the this distribution, we could then solve for each arrival time $\tau_l$.
-We then computed the probability of fixation using \eqref{eq:5}.
+We then computed the probability of fixation using \eqref{eq:6}.
 The integral was numerically approximated up to an appropriate cutoff.
 To determine if a mutation would establish, we sampled a uniform random variable and checked it against the calculated $\pi_{fix}$.
-The next class to establish was chosen accordingly and incorporated into the bulk with a random initial size generated with from \eqref{eq:4}.\par
+The next class to establish was chosen accordingly and incorporated into the bulk with a random initial size generated with from \eqref{eq:5}.\par
 % * <masel@email.arizona.edu> 2017-08-22T22:57:52.494Z:
@@
 The data was used to compute the marginal mean fitness and variances of each trait, as well as their covariance.
-From these quantities we also computed the instantaneous rate of adaptation \eqref{eq:5}, the mean rate of fitness increase in each trait and the overall rate of adaptation.
+From these quantities we also computed the instantaneous rate of adaptation \eqref{eq:6}, the mean rate of fitness increase in each trait and the overall rate of adaptation.
 
@@
 \label{sec:results}
-As expected, we found that an additional trait led to a decrease in the rate of adaptation in the first.
-We found that the reduction was primarily caused by a build of negative covariance between the two traits.
-When the second trait is allowed to involve with the first, the variance of the focal trait increases.
-The magnitude of this increase is shown in Figure \ref{fig:2} in excess of the due to the appearance of genotypes buildup of linkage disequibria which inhibits the purg
-
-addition of a trait leads to 60\% decrease in the rate of adaptation of a trait as a result adding a second.
-The decrease was consistent across three parameter sets which varied either $N$, $s$ and $U$.
-These results also agree with predictions that can be derived from Desai and Fisher's theoretical work ($v(N,s,U)$).
-However, our simulations show that upon adding the second trait, the net change in $v_1$ arises from an increase in the first trait's variance and the formation of negative covariance.
-The first results in an increase in $v_1$, as a direct consequence of \eqref{eq:5}.
-This gain is more than offset by the negative covariance created between the two traits, whose magnitude is always substantially larger than the variance increase.
-Figure 2 presents the calculated scaled mean variance of trait one, and its scaled covariance with the second trait, as a function of each varying population parameter.
-The graph demonstrates that within these parameter ranges, the new rate of adaptation $v_1$ as a percentage of the prior in every case is nearly 60\% less due solely to the large amounts of negative covariance that builds up between the two traits.\par
+In our simulations we found that a trait's rate of adaptation decreased as result of evolving with another.
+The reduction is due to negative covariance that forms between the two traits.
+The addition of a second trait results in an increase of trait one's mean variance as well as the formation of mean negative covariance, both of which are plotted in Figure \ref{fig:2}.
+The change in variance which acts to speed up the rate of adaptation is offset by changes in covariance whose magnitude is in every case 40\% larger and always negative.
+The net change in $v_1$ whose value can be  Equation \eqref{eq:6} is 60\% of the initial variance when evolving without the second trait.
+\par
 
@@
 This results in a net decrease of 40\% in the rate of adaptation (solid line).
-Fixed simulation parameters in (a-c) are: $s=10^{-2}$, $U=10^{-5}$ and $N=10^9$.}\label{fig:2}
+Fixed simulation parameters in (a-c) are: $s=10^{-2}$, $U=10^{-5}$ and $N=10^9$.}
+\label{fig:2}
 \end{figure}
@@
 
-The decreases shown in the scaled mean values of $v_1$ that are plotted in figure 2 are nearly independent of the population parameters picked within ranges characterizing asexual evolution.
-In all three parameter sets, the decrease nearly is always approximately 40\% of the trait's prior variance.
-The insensitivity follows from Desai and Fisher's expression for the rate of adaptation, which exhibits a logarithmic dependence on $N$, $s$ and $U$.
-This leads to a general expression that can be obtained for the expected reduction in a trait's rate of adaptation as a result of adding  $(n-1)$ traits.
-\[
-\% \Delta v_1 
-= 1-\frac{1}{n}\frac{v(N,s,nU)}{v(N,s,U)}
-\]
-Of the three parameters, changes in $s$ affect the expected decrease in scaled $v_1$ most (fig 2d).
-The total number of additional traits also lead to larger and larger reductions in $v_1$.
-However, the extent of this effect is marginally less with each additional trait, but quite substantive with the first few.
-\par
+The reduction in $v_1$ in depends weakly on changes in the population parameters within the ranges examined, while changes in mean variances and covariances do exhibit some dependence on them.
+The constancy of changes in $v_1$ to varying choices of $N$, $s$, and $U$ is due to their appearance within logarithmic terms of $v(N,s,U)$.
+Among the three, varying values of $s$ and $U$ have the most significant effect on the measured mean variance and covariance.
+The percent decrease in $v$ expected from adding $n$ traits can be calculated using Desai and Fisher's expression.
+\begin{equation}\label{eq:7}
+\% \Delta v_1 (n) 
+= 1-\frac{1}{n}\frac{\log^2[s/U]\log[N^2s(nU)]}{\log^2[s/nU]\log[N^2sU]}
+\end{equation}
+From Equation \eqref{eq:7} it is clear that the number of traits added has the largest effect on the expected decrease in $v_1$, but the change is marginally less with each trait added.
+Figure \ref{fig:2d} plots this expression for three different choices of $s$ to depict its influence over $\%\Delta v_1$.
+Changes in $N$ and $U$ have little effect on $\%\Delta v_1$ and produce similar points to $s=10^{-2}$ across their ranges explored in simulation.\par
 
 % *****************************************************************************************************************
-% ***********************************FIGURE*3*COVARIANCE*TIME*SCALES*ARGUMENT*??***********************************
+% ***********************************FIGURE*3*FLUCTUATIONS*IN*COVARIANCE*******************************************
 % *****************************************************************************************************************
 \begin{figure}[h!]
-\begin{subfigure}[b!]{0.48\linewidth}
-\includegraphics[width=1\linewidth, height=6cm]{fig3_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
-\caption{}\label{fig:3a}
-\end{subfigure}
-\begin{subfigure}[b!]{0.48\linewidth}
-\includegraphics[width=1\linewidth, height=6cm]{fig9a.png}
-\caption{}\label{fig:3b}
-\end{subfigure}
-
-\begin{subfigure}[b]{1\linewidth}
-\includegraphics[width=1\linewidth]{fig9b.png}
-\caption{}\label{fig:3c}
-\end{subfigure}
-\caption{\footnotesize Evolution of the \G-matrix over time.
-(a) Time dependent trajectories of variance and covariance with their mean values (constant lines).
-(b) Two dimensional trait distribution with classes having 38 and 39 total beneficial mutations identified.
-(c) Depictions of the associated one dimensional traveling wave at times $t$ and $t+\tau_q$.
-Major transitions in their values occur over a period of $\tau_q \sim v^{-1}$ described by \citet{desai2007beneficial}.
-At time $t$ the mean fitness, marked by the black dot, resides on the line associted with classes with 38 total beneficial mutations (a).
-These classes compose the peak of the one dimensional Gaussian fitness distribution which is exponentially larger than all others bins.
-As a result, their frequencies determine variance and covariance.
-At time $t+tau_q$ the mean shifts to the next line of constant fitness for 39 total beneficial mutations, and their frequencies will then determine variance and covariance.
-In between these times, their values are approximately a convex combinations of those at the time points.}\label{fig:3}
+\includegraphics[width=1\linewidth]{fig3_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
+\caption{\footnotesize Time dependent trajectories of variance and covariance of trait one, along with variance in total fitness.
+Their mean values are shown as lines.
+Fluctuations in variance and covariance can be substantially larger than those in total fitness, and lead to unstable an \G  matrix over relatively short periods of time.}\label{fig:3}
 \end{figure}
@@
 
-Under beneficial mutation-selection balance the time scale governing the dynamics of the \G-matrix components is given by the mean time between establishements $\tau_q \sim 1/v(N,s,2U)$ described by \cite{desai2007beneficial}.
-Figure 3 shows the fluctuations in the variance of trait one, and its covariance with trait two.
-The two are are highly correlated in their dynamics as a result of their adherence to \eqref{eq:5}.
-When the population is beneficial mutation-selection balance, the left hand side of this expression is approximately constant, and as a result, variance and covariance satisfy a linear relationship on average.
-This allows us to describe the dynamics of \G-matrix by understanding the behavior of covariance over time, which is entirely driven by changes in the associated one dimensional fitness distribution depicted in figure 3c.
-The peak of the traveling wave represents the largest portion of the population.
-The frequencies of classes in this group, which are those whose fitness is equal to the population's mean fitness, to a large extent determine the covariance of the two dimensional distribution.
-After $\tau_q$ generations the mean fitness shift forward, and the peak of the one dimensional traveling wave will then coincide with next group of classes that lie on the next line of constant fitness, whose fitness subsequently equals the mean fitness, as shown in figure 3b.
-These major transitions in covariance are governed by the rate of adaptation.
-In between the times of these transitions, the covariance will simply be a convex combination of the two adjacent sets of classes that lie on the lines of constant fitness below and above the mean.
-Minor fluctuations in these dynamics are contributed by the residual portion of abundances not included in these two groups.\par
+% *****************************************************************************************************************
+% ***********************************FIGURE*4*COVARIANCE*TIME*SCALES*ARGUMENT*??***********************************
+% *****************************************************************************************************************
+\begin{figure}[h!]
+\includegraphics[width=0.49\linewidth]{fig4a.pdf}
+\includegraphics[width=0.49\linewidth]{fig4b.pdf}
+\caption{\footnotesize The dynamics of the \G  matrix components are governed by time scales of the one dimensional traveling wave, $\tau_q$.
+Variances and covariance are determined by the set of classes which makeup the peak of the one dimensional traveling wave due to its Gaussian nature.
+At time $t$ the peak is composed of classes shown in blue, whose frequencies determine the components of the \G  matrix.
+At time $t+tau_q$ the mean shifts, and the set of classes shown in green become the dominant group.
+Accordingly, the variance and covariance of the two dimensional distribution transitions to new values that depend on the frequencies of green set of classes.}\label{fig:4}
+\end{figure}
+% *****************************************************************************************************************
+% ***********************************END*FIGURE*4******************************************************************
+% *****************************************************************************************************************
+
+In our model the components of the \G matrix fluctuate significantly over time even while the population is in beneficial mutation-selection balance.
+Figure \ref{fig:3} includes the trajectories of variance in trait one, covariance, and variance in total fitness, as well as their means, for one of our simulations.
+The dynamics of variance and covariance are highly correlated due to their adherence to Equation \eqref{eq:6}, whose left hand side is approximately constant under beneficial mutation-selection balance.
+As a result, fluctuations in variance and covariance can be substantially larger than those of total variance in fitness.\par
+
+The time scales governing changes in the components of the \G  matrix are defined by the dynamics of the one dimensional traveling wave.
+Since its shape is Gaussian, classes belonging to its peak will dominate the population as shown in Figure \ref{fig:4}.
+These are classes that reside on the line of constant fitness passing through the populations mean fitness.
+The variance and covariance will depend on their abundances.
+As the population mean fitness shifts to the next line of constant fitness, a new set of classes becomes dominant and leads to new values of variance and covariance.
+On average, the time required for this to occur is the mean time between establishments at the nose of the of the traveling wave, denoted $\tau_q$ in \citet{desai2007beneficial}.
+\begin{equation}\label{eq:8}
+\tau_q = \frac{\log^2(s/U)}{s(2\log(Ns)-\log(s/U))}
+\end{equation}
+The relative frequencies of classes residing on a line of constant fitness are set by the timing of their establishments when formed part of the stochastic front.
+Since these events depend on the abundances of classes on the line of constant fitness below, the values of variances and covariance will be correlated from one transition to the next.
+However, these correlations vanish after a period of time on the order of that is roughly equal to the time required for classes in the stochastic front become the dominant group.
+This time period can be expressed as the product $q \tau_q$, where $q = 2\log(Ns)/\log(s/U)$ is the lead of the traveling wave nose over its mean.
 
@@
 
-An interesting consequence of the dynamics of covariance invovles the relationship between stochastic front and its role in setting the covariance of the two dimensional trait distribution.
+%An interesting consequence of the dynamics of covariance invovles the relationship between stochastic front and its role in setting the covariance of the two dimensional trait distribution.
 In between establishement events for the one dimensional traveling wave, classes that lie along the line of constant fitness passing through the fittest part of the stochastic front their relative size to one another.
@@
 These major transitions are consequently governed by the rate of adaptation and determined at the stochastic front.
-
-\subsection*{Comparison with Barton and Otto's Negative Linkage Disequilibria Results}
-In our model, covariance between traits is arises entirely from linkage disequilibrium.
+%
+%\subsection*{Comparison with Barton and Otto's Negative Linkage Disequilibria Results}
+%In our model, covariance between traits is arises entirely from linkage disequilibrium.
 We can assume that specific alleles at either locus correspond to the number of beneficial mutations that have accumulated in the corresponding trait.
@@
 By applying \textbf{equation 2}, we can derive a mathematical relationship between covariance and linkage disequilibrium (need to cite chapter 5, Walsh and Lynch).
-% * <masel@email.arizona.edu> 2017-07-29T17:34:47.794Z:
-% 
-% > specific alleles at either locus correspond to the number of beneficial mutations that have accumulated in the corresponding trait
-% Multiple loci can contribute to the same trait.
+%% * <masel@email.arizona.edu> 2017-07-29T17:34:47.794Z:
+%% 
+%% > specific alleles at either locus correspond to the number of beneficial mutations that have accumulated in the corresponding trait
+%% Multiple loci can contribute to the same trait.
 An allele cannot be a "number of beneficial mutations", because an allele must pertain to one locus.
@@
 Because of this confusion, I got a bit lost here and so skipped Equation 6.
+%% 
+%% ^.
+%\begin{equation}
+%\sigma_{12}=\sum_{ij}D_{ij}[(i-\bar{i})s_1][(j-\bar{j})s_2]. 
+%\end{equation} 
+%As the population adapts, mutations and selection  increase and decrease the amounts LD which ultimately determine how traits interact in their evolution.\par
 %
-% ^.
-\begin{equation}
-\sigma_{12}=\sum_{ij}D_{ij}[(i-\bar{i})s_1][(j-\bar{j})s_2]. 
-\end{equation} 
-As the population adapts, mutations and selection  increase and decrease the amounts LD which ultimately determine how traits interact in their evolution.\par
-
-We examined the resulting dynamics of trait evolution using numerical simulations of our model using Mathematica.
+%We examined the resulting dynamics of trait evolution using numerical simulations of our model using Mathematica.
 We focused on modeling the symmetric case, in which the two traits had identical mutation rate and selection coefficients.

### 2017-10-04 plotfunctions.py

@@ -272,40 +272,39 @@ def generate_figure(figNum,data_name,folder_location,sim_start,sim_end,pop_param
         var1_avg = np.mean(fit_var[:,0])*np.ones(np.shape(times))
         var2_avg = np.mean(fit_var[:,1])*np.ones(np.shape(times))
         cov_avg = np.mean(fit_cov[:])*np.ones(np.shape(times))
-        rate_adpt1_avg = var1_avg + cov_avg
-        rate_adpt2_avg = var2_avg + cov_avg
+        rate_adpt_avg = var1_avg + var2_avg + 2*cov_avg
         
         # plot data for figure 2a and 2b 
-        fig3a, ax3a = plt.subplots(1,1,figsize=[8,10])
-        fig3a.subplots_adjust(bottom=0.35)
+        fig3a, ax3a = plt.subplots(1,1,figsize=[8,8])
+        fig3a.subplots_adjust(bottom=0.2)
         ax3b = plt.twinx(ax3a)
-        ax3a.plot(times,var1_avg,c="black",label='var($r_1|r_2$)=' + str(round(var1_avg[0],7)),linewidth=3.0,linestyle = '--')
-        ax3a.plot(times,cov_avg,c="black",label='cov($r_1$,$r_2$)=' + str(round(cov_avg[0],7)),linewidth=3.0,linestyle = ':')
-#        ax3a.plot(times,vU_thry*np.ones(np.shape(var1_avg)),c="black",label='var($r_1$)=' + str(round(vU_thry,7)),linewidth=2.0,linestyle = '-')        
+        ax3a.plot(times,rate_adpt_avg,c="black",label='var$_{fitness}$',linewidth=2.0,linestyle = '-')                
+        ax3a.plot(times,var1_avg,c="black",label='var($r_1|r_2$)',linewidth=3.0,linestyle = '--')
+        ax3a.plot(times,cov_avg,c="black",label='cov($r_1$,$r_2$)',linewidth=3.0,linestyle = ':')
         ax3a.axhline(linewidth=0.5, color = 'k')
         ax3a.set_ylabel('Fitness Variances & Covariance',fontsize=18)
         ax3a.set_xlabel('Time (Generations)',fontsize=18)
         ax3a.set_ylim((-3e-4,4e-4))
         ax3a.set_xlim((1e4,2e4))
-        ax3a.tick_params(axis='both',labelsize=18)
[…]

### 2017-10-04 plots.py

@@ -12,8 +12,8 @@ Script for plots used in "Modeling How Evolution in One Trait is Affected by Ada
 import plotfunctions as pltfun
 
 # load data stored in file generated by plotdata
-#data_name = '_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1'
-data_name = '5'
+data_name = '_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1'
+#data_name = '5'
 #folder_location = 'Documents/kgrel2d/'
 folder_location = ''
 
@@ -25,8 +25,8 @@ num_exp = 33
 ## generate figure 2. (parameters N=1e9, s1=s2=1e-2, U1=U2=1e-5)
 #pltfun.generate_figure(2,data_name,folder_location,snapshot,snapshot,pop_param,'fig2-'+str(snapshot),traitno,num_exp)
 #
-## generate figure 3. (parameters N=1e9, s1=s2=1e-2, U1=U2=1e-5)
-#pltfun.generate_figure(3,data_name,folder_location,sim_start,sim_end,pop_param,'fig3',traitno,num_exp)
+# generate figure 3. (parameters N=1e9, s1=s2=1e-2, U1=U2=1e-5)
+pltfun.generate_figure(3,data_name,folder_location,sim_start,sim_end,pop_param,'fig3',traitno,num_exp)
 
 ## generate figure 4. (parameters N=1e9, s1=s2=1e-2, U1=U2=1e-5)
 #timeframe = 30
@@ -45,4 +45,4 @@ num_exp = 33
 #   
    
 # generate set for figure 5
-pltfun.generate_figure(5,data_name,folder_location,sim_start,sim_end,pop_param,'fig',traitno,num_exp)
+#pltfun.generate_figure(5,data_name,folder_location,sim_start,sim_end,pop_param,'fig',traitno,num_exp)

### 2018-01-29 manuscript.tex
made portable versions of the manuscript and outline.

@@
-\documentclass[11pt,twocolumn]{article} 
-\usepackage{natbib}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{graphicx}
+\documentclass[9pt,twocolumn,twoside]{gsajnl}
+% Use the documentclass option 'lineno' to view line numbers
 \usepackage{caption}
 \usepackage{subcaption}
-\usepackage{hyperref}
-\usepackage[margin=.5in]{geometry}
-
-\graphicspath{{figures/}}
+\usepackage[nameinlink,capitalise]{cleveref}
 \newcommand{\G}{\textbf{G }}
 
-\title{Modeling How Evolution in One Trait is Affected by Adaptation in Another}
-\date{Fall 2017}
-\author{Kevin Gomez,Jason Bertram,Joanna Masel}
+\articletype{inv} % article type
+% {inv} Investigation 
+% {gs} Genomic Selection
+% {goi} Genetics of Immunity 
+% {gos} Genetics of Sex 
+% {mp} Multiparental Populations
 
-\begin{document}
-\maketitle
-\newpage
+\title{Linkage disequilibrium can drive strong and unstable variances and covariances in fitness-associated traits}
+
+\author[$\ast$,1]{Kevin Gomez}
+\author[$\dagger$]{Jason Bertram}
+\author[$\dagger$]{Joanna Masel}
+
+\affil[$\ast$]{Graduate Interdisciplinary Program in Applied Mathematics, University of Arizona, and}
+% * <masel@email.arizona.edu> 2017-10-24T23:03:12.117Z:
+%
+% ^.
+\affil[$\dagger$]{Department of Evolution and Ecology, University of Arizona}
+
+\keywords{Keyword; Keyword2; Keyword3; ...}
+
+\runningtitle{GENETICS Journal Template on Overleaf} % For use in the footer
+
+%% For the footnote.
+%% Give the last name of the first author if only one author;
+% \runningauthor{FirstAuthorLastname}
+%% last names of both authors if there are two authors;
+% \runningauthor{FirstAuthorLastname and SecondAuthorLastname}
+%% last name of the first author followed by et al, if more than two authors.
+\runningauthor{Gomez \textit{et al.}}
 
@@
 
-\section*{Introduction}
-\label{sec:introduction}
+\setboolean{displaycopyright}{true}
 
-Natural selection frequently acts on multiple traits simultaneously.
-A single trait's response to selection leads to a shift in the population mean towards the fittest phenotype.
-With multiple traits the response involves both direct and indirect effects that together determine the evolution each trait \citep{lande1983measurement,Lovell2013,Wagner2011}.
-Direct effects are those induced by selection favoring fitter phenotypic values in each of the traits.
-However, traits may be associated at the genetic level in a manner that permits only certain combinations of their phenotypic values.
-As a result, selection on any one trait produces to changes in the others due to the constraints on their possible combinations.
-For example, we can consider two traits that are controlled by a single pleiotropic locus, but with only the first under selection.
-Allele frequencies at the locus change in response to selection against the first trait, and they also induce a shift in the mean of the second.
-These are indirect effects experienced by the second trait, which are caused by selection on the first.
-Associations such as these can alter fundamental features of adaptation and have significant evolutionary consequences \citep{Felsenstein1979, Arnold2001, Arnold2008}.\par
-% * <masel@email.arizona.edu> 2017-08-22T16:40:07.527Z:
+\begin{document}
+
+\maketitle
+\thispagestyle{firststyle}
+\marginmark
+\firstpagefootnote
+\correspondingauthoraffiliation{Please insert the affiliation correspondence address and email for the corresponding author.
+The corresponding author should be marked with a `1' in the author list, as shown in the example.}
+\vspace{-11pt}%
+
+\section{Introduction}
+
+Natural selection acts on multiple traits simultaneously.
+The mean trait value in a population can change either because of direct selection on trait X, or because of selection on trait Y plus a genetic correlation between X and Y \citep{lande1979quantitative,lande1983measurement}.
+The role of genetic correlations in the response of quantitative traits to natural selection is described by Lande's equation ($\Delta \bar{z} = \G \hspace{.05in}\beta$), where $\Delta \bar{z}$ is the expected change in the population mean of a trait $\Delta \bar{z}$, $\beta$ is the selection gradient, and the components of the \G  matrix are the additive genetic covariances between traits.
+Lande's equation governs the short-term evolution of correlated quantitative traits under natural selection, but it may play a role in bridging the gap between micro- and macro-evolution \citep{Arnold2001,Arnold2008,Schluter1996,Schluter2000}.\par
+
+When the \G matrix is stable, it can be used to extrapolate future trait evolution from historical patterns, and to infer past selection gradients.
+The requirement for a stable \G matrix has motivated empirical work to measure and compare \G in related species.
+At least portions of the \G matrix may be conserved among certain phylogenetic groups \citep{Arnold1999,Roff2000,Steppan2002}, but other studies also indicate that \G is unstable.
+***
+
+Theoretical quantitative genetic models have also been used to examine the conditions under which the stability of the \G matrix can be expected \citep{Turelli1988}.  \par
+% * <masel@email.arizona.edu> 2017-10-06T23:52:45.315Z:
 % 
-% > permits only certain combinations of their phenotypic values
-% The rest of this paragraph seems to be about correlated traits, but here you talk about hard constraints.
-The two are not equivalent, indeed you can have either one without the other.
-I'm not totally sure what you want to achieve in this paragraph, other than giving a non-mathematical version of the next paragraph on Lande's equation.
-If so, I wonder if it is better to merge the two paragraphs, while removing the language about absolute constraints.
-% 
-% ^ <kgomez81@math.arizona.edu> 2017-08-22T20:17:12.979Z:
-% 
-% 
-% 
-% ^.
-% * <masel@email.arizona.edu> 2017-07-28T22:53:26.215Z:
-% 
-% > trait expression,
-% "Expression" is a funny word to use.
-For most biologists, it means gene expression.
-By calling it "trait expression", you are signalling that you mean something else, but I think readers might still find it confusing.
-The term "expression" pre-supposes that the genetic basis for something is there, and focuses on whether it is somehow realized or not; I don't think that is what you mean.
-% 
-% ^ <kgomez81@math.arizona.edu> 2017-08-22T14:05:36.066Z.
-% * <masel@email.arizona.edu> 2017-07-28T22:52:47.005Z:
-% 
-% > the response involves both direct and indirect effects
-% You need to define what you mean by direct and indirect effects: the reader doesn't know.
-% 
-% ^ <kgomez81@math.arizona.edu> 2017-08-07T14:10:25.416Z.
-
-For quantitative traits, genetic associations and their effects on adaptation are mathematically described by Lande's equation ($\Delta \bar{z} = \G \hspace{.05in}\beta$).
-The expression provides the expected change in the population mean of a trait $\Delta \bar{z}$ as the product of the selection gradient $\beta$ and what is known as the \G  matrix, whose components are the additive genetic covariances between traits.
-Lande’s result has been used to extend much of our understanding of how phenotypes evolve under selection, yielding a rich set of ideas that emphasize the role of the \G  matrix in directing adaptive evolution at both micro- and macro-evolutionary scales.\par
-% * <masel@email.arizona.edu> 2017-07-29T12:43:16.916Z:
-% 
-% > When combined with the notions of adaptive landscapes
-% The link to adaptive landscapes is not obvious, including to me.
-G-matrix theory is all continuous, and the only possible landscape is a phenotype-fitness map.
-The adaptive landscape you are going to deal with is a genotype-fitness map, ie a completely different object.
-% 
-% ^ <kgomez81@math.arizona.edu> 2017-08-07T14:10:30.176Z.
-
-The behavior of \G over time largely depends on what are the causes of the genetic correlations between traits.
-These can arise for different reasons, but often have a genetic basis \citep{Saltz2017}.
-Pleiotropy and linkage disequilibria are by far the most common causes.
-Historically, there have been two schools of thought on which of these underlie most of the genetic correlations detected.
-Pleiotropy was favored by the Edinburgh school, which adopted the view that recombination could quickly eliminate linkage disequilibrium and remove its effects \citep[Chapter~20]{fox2006evolutionary}.
-This belief was largely motivated by their interests in relationships between parent-offspring phenotypes within randomly mating animal populations.
-In such groups pleiotropy is indeed the most prominent cause of genetic correlations that can persist over evolutionary time scales.
-An alternative view was put forth by the Birmingham school.
-Quantitative geneticists in this group largely worked on inbred lines of plants, and as result, where often confronted with the effects of linkage disequilibria in the results of their experiments.
-Presently, the Edinburgh view is more widely accepted, and as a result, most models for the evolution of the \G  matrix assume genetic correlations are pleiotropic and proceed with examining how evolutionary processes influence them.
-We know far less about the about the dynamics of genetic correlations arising from linkage disequilibria.\par
-% * <masel@email.arizona.edu> 2017-08-22T16:50:23.214Z:
-% 
-% >  most models for the evolution of the \G  matrix assume genetic correlations are pleiotropic and proceed with examining how evolutionary processes influence them
-% An important exception to note is the "Bulmer effect", which I believe describes how selection can cause LD even in the absence of linkage, which then of course appears in G.
-This is distinct from our work where linkage to the background on which a mutation first appear is the cause of LD, albeit later shaped by selection.
-I believe there is some material on the Bulmer effect in Bruce Walsh's book.
-It is important to review it just enough to distinguish what you are doing from what Bulmer already did.
-% 
-% ^.
-% * <masel@email.arizona.edu> 2017-08-22T16:47:43.713Z:
-% 
-% > the product of trans acting elements
-% This is not self-explanatory, and insofar as I can guess what it means, it seems to me to be a special case of pleiotropy rather than a third distinct possibility.
+% > other studies also indicate that \G is unstable
+% Citations(s) needed for empirical work for instability of G and theoretical work either deriving or assuming stability.
 % 
@@
 
-For most asexuals, recombination is limited or absent, which makes the Birmingham view appropriate when considering the source of genetic correlations.
-Linkage disequilibrium certainly persists long enough to affect adaptation in these organisms.
-This fact has been the basis of arguments favoring the evolutionary advantages of sex and recombination \citep{Barton2005,Otto2009}.
-Consequently, accounting for the effects of genetic correlations from linkage disequilibrium becomes crucial for studying how traits in asexuals evolve.
-Doing so could potentially lead to insights about evolution of the oldest and most abundant life forms on the planet, such as bacteria, and prove useful to future evolutionary experiments.\par
-
-Adaptation in asexuals and the resulting evolution of their traits through selection will be governed by Lande’s equation.
-However, by considering genetic correlations arising from linkage disequilibria, key questions arise about the dynamics of variances and covariances which have important implications for the use of Lande's framework to answer questions about long term features of adaptation.
-In particular, the stability of the \G matrix is chief among these.
-When the \G matrix is stable, it can be used to extrapolate historical patterns of selection.
-This assumption is critical to the use of \G for selection studies and any macro-evolutionary conclusions that can be drawn from them.
-Empirical evidence suggests that at least portions of the \G matrix may be conserved among certain phylogenetic groups \citep{Arnold1999,Roff2000,Steppan2002}, but other studies also indicate that \G is unstable.
-Models for the evolution of the \G matrix have been used to provide insights on this matter, but these assume genetic correlations arise from pleiotropy \citep{Turelli1988}.
-The stability of \G when additive genetic variances and covariance arise from linkage disequilibria have not been explored, but deserves exploration.
-
-Population genetic models based on traveling waves have been used to study the dynamics of adaptation in asexuals \citep{rouzine2003solitary,desai2007beneficial}.
-One of their chief accomplishments is their ability to provide details  about the relationship between the rate of adaptation while accounting for the effects of clonal interference.
-The traveling wave model by Desai and Fisher provides useful estimates of the rate of adaptation using a relatively simple framework.
-Their model treats asexual populations of fixed size $N$, in which only beneficial mutation occur at the rate $U$, with each having selection coefficient $s$.
-The expression they obtain for the rate of adaptation in terms of these parameters is given by \eqref{eq:1}.
-\begin{equation}\label{eq:1}
-v(N,s,U) = s^2\frac{2\log(N s)-\log(s/U)}{\log^2(s/U)}
-\end{equation} 
-This is an important result not only because it elucidates on the effects that the population parameters have on the rate of adaptation, but also because Equation \eqref{eq:1} is also the additive variance in fitness $\sigma^2$ of the population by Fisher's fundamental theorem.
-If we regard the population as having one adaptive trait, then $\sigma^2$ is in fact the trait's additive variance and we obtain a quantitative genetics result as well.\par
-% * <masel@email.arizona.edu> 2017-08-22T17:28:18.736Z:
+The genetic correlations described by \G are due to some combination of pleiotropy and linkage disequilibrium \citep{Saltz2017}.
+Historically, pleiotropy was favored by the Edinburgh school, who argued that recombination would quickly eliminate linkage disequilibrium \citep[Chapter~20]{fox2006evolutionary}, particularly in the randomly mating animal populations that the Edinburgh school focused on.
+In contrast, quantitative geneticists in the alternative, Birmingham school largely worked on inbred lines of plants, where the effects of linkage disequilibrium were impossible to ignore.
+The Edinburgh view came to be widely accepted, and as a result, most models for the evolution of the \G  matrix assume that genetic correlations are pleiotropic and go on to examine how evolutionary processes influence them \citep{Turelli1988}.
+We know far less about the about the dynamics of genetic correlations arising from linkage disequilibrium.\par
+% * <masel@email.arizona.edu> 2017-10-24T22:58:18.095Z:
 % 
-% > the model developed by \cite{desai2007beneficial}
-% This is the place to briefly summarize the core assumptions of the model.
-You have already given the assumption of strict asexuality, the others are constant N, absence of deleterious mutations, and lack of variation in s among beneficial mutants.
-The latter (combined with asexuality) is the key innovation that makes the model work, because it relieves the need to track the identity of individual mutations.
-% 
-% ^ <kgomez81@math.arizona.edu> 2017-08-27T02:56:42.167Z:
-% 
-% I've added a statement about Desai and Fisher's assumptions following the citation, and I've also slightly changed the sentences after.
+% > most models for the evolution of the \G  matrix assume that genetic correlations are pleiotropic and go on to examine how evolutionary processes influence them \citep{Turelli1988
+% More citations needed
 % 
@@
 
-Unfortunately, if we apply Desai and Fisher's result to the case of two adaptive traits, we still can still solve for the overall fitness variance $\sigma^2$ of the population, but nothing can be said about the variances and covariance of the two traits.
-This highlights the shortcomings of the traveling wave model in the case of multiple traits, for which we should expect Lande's equation to hold.
-As an example, consider a single adaptive trait evolving with parameters $N$, $s$ and $U$, then its variance and rate of adaption are $v(U;N,s)$ according to the formula given above.
-If the population acquires a second adaptive trait with identical parameters $s$ and $U$, then the new rate of adaptation is clearly $v(2U;N,s)$.
-By symmetry we know that rate of adaptation in each trait alone must be $0.5 \hspace{.02 in} v(2U;N,s)$, and furthermore, the rate of adaptation in trait one is now less with the second trait.
-Lande's result implies that trait one's rate of adaptation, when evolving with the second, must equal the sum of its variances and covariance ($v_1 =\sigma_1^2 +\sigma_{12}$).
-Clearly, adding the second trait produces a new covariance term that affects $v_1$, but it also induces a change in the variance of the first trait.
-What we do not know is the amount that each of these contributes to the observed reduction in trait one's rate of adaptation.
-This is exactly what we aim to determine.
+Biology today has come to recognize the dominance of relatively asexual microbes \cite{mcfall2013animals}, for whom the Birmingham emphasis on linkage disequilibrium as a source of genetic correlations cannot be ignored.
+Linkage disequilibria certainly persist long enough to affect adaptation in asexuals; this fact has, for example, been the basis of explanations for the evolutionary advantages of sex and recombination \citep{Barton2005,Otto2009}.
 \par
+% * <masel@email.arizona.edu> 2017-10-06T22:23:25.138Z:
+% 
+% > Biology today has come to recognize the dominance of relatively asexual microbes. 
+% Cite www.pnas.org/cgi/doi/10.1073/pnas.1218525110 and maybe others making complementary points.
+% 
+% ^ <kgomez81@math.arizona.edu> 2017-12-05T18:40:59.153Z:
+% 
+% I added the citation.
+%
+% ^.
+Asexual evolution is normally approached using a population genetic rather than a quantitative genetic approach.
+In particular, traveling wave models have been successfully used to describe the rate of multi-locus adaptation in asexuals given clonal interference \citep{rouzine2003solitary,desai2007beneficial}.
+The basic model has only three parameters: fixed population size $N$, beneficial mutation rate $U$, and per-beneficial-mutation selection coefficient $s$.
+Adaptation, i.e.
+the increase in mean population fitness, occurs at the rate given by Desai and Fisher \citep[Equation 41]{desai2007beneficial}, expressed as
+% * <masel@email.arizona.edu> 2017-10-14T17:09:10.651Z:
+% 
+% > Adaptation, i.e. the increase in mean population fitness, occurs at rate
+% Cite the specific equation number in Desai & Fisher. Also give the parameter value conditions for which this equation holds.
+% 
+% ^ <kgomez81@math.arizona.edu> 2017-12-06T20:47:24.968Z:
+% 
+% I added the equation number to the citation.
+%
+% ^.
+\begin{equation}\label{eq:1}
+v(U,N,s) = s^2\frac{2\log(N s)-\log(s/U)}{\log^2(s/U)}.
+\end{equation}
+The expression holds when the asexual population is adapting in what is defined by Desai and Fisher as the ``concurrent mutations regime.'' In a monoclonal population, selection will cause a beneficial mutation to spread until all individuals carry the new mutation, at which point, it is said to have fixed in the population.
+The time required for beneficial mutation to fix is called its fixation time.
+For asexuals adapting in the concurrent mutations regime, beneficial mutations appear in rapid succession relative to the fixation time.
+This produces many mutant lineages that vary in the numbers of beneficial mutations that have accumulated in their genome.
+Beneficial mutations that appear on less fit lineages are lost due to clonal interference.
+Increasing the number of beneficial mutations that enter the population increases the proportion of beneficial mutations that are wasted.
+The logarithmic dependence of $v(U,N,s)$ on the parameters captures these effects.
+An asexual population is in the concurrent mutations regime when the parameters satisfy the inequality $NU \gtrsim 1/\ln(Ns)$.
+\par
+
+Equation \ref{eq:1}, combined with Fisher's fundamental theorem, also gives the additive variance in fitness $\sigma^2$.
+With only one adaptive trait,  $\sigma^2$ also gives that trait's additive variance.
+With two adaptive traits, Equation \ref{eq:1} still gives overall fitness variance $\sigma^2$, but nothing can be said about the variances and covariance of the two traits.\par
+
+Adaptation in one trait proceeds more slowly when there is clonal interference with mutations that are adaptive in a second trait.
+Quantitatively, when the two adaptive traits each have mutations with fitness benefit $s$ appearing at rate $U$, the overall adaptation rate is $v(2U,N,s)$, and so by symmetry, adaptation in each trait alone occurs at rate $0.5 \hspace{.02 in} v(2U,N,s)$.
+This is lower than the rate $v(U,N,s)$ when the other trait is not also evolving.
+The reduction in trait-specific adaptation rate must be distributed between the two terms of Lande's equation $v_1 =\sigma_1^2 +\sigma_{1,2}$.
+Here, we will calculate how the reduction in trait-specific adaptation partitions between the variance in that trait and its covariance with the other.\par
+
 % * <masel@email.arizona.edu> 2017-07-29T13:00:14.234Z:
@@
 
-To explore how adaptation in one trait is affected by evolution in another, we extend the work of Desai and Fisher's traveling wave model and consider a two dimensional trait space.
-We examine the simplest case where individuals have two non-pleiotropic traits that contribute equally to fitness, each controlled by separate loci.
-We assume that beneficial mutations occur with identical rates in each trait, affect each independently, and exhibit no epistatic interactions.
-Lastly, we will focus on a population whose size remains fixed.
-Our aim is to examine the effects of genetic correlations arising solely from linkage disequilibrium and determine how they influence trait evolution.
+To explore how adaptation in one trait is affected by evolution in another, we extend the traveling wave model approach to a two dimensional trait space.
+We assume that each of the two fitness-determining traits is controlled by different loci, i.e.
+there is no pleiotropy.
+With pleiotropic contributions to \G excluded by design , we use this model to examine the dynamics of genetic correlations arising solely from linkage disequilibrium.
 
-\section*{Materials and Methods}
+\section{Materials and Methods}
 \label{sec:materials:methods}
 
-We considered an asexual population at carrying capacity $N$ in which haploid individuals had two traits contributing to their relative fitness.
-Each trait was assumed to be under the control of a single locus.
-We assumed that beneficial mutations improving each trait appeared at a rate of $U$ and increase th fitness by $s$; deleterious mutations were excluded.
-Individuals that had accumulated $i$ mutations in the first and $j$ in the second were taken to have an overall fitness of $r_{ij} = i s+j s$.
-Their relative fitness is given by the difference $(r_{ij}-\bar{r})$, where  $\bar{r} $  is the population mean fitness.
-We will often right $r_i$ in place of $i s$ and $r_j$ in place of $j s$, so that $r_{ij} = r_i + r_j$.
-% * <masel@email.arizona.edu> 2017-08-22T19:12:14.027Z:
-% 
-% > appeared at a rate of $U_k$ ($k=1,2$) and increase th fitness by $s_k$
-% Since you are going to make U and s independent of k, it's easier on the reader if you do that from the outset, rather than set up the general case and then  collapse it.
-% 
-% ^ <kgomez81@math.arizona.edu> 2017-08-23T16:45:52.343Z:
-% 
-% I will change the notation the instances of U_k and s_k to just U and s here and in the rest of the document. 
-%
-% ^.
+We consider an asexual population of haploids with fixed total population size $N$.
+To keep the model simple, we assume that beneficial mutations occur with the same rate of $U$ mutations per birth per trait with fitness benefit $s$ in each trait and no epistasis.
+Deleterious mutations are not considered.
+Individuals with $i$ mutations in the first trait and $j$ in second, have relative fitness $r_{i,j}= i s+j s$ and absolute fitness $r_{i,j}-\bar{r}$, where $\bar{r}$ is the population mean fitness.
+We refer to each such set of individuals as a ``class'' and denote their abundances and frequencies by $n_{i,j}$ and $p_{i,j}$.
+\par
 
-In an analogous approach to that of \citet{desai2007beneficial}, we divided our population into classes based on the number of beneficial mutations in each trait.
-Abundances are accordingly denoted $n_{ij}$ and their frequencies by $p_{ij}$.
-We assumed that $1/N \ll s \ll 1$ to ensure that the set of classes could be neatly separated into two groups based on whether their behavior was deterministic and stochastic.
-Following the convention of Desai and Fisher, we refer to the set of deterministically behaving classes as the bulk, while those that grow stochastically and have a positive selective advantage ($r_{ij}>\bar{r}$) will be referred to as the stochastic front.
-Unlike in the case of Desai and Fisher's  traveling wave whose stochastic front is a single point, our stochastic front is the boundary of the bulk's distribution above the line crossing the population's mean fitness.
-This is shown in figure 1, where classes in the bulk are shown in shades of blue, while those belonging to the stochastic front are shown in shades of red.
-One reason why the growth of these two groups is treated differently has to do with the fact that only beneficial mutations in the stochastic front are consequential to adaptation.
-In asexuals, a large percentage of beneficial mutations are wasted since most occur poor genetic backgrounds, and their lineages remain relatively insignificant size and as a result, they do not contribute to the population's adaptive evolution.
-These are the beneficial mutations that occur within the bulk.
-On the other hand, mutations that do manage to occur on good genetic backgrounds will eventually grow large enough in size to influence the population's mean relative fitness over time.
-These are exactly the beneficial mutations produced at the edge of the bulk that give rise to classes in the stochastic front, marked by arrows in figure 1.
-Thus, the bulk's distribution determines the position and behavior of the stochastic front, but in time, the latter also shapes how the former shifts forward (fix this!).\par
+We assume $1/N \ll s \ll 1$.
+Following the convention of Desai and Fisher, we refer to the set of classes abundant enough to behave deterministically $(n_{i,j}>1/s)$ as the bulk, and to smaller adaptive classes ($r_{i,j}>\bar{r}$) as the stochastic front (Figure \ref{fig:1}, solid and patterned, respectively).
+The stochastic front shares the same role of the nose in Desai and Fisher's traveling wave model \cite{desai2007beneficial}.
+Like the nose, the stochastic front contains the beneficial mutations (arrows in Figure \ref{fig:1}) that will drive future adaptation in the bulk; mutations within the bulk are wasted.
+However, the front in our model is evolving boundary rather than a point.
+Classes in the front become part of the bulk once they have ``established'' (grown sufficiently large).
+This shifts the position of  the bulk, with the stochastic front shifting also, to form a two dimensional traveling wave.
+\par
 % * <masel@email.arizona.edu> 2017-08-22T19:17:50.152Z:
 % 
-% > those that grow stochastically and have a positive selective advantage ($r_{ij}>\bar{r}$) will be referred to as the stochastic front
+% > those that grow stochastically and have a positive selective advantage ($r_{i,j}>\bar{r}$) will be referred to as the stochastic front
 % This is a good place to point a key difference between you and prior work, namely that your front is a curve not a point.
@@
 % 
-% I added a bit more after this sentence, and changed the subsequent one.
-% "Unlike in the case of Desai and Fisher's  traveling wave whose stochastic front is a single point, our stochastic front is the boundary of the bulk's distribution above the line crossing the population's mean fitness.
-This is shown in figure 1, where classes in the bulk are shown in shades of blue, while those belonging to the stochastic front are shown in shades of red.
-"
+% I added a bit more after this sentence, and changed the subsequent one one more sentence to indicate the differences in between Desai and Fisher's traveling wave nose.
+"The stochastic front shares the same role of the nose in Desai and Fisher's traveling wave model\cite{desai2007beneficial}.
+Like the nose, the stochastic front contains the beneficial mutations (arrows in Figure \ref{fig:1}) that will drive future adaptation in the bulk; mutations within the bulk are wasted."
 % 
@@
 
-% *****************************************************************************************************************
-% ***********************************FIGURE 1*2D*Travelling*Wave***************************************************
-% *****************************************************************************************************************
-\begin{figure}[h!]
-\includegraphics[width=1\linewidth]{fig1a.pdf}
-\caption{\footnotesize Illustration of a two dimensional trait distribution taken from simulation.
-Boxes correspond to classes whose members have identical numbers of beneficial mutations in each trait.
-The bulk is composed of classes with solid interiors, while those with patterned interiors form part of the stochastic front.
-The red dot marks the number of beneficial mutations in the average individual.
-Classes are grouped with respect to their total relative fitness, but vary in the number of beneficial mutations that have accumulated in each trait.
-These groups are represented by the lines passing through their centers that are parallel to the dashed line.
-These are refered to lines of constant fitness.
-The red line marks the set of classes whose fitness equals the population mean fitness.
-Fitness increases along the perpendicular direction to this line.
-Only beneficial mutations occuring on fit backgrounds produce lineages that grow sufficiently large to affect adaptation.
-These are represented by arrows from the bulk into the stochastic front.
-Other mutant lineages are out-competed and do not significantly contribute to the population's adapatation.
-The stochastict front expands in the direction of increasing relative fitness, while classes in the bulk below the red line disappear.
-These two processes continually shift the two dimensional trait distribution in the direction of increasing fitness over time.
-(Simulation parameters: $N=10^9$, $s=10^{-2}$, and $U=10^{-5}$)}\label{fig:1}
-\end{figure}
-% *****************************************************************************************************************
-% *******************************END*FIGURE 1**********************************************************************
-% *****************************************************************************************************************
-
-The dynamics of the bulk was modeled by a system of ordinary differential equations, in which abundances changed in proportion to their selective advantage \eqref{eq:2}. 
-\begin{equation} \label{eq:2}
-\dot{n}_{ij}(t) = (r_{ij}-\bar{r}) n_{ij}(t). 
-\end{equation}
-Abundances in the stochastic front were negligible in size relative to $N$ due to our parameter assumptions stated earlier.
-Since the bulk represents the majority of population, we assumed that the frequencies of the bulk followed an analogous ODE.
-\begin{equation} \label{eq:3}
-\dot{p}_{ij}(t) = (r_{ij}-\bar{r}) p_{ij}(t) 
-\end{equation}
-These frequencies determine all of the quantitative genetic measures that are of interest to us.\par
-
-Classes in the stochastic front intermittently transition from stochastic to deterministic behavior and become part of the bulk once they have grown sufficiently large.
-Such classes are said to have established.
-Prior to establishment, random fluctuations due to drift drive the dynamics of abundances.
-The behavior of these classes can be modeled as a branching process fed by incoming mutants generated in adjacent, less fit, classes within the bulk.
-
-Each mutant lineage that appears in the stochastic front either goes establishes or goes extinct.
-The probability of fixation measures the chances of the former outcome, and it depends on the selective advantage of the mutant.
-If its genotype is $(i,j)$, then its selective advantage is $s_{ij}=(r_{ij}-\bar{r})$.
-Standard mathematical results for branching processes can be used to solve for the lineage's probability of fixation which is given by equation 3.
-% * <masel@email.arizona.edu> 2017-08-22T21:57:02.810Z:
+\begin{figure}[!ht]
+\includegraphics[width=1\linewidth]{figures/fig1a.pdf}
+\caption{Illustration of a two dimensional trait distribution taken from a simulation.
+Individuals having an equal number of beneficial mutations in each trait are grouped into classes (boxes).
+Classes growing deterministically form part of the bulk (solid interiors), and those growing stochastically belong to stochastic front (patterned interiors).
+The red dot marks the average number of beneficial mutations in each trait.
+Fitness-isoclines identify the set of equally fit classes and are parallel the population mean fitness isocline (dashed red line).
+In asexuals, beneficial mutations must occur on the fittest backgrounds in order to contribute to the adaptive process.
+These mutations are represented by arrows from the bulk into the stochastic front.
+Over time, classes in stochastic front become part of the bulk once their abundances have become sufficiently large, while new ones subsequently emerge and become part of the stochastic front.
+Classes in the bulk below the population mean fitness isocline dwindle in size by selection.
+The distribution shifts direction of increasing relative fitness as result of these two processes and produce a two-dimensional traveling wave.
+(Simulation parameters: $N=10^9$, $s=0.02$, and $U=10^{-5}$)}\label{fig:1}
+% * <masel@email.arizona.edu> 2017-10-14T18:05:49.152Z:
 % 
-% > Standard mathematical results for branching processes can be used to solve for the lineage's probability of fixation which is given by equation 3
-% Eq.
-3 is not in a familiar form to me, and you don't give a citation.
-The result we discussed at your oral exams was pi = s/sigma^2 where sigma is the standard deviation of offspring number given a particular fitness/expectation.
-D&F for-convenience choice was sigma=1, so pi=s.
+% > These groups are represented by the lines passing through their centers that are parallel to the dashed line.
+% Do you want to put these lines into the figure, or take this sentence out of the figure legend?
 % 
-% ^ <kgomez81@math.arizona.edu> 2017-08-23T17:43:47.591Z:
-% 
-% Their initial discussion of the branching process uses "s" to state the general results of mutant lineages establishing, but they later use "qs" for their analysis of establishments at the nose.
-Equation 3 is taken from equation 16 in Desai and Fisher, but is also a general result found in Allen's book "intro to stoch proc" page 262.
 % 
 % ^.
-\begin{equation} \label{eq:4}
-\pi_{fix} = \frac{s_{ij}}{ 1+s_{ij}}
+\end{figure}
+
+Changes in abundances within the bulk follow a system of ordinary differential equations given in \cite[page 27]{Crow1970} :
+\begin{equation} \label{eq:2}
+\dot{n}_{i,j}(t) =\left (r_{i,j}-\frac{N(t)}{N}\bar{r}\right) n_{i,j}(t),
 \end{equation}
-The lineage establishes when it has reached a size if roughly $1/s_{ij}$, and its subsequent growth is exponential.
-\cite{desai2007beneficial} expressed the abundances of these lineages as $n_{ij}(t) = \frac{1}{s_{ij}} \exp[{s_{ij}(t-\tau)}]$.
-The random variable $\tau$ is the time of establishment and it captures the cumulative effects of the stochastic fluctuations.
-In their initial analysis, Desai and Fisher solve for the probability density function of $\tau$ for lineages destined to establish.
-In our model, we find it more useful to consider the related random variable $\nu = \frac{1}{s_{ij}}\exp[-\Delta r_{ij}\tau]$ representing an initial size that produces the deterministic trajectory observed after establishment.
-Using the expression given by Desai and Fisher for the PDF of $\tau$, we can arrive at the cumulative distribution of  $\nu$ (equation 5) by following the analysis of \citet{Uecker2011}.
-% * <masel@email.arizona.edu> 2017-08-22T22:14:21.200Z:
+where time $t$ is measured in generations.
+At capacity, Equation \ref{eq:2} gives the selective advantage of a class relative to population mean fitness, denoted $s_{i,j}$, whose value is approximately  $s_{i,j} \approx (r_{i,j}-\bar{r})$.
+The stochastic front contains so little of the total population that variances and covariance are effectively determined only by the frequencies of classes within the bulk of two dimensional distribution.
+Each iteration began with solving (using NDSolve in Mathematica) for the time-dependent abundances in the bulk governed by \ref{eq:2} over a time interval of one thousand generations.\par
+
+Growth of classes in the front can be modeled by branching processes fed by mutants from the adjacent (and exponentially growing) classes in the bulk.
+These mutations are shown by black arrows in Figure \ref{fig:1}.
+We sample the next mutation at the front by sampling random variable $x$ from a standardized exponential and solving $ x=\int_0^{\tau} b_{i,j} n_{i,j}(t)U dt$ for $\tau$.
+The birth rate in our model is set to $b_{i,j}=1+s_{i,j}$ per individual per generation.
+
+We then sampled whether the mutation would establish, by comparing a uniform random variable to the probability of fixation from Equation 16 of Desai and Fisher:
+
+\begin{equation} \label{eq:3}
+\pi_{fix} = \frac{s_{i,j}}{ 1+s_{i,j}}.
+\end{equation}
+After establishment of a mutant lineage at abundance $1/s_{i,j}$, it is part of the bulk, and subsequent growth is deterministic and exponential. \par
+
+We update our system based on the first mutation to appear that is destined for establishment.
+The timing of establishment events is also stochastic, with later deterministic growth described by $n_{i,j}(t) = \frac{1}{s_{i,j}} \exp[{s_{i,j}(t-\tau)}]$ with the random variable $\tau$ measuring the time of establishment.
+We consider the related random variable $\nu = \frac{1}{s_{i,j}}\exp[-s_{i,j}\tau]$ representing the initial size at the time of first mutation that would produce the deterministic trajectory observed after establishment.
+Using the expression given by Desai and Fisher for the PDF of $\tau$ \citep[Equation 12]{desai2007beneficial}, we can arrive at the cumulative distribution of $\nu$:
+% * <masel@email.arizona.edu> 2017-10-23T22:46:00.814Z:
 % 
-% > useful to consider the related random variable $\nu = \frac{1}{s_{ij}}\exp[-\Delta r_{ij}\tau]$ representing an initial size that produces the deterministic trajectory observed after establishment
-% If doing this was Uecker & Hermisson's idea, then this is the place to give them credit/citation.
+% > Using the expression given by Desai and Fisher for the PDF of $\tau$
+% Give an Equation number
 % 
-% ^ <kgomez81@math.arizona.edu> 2017-08-23T17:49:29.395Z:
+% ^ <kgomez81@math.arizona.edu> 2017-12-05T19:58:30.000Z:
 % 
-% I modified the last sentence, but this is probably not clear.  I'll have to go back to Uecker and Hermisson to come up with a more detailed explanation concerning their analysis.
-% 
-% Using the expression given by Desai and Fisher for the PDF of $\tau$, we can arrive at the cumulative distribution of  $\nu$ (equation 5) by following the analysis of \cite{Uecker2011}. 
+% Added the citation for the PDF of \tau needed to get the CDF of \nu, and removed the citation to Uecker and Hermisson previously included.
 % 
 % ^.
-\begin{equation} \label{eq:5}
+
+\begin{equation} \label{eq:4}
 P(\nu \leq \nu_0) = 1- e^{-\pi_{fix} \nu_0}
-\end{equation}
-Each class in the stochastic front may consist of many mutant lineages.
+\end{equation}and used it to set the initial abundance of  the establishing class equal to $\nu$.
+We update the other abundances of the bulk at this time.
+Any classes with less than one individual were removed.
+The next iteration began with solving for the timecourse of the new bulk, now including the newly established class, followed by the subsequent steps outlined above.\par
+
+\begin{figure*}[!ht]
+\centering
+\includegraphics[width=1\linewidth]{figures/fig2.pdf} 
+\caption{Measured averages of variance (open diamonds) and covariance (closed circles) for trait one obtained from simulations, along with the theoretically predicted drop in its rate of adaptation (solid line).
+In each graph, two population parameters are kept constant while the third is varied to determine its effects on the mean variance and covariance of trait one induced the by evolution with the second trait.
+In every case, the variance of trait one increases ($\sigma_1^2 >1$), while trait interactions result in negative covariance whose magnitude is larger than the change in variance ($|\sigma_{1,2}| > \sigma_1^2-1$), indicating that the reduction in the rate of adaptation of trait one is the result of large negative covariances.
+Mean variance and covariance are insensitive to changes in the population size $N$ (pane c), but exhibit some dependence on the values of $U$ and $s$ (panes a,b).
+The values  $s=0.02$, $U=10^{-5}$ and $N=10^9$ where used for parameters that were kept constant.}\label{fig:2}
+\end{figure*}
+
+Each class in the stochastic front may be a mixture of lineages, each arising from a different mutation event.
+However, we assume that the arrival time of the first mutant lineage destined to establish is the one determines which class will join the bulk next.
+With some probability, the  \par
+% * <masel@email.arizona.edu> 2017-10-23T22:52:29.988Z:
+% 
+% > Each class in the stochastic front may be a mixture of lineages, each arising from a different mutation event.
 However, it is the arrival time of the first mutant lineage destined to establish that determines which class will join the bulk next.
-Once the genotype of this mutant lineage is obtained, its initial size $\nu$ can be sampled.
-The class can then be regarded as belonging to the bulk at the determined arrival time with initial size $\nu$.
 \par
-% * <masel@email.arizona.edu> 2017-08-22T22:18:30.247Z:
-% 
-% >  its initial size $\nu$ can be sampled
-% I think this is the first time in the Methods that you talk about sampling.
-The rest could be read as a general mathematical description, rather than a methodology for simulation, so this could cause confusion.
-Right now you begin talking about numerical implementation two paragraphs down from here, and this reference to sampling doesn't have that context.
+% Requires discussion. Errors introduced when the first establishing is earlier than the establishing time of the first arrival destined to establish.
 % 
 % ^.
-% * <masel@email.arizona.edu> 2017-07-29T16:52:06.887Z:
+
+Simulations were initialized with monoclonal populations at carrying capacity consisting of individuals having no beneficial mutations in either trait and relative fitness set to zero.
+We varied either $N$, $s$ and $U$ while fixing the others.
+Each was varied across a range constructed from values reported by evolutionary experiments involving either E.
+coli or S.
+cerevisiae \citep{desai2007speed,Levy2015,Perfeito2007}.
+\par
+% * <masel@email.arizona.edu> 2017-10-23T22:58:03.572Z:
 % 
-% > This was obtained using the expression below derived by \citep{Uecker2011} [eqt. 16b].
-% Do you use the full generality of Uecker & Hermisson, or are earlier expressions, eg as used in Desai & Fisher, sufficient for your work?
+% > We varied either $N$, $s$ and $U$ while fixing the others. Each was varied across a range constructed from values reported by evolutionary experiments involving either E. coli or S. cerevisiae
+% More detail on what exactly we learned from each paper, and also how we kept within concurrent-mutations regime, including saying what the concurrent mutations regime, or do the latter around Equation 1, and refer to it here.
 % 
+% ^ <kgomez81@math.arizona.edu> 2017-12-06T20:49:15.515Z:
 % 
-% ^ <kgomez81@math.arizona.edu> 2017-08-04T23:43:18.217Z.
-In order to investigate how of traits affect one another, we identified the contributions of each trait to the evolution of the population.
-The rate of adaptation can obtained by differentiating the population's mean relative fitness $\bar{r}=\sum_{ij} p_{ij} r_{ij}$, which is itself the sum of the two marginal means, $\bar{r_1}=\sum_{ij} p_{ij} is_1$ and $\bar{r_2}=\sum_{ij} p_{ij} j s_2$.
-It follows that the total rate of adaptation is sum of rates of adaptation corresponding to each trait ($\dot{\bar{r}}=\dot{\bar{r_1}} +\dot{\bar{r_2}}$).
-We can apply Lande's result to the two components $\dot{\bar{r_1}} $ and $\dot{\bar{r_2}} $ and rewrite them in terms of the trait variances $\sigma_1^2 = \sum_{ij} p_{ij} (r_i-\bar{r_1})^2$ and $\sigma_2^2 = \sum_{ij} p_{ij} (r_j-\bar{r_2})^2$, and their covariance $\sigma_{12} =\sum_{ij} p_{ij} (r_i-\bar{r_1})(r_j-\bar{r_2})$.
-In doing so, we obtain the system expressed in \eqref{eq:6}.
-% * <masel@email.arizona.edu> 2017-08-22T22:41:29.374Z:
-% 
-% >  equation 6
-% Instead of hard-coding equation numbers, you can create auto-numbered links - ask Jason how.
-Right now the number doesn't match.
-You should also link this more clearly to the way it was presented in the introduction, ie you should explain that this is simply $\Delta \bar{z} = \G \hspace{.05in}\beta$ in the case of two traits both contributing directly to fitness
+% I added a small discussion of the concurrent mutations regime near equation 1.
 %
 % ^.
-% * <masel@email.arizona.edu> 2017-07-29T17:21:52.605Z:
-% 
-% >  We found that each of these rates could then be rewritten in terms of trait variances and covariance
-% This is not something that "we found" or "obtained".
-It is simply a statement of Lande's equation for an appropriate (normalized) definition of what a trait is.
-Also, remember to define the various sigma terms.
-% 
-% ^.
-% * <masel@email.arizona.edu> 2017-07-29T17:18:22.431Z:
-% 
-% > arrived at the expression $\dot{\bar{r}}=\dot{\bar{r_1}} +\dot{\bar{r_2}}$
-% I'm OK going straight to the expression. Your derivation of it is hard to follow, but I'm not convinced it is necessary.
-% 
-% ^.
-\begin{equation}\label{eq:6}
+
+The evolution of the population can be broken down into the evolution of the two traits, according to Lande's equation:
+\begin{equation}\label{eq:5}
 \left[
@@
 \begin{array}{cc}
-\sigma_1^2 & \sigma_{12} \\
-\sigma_{12} & \sigma_2^2 
+\sigma_1^2 & \sigma_{1,2} \\
+\sigma_{1,2} & \sigma_2^2 
 \end{array}
@@
 \right]
-\end{equation}
-The components of the selection gradient and trait response have been normalized in the expression above. \par
-
-Our numerical implementation consisted of two routines which ran iteratively for the duration of the simulation.
-Each iteration began with solving for the time-dependent abundances in the bulk governed by \eqref{eq:2} over a time interval of one-thousand generations.
-We used Mathematica's standard numerical ODE solver to obtain the solutions and used them to determine the appearance time of the next mutant lineage destined to establish.
-Having computed the timing of this event, we evaluated the solutions and accordingly updated the abundances of the bulk.
-Any classes with less than one individual were removed.
-The establishing class was then added to the bulk with an appropriate initial size.
-Following this step, the next iteration began with solving for the abundances of the new bulk, followed by the subsequent steps outlined above.\par
-
-The contributions of the stochastic front were incorporated into the adaptive evolution of the population by simulating each establishment event.
-We did this by identifying the first mutant lineage destined to establish among all mutants appearing in the stochastic front.
-Mutants awhich are parallel to the line shownwhich are parallel to the line shown appear according to a time-inhomogeneous Poisson process with rate functions given by the abundances of the adjacent classes from which they originated.
-Consequently, the arrival time $\tau_l$ of the $l^{th}$ mutant generated from the class $n_{ij}(t)$ having a mutation in trait $k$ can be used to form the random variable
-% * <masel@email.arizona.edu> 2017-08-22T22:55:54.046Z:
+\end{equation}where  $\bar{r_1}=\sum_{i,j} p_{i,j} is_1$, $\bar{r_2}=\sum_{i,j} p_{i,j} j s_2$, $\sigma_1^2 = \sum_{i,j} p_{i,j} (r_i-\bar{r_1})^2$, $\sigma_2^2 = \sum_{i,j} p_{i,j} (r_j-\bar{r_2})^2$, and  $\sigma_{1,2} =\sum_{i,j} p_{i,j} (r_i-\bar{r_1})(r_j-\bar{r_2})$.
+Lande's equation above is expressed as the change in the population's mean trait values over time intervals of size $s$.
+\par
+% * <masel@email.arizona.edu> 2017-10-21T16:44:31.244Z:
 % 
-% > can be used to form the random variable
-% This is confusing. The equation is for the cumulant for the inhomogeneous rate, the equation does not describe a random variable.
+% > have been normalized
+% How have they been normalized?
 % 
-% ^.
-\[ 
-\int_0^{\tau_l} U_k n_{ij}(t) dt ,
-\] 
-which has the distribution of $l$ i.i.d exponentially distributed summed random variables.
-By randomly sampling from the this distribution, we could then solve for each arrival time $\tau_l$.
-We then computed the probability of fixation using \eqref{eq:6}.
-The integral was numerically approximated up to an appropriate cutoff.
-To determine if a mutation would establish, we sampled a uniform random variable and checked it against the calculated $\pi_{fix}$.
-The next class to establish was chosen accordingly and incorporated into the bulk with a random initial size generated with from \eqref{eq:5}.\par
-% * <masel@email.arizona.edu> 2017-08-22T22:57:52.494Z:
+% ^ <kgomez81@math.arizona.edu> 2017-12-05T20:31:15.198Z:
 % 
-% >  \textbf{equation 5}
-% Again, replace all these with auto-numbered links to specific equations, not to hard-numbered text.
+% Sorry, its time that's scaled (not normalized) to Lande's expression above.
 % 
@@
 
-Simulations were initialized with monoclonal populations at carrying capacity consisting of individuals having no beneficial mutations in either trait and relative fitness set zero.
-We constructed three parameter sets, each one varying either $N$, $s$ and $U$ while fixing the others.
-Each was varied across a range constructed from values reported by evolutionary experiments involving either E.
-coli or S.
-cerevisiae \citep{desai2007speed,Levy2015,Perfeito2007}.
-We ran three sets of simulations, each set constructed with one varying parameter among the three.
-The mean values for trait one's variance and covariance were recorded in proportion to the variance of trait one when evolving independently.
-
 We allowed the simulation to run for 5,000 generations to allow the population to achieve beneficial mutation-selection balance before collecting data.
 This ensured that the transient effects from the initializing would not distort the statistics collected for the distribution.
-Following the burn period of the simulation, we recorded the existing classes in the bulk and their respective abundances.
+Following the burn-in period of the simulation, we recorded the existing classes in the bulk and their respective abundances.
 The data was used to compute the marginal mean fitness and variances of each trait, as well as their covariance.
-From these quantities we also computed the instantaneous rate of adaptation \eqref{eq:6}, the mean rate of fitness increase in each trait and the overall rate of adaptation.
+From these quantities we also computed the instantaneous rate of adaptation (Equation \ref{eq:5}) for each trait and overall.\par
+% * <masel@email.arizona.edu> 2017-10-23T23:12:23.801Z:
+% 
+% >  5,000 generation
+% Need to define/explain what a generation is in the model.
+% 
+% ^.
 
@@
 \label{sec:results}
-In our simulations we found that a trait's rate of adaptation decreased as result of evolving with another.
-The reduction is due to negative covariance that forms between the two traits.
-The addition of a second trait results in an increase of trait one's mean variance as well as the formation of mean negative covariance, both of which are plotted in Figure \ref{fig:2}.
-The change in variance which acts to speed up the rate of adaptation is offset by changes in covariance whose magnitude is in every case 40\% larger and always negative.
-The net change in $v_1$ whose value can be  Equation \eqref{eq:6} is 60\% of the initial variance when evolving without the second trait.
+
+The reduction in trait-specific adaptation $v_1$ from $v(U,N,s)$ down to $0.5 \hspace{.02 in} v(2U,N,s)$ (Figure \ref{fig:2}, solid lines), caused by clonal interference with the second trait, is driven by high levels of negative covariance (Figure \ref{fig:2}, closed circles).
+This is partly compensated for by a substantial increase in additive genetic variance (Figure \ref{fig:2}, open diamonds), because the negative covariance slows the removal of additive variance from the population.
+
+In other words, clonal interference in a rapidly adapting population can explain the common observation of high genetic variance in a fitness-associated trait combined with strong negative covariance with other fitness-associated traits, even in the complete absence of pleiotropic trade-offs.
+This is striking, because this kind of evidence is often interpreted as evidence for trade-offs.
 \par
 
-% *****************************************************************************************************************
-% ***********************************FIGURE 2*Scaled Var/Cov Changes & Trait Reduction*****************************
-% *****************************************************************************************************************
-\begin{figure}[h!]
-\begin{subfigure}[b!]{0.48\linewidth}
+\begin{figure}[!ht]
 \centering
-\includegraphics[width=1\linewidth]{fig5d.pdf}
-\caption{}\label{fig:2a}
-\end{subfigure}
-\begin{subfigure}[b!]{0.48\linewidth}
-\centering
-\includegraphics[width=1\linewidth]{fig5e.pdf}
-\caption{}\label{fig:2b}
-\end{subfigure}
+\includegraphics[width=1\linewidth]{figures/fig3.pdf}
+\caption{The rate of adaptation of a focal trait as a function of the number of traits undergoing adaptation.
+.
+decays to zero in the Equation \ref{eq:6}, but in practice the expression is only valid when $kU/s \ll 1$.
+The effects of clonal interference on trait evolution are most significant in the case of two traits.
+The values of the parameters are taken as $s=0.02$, $U=10^{-5}$ and $N=10^9$.}
+% * <masel@email.arizona.edu> 2017-10-28T22:59:26.669Z:
+% 
+% > Each plot includes the measured values produced by simulations with either varying $s$, $U$, or $N$. 
+% Aren't var and covar from simulations and v from the equation?
+% 
+% ^.
+\label{fig:3}
+\end{figure}
 
-\begin{subfigure}[b]{0.48\linewidth}
-\centering
-\includegraphics[width=1\linewidth]{fig5f.pdf}
-\caption{}\label{fig:2c}
-\end{subfigure}
-\begin{subfigure}[b]{0.48\linewidth}
-\centering
-\includegraphics[width=1\linewidth]{fig5g.pdf}
-\caption{}\label{fig:2d}
-\end{subfigure}
-\caption{\footnotesize 
-(a-c) Magnitudes of changes in mean variance, covariance, and rate of adaptation of trait one divided by its expected rate of adaptation when evolving without the second.
-Each plot includes the measured values produced by simulations with either varying $s$, $U$, or $N$.
-(d) The expected decrease in trait one's rate of adapatation for various numbers of traits added.
-Three choices of $s$ are shown to depict its effect.
-Adaptation in a single trait is affected in two ways when evolving with a second trait.
-First, its variance increases due to classes with poorer backgrounds in the first trait persisting longer when they are fit in the second.
-In (a-c), the increase (dashed line) ranges from 60\% to 100\% depending on the parameter.
-The second effect, involves the formation of negative covariance between the two traits.
-Its magnitude (dotted line) is substantially larger, ranging from 80\% to 140\%, and always 40\% larger than the increase in variance.
-This results in a net decrease of 40\% in the rate of adaptation (solid line).
-Fixed simulation parameters in (a-c) are: $s=10^{-2}$, $U=10^{-5}$ and $N=10^9$.}
-\label{fig:2}
-\end{figure}
-% *****************************************************************************************************************
-% ***********************************END*FIGURE*2******************************************************************
-% *****************************************************************************************************************
+The reduction in $v_1$ (Figure \ref{fig:2}, solid lines) is insensitive to changes in the values of  $U$, $N$, or $s$, due to the effects of clonal interference.
+When there are $k$ traits with equal $U$ and $s$, the expected $v_1$ is:
+\begin{equation}\label{eq:6}
+ \tilde{v}_1 (U,s,N;k) 
+= \frac{1}{k}\frac{\log^2[s/U]\log[N^2s(kU)]}{\log^2[s/(kU)]\log[N^2sU]},
+\end{equation}
+in units of multiples of the rate of adaptation in isolation, $v(U,N,s)$.
+This expression is obtained from Equation \ref{eq:1} and plotted in Figure \ref{fig:3}.
+Each additional trait increases the total beneficial mutation rate by $U$, which intensifies the effects of clonal interference from trait interactions.
+The total rate of adaptation that is less than the sum of the individual rates of each trait in isolation.
+The shortfall is measured by the departure of $\tilde{v}_1$ from unity.
+\par
 
-The reduction in $v_1$ in depends weakly on changes in the population parameters within the ranges examined, while changes in mean variances and covariances do exhibit some dependence on them.
-The constancy of changes in $v_1$ to varying choices of $N$, $s$, and $U$ is due to their appearance within logarithmic terms of $v(N,s,U)$.
-Among the three, varying values of $s$ and $U$ have the most significant effect on the measured mean variance and covariance.
-The percent decrease in $v$ expected from adding $n$ traits can be calculated using Desai and Fisher's expression.
-\begin{equation}\label{eq:7}
-\% \Delta v_1 (n) 
-= 1-\frac{1}{n}\frac{\log^2[s/U]\log[N^2s(nU)]}{\log^2[s/nU]\log[N^2sU]}
-\end{equation}
-From Equation \eqref{eq:7} it is clear that the number of traits added has the largest effect on the expected decrease in $v_1$, but the change is marginally less with each trait added.
-Figure \ref{fig:2d} plots this expression for three different choices of $s$ to depict its influence over $\%\Delta v_1$.
-Changes in $N$ and $U$ have little effect on $\%\Delta v_1$ and produce similar points to $s=10^{-2}$ across their ranges explored in simulation.\par
-
-% *****************************************************************************************************************
-% ***********************************FIGURE*3*FLUCTUATIONS*IN*COVARIANCE*******************************************
-% *****************************************************************************************************************
-\begin{figure}[h!]
-\includegraphics[width=1\linewidth]{fig3_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pdf}
-\caption{\footnotesize Time dependent trajectories of variance and covariance of trait one, along with variance in total fitness.
-Their mean values are shown as lines.
-Fluctuations in variance and covariance can be substantially larger than those in total fitness, and lead to unstable an \G  matrix over relatively short periods of time.}\label{fig:3}
-\end{figure}
-% *****************************************************************************************************************
-% ***********************************END*FIGURE*3******************************************************************
-% *****************************************************************************************************************
-
-% *****************************************************************************************************************
-% ***********************************FIGURE*4*COVARIANCE*TIME*SCALES*ARGUMENT*??***********************************
-% *****************************************************************************************************************
-\begin{figure}[h!]
-\includegraphics[width=0.49\linewidth]{fig4a.pdf}
-\includegraphics[width=0.49\linewidth]{fig4b.pdf}
-\caption{\footnotesize The dynamics of the \G  matrix components are governed by time scales of the one dimensional traveling wave, $\tau_q$.
-Variances and covariance are determined by the set of classes which makeup the peak of the one dimensional traveling wave due to its Gaussian nature.
-At time $t$ the peak is composed of classes shown in blue, whose frequencies determine the components of the \G  matrix.
-At time $t+tau_q$ the mean shifts, and the set of classes shown in green become the dominant group.
-Accordingly, the variance and covariance of the two dimensional distribution transitions to new values that depend on the frequencies of green set of classes.}\label{fig:4}
-\end{figure}
-% *****************************************************************************************************************
-% ***********************************END*FIGURE*4******************************************************************
-% *****************************************************************************************************************
-
-In our model the components of the \G matrix fluctuate significantly over time even while the population is in beneficial mutation-selection balance.
-Figure \ref{fig:3} includes the trajectories of variance in trait one, covariance, and variance in total fitness, as well as their means, for one of our simulations.
-The dynamics of variance and covariance are highly correlated due to their adherence to Equation \eqref{eq:6}, whose left hand side is approximately constant under beneficial mutation-selection balance.
+% * <masel@email.arizona.edu> 2017-10-28T23:18:33.323Z:
+% 
+% > decreases in $v_1$ from additional traits are larger with increasing $s$
+% I don't see this clearly in the figure - are you sure this is worth commenting on?
+% 
+% ^.
+% * <masel@email.arizona.edu> 2017-10-28T23:09:27.018Z:
+% 
+% > Changes in $s$ are the most influential at smaller effect sizes.
+% Needs more explanation. Why do you say this? Are you just observing from Figure 2a? The more natural axis for Figure 2a, given Equation 6, would be a log-scale.
+% 
+% Isn't the more important point that s is more important than U or N, at least over the parameter range considered?
+% 
+% ^.
+Under beneficial mutation-selection balance, the components of the \G matrix fluctuate significantly over time, far more than the instantaneous rate of adaptation, with transient spikes in additive genetic variance $\sigma_1^2$ mostly canceled out by transient spikes in negative covariance $\sigma_{1,2}$ (representative simulation shown in Figure \ref{fig:3}).
 As a result, fluctuations in variance and covariance can be substantially larger than those of total variance in fitness.\par
 
-The time scales governing changes in the components of the \G  matrix are defined by the dynamics of the one dimensional traveling wave.
-Since its shape is Gaussian, classes belonging to its peak will dominate the population as shown in Figure \ref{fig:4}.
-These are classes that reside on the line of constant fitness passing through the populations mean fitness.
-The variance and covariance will depend on their abundances.
-As the population mean fitness shifts to the next line of constant fitness, a new set of classes becomes dominant and leads to new values of variance and covariance.
-On average, the time required for this to occur is the mean time between establishments at the nose of the of the traveling wave, denoted $\tau_q$ in \citet{desai2007beneficial}.
+\begin{figure}[!ht]
+\includegraphics[width=1\linewidth]{figures/fig4.pdf}
+% * <masel@email.arizona.edu> 2017-10-28T23:28:47.376Z:
+% 
+% > fig3.pdf
+% Need tick labels on the x-axis.
+% 
+% ^.
+\caption{Trajectories of the variance of trait one (dashed line), covariance (dotted line), and the total variance in fitness of the population (solid line), plotted over 10,000 generations.
+The mean values of each are included as lines with matching line-styles.
+The \G matrix over this period of time is unstable.
+Its two components, trait one variance and covariance, fluctuate significantly relative to the variance in total fitness over identical time periods.
+The trajectories were sampled from one simulation with parameters $s=0.02$, $U=10^{-5}$ and $N=10^9$.}\label{fig:4}
+% * <masel@email.arizona.edu> 2017-10-28T23:32:29.075Z:
+% 
+% > along with variance in total fitness
+% I thought you were plotting sigma_1^2+sigma_12 which is the instantaneous adaptation rate of trait 1, but here you say that you are plotting something else. Please confirm.
+% 
+% ^.
+\end{figure}
+
+The two-dimensional traveling wave of Figure \ref{fig:1} can be projected onto a one-dimension traveling wave of fitness, where classes along a fitness isocline such as those along the red line in Figure \ref{fig:1}, are aggregated into fitness classes (Figure \ref{fig:4}).
+The shape of the one-dimensional fitness distribution is Gaussian and highly peaked near the population's mean fitness \citep{desai2007beneficial}.
+This means that the \G matrix will dominated by the distribution of genotypes in this fitness class (red line in Figure \ref{fig:1}, mode in (Figure \ref{fig:4}).
+As adaptation continues, a fitter class becomes the most abundant that will have a different distribution of genotypes and hence different trait variances and covariance.
+On average, the time required for this shift is the mean time between establishments at the nose of the of the traveling wave, denoted $\tau_q$ in \citet{desai2007beneficial}.
+
 \begin{equation}\label{eq:8}
 \tau_q = \frac{\log^2(s/U)}{s(2\log(Ns)-\log(s/U))}
-\end{equation}
-The relative frequencies of classes residing on a line of constant fitness are set by the timing of their establishments when formed part of the stochastic front.
-Since these events depend on the abundances of classes on the line of constant fitness below, the values of variances and covariance will be correlated from one transition to the next.
-However, these correlations vanish after a period of time on the order of that is roughly equal to the time required for classes in the stochastic front become the dominant group.
+\end{equation}\par
+
+Genotypes frequencies within the peak of the traveling wave are set during the stochastic growth phase of the fitness class prior to establishing \citep{Desai2013}.
+As previously discussed, beneficial mutations that give rise to mutants in the nose produce distinct but equally fit mutant lineages.
+Many of lineages become extinct, but those that manage to survive are subsequently and equally amplified by selection following establishment.
+Since the relative sizes of the surviving lineages with respect to one do not change once they begin growing deterministically, the frequencies of genotypes in a fitness class become ``frozen'' at the time of establishment.
+Beneficial mutations that occur after that time contribute very little to the size and composition of fitness class, since their corresponding lineages will be exponentially smaller than those which initially appeared.
+After a fitness class has established, it eventually goes on to become the dominant group of the fitness distribution, or rather, it forms the peak of the traveling wave.
+The time between its establishment and its succession as the dominant fitness class is called the sweep time (denoted $\tau_{sw}$).
+Consequently, the frequencies of genotypes in the nose determine the values of variance and covariances approximately $\tau_{sw}$ generations later.
+Figure \ref{fig:6} plots the correlation between trait covariances of the bulk and nose, as a function of a time offset.
+The signal peeks at approximately $\bar{\tau}_{fix}$ (mean time to fixation).
+It shows that covariances between traits are largely set by the behavior at the nose of the one dimensional traveling wave, which consists of genotypes along the lead fitness isocline through the stochastic front.
+
+\begin{figure}[!ht]
+\includegraphics[width=0.49\linewidth]{figures/fig5a.pdf}
+\includegraphics[width=0.49\linewidth]{figures/fig5b.pdf}
+\caption{\footnotesize The dynamics of the \G matrix components are governed by time scale $\tau_q$ of the one-dimensional traveling wave of fitness classes.
+Variances and covariance are dominated by the set of genotypes that make up the peak of the one-dimensional traveling wave, whose shape is approximately Gaussian.
+At time $t$ the peak is composed of classes shown in blue, whose frequencies determine the components of the \G  matrix.
+At time $t+\tau_q$ the mean shifts (relatively abruptly, due to the exponential growth of classes fitter than the mean), and the set of genotypes shown in green become the dominant group, whose distribution dominates the new trait variances and covariance.}\label{fig:5}
+% * <masel@email.arizona.edu> 2017-10-28T23:57:21.626Z:
+% 
+% > whose shape is approximately Gaussian
+% The text hasn't really explained or provided a citation for this point.
+% 
+% ^.
+\end{figure}
+
+\begin{figure}[!ht]
+\includegraphics[width=1\linewidth]{figures/fig6a.pdf}
+\caption{Correlation between covariance of stochastic front and bulk.
+The correlation between the two peaks when their signals are offset by the the mean time to fixation $\bar{\tau}_{fix}$ (marked by the blue line), indicating that the dynamics of the stochastic front explains 69\% of the covariance lated detected in the bulk $\bar{\tau}_{fix}$ generation after.}\label{fig:6}
+% * <masel@email.arizona.edu> 2017-10-28T23:57:21.626Z:
+% 
+% > whose shape is approximately Gaussian
+% The text hasn't really explained or provided a citation for this point.
+% 
+% ^.
+\end{figure}
+
+Because mutational influx depends on the abundances of adjacent less-fit genotypes, the values of variances and covariance will be correlated from one transition to the next.
+However, these correlations vanish after a period of time roughly equal to the time required for classes in the stochastic front become the dominant group.
 This time period can be expressed as the product $q \tau_q$, where $q = 2\log(Ns)/\log(s/U)$ is the lead of the traveling wave nose over its mean.
+Over these periods of time $q \tau_q$, fluctuations in covariance can be significantly larger than those occurring over periods equal to $\tau_q$, resulting in an unstable \G-matrix over the time scale of adaptation.
+\par
+% * <masel@email.arizona.edu> 2017-10-29T00:04:55.206Z:
+% 
+% > these correlations vanish after a period of time roughly equal to the time required for classes in the stochastic front become the dominant group
+% Why is this?
+% 
+% ^.
 
@@
 
-%An interesting consequence of the dynamics of covariance invovles the relationship between stochastic front and its role in setting the covariance of the two dimensional trait distribution.
+%An interesting consequence of the dynamics of covariance involves the relationship between stochastic front and its role in setting the covariance of the two dimensional trait distribution.
 In between establishement events for the one dimensional traveling wave, classes that lie along the line of constant fitness passing through the fittest part of the stochastic front their relative size to one another.
@@
 We can assume that specific alleles at either locus correspond to the number of beneficial mutations that have accumulated in the corresponding trait.
-It follows that the $p_{ij}$ are the gametic frequencies for the genome consisting of allele $i$ paired with allele $j$, and $D_{ij} = p_{ij}-p_{i\cdot} p_{\cdot j}$ is their coefficient of linkage disequilibrium ( $p_{i\cdot}$ and $p_{\cdot j}$ are marginal distributions).
+It follows that the $p_{i,j}$ are the gametic frequencies for the genome consisting of allele $i$ paired with allele $j$, and $D_{i,j} = p_{i,j}-p_{i\cdot} p_{\cdot j}$ is their coefficient of linkage disequilibrium ( $p_{i\cdot}$ and $p_{\cdot j}$ are marginal distributions).
 By applying \textbf{equation 2}, we can derive a mathematical relationship between covariance and linkage disequilibrium (need to cite chapter 5, Walsh and Lynch).
@@
 An allele cannot be a "number of beneficial mutations", because an allele must pertain to one locus.
-Stick to the word "genotype" not "allele", eg "genotype ij has i beneficial mutations in trait 1 and j beneficial mutations in trait 2 compared to the ancestral population".
+Stick to the word "genotype" not "allele", eg "genotype i,j has i beneficial mutations in trait 1 and j beneficial mutations in trait 2 compared to the ancestral population".
 This means that you cannot calculate linkage disequilibrium as you do: linkage disequilibrium is generally defined as being between two particular beneficial mutations.
@@
 %\begin{equation}
-%\sigma_{12}=\sum_{ij}D_{ij}[(i-\bar{i})s_1][(j-\bar{j})s_2]. 
+%\sigma_{1,2}=\sum_{i,j}D_{i,j}[(i-\bar{i})s_1][(j-\bar{j})s_2]. 
 %\end{equation} 
-%As the population adapts, mutations and selection  increase and decrease the amounts LD which ultimately determine how traits interact in their evolution.\par
+%As the population adaptsAcknowledgments, mutations and selection  increase and decrease the amounts LD which ultimately determine how traits interact in their evolution.\par
 %
@@
 \label{sec:acknowledgments}
+We thank Patrick Phillips for helpful discussions. Funding was provided by the National Science Foundation (DEB-1348262) and the National Institutes of Health (T32 GM084905).
+\bibliography{biblio}
 
-\bibliography{biblio}
-\bibliographystyle{plainnat}
+\section*{Appendix}
+The derivation of the expression given in \ref{eq:4} from Lande's equation is as follows.
+The derivative measuring the change in the mean of a trait $\dot{\bar{r_1}} \approx \Delta \bar{r_1}/\Delta t$ for a sufficiently small interval of time.
+Lande's original expression yields,
+\[
+\left[
+\begin{array}{c}
+\Delta \bar{r_1} \\
+\Delta \bar{r_2} 
+\end{array}
+\right]
+=
+\left[
+\begin{array}{cc}
+\sigma_1^2 & \sigma_{1,2} \\
+\sigma_{1,2} & \sigma_2^2 
+\end{array}
+\right]
+\left[
+\begin{array}{c}
+\beta_1 \\
+\beta_2 
+\end{array}
+\right]
+\]
+The selection gradient on the right hand side is the change in fitness with respect to change in the trait value, indicating that $\beta_1=\beta_2=s$.
+Dividing both sides by $\Delta t$ allows us to equate the left side to the column vector $(\dot{\bar{r_1}},\dot{\bar{r_2}})^T$, while the scaled selection gradient becomes $(s/\Delta t,s/\Delta t)$.
+Setting $\Delta t = s$ provides equation 5.
 
-\end{document}
+\end{document}Acknowledgments

### 2018-02-12 manuscript.tex
added new references, updated local copy of the manuscript, and changed figure 4 in the plots code.

@@
 \usepackage[nameinlink,capitalise]{cleveref}
+\graphicspath{{figures/}}
 \newcommand{\G}{\textbf{G }}
@@
 The role of genetic correlations in the response of quantitative traits to natural selection is described by Lande's equation ($\Delta \bar{z} = \G \hspace{.05in}\beta$), where $\Delta \bar{z}$ is the expected change in the population mean of a trait $\Delta \bar{z}$, $\beta$ is the selection gradient, and the components of the \G  matrix are the additive genetic covariances between traits.
-Lande's equation governs the short-term evolution of correlated quantitative traits under natural selection, but it may play a role in bridging the gap between micro- and macro-evolution \citep{Arnold2001,Arnold2008,Schluter1996,Schluter2000}.\par
+Lande's equation governs the short-term evolution of correlated quantitative traits under natural selection.
+Genetic variances and covariances can also evolve over appreciable time scales, and as a result, Lande's expression alone cannot tell us how correlated quantitative traits will change in the long-run.
 
-When the \G matrix is stable, it can be used to extrapolate future trait evolution from historical patterns, and to infer past selection gradients.
-The requirement for a stable \G matrix has motivated empirical work to measure and compare \G in related species.
-At least portions of the \G matrix may be conserved among certain phylogenetic groups \citep{Arnold1999,Roff2000,Steppan2002}, but other studies also indicate that \G is unstable.
-***
+When the \G matrix is stable, it is possible to infer past selection gradients and extrapolate future trait evolution (both direction and rate) from historical patterns.
+For example, studies of past macro-evolution in populations and species have used net selection gradients to draw connections between morphological evolution and reproductive isolation \citep{price1984genetic}, infer morphological and phylogenetic relations \citep{schluter1984morphological}, and determine the extent of local adaptation between populations \citep{dudley1996}.
+Similar work has looked the impact of genetic constraints on rates of adaptation \citep{teplitsky2014assessing,teplitsky2011quantitative}, their ability to bias phenotypic evolution over millions of years \cite{Schluter1996}.
 
-Theoretical quantitative genetic models have also been used to examine the conditions under which the stability of the \G matrix can be expected \citep{Turelli1988}.  \par
-% * <masel@email.arizona.edu> 2017-10-06T23:52:45.315Z:
-% 
-% > other studies also indicate that \G is unstable
-% Citations(s) needed for empirical work for instability of G and theoretical work either deriving or assuming stability.
-% 
-% ^.
+The importance of \G's stability in evolutionary inferences such as these has motivated a large body of empirical work focused on measuring and comparing it between populations.
+Comparative studies over the decades have used different metrics and statistical procedures to check whether populations have similar \G matrices.
+Each of these metrics vary in their ability to detect conserved structure in genetic variances and covariances \cite{Steppan2002}.
+Matrix correlations and maximum liklihood tests, for example, often test for correlations between \G matrices or and their components, but do not give additional information about other important characteristics that can be shared between \G matrices.
+Methods based on the use of principle components, such as common principle component analysis (CPCA), in contrast, can compare features between \G matrices such as shape, size, and orientation.
+These methods have gained popularity for this reason, and because they are capable of revealing which evolutionary forces may have significantly influenced the past dynamics of the population's \G matrix \cite{Phillips1999,Arnold2008}.
 
-The genetic correlations described by \G are due to some combination of pleiotropy and linkage disequilibrium \citep{Saltz2017}.
+A significant number of comparative studies examining either natural populations or laboratory experiments show that at least some features of the \G matrix are relatively stable.
+In over 70\% of comparisons between conspecific groups and species find that \G matrices have similar shape and orientation \cite{Arnold2008}.
+These include populations that have diverged over millions of years \cite{Arnold1999}, and across distinct environments \cite{puentes2016similarity,teplitsky2014assessing}, suggesting that aspects of genetic constraints can be maintained over significant time scales and across a diverse set of environmental factors.
+There are also comparative studies that find no related structure between \G matrices \cite{wilkinson1990resistance,roff1999does,Steppan1997,Waldmann2000}, and others that reveal large instabilities in \G over very short time scales  \cite{bjorklund2013rapid,Siren2017}.
+% \cite{roff1999does} \cite{Begin2003,Begin2004}
+
+Quantitative genetic models of genetic variances and covariances provide some indication of how ecological and evolutionary forces contribute to stability in \G's dynamics \citep{Turelli1988,barton1989evolutionary,reeve2000predicting}.
+For example, changes in genotype $\times$ environment interactions directly affect genetic variances and covariances and lead to significant changes in \G over few generations.
+Other critical stability conditions involve the strength of selection and mutational covariances.
+Further analytical insights from these models require knowledge of unknown genetic details, such as the distribution of allelic effects \cite{burger2000mathematical}.
+Simulations have been used to bypass this problem and examine the dynamics of \G in more detail \cite{Jones2003,Jones2007}.
+In both theoretical analysis and numerical simulations, selection and mutations play an important role in ensuring the stability of the \G matrix, but the effects of the latter are largely assumed by most quantitative geneticists to be the result of pleiotropic effects \cite{lande1979quantitative,lande1980genetic,Arnold2008}(I have more citations, and will insert them).
+
+The genetic correlations described by \G are due to some combination of pleiotropy and linkage disequilibrium \cite{Saltz2017}.
 Historically, pleiotropy was favored by the Edinburgh school, who argued that recombination would quickly eliminate linkage disequilibrium \citep[Chapter~20]{fox2006evolutionary}, particularly in the randomly mating animal populations that the Edinburgh school focused on.
@@
 We know far less about the about the dynamics of genetic correlations arising from linkage disequilibrium.\par
-% * <masel@email.arizona.edu> 2017-10-24T22:58:18.095Z:
+
+To study linkage disequilibrium, we have to consider explicit alleles or genotypes, not just quantitative traits.
+Beneficial mutations are discrete not infinitesimal, and appear on distinct genetic backgrounds.
+Selective sweeps compete with one another, and so many beneficial mutations are lost in a process known as ``clonal interference''.
+This will slow down adaptation \citep{hill1966effect} unless recombination brings them together, reducing the negative linkage disequilibrium between competing beneficial mutations \cite{fisher1930genetical,muller1932some}.
+Classical models of clonal interference considered only two loci each with two alleles.
+More recently, traveling wave models have been developed to describe clonal interference among large numbers of loci, arising at an arbitrary mutation rate \citep{rouzine2003solitary,desai2007beneficial,park2010speed,good2012distribution,neher2010rate,fisher2013asexual,rouzine2010multi,rouzine2007highly,rouzine2008traveling}.
+These models have documented that clonal interference can have an enormous impact on adaptation rates.
+However, to date, most of these models have treated only the evolution of ``fitness'' rather than the evolution of individual traits and their correlations \cite{}.
+\par
+
+% * <masel@email.arizona.edu> 2018-01-24T22:40:30.992Z:
 % 
-% > most models for the evolution of the \G  matrix assume that genetic correlations are pleiotropic and go on to examine how evolutionary processes influence them \citep{Turelli1988
-% More citations needed
+% > owever, to date, these models have treated only the evolution of ``fitness'' rather than the evolution of individual traits and their correlations. 
+% You mentioned one paper that might be an exception - find it!
+% 
+% ^ <kgomez81@math.arizona.edu> 2018-02-10T00:01:15.252Z:
+% 
+% I am still looking for this reference.
+%
+% ^.
+% * <masel@email.arizona.edu> 2018-01-24T22:37:04.241Z:
+% 
+% >  \citep{rouzine2003solitary,desai2007beneficial,park2010speed,good2012distribution}
+% You can include sexual models here too, eg by Neher
+% 
+% ^ <kgomez81@math.arizona.edu> 2018-02-09T23:38:56.909Z:
+% 
+% I added the Neher paper, and other references. I'll conduct another search for other sources. 
 % 
@@
 
-Biology today has come to recognize the dominance of relatively asexual microbes \cite{mcfall2013animals}, for whom the Birmingham emphasis on linkage disequilibrium as a source of genetic correlations cannot be ignored.
-Linkage disequilibria certainly persist long enough to affect adaptation in asexuals; this fact has, for example, been the basis of explanations for the evolutionary advantages of sex and recombination \citep{Barton2005,Otto2009}.
-\par
-% * <masel@email.arizona.edu> 2017-10-06T22:23:25.138Z:
-% 
-% > Biology today has come to recognize the dominance of relatively asexual microbes. 
-% Cite www.pnas.org/cgi/doi/10.1073/pnas.1218525110 and maybe others making complementary points.
-% 
-% ^ <kgomez81@math.arizona.edu> 2017-12-05T18:40:59.153Z:
-% 
-% I added the citation.
-%
-% ^.
-Asexual evolution is normally approached using a population genetic rather than a quantitative genetic approach.
-In particular, traveling wave models have been successfully used to describe the rate of multi-locus adaptation in asexuals given clonal interference \citep{rouzine2003solitary,desai2007beneficial}.
-The basic model has only three parameters: fixed population size $N$, beneficial mutation rate $U$, and per-beneficial-mutation selection coefficient $s$.
-Adaptation, i.e.
-the increase in mean population fitness, occurs at the rate given by Desai and Fisher \citep[Equation 41]{desai2007beneficial}, expressed as
-% * <masel@email.arizona.edu> 2017-10-14T17:09:10.651Z:
-% 
-% > Adaptation, i.e. the increase in mean population fitness, occurs at rate
-% Cite the specific equation number in Desai & Fisher. Also give the parameter value conditions for which this equation holds.
-% 
-% ^ <kgomez81@math.arizona.edu> 2017-12-06T20:47:24.968Z:
-% 
-% I added the equation number to the citation.
-%
-% ^.
+Here, we focus on asexual populations, because they have the strongest clonal interference and resulting linkage disequilibrium.
+Moreover, biology today has come to recognize the numerical prevalence and ubiquitous significance of relatively asexual microbes \cite{mcfall2013animals}.\par
+
+We begin with Desai and Fisher's \citep{desai2007beneficial} framework of fixed population size $N$, beneficial mutations at rate $U$, and per-beneficial-mutation selection coefficient $s$.
+This yields the rate of adaptation \citep[Equation 41]{desai2007beneficial}
 \begin{equation}\label{eq:1}
@@
 \end{equation}
-The expression holds when the asexual population is adapting in what is defined by Desai and Fisher as the ``concurrent mutations regime.'' In a monoclonal population, selection will cause a beneficial mutation to spread until all individuals carry the new mutation, at which point, it is said to have fixed in the population.
-The time required for beneficial mutation to fix is called its fixation time.
-For asexuals adapting in the concurrent mutations regime, beneficial mutations appear in rapid succession relative to the fixation time.
-This produces many mutant lineages that vary in the numbers of beneficial mutations that have accumulated in their genome.
-Beneficial mutations that appear on less fit lineages are lost due to clonal interference.
-Increasing the number of beneficial mutations that enter the population increases the proportion of beneficial mutations that are wasted.
-The logarithmic dependence of $v(U,N,s)$ on the parameters captures these effects.
-An asexual population is in the concurrent mutations regime when the parameters satisfy the inequality $NU \gtrsim 1/\ln(Ns)$.
+Equation \ref{eq:1} holds when the parameters satisfy $NU \gtrsim 1/\ln(Ns)$ and $U/s \ll 1$ (the ``concurrent mutations regime''  where beneficial mutations appear rapidly relative to the time required for any one of them to fix).
+When combined with Fisher's fundamental theorem, Equation \eqref{eq:1} also gives the additive variance in fitness $\sigma^2$.
+When there is only one adaptive trait, $\sigma^2$ is that trait's additive variance.
 \par
 
-Equation \ref{eq:1}, combined with Fisher's fundamental theorem, also gives the additive variance in fitness $\sigma^2$.
-With only one adaptive trait,  $\sigma^2$ also gives that trait's additive variance.
-With two adaptive traits, Equation \ref{eq:1} still gives overall fitness variance $\sigma^2$, but nothing can be said about the variances and covariance of the two traits.\par
+With two adaptive traits, Equation \eqref{eq:1} still provides overall fitness variance $\sigma^2$, but it does not give its decomposition into the variances and covariance expressed in Lande's equation.
+Consider just two traits, each experiencing beneficial mutations at rate $U$ which each increase fitness by $s$.
+Equation \ref{eq:1} indicates that the overall adaptation rate is $v(2U,N,s)$, and so by symmetry, adaptation in each trait alone occurs at rate $0.5 \hspace{.02 in} v(2U,N,s)$.
+This is lower than the rate $v(U,N,s)$ when the other trait is not evolving.
+The reduction in trait-specific adaptation rate must be distributed between the two terms of Lande's equation $v_1 =\sigma_1^2 +\sigma_{1,2}$.
+More importantly, the contributions of each term will over time change if the \G matrix is not constant.\par
 
-Adaptation in one trait proceeds more slowly when there is clonal interference with mutations that are adaptive in a second trait.
-Quantitatively, when the two adaptive traits each have mutations with fitness benefit $s$ appearing at rate $U$, the overall adaptation rate is $v(2U,N,s)$, and so by symmetry, adaptation in each trait alone occurs at rate $0.5 \hspace{.02 in} v(2U,N,s)$.
-This is lower than the rate $v(U,N,s)$ when the other trait is not also evolving.
-The reduction in trait-specific adaptation rate must be distributed between the two terms of Lande's equation $v_1 =\sigma_1^2 +\sigma_{1,2}$.
-Here, we will calculate how the reduction in trait-specific adaptation partitions between the variance in that trait and its covariance with the other.\par
-
-% * <masel@email.arizona.edu> 2017-07-29T13:00:14.234Z:
-% 
-% > Clearly, we lack any information about how trait two is affecting trait one's evolution,
-% Spell out what exactly it is that we don't know, ie the degree to which the reduction in adaptation rate of trait 1 is due to a change in variance or a change in covariance.
-% 
-% 
-% ^ <kgomez81@math.arizona.edu> 2017-08-22T14:11:45.747Z.
-
-To explore how adaptation in one trait is affected by evolution in another, we extend the traveling wave model approach to a two dimensional trait space.
-We assume that each of the two fitness-determining traits is controlled by different loci, i.e.
-there is no pleiotropy.
-With pleiotropic contributions to \G excluded by design , we use this model to examine the dynamics of genetic correlations arising solely from linkage disequilibrium.
+Here we analyze a two-dimensional traveling wave model of an asexual population, and find that clonal interference alone, in the absence of pleiotropy, leads on average to greatly elevated variance of fitness-associated traits and to negative covariance between them.
+However, the \G matrix is highly unstable over the timescale of selective sweeps.
 
@@
 
-We consider an asexual population of haploids with fixed total population size $N$.
-To keep the model simple, we assume that beneficial mutations occur with the same rate of $U$ mutations per birth per trait with fitness benefit $s$ in each trait and no epistasis.
+We consider an asexual population of haploids with fixed total population size $N$ evolving in continuous time.
+To keep the model simple, we assume that beneficial mutations occur with the same rate of $U$ mutations per trait per birth giving a fitness benefit $s$ with no epistasis.
 Deleterious mutations are not considered.
-Individuals with $i$ mutations in the first trait and $j$ in second, have relative fitness $r_{i,j}= i s+j s$ and absolute fitness $r_{i,j}-\bar{r}$, where $\bar{r}$ is the population mean fitness.
-We refer to each such set of individuals as a ``class'' and denote their abundances and frequencies by $n_{i,j}$ and $p_{i,j}$.
+Since there is no epistasis, all individual with $i$ mutations in the first trait and $j$ in the second have the same Malthusian relative fitness $r_{i,j}= i s+j s$.
+We refer to each such set of individuals with the same $i$ and $j$ as a ``class'' and denote their abundances and frequencies by $n_{i,j}$ and $p_{i,j}$ respectively.
+The population's mean fitness is denoted $\bar{r}$.
+
+We assume that $N$ is so large that selection is strong relative to drift ($1/N \ll s$).
+In this case, most classes are so large that their frequency grows or declines approximately deterministically due to selection.
+We refer to these classes as the ``bulk''  (Figure \ref{fig:1}, solid cells).
+However, adaptive mutant lineages arising along the high-fitness ``front'' of the population will start with a single individual, and initially behave stochastically (Figure \ref{fig:1}, patterned cells; arrows show mutations).
+Many of these lineages will go extinct before attaining appreciable frequencies, but some will make it to high enough abundances that they begin to grow deterministically.
+This transition from stochastic to deterministic behavior is called ``establishment'' and occurs if the lineage reaches an abundance of order $1/s$.
 \par
 
-We assume $1/N \ll s \ll 1$.
-Following the convention of Desai and Fisher, we refer to the set of classes abundant enough to behave deterministically $(n_{i,j}>1/s)$ as the bulk, and to smaller adaptive classes ($r_{i,j}>\bar{r}$) as the stochastic front (Figure \ref{fig:1}, solid and patterned, respectively).
-The stochastic front shares the same role of the nose in Desai and Fisher's traveling wave model \cite{desai2007beneficial}.
-Like the nose, the stochastic front contains the beneficial mutations (arrows in Figure \ref{fig:1}) that will drive future adaptation in the bulk; mutations within the bulk are wasted.
-However, the front in our model is evolving boundary rather than a point.
-Classes in the front become part of the bulk once they have ``established'' (grown sufficiently large).
-This shifts the position of  the bulk, with the stochastic front shifting also, to form a two dimensional traveling wave.
+The evolution of population follows an iterative process with two steps.
+First, the next class to establish is determined, and its depends on the growth of classes producing adaptive mutants in the front (\ref{fig:1}, arrows).
+Once a class has established, it begins to alter the growth dynamics of the other classes in the bulk.
+The second step involves producing the solutions for the dynamics of the bulk with the newly established class.
+Having done so, the process then repeats with determining the next class from the front to establish using the solutions obtained.
+
+To simulate the timing of the next establishment, stochastic fluctuations induced by drift and mutations must be accounted for in the growth of classes at the front.
+These growth of classes in the front follow a branching processes fed by mutants from the adjacent (and exponentially growing) classes in the bulk.
+Each of the mutant lineages that arise will either go extinct, or grow sufficiently large to experience deterministic growth from selection.
+The chances of the latter happening is given by the probability of establishment, denoted $\pi$.
+We use the value of $\pi$ given by the expression:
+\begin{equation} \label{eq:2}
+\pi = \frac{s_{i,j}}{ 1+s_{i,j}},
+\end{equation}and derived from \cite[Equation 16]{desai2007beneficial}.
+Here $s_{i,j}=(r_{i,j}-\bar{r})$ is the selective advantage of an individual in the class with $i$ mutation in the first trait, and $j$ mutations in the second, with respect to an average individual in the population.
+Although classes in the front can consist of many mutant lineages that contribute to their size during the stochastic growth phase, their establishment  is driven by the size of the first lineage destined to establish.
+In particular, the time required for this single lineage to transition into the deterministic growth phase is a good approximation for the actual time of establishment of its class.
+The justification for this approximation is discussed in detail by \cite[p.1773]{desai2007beneficial}, and intuitively has to do with the fact that lineages from early mutation events often quickly die out.
 \par
-% * <masel@email.arizona.edu> 2017-08-22T19:17:50.152Z:
-% 
-% > those that grow stochastically and have a positive selective advantage ($r_{i,j}>\bar{r}$) will be referred to as the stochastic front
-% This is a good place to point a key difference between you and prior work, namely that your front is a curve not a point.
-% 
-% ^ <kgomez81@math.arizona.edu> 2017-08-23T17:03:48.195Z:
-% 
-% I added a bit more after this sentence, and changed the subsequent one one more sentence to indicate the differences in between Desai and Fisher's traveling wave nose.
-"The stochastic front shares the same role of the nose in Desai and Fisher's traveling wave model\cite{desai2007beneficial}.
-Like the nose, the stochastic front contains the beneficial mutations (arrows in Figure \ref{fig:1}) that will drive future adaptation in the bulk; mutations within the bulk are wasted."
-% 
-% ^.
 
-\begin{figure}[!ht]
-\includegraphics[width=1\linewidth]{figures/fig1a.pdf}
-\caption{Illustration of a two dimensional trait distribution taken from a simulation.
-Individuals having an equal number of beneficial mutations in each trait are grouped into classes (boxes).
-Classes growing deterministically form part of the bulk (solid interiors), and those growing stochastically belong to stochastic front (patterned interiors).
-The red dot marks the average number of beneficial mutations in each trait.
-Fitness-isoclines identify the set of equally fit classes and are parallel the population mean fitness isocline (dashed red line).
-In asexuals, beneficial mutations must occur on the fittest backgrounds in order to contribute to the adaptive process.
-These mutations are represented by arrows from the bulk into the stochastic front.
-Over time, classes in stochastic front become part of the bulk once their abundances have become sufficiently large, while new ones subsequently emerge and become part of the stochastic front.
-Classes in the bulk below the population mean fitness isocline dwindle in size by selection.
-The distribution shifts direction of increasing relative fitness as result of these two processes and produce a two-dimensional traveling wave.
-(Simulation parameters: $N=10^9$, $s=0.02$, and $U=10^{-5}$)}\label{fig:1}
-% * <masel@email.arizona.edu> 2017-10-14T18:05:49.152Z:
-% 
-% > These groups are represented by the lines passing through their centers that are parallel to the dashed line.
-% Do you want to put these lines into the figure, or take this sentence out of the figure legend?
-% 
-% 
-% ^.
-\end{figure}
+The next establishment event and its corresponding class  is determined by sampling the each of the sojourn times of mutations in the front that are destined to establish and identifying the first to appear.
+Beneficial mutations within the bulk are ignored because these are lost to clonal interference.
+At the front, the number of beneficial mutations from a class $(i,j)$ in the bulk follows an inhomogeneous Poisson  process with a rate $\lambda_{i,j}(t)$.
+The rate will depend on the birth rate of individuals in the class $(i,j)$, their time-dependent abundances, and the beneficial mutation rate (see appendix).
+The sojourn time $\tau$ for the appearance of the first mutation destined to establish can be transformed into the random variable $x=\int_0^\tau \pi_{fix} \lambda_{i,j}(t) dt$ that has a standard exponential distribution.
+The sampling of $\tau$ is be done by generating a value for  $x$ and numerically solving for $\tau$ in the integral equation.
+This is repeated for each class of the bulk adjacent to the stochastic front, and the next class to establish is picked to be the one with the smallest sojourn time.
+\par
 
-Changes in abundances within the bulk follow a system of ordinary differential equations given in \cite[page 27]{Crow1970} :
-\begin{equation} \label{eq:2}
-\dot{n}_{i,j}(t) =\left (r_{i,j}-\frac{N(t)}{N}\bar{r}\right) n_{i,j}(t),
+A class that is chosen to establish is immediately incorporated into the bulk with an initial size that reflects stochastic effects on the growth of the single mutant lineage that drives its establishment.
+If a mutant lineage that is destined to establish appears at time $t=0$, it requires $T$ generations to grow to size $1/s_{i,j}$.
+The variable $T$ is random due to the effects of drift on the growth of the lineage, but its distribution is known and given by \cite[Equation 12]{desai2007beneficial}.
+Once this lineage establishes $T$ generations later, the time dependent growth of its class is approximately $n_{i,j}(t) =s_{i,j}^{-1} \exp[{s_{i,j}(t-T)}]$.
+We consider instead an initial size  $\nu = s_{i,j}^{-1}\exp[-s_{i,j}T]$, and equivalently regard $n_{i,j}(t) = \nu s_{i,j}^{-1} \exp[{s_{i,j}t}] $ as describing the deterministic growth of the class beginning at time $t=0$.
+Consequently, each class that establishes can be incorporated into the bulk from the moment at which mutation driving establishment appears, but its initial size will be $\nu$.
+The random variable $\nu$ can be sampled by using its cumulative distribution function:
+\begin{equation}\label{eq:3}
+P(\nu \leq \nu_0) = 1- e^{-\pi_{fix} \nu_0},
 \end{equation}
-where time $t$ is measured in generations.
-At capacity, Equation \ref{eq:2} gives the selective advantage of a class relative to population mean fitness, denoted $s_{i,j}$, whose value is approximately  $s_{i,j} \approx (r_{i,j}-\bar{r})$.
-The stochastic front contains so little of the total population that variances and covariance are effectively determined only by the frequencies of classes within the bulk of two dimensional distribution.
-Each iteration began with solving (using NDSolve in Mathematica) for the time-dependent abundances in the bulk governed by \ref{eq:2} over a time interval of one thousand generations.\par
+which is obtained from the probability density function of  $T$ in \cite[Equation 12]{desai2007beneficial}.
 
-Growth of classes in the front can be modeled by branching processes fed by mutants from the adjacent (and exponentially growing) classes in the bulk.
-These mutations are shown by black arrows in Figure \ref{fig:1}.
-We sample the next mutation at the front by sampling random variable $x$ from a standardized exponential and solving $ x=\int_0^{\tau} b_{i,j} n_{i,j}(t)U dt$ for $\tau$.
-The birth rate in our model is set to $b_{i,j}=1+s_{i,j}$ per individual per generation.
+Following establishment, we update the set of classes in the bulk and calculate solutions to their time-dependent growth.
+Initially, classes with less than one individual are regarded as having gone extinct and removed from the bulk.
+Abundances of the remaining class and the newly established one change at a rate of  $\dot{n}_{i,j}(t) =(r_{i,j}-\bar{r})n_{i,j}(t)$ due to selection.
+Initial values are the size of each class at the time of establishment.
+Changes in the set of frequencies $p_{i,j}$ obey the same ordinary differential equation because $N$ is constant.
+Solutions of the coupled system of ODE's are then obtained using an ODE solver (NDSolve in Mathematica) for a period of one thousand generations, and are then used to determine the next establishment event.
+The process then continues for the duration of the simulation.
+Note that in our implementation the population size $N(t)=\sum n_{i,j}(t)$ can deviate slightly from the value $N$ as establishments take place and classes in the bulk become extinct.
+To account for this and maintain a total population size as close to $N$ as possible, we instead use the ODE
+\[ \dot{n}_{i,j} = (r_{i,j}-\bar{r})n_{i,j} - \bar{r}\left (\frac{N(t)}{N}-1 \right)n_{i,j} \] 
+given by \cite[p.133]{Crow1970} in our simulations.\par
 
-We then sampled whether the mutation would establish, by comparing a uniform random variable to the probability of fixation from Equation 16 of Desai and Fisher:
-
-\begin{equation} \label{eq:3}
-\pi_{fix} = \frac{s_{i,j}}{ 1+s_{i,j}}.
-\end{equation}
-After establishment of a mutant lineage at abundance $1/s_{i,j}$, it is part of the bulk, and subsequent growth is deterministic and exponential. \par
-
-We update our system based on the first mutation to appear that is destined for establishment.
-The timing of establishment events is also stochastic, with later deterministic growth described by $n_{i,j}(t) = \frac{1}{s_{i,j}} \exp[{s_{i,j}(t-\tau)}]$ with the random variable $\tau$ measuring the time of establishment.
-We consider the related random variable $\nu = \frac{1}{s_{i,j}}\exp[-s_{i,j}\tau]$ representing the initial size at the time of first mutation that would produce the deterministic trajectory observed after establishment.
-Using the expression given by Desai and Fisher for the PDF of $\tau$ \citep[Equation 12]{desai2007beneficial}, we can arrive at the cumulative distribution of $\nu$:
-% * <masel@email.arizona.edu> 2017-10-23T22:46:00.814Z:
-% 
-% > Using the expression given by Desai and Fisher for the PDF of $\tau$
-% Give an Equation number
-% 
-% ^ <kgomez81@math.arizona.edu> 2017-12-05T19:58:30.000Z:
-% 
-% Added the citation for the PDF of \tau needed to get the CDF of \nu, and removed the citation to Uecker and Hermisson previously included.
-% 
-% ^.
-
-\begin{equation} \label{eq:4}
-P(\nu \leq \nu_0) = 1- e^{-\pi_{fix} \nu_0}
-\end{equation}and used it to set the initial abundance of  the establishing class equal to $\nu$.
-We update the other abundances of the bulk at this time.
-Any classes with less than one individual were removed.
-The next iteration began with solving for the timecourse of the new bulk, now including the newly established class, followed by the subsequent steps outlined above.\par
-
-\begin{figure*}[!ht]
-\centering
-\includegraphics[width=1\linewidth]{figures/fig2.pdf} 
-\caption{Measured averages of variance (open diamonds) and covariance (closed circles) for trait one obtained from simulations, along with the theoretically predicted drop in its rate of adaptation (solid line).
-In each graph, two population parameters are kept constant while the third is varied to determine its effects on the mean variance and covariance of trait one induced the by evolution with the second trait.
-In every case, the variance of trait one increases ($\sigma_1^2 >1$), while trait interactions result in negative covariance whose magnitude is larger than the change in variance ($|\sigma_{1,2}| > \sigma_1^2-1$), indicating that the reduction in the rate of adaptation of trait one is the result of large negative covariances.
-Mean variance and covariance are insensitive to changes in the population size $N$ (pane c), but exhibit some dependence on the values of $U$ and $s$ (panes a,b).
-The values  $s=0.02$, $U=10^{-5}$ and $N=10^9$ where used for parameters that were kept constant.}\label{fig:2}
-\end{figure*}
-
-Each class in the stochastic front may be a mixture of lineages, each arising from a different mutation event.
-However, we assume that the arrival time of the first mutant lineage destined to establish is the one determines which class will join the bulk next.
-With some probability, the  \par
-% * <masel@email.arizona.edu> 2017-10-23T22:52:29.988Z:
-% 
-% > Each class in the stochastic front may be a mixture of lineages, each arising from a different mutation event.
-However, it is the arrival time of the first mutant lineage destined to establish that determines which class will join the bulk next.
-\par
-% Requires discussion. Errors introduced when the first establishing is earlier than the establishing time of the first arrival destined to establish.
-% 
-% ^.
-
-Simulations were initialized with monoclonal populations at carrying capacity consisting of individuals having no beneficial mutations in either trait and relative fitness set to zero.
+Simulations were initialized with monoclonal populations at carrying capacity consisting of individuals having one beneficial mutation in either trait and relative fitness set to zero.
 We varied either $N$, $s$ and $U$ while fixing the others.
@@
 cerevisiae \citep{desai2007speed,Levy2015,Perfeito2007}.
+We allowed the simulation to run for 5,000 generations, where one generation  This allowed the population to achieve beneficial mutation-selection balance before collecting data, and ensured that the transient effects from the initializing would not distort the statistics collected for the distribution.
+Following the burn-in period of the simulation, we recorded the existing classes in the bulk and their respective abundances.
 \par
-% * <masel@email.arizona.edu> 2017-10-23T22:58:03.572Z:
-% 
-% > We varied either $N$, $s$ and $U$ while fixing the others. Each was varied across a range constructed from values reported by evolutionary experiments involving either E. coli or S. cerevisiae
-% More detail on what exactly we learned from each paper, and also how we kept within concurrent-mutations regime, including saying what the concurrent mutations regime, or do the latter around Equation 1, and refer to it here.
-% 
-% ^ <kgomez81@math.arizona.edu> 2017-12-06T20:49:15.515Z:
-% 
-% I added a small discussion of the concurrent mutations regime near equation 1.
-%
-% ^.
 
+The data was used to compute the marginal mean fitness and variances of each trait, as well as their covariance.
+From these quantities we also computed the instantaneous rate of adaptation (Equation \eqref{eq:5}) for each trait and overall.
 The evolution of the population can be broken down into the evolution of the two traits, according to Lande's equation:
 \begin{equation}\label{eq:5}
-\left[
-\begin{array}{c}
+\left( \begin{array}{c}
 \dot{\bar{r_1}} \\
 \dot{\bar{r_2}} 
-\end{array}
-\right]
+\end{array} \right)
 =
-\left[
-\begin{array}{cc}
+\left( \begin{array}{cc}
 \sigma_1^2 & \sigma_{1,2} \\
 \sigma_{1,2} & \sigma_2^2 
-\end{array}
-\right]
-\left[
-\begin{array}{c}
+\end{array}\right)
+\left(\begin{array}{c}
 1 \\
 1 
-\end{array}
-\right]
-\end{equation}where  $\bar{r_1}=\sum_{i,j} p_{i,j} is_1$, $\bar{r_2}=\sum_{i,j} p_{i,j} j s_2$, $\sigma_1^2 = \sum_{i,j} p_{i,j} (r_i-\bar{r_1})^2$, $\sigma_2^2 = \sum_{i,j} p_{i,j} (r_j-\bar{r_2})^2$, and  $\sigma_{1,2} =\sum_{i,j} p_{i,j} (r_i-\bar{r_1})(r_j-\bar{r_2})$.
+\end{array}\right),
+\end{equation}
+where  $\bar{r_1}=\sum_{i,j} p_{i,j} is_1$, $\bar{r_2}=\sum_{i,j} p_{i,j} j s_2$, $\sigma_1^2 = \sum_{i,j} p_{i,j} (r_i-\bar{r_1})^2$, $\sigma_2^2 = \sum_{i,j} p_{i,j} (r_j-\bar{r_2})^2$, and  $\sigma_{1,2} =\sum_{i,j} p_{i,j} (r_i-\bar{r_1})(r_j-\bar{r_2})$.
+Here, it is the frequencies corresponding to the bulk that are relevant, since classes in the front are assumed to be too small to affect the means, variances and covariances of the distribution  ($1/s \ll N$).
 Lande's equation above is expressed as the change in the population's mean trait values over time intervals of size $s$.
 \par
-% * <masel@email.arizona.edu> 2017-10-21T16:44:31.244Z:
-% 
-% > have been normalized
-% How have they been normalized?
-% 
-% ^ <kgomez81@math.arizona.edu> 2017-12-05T20:31:15.198Z:
-% 
-% Sorry, its time that's scaled (not normalized) to Lande's expression above.
-% 
-% ^.
-
-We allowed the simulation to run for 5,000 generations to allow the population to achieve beneficial mutation-selection balance before collecting data.
-This ensured that the transient effects from the initializing would not distort the statistics collected for the distribution.
-Following the burn-in period of the simulation, we recorded the existing classes in the bulk and their respective abundances.
-The data was used to compute the marginal mean fitness and variances of each trait, as well as their covariance.
-From these quantities we also computed the instantaneous rate of adaptation (Equation \ref{eq:5}) for each trait and overall.\par
-% * <masel@email.arizona.edu> 2017-10-23T23:12:23.801Z:
-% 
-% >  5,000 generation
-% Need to define/explain what a generation is in the model.
-% 
-% ^.
 
@@
 
-The reduction in trait-specific adaptation $v_1$ from $v(U,N,s)$ down to $0.5 \hspace{.02 in} v(2U,N,s)$ (Figure \ref{fig:2}, solid lines), caused by clonal interference with the second trait, is driven by high levels of negative covariance (Figure \ref{fig:2}, closed circles).
+The reduction in trait-specific adaptation $v_1$ from $v(U,N,s)$ down to $0.5 \hspace{.02 in} v(2U,N,s)$ (Figure \ref{fig:2}, extent to which solid lines are below 1), caused by clonal interference with the second trait, is driven by high levels of negative covariance (Figure \ref{fig:2}, closed circles).
 This is partly compensated for by a substantial increase in additive genetic variance (Figure \ref{fig:2}, open diamonds), because the negative covariance slows the removal of additive variance from the population.
@@
 
-\begin{figure}[!ht]
-\centering
-\includegraphics[width=1\linewidth]{figures/fig3.pdf}
-\caption{The rate of adaptation of a focal trait as a function of the number of traits undergoing adaptation.
-.
-decays to zero in the Equation \ref{eq:6}, but in practice the expression is only valid when $kU/s \ll 1$.
-The effects of clonal interference on trait evolution are most significant in the case of two traits.
-The values of the parameters are taken as $s=0.02$, $U=10^{-5}$ and $N=10^9$.}
-% * <masel@email.arizona.edu> 2017-10-28T22:59:26.669Z:
-% 
-% > Each plot includes the measured values produced by simulations with either varying $s$, $U$, or $N$. 
-% Aren't var and covar from simulations and v from the equation?
-% 
-% ^.
-\label{fig:3}
-\end{figure}
-
-The reduction in $v_1$ (Figure \ref{fig:2}, solid lines) is insensitive to changes in the values of  $U$, $N$, or $s$, due to the effects of clonal interference.
-When there are $k$ traits with equal $U$ and $s$, the expected $v_1$ is:
+The reduction in $v_1$ due to the effects of clonal interference is insensitive to changes in the values of $s$ (Figure \ref{fig:2}a, solid lines),  $U$ (Figure \ref{fig:2}b), or $N$ (Figure \ref{fig:2}c).
+Generalizing to $k$ traits, each with equal $U$ and $s$, from Equation \eqref{eq:1}  we derive
 \begin{equation}\label{eq:6}
@@
 in units of multiples of the rate of adaptation in isolation, $v(U,N,s)$.
-This expression is obtained from Equation \ref{eq:1} and plotted in Figure \ref{fig:3}.
-Each additional trait increases the total beneficial mutation rate by $U$, which intensifies the effects of clonal interference from trait interactions.
-The total rate of adaptation that is less than the sum of the individual rates of each trait in isolation.
-The shortfall is measured by the departure of $\tilde{v}_1$ from unity.
-\par
+Each additional trait increases clonal interference on the focal trait by a diminishing amount, with curvature apparent even with respect to the logarithm of the number of traits (Figure \ref{fig:3}).
+This suggests that much can be learned even from the simplest case of clonal interference between only two fitness-associated traits.\par
 
-% * <masel@email.arizona.edu> 2017-10-28T23:18:33.323Z:
-% 
-% > decreases in $v_1$ from additional traits are larger with increasing $s$
-% I don't see this clearly in the figure - are you sure this is worth commenting on?
-% 
-% ^.
-% * <masel@email.arizona.edu> 2017-10-28T23:09:27.018Z:
-% 
-% > Changes in $s$ are the most influential at smaller effect sizes.
-% Needs more explanation. Why do you say this? Are you just observing from Figure 2a? The more natural axis for Figure 2a, given Equation 6, would be a log-scale.
-% 
-% Isn't the more important point that s is more important than U or N, at least over the parameter range considered?
-% 
-% ^.
-Under beneficial mutation-selection balance, the components of the \G matrix fluctuate significantly over time, far more than the instantaneous rate of adaptation, with transient spikes in additive genetic variance $\sigma_1^2$ mostly canceled out by transient spikes in negative covariance $\sigma_{1,2}$ (representative simulation shown in Figure \ref{fig:3}).
+Under beneficial mutation-selection balance, the components of the \G matrix fluctuate far more than the instantaneous rate of adaptation, with transient spikes in additive genetic variance $\sigma_1^2$ mostly canceled out by transient spikes in negative covariance $\sigma_{1,2}$ (representative simulation shown in Figure \ref{fig:4}).
 As a result, fluctuations in variance and covariance can be substantially larger than those of total variance in fitness.\par
-
-\begin{figure}[!ht]
-\includegraphics[width=1\linewidth]{figures/fig4.pdf}
-% * <masel@email.arizona.edu> 2017-10-28T23:28:47.376Z:
-% 
-% > fig3.pdf
-% Need tick labels on the x-axis.
-% 
-% ^.
-\caption{Trajectories of the variance of trait one (dashed line), covariance (dotted line), and the total variance in fitness of the population (solid line), plotted over 10,000 generations.
-The mean values of each are included as lines with matching line-styles.
-The \G matrix over this period of time is unstable.
-Its two components, trait one variance and covariance, fluctuate significantly relative to the variance in total fitness over identical time periods.
-The trajectories were sampled from one simulation with parameters $s=0.02$, $U=10^{-5}$ and $N=10^9$.}\label{fig:4}
-% * <masel@email.arizona.edu> 2017-10-28T23:32:29.075Z:
-% 
-% > along with variance in total fitness
-% I thought you were plotting sigma_1^2+sigma_12 which is the instantaneous adaptation rate of trait 1, but here you say that you are plotting something else. Please confirm.
-% 
-% ^.
-\end{figure}
 
@@
 Figure \ref{fig:6} plots the correlation between trait covariances of the bulk and nose, as a function of a time offset.
-The signal peeks at approximately $\bar{\tau}_{fix}$ (mean time to fixation).
+The signal peaks at approximately $\bar{\tau}_{fix}$ (mean time to fixation).
 It shows that covariances between traits are largely set by the behavior at the nose of the one dimensional traveling wave, which consists of genotypes along the lead fitness isocline through the stochastic front.
+The values of variances and covariance will be correlated from one transition to the next, because mutational influx depends on the abundances of adjacent less-fit genotypes.
+
+% (Dynamics of structures withing the \G matrix)
+Principle components can be used to examine the structural features of our \G matrix.
+Although Fluctuations in the individual components $\sigma_1^2$, $\sigma_1^2$, $\sigma_{1,2}$ can be large relative to the rate of adaptation, but there sum has lower overall variance (Figure \ref{fig:5}, black line).
+The
+\[
+E\left[
+\left( \begin{array}{cc}
+\sigma_1^2 & \sigma_{1,2} \\
+\sigma_{1,2} & \sigma_2^2 
+\end{array}\right)
+\right] = 
+\]
 
 \begin{figure}[!ht]
-\includegraphics[width=0.49\linewidth]{figures/fig5a.pdf}
-\includegraphics[width=0.49\linewidth]{figures/fig5b.pdf}
-\caption{\footnotesize The dynamics of the \G matrix components are governed by time scale $\tau_q$ of the one-dimensional traveling wave of fitness classes.
+\includegraphics[width=1\linewidth]{fig1a.pdf}
+\caption{Representative two-dimensional genotype distribution.
+Individuals with equal numbers of beneficial mutations in each trait are combined into classes (boxes).
+Abundances of the bulk (solid boxes) behave deterministically, while abundance of the stochastic front (patterned boxes) behave stochastically.
+The red dot marks the average number of beneficial mutations in each trait, and equally fit classes are found along the red fitness isocline).
+In asexuals, beneficial mutations must occur on the fittest genetic backgrounds in order to contribute to the adaptive process.
+These mutations are represented by arrows from the bulk into the stochastic front.
+Classes in the stochastic front become part of the bulk once their abundances have become sufficiently large, pushing the stochastic front to include new classes.
+As classes below the population mean fitness isocline decline and those above increase exponentially, a two-dimensional traveling wave is produced.
+(Simulation parameters: $N=10^9$, $s=0.02$, and $U=10^{-5}$)}\label{fig:1}
+\end{figure}
+
+\begin{figure*}[!ht]
+\centering
+\includegraphics[width=1\linewidth]{fig2.pdf} 
+\caption{Variance (open diamonds) of one trait and its covariance (closed circles) with the other, averaged across simulations, whose expected sum is equal to the analytically calculated adaptation rate (solid line).
+Variance of the focal trait is always higher than it would be in the absence of the second trait ($\sigma_1^2 >1$); negative covariance more than cancels this out ($|\sigma_{1,2}| > \sigma_1^2-1$) to reduce the adaptation rate.
+Mean variance and covariance are insensitive to changes in the population size $N$ (c), but exhibit some dependence on the values of $s$ (a) and $U$ (b).
+For the parameter values not being varied on the x-axis, $s=0.02$, $U=10^{-5}$ and $N=10^9$.
+}\label{fig:2}
+% * <masel@email.arizona.edu> 2018-01-22T19:36:50.877Z:
+% 
+% > averaged across simulations
+% How many simulations were averaged?
+% 
+% ^.
+\end{figure*}
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=1\linewidth]{fig3.pdf}
+\caption{Rate of adaptation of a focal trait as a function of the number of total traits undergoing adaptation.
+Each additional trait reduces the rate of adaptation, but by marginally less as the trait number grows.
+The decrease decays to zero in the Equation \eqref{eq:6}, but in practice the expression is only valid when $kU/s \ll 1$.
+The effects of clonal interference on trait evolution are most significant in the case of two traits.
+The values of the parameters are taken as $s=0.02$, $U=10^{-5}$ and $N=10^9$.}
+% * <masel@email.arizona.edu> 2017-10-28T22:59:26.669Z:
+% 
+% > Each plot includes the measured values produced by simulations with either varying $s$, $U$, or $N$. 
+% Aren't var and covar from simulations and v from the equation?
+% 
+% ^.
+\label{fig:3}
+\end{figure}
+
+\begin{figure}[!ht]
+\includegraphics[width=1\linewidth]{fig4.pdf}
+% * <masel@email.arizona.edu> 2017-10-28T23:28:47.376Z:
+% 
+% > fig3.pdf
+% Need tick labels on the x-axis.
+% 
+% ^.
+\caption{Trajectories of the variance of trait one (dashed line), covariance (dotted line), and the total variance in fitness of the population (solid line), plotted over 10,000 generations.
+The mean values of each are included as lines with matching line-styles.
+The \G matrix over this period of time is unstable.
+Its two components, trait one variance and covariance, fluctuate significantly relative to the variance in total fitness over identical time periods.
+The trajectories were sampled from one simulation with parameters $s=0.02$, $U=10^{-5}$ and $N=10^9$.}\label{fig:4}
+% * <masel@email.arizona.edu> 2017-10-28T23:32:29.075Z:
+% 
+% > along with variance in total fitness
+% I thought you were plotting sigma_1^2+sigma_12 which is the instantaneous adaptation rate of trait 1, but here you say that you are plotting something else. Please confirm.
+% 
+% ^.
+\end{figure}
+
+\begin{figure}[!ht]
+\includegraphics[width=0.49\linewidth]{fig5a.pdf}
+\includegraphics[width=0.49\linewidth]{fig5b.pdf}
+\caption{The dynamics of the \G matrix components are governed by time scale $\tau_q$ of the one-dimensional traveling wave of fitness classes.
 Variances and covariance are dominated by the set of genotypes that make up the peak of the one-dimensional traveling wave, whose shape is approximately Gaussian.
@@
 \begin{figure}[!ht]
-\includegraphics[width=1\linewidth]{figures/fig6a.pdf}
+\includegraphics[width=1\linewidth]{fig6a.pdf}
 \caption{Correlation between covariance of stochastic front and bulk.
@@
 
-Because mutational influx depends on the abundances of adjacent less-fit genotypes, the values of variances and covariance will be correlated from one transition to the next.
-However, these correlations vanish after a period of time roughly equal to the time required for classes in the stochastic front become the dominant group.
-This time period can be expressed as the product $q \tau_q$, where $q = 2\log(Ns)/\log(s/U)$ is the lead of the traveling wave nose over its mean.
-Over these periods of time $q \tau_q$, fluctuations in covariance can be significantly larger than those occurring over periods equal to $\tau_q$, resulting in an unstable \G-matrix over the time scale of adaptation.
-\par
-% * <masel@email.arizona.edu> 2017-10-29T00:04:55.206Z:
+\section*{Discussion}
+
+% * <masel@email.arizona.edu> 2018-01-17T17:50:06.655Z:
 % 
-% > these correlations vanish after a period of time roughly equal to the time required for classes in the stochastic front become the dominant group
-% Why is this?
+% > Discussion}
+% Include speculation about the effects of adding deleterious mutations.
+% 
+% ^ <masel@email.arizona.edu> 2018-01-24T22:30:26.120Z:
+% 
+% Compare to all other theoretical sources of G-matrix instability eg. mutation drift migration etc. and argue it is strong relative to others.
 %
 % ^.
+\label{sec:discussion}
+% (Genetic Constraints on the Rate of Adaptation)
+% \cite{phillips1989visualizing} discusses a unifying framework for micro- and macro-evolution based on a unification of Lande's quantitative genetics theory \cite{lande1979quantitative} and Simpson's adaptive landscapes \cite{simpson1944tempo}.
+His discussion centers on the role of the |G matrix, which encodes genetic constraints, in defining the local behavior of evolutionary trajectories on the fitness landscape.
+Aspects of this framework has been tested emprically by many.
+\cite{Schluter1996} analysis of empirical data on threespine stickleback populations showed that phenotypic responses over the past four million were biased towards \G's leading eigenvector ($g_{max}$); similar patterns were found in data on mice, sparrows, and finches.
+Studies like these have prompted subsequent investigations of the impact that genetic constraints might have on the rate of adaptation.
+\cite{teplitsky2014assessing} for example, estimated population-specific \G matrices and multivariate selection coefficients form long-term data for seven bird species (10 populations).
+He found that genetic correlations reduced the rate of adaptation by 28\% on average, and multivariate evolvability tended to be systematically lower in the direction of selection compared to others (see also Teplitsky et al.
+2011).
+\cite{kopp2014rapid} provides a review of this subject in relation to the potential for rapid adaptation to environmental changes given genetic constraints, phenotypic plasticity, and evolvability, based on different models.
+Their discussion of genetic constraints in quantitative genetic models review most of the finding expressed in past papers (see Schluter 1996, Phillips et al 1989, Hansen and Houle 2008).
+It is still unclear whether genetic constraints do in general limit the rate of adaptation, since other related studies have not been so conclusive on the issue \citep{Agrawal2009,hine2014evolutionary}.
+In addition, different metrics have been used measure the impact of genetic constraints on rates of adaptation using \G \citep{HANSEN2008,Kirkpatrick2009,Agrawal2009}, and there subtle differences in what they indicate about constraints (see Kopp et al.
+2014).
+Some of the issues may in fact be related to the use of \G itself \citep{chebib2017affects}, which may not truly measure genetic constraints at all.
+The role of genetic constraints must be addressed.
+\cite{matuszewski2015catch} showed in their work that standing genetic variation has greater potential for facilitating adaptation to environmental change than de novo beneficial mutations.
+In their approach, which combined those of adaptive walks models (discrete substitutions from new mutations) and quantitative genetic models (adaptation from standing genetic variation), and they found that populations that adapt from standing genetic variation are capable of significantly more phenotype change over time, and thus, have a higher potential to adapt when faced with environmental change that is fast, rather than slow.
 
-\section*{Discussion}
-\label{sec:discussion}
+% ((General theory for changes in G))
+Quantitative genetic models for the evolution of \G indicate that selection, drift, mutation and recombination each induce changes in genetic variances and covariances over time.
+Lande's Gaussian framework in particular is often used to specify a decomposition of changes in the \G matrix into three parts represented in the expression:
+\[\Delta G = G(\gamma - \beta \cdot \beta^T)G + 2 M + 2\sum_{i,j=1}^n r_{i,j}(C_{i,j}-C_{i,j}'),\]
+which is given in \cite[Chapter 20, Box 3]{fox2006evolutionary}.
+The first term describes the action of selection, which includes directional selection eroding genetic variation ($\beta$ terms), as well as stabilizing and correlational selection which reorienting the G matrix ($\gamma$ term).
+The second term involves the mutational covariance matrix \textbf{M}, and gives the sum of mutational covariances generated by the effects of pleiotropic mutations at each locus.
+The third term gives the rate of degradation in covariance caused by linkage disequilibrium, which depends on the recombination rate.
+More terms may be needed depending on whether the distribution of allelic effects are normal (Barton and Turelli 1987).
+More detailed explanations of each terms are given in  \cite[Chp34,35,36]{lynch2018evolution}, but the last term is discussed under assumptions that are valid for infinitesimal model.
 
-%An interesting consequence of the dynamics of covariance involves the relationship between stochastic front and its role in setting the covariance of the two dimensional trait distribution.
-In between establishement events for the one dimensional traveling wave, classes that lie along the line of constant fitness passing through the fittest part of the stochastic front their relative size to one another.
-OneThis process is exactly  manner describe by \cite{Desai2013}.
-Thus changes in the value of covariance in fact follow the transitions of the peaks the one dimensional traveling wave shift.
-These major transitions are consequently governed by the rate of adaptation and determined at the stochastic front.
-%
-%\subsection*{Comparison with Barton and Otto's Negative Linkage Disequilibria Results}
-%In our model, covariance between traits is arises entirely from linkage disequilibrium.
+% (Overview of Genetic Drift and its relevance in our model)
+With respect to the importance of mutations, pleiotropy is largely emphasized \cite{reeve2000predicting,Steppan2002,Arnold2008}.
+Changes in genetic correlaitons due to linkage disequilibria enter into the dynamics of \G through the mutation matrix \textbf{M}.
+\cite{Phillips2001}
+\cite{lande1979quantitative}
+
+% (Fluctuations in G due to Mutations)
+\cite{Arnold2008} has a section on stability of the G matrix when mutational effects evolve (Jones papers cited). 
+\cite{Jones2003} The nature of mutations at pleiotropic loci can dramatically influence stability of G, i.e.
+when a mutation at a single locus simultaneously changes the value of the two traits (due to pleiotropy) and these effects are correlated, mutation can generate extreme stability of G.
+\cite{Jones2007} examine a two trait system and find that mutational correlation does evolve in response to selection on the bivariate phenotype (mutation is third quant-trait).\par
+
+% Quantitative genetic models for the dynamics of \G attribute changes in size are thought to be driven by drift \cite{lande1979quantitative,Roff2000,Arnold2008}.
+In our model, changes in the size of the \G matrix are induced by stochastic fluctuations in the front resulting from the effects of drift.
+% (Fluctations in G: Nose Mutations vs Mutation Matrix)
+% \cite{Jones2007} 
+% \cite{Arnold2008}
+% \cite{HANSEN2008}
+
+% (Genetic correlations as constraints to trait evolution)
+% \cite{hine2014evolutionary}
+
+% (Comparison with Barton and Otto's Negative Linkage Disequilibria Results, and other related quantitative genetics)
+% \cite{Barton2005} 
+% \cite{Bulmer1971} Selection induces changes in variance contributions from linkage disequilbrium are shaped by selection, since the infinitesimal model, frequencies 
+% \cite{}
+% In our model, covariance between traits is arises entirely from linkage disequilibrium.
 We can assume that specific alleles at either locus correspond to the number of beneficial mutations that have accumulated in the corresponding trait.
@@
 By applying \textbf{equation 2}, we can derive a mathematical relationship between covariance and linkage disequilibrium (need to cite chapter 5, Walsh and Lynch).
-%% * <masel@email.arizona.edu> 2017-07-29T17:34:47.794Z:
-%% 
-%% > specific alleles at either locus correspond to the number of beneficial mutations that have accumulated in the corresponding trait
-%% Multiple loci can contribute to the same trait.
+% % * <masel@email.arizona.edu> 2017-07-29T17:34:47.794Z:
+% % 
+% % > specific alleles at either locus correspond to the number of beneficial mutations that have accumulated in the corresponding trait
+% % Multiple loci can contribute to the same trait.
 An allele cannot be a "number of beneficial mutations", because an allele must pertain to one locus.
@@
 Because of this confusion, I got a bit lost here and so skipped Equation 6.
-%% 
-%% ^.
-%\begin{equation}
-%\sigma_{1,2}=\sum_{i,j}D_{i,j}[(i-\bar{i})s_1][(j-\bar{j})s_2]. 
-%\end{equation} 
-%As the population adaptsAcknowledgments, mutations and selection  increase and decrease the amounts LD which ultimately determine how traits interact in their evolution.\par
-%
-%We examined the resulting dynamics of trait evolution using numerical simulations of our model using Mathematica.
+% % 
+% % ^.
+% \begin{equation}
+% \sigma_{1,2}=\sum_{i,j}D_{i,j}[(i-\bar{i})s_1][(j-\bar{j})s_2]. 
+% \end{equation} 
+% As the population adaptsAcknowledgments, mutations and selection  increase and decrease the amounts LD which ultimately determine how traits interact in their evolution.\par
+
+% We examined the resulting dynamics of trait evolution using numerical simulations of our model using Mathematica.
 We focused on modeling the symmetric case, in which the two traits had identical mutation rate and selection coefficients.
@@
 
+% The behavior of the stochastic front is an evolving boundary in two-dimensional trait space.
+\cite{pearce2017rapid} derive features of the steady state two-dimensional front in the asymptotic limit $\log(N s) \rightarrow \infty$, but these do not characterize its stochasticity.
+As we will later show, the dynamics variances and covariance depend significantly on these fluctuations.
+Thus, we proceed by simulation.
+[Polish]
+
 \section*{Acknowledgments}
 \label{sec:acknowledgments}
-We thank Patrick Phillips for helpful discussions. Funding was provided by the National Science Foundation (DEB-1348262) and the National Institutes of Health (T32 GM084905).
+We thank Patrick Phillips and Bruce Walsh for helpful discussions. Funding was provided by the National Science Foundation (DEB-1348262) and the National Institutes of Health (T32 GM084905).
 \bibliography{biblio}
@@
 
+\section{Relevant That Needs Discussion}
+in asexuals; this fact has, for example, been the basis of explanations for the evolutionary advantages of sex and recombination \citep{Barton2005,Otto2009}
 \end{document}Acknowledgments

### 2018-02-12 plots.py

@@ -235,11 +235,11 @@ fit_var = fit_var[start_indx:end_indx]
 fit_cov = fit_cov[start_indx:end_indx]
 
 # compute means of the time series data and store as constant arry
-rate_adpt = np.asarray(fit_var[:,0])+np.asarray(fit_var[:,1])+2*np.asarray(fit_cov[:])
+rate_adpt_t1 = np.asarray(fit_var[:,0])+np.asarray(fit_cov[:])
 var1_avg = np.mean(fit_var[:,0])*np.ones(np.shape(times))
 var2_avg = np.mean(fit_var[:,1])*np.ones(np.shape(times))
 cov_avg = np.mean(fit_cov[:])*np.ones(np.shape(times))
-rate_adpt_avg = var1_avg + var2_avg + 2*cov_avg
+rate_adpt_t1_avg = var1_avg + cov_avg
 
 my_xticks = [10000+i*1000 for i in range(11)]
 my_xlabel = ['10', '', '12', '', '14', '', '16', '', '18', '', '20']
@@ -248,10 +248,10 @@ my_ylabel = ['', '-4', '', '-2', '', '0', '', '2', '', '4', '', '6', '']
 
 # plot data for figure 3a and 3b 
 fig, ax = plt.subplots(1,1,figsize=[8,8])
-ax.plot(times,rate_adpt_avg,c="black",label=r'$\sigma^2$',linewidth=2.0,linestyle = '-')                
+ax.plot(times,rate_adpt_t1_avg,c="black",label=r'$\sigma^2$',linewidth=2.0,linestyle = '-')                
 ax.plot(times,var1_avg,c="black",label=r'$\sigma_1^2$',linewidth=3.0,linestyle = '--')
 ax.plot(times,cov_avg,c="black",label=r'$\sigma_{1,2}$',linewidth=3.0,linestyle = ':')
-ax.plot(times,rate_adpt,c="black",linestyle="-",linewidth=3.0)        
+ax.plot(times,rate_adpt_t1,c="black",linestyle="-",linewidth=3.0)        
 ax.plot(times,fit_var[:,0],c="black",linestyle="--",linewidth=3.0)
 ax.plot(times,fit_cov[:],c="black",linestyle=":",linewidth=3.0)
 ax.axhline(linewidth=0.5, color = 'k')

### 2018-02-19 plots.py
corrected some small problems with figure 4 and figure 6. I also added some new references for Lande's earlier work related to the use of his QG theory for macroevolution.

@@ -248,7 +248,7 @@ my_ylabel = ['', '-4', '', '-2', '', '0', '', '2', '', '4', '', '6', '']
 
 # plot data for figure 3a and 3b 
 fig, ax = plt.subplots(1,1,figsize=[8,8])
-ax.plot(times,rate_adpt_t1_avg,c="black",label=r'$\sigma^2$',linewidth=2.0,linestyle = '-')                
+ax.plot(times,rate_adpt_t1_avg,c="black",label=r'$v_1$',linewidth=2.0,linestyle = '-')                
 ax.plot(times,var1_avg,c="black",label=r'$\sigma_1^2$',linewidth=3.0,linestyle = '--')
 ax.plot(times,cov_avg,c="black",label=r'$\sigma_{1,2}$',linewidth=3.0,linestyle = ':')
 ax.plot(times,rate_adpt_t1,c="black",linestyle="-",linewidth=3.0)        
@@ -301,8 +301,8 @@ my_ylabel = ['0', '', '0.2', '', '0.4', '', '0.6', '', '0.8', '', '1']
 fig,ax = plt.subplots(figsize=[8,8])
 ax.plot((1/tau_fix_avg)*np.asarray(t_off),np.asarray(t_cov),c="black",linestyle="-",linewidth=3.0)
 ax.plot(xline,yline,c="black",linestyle="--",linewidth=3.0)
-ax.set_ylabel(r'Font-Bulk $\sigma_{1,2}$ Correlation',fontsize=20)
-ax.set_xlabel(r'Time offset (Multiples of $\bar{\tau}_{fix}$)',fontsize=20)
+ax.set_ylabel(r'Front-Bulk $\sigma_{1,2}$ Correlation',fontsize=20)
+ax.set_xlabel(r'Time offset (Multiples of $\bar{\tau}_{SW}$)',fontsize=20)
 ax.axhline(linewidth=0.5, color = 'k')        
 ax.set_ylim((0,1))
 ax.set_xlim((0,1.3e3/tau_fix_avg))        

### 2018-02-21 manuscript.tex
updates to manuscript introduction and included additional comments in of the references on Lande's work.

@@
 \newcommand{\G}{\textbf{G }}
+\newcommand{\M}{\textbf{M }}
 
@@
 \affil[$\ast$]{Graduate Interdisciplinary Program in Applied Mathematics, University of Arizona, and}
-% * <masel@email.arizona.edu> 2017-10-24T23:03:12.117Z:
-%
-% ^.
 \affil[$\dagger$]{Department of Evolution and Ecology, University of Arizona}
@@
 Lande's equation governs the short-term evolution of correlated quantitative traits under natural selection.
-Genetic variances and covariances can also evolve over appreciable time scales, and as a result, Lande's expression alone cannot tell us how correlated quantitative traits will change in the long-run.
 
-When the \G matrix is stable, it is possible to infer past selection gradients and extrapolate future trait evolution (both direction and rate) from historical patterns.
-For example, studies of past macro-evolution in populations and species have used net selection gradients to draw connections between morphological evolution and reproductive isolation \citep{price1984genetic}, infer morphological and phylogenetic relations \citep{schluter1984morphological}, and determine the extent of local adaptation between populations \citep{dudley1996}.
+Genetic variances and covariances also evolve over time, and as result, the long term evolution of correlated traits will depend on both selection gradients and the dynamics of the \G matrix.
+As the mean phenotype of a populations changes, mean fitness increases.
+The process can be visualized as movement along a fitness surface (Simpson's Adaptive Landscape) defined over the space of phenotypes \cite{phillips1989visualizing}.
+Lande's expression characterizes how the \G matrix influences the local behavior of trajectories on this surface.
+The trajectories followed by populations depict their macroevolution, and as result, interest in the \G matrix has grown among evolutionary biologist \cite{Arnold2001,Arnold2008}.
+
+When the \G matrix is stable, it is possible to infer past selection gradients (need citations) and forecast future trait evolution (both direction and rate) \citep{teplitsky2014assessing,teplitsky2011quantitative,Schluter1996}.
+However, it has not come into widespread use for this purpose, perhaps in part because its stability cannot be assumed a priori.
+\cite[p.407]{lande1979quantitative} suggested that the \G matrix might be near an equilibrium determined by a balance of mutation, selection and recombination \cite{lande1977influence, lande1980genetic}, and subsequent work has examined this claim in more detail.
+On one front, theoretical models based on quantitative genetics have developed to try and determine conditions under which the \G matrix is stable.
+On the other, empirical work focused on
+
+A large body of work measuring and comparing it between populations.
+There are several approaches to comparing the \G matrices of different populations \citep{Steppan2002}.
+Some compare the magnitudes of the individual variance and covariance elements \citep{atchley1992evolutionary,lofsvold1986quantitative}.
+However, this might not capture important characteristics of \G matrices that are shared between populations.
+Methods based on the use of principal components, such as common principle component analysis (CPCA), in contrast, can compare features between \G matrices such as shape, size, and orientation.
+These latter methods have gained popularity, and have the potential to reveal which evolutionary forces significantly influenced the past dynamics of the population's \G matrix \cite{Phillips1999,Arnold2008}.
+% * <masel@email.arizona.edu> 2018-02-14T18:34:10.127Z:
+% 
+% > Some compare the magnitudes of the individual variance and covariance elements \citep{atchley1992evolutionary,lofsvold1986quantitative}.
+% Add more citations beyond the statistically flawed element by element.
+% 
+% 
+% ^.
+
+% (Lande Notes): Lande (1979) indicated that if the \G matrix were stable, it would be possible to measure selective forces leading to differentiation between two contemporary species, assuming other effects polyploidy, macro-mutation, and the direct effects of the environment can be ignored.
+Consequently, phyletic changes caused by responses to natural selection; assuming that the appearance of new forms is due to the accumulation of genetic variation, and ignoring polyploidy, macromutation, and the direct effects of the environment.
+Lande later points out that if genetic variances and covariances remain constant in a lineage, its net selection gradient can be obtained using \citep[Equation 9]{lande1979quantitative}.
+
+%Work by \cite{price1984genetic} examined this in two small populations of Darwin's ground finches (Geospiza), where he used measurements of genetic and ecological parameters obtained during a 6-year study of their morphological evolution.
+This included data on variances and covariances for weight, beak length, beak width, beak depth, population size, and estimates of the selection surface for local shape of the adaptive landscape(directional beta and quadratic term gamma using in Lande and Arnold's 1983 approach with regression).
+Observations of sexual selection indicated a connection between morphological evolution and reproductive isolation, hence speciation.
+Studies by Ratcliffe and Grant (1983) have shown that body size and beak characteristics are used by sexes to distinguish between con- and heterospecifics, and these may have been brought about by differences in sex ratios (observed in 1977 drought).
+The coupling of morphological evolution and reproductive isolation is unique in these species.
+The use of Lande's ideas involving selection leading to species transition are discussed page 58 \citep{lande1979quantitative}.
+
+%Lande's 1977 paper demonstrated that mating systems had no effect on the genetic variation of additive genes in large populations, and ``found'' that genetic variance is determined entirely by mutation and selection parameters, as well as environmental variance (Equation 9).
+The implication from this work is that in natural populations mutation and selection can maintain constant genetic variances (also depends on environmental variance).
+Lande's 1980 paper deals with finding expressions for the dynamics of the \G matrix under selection, pleiotropic mutations, and recombination.
+With selection weak relative to recombination (eq 18), the dynamics of the \G matrix depend entirely on mutations, selection and environmental variances (eq 21c, 22a); this assumption of weak selection is made clear below Equation 17.
+\par
+
+% * <masel@email.arizona.edu> 2018-02-14T18:17:58.918Z:
+% 
+% > When the \G matrix is stable, it is possible to infer past selection gradients
+% Need citation
+% 
+% ^.
+
+%For example, studies of past macro-evolution in populations and species have used net selection gradients to draw connections between morphological evolution and reproductive isolation \citep{price1984genetic}, infer morphological and phylogenetic relations \citep{schluter1984morphological}, and determine the extent of local adaptation between populations \citep{dudley1996}.
 Similar work has looked the impact of genetic constraints on rates of adaptation \citep{teplitsky2014assessing,teplitsky2011quantitative}, their ability to bias phenotypic evolution over millions of years \cite{Schluter1996}.
 
-The importance of \G's stability in evolutionary inferences such as these has motivated a large body of empirical work focused on measuring and comparing it between populations.
-Comparative studies over the decades have used different metrics and statistical procedures to check whether populations have similar \G matrices.
-Each of these metrics vary in their ability to detect conserved structure in genetic variances and covariances \cite{Steppan2002}.
-Matrix correlations and maximum liklihood tests, for example, often test for correlations between \G matrices or and their components, but do not give additional information about other important characteristics that can be shared between \G matrices.
-Methods based on the use of principle components, such as common principle component analysis (CPCA), in contrast, can compare features between \G matrices such as shape, size, and orientation.
-These methods have gained popularity for this reason, and because they are capable of revealing which evolutionary forces may have significantly influenced the past dynamics of the population's \G matrix \cite{Phillips1999,Arnold2008}.
+% (Needs Work)
+At least some features of the \G matrix can be relatively stable.
+Over 70\% of the \G matrices comparisons between conspecific populations or between related species find similar shape and orientation \cite{Arnold2008}, suggesting that aspects of genetic constraints can be maintained over significant time scales.
+However, comparative studies of \G also indicate that selection and drift can produce instabilities even over very short time scales \cite{bjorklund2013rapid,Siren2017,Phillips2001,wilkinson1990resistance,roff1999does,Steppan1997,Waldmann2000}.
+% \cite{roff1999does} \cite{Begin2003,Begin2004} 
+5
+%(Needs Work)
+Quantitative genetic models confirm that the \G matrix can be changed by these and other factors, and they have been used to identify other conditions that contribute stability of the \G matrix \citep{Turelli1988,barton1989evolutionary,reeve2000predicting}.
 
-A significant number of comparative studies examining either natural populations or laboratory experiments show that at least some features of the \G matrix are relatively stable.
-In over 70\% of comparisons between conspecific groups and species find that \G matrices have similar shape and orientation \cite{Arnold2008}.
-These include populations that have diverged over millions of years \cite{Arnold1999}, and across distinct environments \cite{puentes2016similarity,teplitsky2014assessing}, suggesting that aspects of genetic constraints can be maintained over significant time scales and across a diverse set of environmental factors.
-There are also comparative studies that find no related structure between \G matrices \cite{wilkinson1990resistance,roff1999does,Steppan1997,Waldmann2000}, and others that reveal large instabilities in \G over very short time scales  \cite{bjorklund2013rapid,Siren2017}.
-% \cite{roff1999does} \cite{Begin2003,Begin2004}
+Environmental change affects genotype $\times$ environment interactions and hence changes the \G matrix stability, so to assume a constant \G matrix, one must assume the absence of GxE interactions.
+Simulations have revealed that large population sizes and strong correlational selection can induce stability in \G's dynamics \cite{Jones2003}.
+When mutational covariances can respond to selection, the \M matrix and hence the \G matrix is more stable \cite{Jones2007}.
 
-Quantitative genetic models of genetic variances and covariances provide some indication of how ecological and evolutionary forces contribute to stability in \G's dynamics \citep{Turelli1988,barton1989evolutionary,reeve2000predicting}.
-For example, changes in genotype $\times$ environment interactions directly affect genetic variances and covariances and lead to significant changes in \G over few generations.
-Other critical stability conditions involve the strength of selection and mutational covariances.
-Further analytical insights from these models require knowledge of unknown genetic details, such as the distribution of allelic effects \cite{burger2000mathematical}.
-Simulations have been used to bypass this problem and examine the dynamics of \G in more detail \cite{Jones2003,Jones2007}.
-In both theoretical analysis and numerical simulations, selection and mutations play an important role in ensuring the stability of the \G matrix, but the effects of the latter are largely assumed by most quantitative geneticists to be the result of pleiotropic effects \cite{lande1979quantitative,lande1980genetic,Arnold2008}(I have more citations, and will insert them).
+Historically, pleiotropic effects have been emphasized over linkage disequilibrium in their contribution to \G's dynamics, since recombination is assumed to remove the latter \citep{lande1979quantitative,lande1980genetic,Arnold2008} (more).
+As a result, we know relatively little about the possible effects that linkage disequilibrium might have on the long-term evolution of genetic variances and covariances.
+% * <masel@email.arizona.edu> 2018-02-14T18:56:02.901Z:
+% 
+% > When mutational covariances can respond to selection, the \M matrix and hence the \G matrix is more stable \cite{Jones2007}.
+% Look up this paper. This might not be the mechanism. If the paper is important, spell out the mechanism. Otherwise, leave it out.
+% 
+% 
+% ^.
+% * <masel@email.arizona.edu> 2018-02-14T18:53:18.109Z:
+% 
+% > large population sizes and strong correlational selection can induce stability in \G's dynamics \cite{Jones2003}
+% Rephrase to be about how drift changes G, not how large N promotes G stability.
+% 
+% ^.
+% * <masel@email.arizona.edu> 2018-02-14T18:42:01.952Z:
+% 
+% > However, comparative studies of \G also indicate that selection and drift can produce instabilities even over very short time scales \cite{bjorklund2013rapid,Siren2017,Phillips2001,wilkinson1990resistance,roff1999does,Steppan1997,Waldmann2000}.
+% > % \cite{roff1999does} \cite{Begin2003,Begin2004} 
+% > Quantitative genetic models confirm that the \G matrix can be changed by these and other factors, and they have been used to identify other conditions that contribute stability of the \G matrix \citep{Turelli1988,barton1989evolutionary,reeve2000predicting}.
+% Not specific enough: unpack. Can also be done in the Discussion instead. Following text does make more specific points: add citations of who made them to these.
+% 
+% ^.
+
+% * <masel@email.arizona.edu> 2018-02-12T00:03:04.529Z:
+% 
+% > such as the distribution of allelic effects \cite{burger2000mathematical}. Simulations have been used to bypass this problem
+% It seems unlikely that simulations would bypass the problem of not knowing the distribution of allelic effects.
+% 
+% ^ <kgomez81@math.arizona.edu> 2018-02-14T17:55:06.818Z:
+% 
+% I removed this sentence.
+I know that quantitative genetic models depend on frequencies of alleles and the distribution of their effects, and this has limited the use of theory to gain further insights on G's stability.
+Simulations, such as those of Jones (mentioned later), have been used instead, but I believe that they tackle mostly the issue of unknown frequencies.
+I am not sure that they address the distribution of allelic effects in any way that solves the issues identified in the quantitative genetic theory.
+I would have to fish out the details from Bruce's book.
+%
+% ^.
+% * <masel@email.arizona.edu> 2018-02-12T00:01:20.366Z:
+% 
+% > Other critical stability conditions involve the strength of selection and mutational covariances.
+% This sentence doesn't tell me what the stability condition is. You don't need to go into details, but you do need to give a heuristic.
+% 
+% ^ <kgomez81@math.arizona.edu> 2018-02-14T17:57:51.147Z:
+% 
+% I added a heuristic explanation, but I will work on improving it.  I think I can do better.
+%
+% ^.
+% * <masel@email.arizona.edu> 2018-02-12T00:00:05.266Z:
+% 
+% > changes in genotype $\times$ environment interactions directly affect genetic variances and covariances
+% I can't easily follow the logic of this just by reading; more explanation is in order.
+% 
+% 
+% ^ <kgomez81@math.arizona.edu> 2018-02-13T23:33:02.198Z:
+% 
+% For example, genotype $\times$ environment interactions must be absent since they can lead to changes in genetic variances and covariances with new environments.
+%
+% ^.
 
@@
 
-To study linkage disequilibrium, we have to consider explicit alleles or genotypes, not just quantitative traits.
+To study how linkage disequilibrium changes genetic variances and covariance over time [this is not what you're studying, need to define a purpose here related to your quesiton(s)], we have to consider explicit alleles or genotypes, not just quantitative traits.
 Beneficial mutations are discrete not infinitesimal, and appear on distinct genetic backgrounds.
@@
 Here we analyze a two-dimensional traveling wave model of an asexual population, and find that clonal interference alone, in the absence of pleiotropy, leads on average to greatly elevated variance of fitness-associated traits and to negative covariance between them.
-However, the \G matrix is highly unstable over the timescale of selective sweeps.
+However, the \G matrix is highly unstable over the timescale of selective sweeps.\par
 
@@
 \label{sec:materials:methods}
-
 We consider an asexual population of haploids with fixed total population size $N$ evolving in continuous time.
@@
 We refer to each such set of individuals with the same $i$ and $j$ as a ``class'' and denote their abundances and frequencies by $n_{i,j}$ and $p_{i,j}$ respectively.
-The population's mean fitness is denoted $\bar{r}$.
+The population's mean fitness is denoted $\bar{r}$.\par
+
+\begin{figure}[!ht]
+\includegraphics[width=1\linewidth]{fig1a.pdf}
+\caption{Representative two-dimensional genotype distribution.
+Individuals with equal numbers of beneficial mutations in each trait are combined into classes (boxes).
+Abundances of the bulk (solid boxes) behave deterministically, while abundance of the stochastic front (patterned boxes) behave stochastically.
+The red dot marks the average number of beneficial mutations in each trait, and equally fit classes are found along the red fitness isocline).
+In asexuals, beneficial mutations must occur on the fittest genetic backgrounds in order to contribute to the adaptive process.
+These mutations are represented by arrows from the bulk into the stochastic front.
+Classes in the stochastic front become part of the bulk once their abundances have become sufficiently large, pushing the stochastic front to include new classes.
+As classes below the population mean fitness isocline decline and those above increase exponentially, a two-dimensional traveling wave is produced.
+(Simulation parameters: $N=10^9$, $s=0.02$, and $U=10^{-5}$)}\label{fig:1}
+\end{figure}
 
@@
 This transition from stochastic to deterministic behavior is called ``establishment'' and occurs if the lineage reaches an abundance of order $1/s$.
+Beneficial mutations within the bulk are ignored because these are lost to clonal interference.
 \par
 
-The evolution of population follows an iterative process with two steps.
-First, the next class to establish is determined, and its depends on the growth of classes producing adaptive mutants in the front (\ref{fig:1}, arrows).
+The coupled dynamics of the bulk and front determine genetic variances and covariances in the population.
+The genetic structure of the bulk influences which new adaptive mutants will establish in the future, but  the bulk itself is the outcome of previous mutant establishments.
+The steady-state patterns of genetic variation are determined by beneficial mutation-selection balance: while selection in the bulk continually acts to eliminate genetic variation, beneficial mutations create new genetic variation.
+However, the resulting patterns of genetic variation are not simple to analyze mathematically, particularly because we are interested in the way that genetic variation is  partitioned between two traits \cite{pearce2017rapid,Desai2013}.
+Moreover, the fluctuations around this steady state due to the stochasticity of the front have important consequences for patterns of genetic variation, and these are even harder to analyze \cite{fisher2013asexual} [cite hallatschek].
+Thus, we performed numerical simulations.
+
+Our simulations follow an iterative process with two steps.
+First, we determine the next class to establish, which depends on the growth of classes producing adaptive mutants in the front.
 Once a class has established, it begins to alter the growth dynamics of the other classes in the bulk.
 The second step involves producing the solutions for the dynamics of the bulk with the newly established class.
-Having done so, the process then repeats with determining the next class from the front to establish using the solutions obtained.
-
-To simulate the timing of the next establishment, stochastic fluctuations induced by drift and mutations must be accounted for in the growth of classes at the front.
-These growth of classes in the front follow a branching processes fed by mutants from the adjacent (and exponentially growing) classes in the bulk.
-Each of the mutant lineages that arise will either go extinct, or grow sufficiently large to experience deterministic growth from selection.
-The chances of the latter happening is given by the probability of establishment, denoted $\pi$.
-We use the value of $\pi$ given by the expression:
-\begin{equation} \label{eq:2}
-\pi = \frac{s_{i,j}}{ 1+s_{i,j}},
-\end{equation}and derived from \cite[Equation 16]{desai2007beneficial}.
-Here $s_{i,j}=(r_{i,j}-\bar{r})$ is the selective advantage of an individual in the class with $i$ mutation in the first trait, and $j$ mutations in the second, with respect to an average individual in the population.
-Although classes in the front can consist of many mutant lineages that contribute to their size during the stochastic growth phase, their establishment  is driven by the size of the first lineage destined to establish.
-In particular, the time required for this single lineage to transition into the deterministic growth phase is a good approximation for the actual time of establishment of its class.
-The justification for this approximation is discussed in detail by \cite[p.1773]{desai2007beneficial}, and intuitively has to do with the fact that lineages from early mutation events often quickly die out.
+The process then repeats.
 \par
 
-The next establishment event and its corresponding class  is determined by sampling the each of the sojourn times of mutations in the front that are destined to establish and identifying the first to appear.
-Beneficial mutations within the bulk are ignored because these are lost to clonal interference.
-At the front, the number of beneficial mutations from a class $(i,j)$ in the bulk follows an inhomogeneous Poisson  process with a rate $\lambda_{i,j}(t)$.
-The rate will depend on the birth rate of individuals in the class $(i,j)$, their time-dependent abundances, and the beneficial mutation rate (see appendix).
-The sojourn time $\tau$ for the appearance of the first mutation destined to establish can be transformed into the random variable $x=\int_0^\tau \pi_{fix} \lambda_{i,j}(t) dt$ that has a standard exponential distribution.
+To determine the next class to establish, stochastic fluctuations induced by drift and mutations must be accounted for in the growth of classes at the front.
+The growth of classes in the front follows a branching processes fed by mutants from the adjacent (and exponentially growing) classes in the bulk.
+The next establishment event and its corresponding class  is determined by sampling each of the sojourn times of mutations in the front that are destined to establish and identifying the first to appear.
+The number of beneficial mutations from a class $(i,j)$ in the bulk follows an inhomogeneous Poisson  process with a rate $\lambda_{i,j}(t)=U (1+s_{i,j}) n_{i,j}(t)$, where $s_{i,j}=(r_{i,j}-\bar{r})$ is the selective advantage of an individual in class $(i,j)$, with respect to an average individual in the population.
+The probability that a new mutant lineage establishes is given by,
+\begin{equation} \label{eq:2}
+\pi = \frac{s_{i,j}}{ 1+s_{i,j}}
+\end{equation}
+(this is derived from \cite[Equation 16]{desai2007beneficial}).
+[Check i+1] The sojourn time $\tau$ for the appearance of the first mutation destined to establish can be transformed into the random variable $x=\int_0^\tau \pi \lambda_{i,j}(t) dt$ that has a standard exponential distribution.
 The sampling of $\tau$ is be done by generating a value for  $x$ and numerically solving for $\tau$ in the integral equation.
 This is repeated for each class of the bulk adjacent to the stochastic front, and the next class to establish is picked to be the one with the smallest sojourn time.
+Note that establishment in our simulation is solely based on the time to the first mutant establishment.
+In reality, an establishing class consists of multiple mutant lineages \cite{Desai2013}, but in practice this makes little quantitative difference (see \cite[p.1773]{desai2007beneficial}).
 \par
@@
 \end{equation}
-which is obtained from the probability density function of  $T$ in \cite[Equation 12]{desai2007beneficial}.
+which is obtained from the probability density function of  $T$ in \cite[Equation 12]{desai2007beneficial}.\par
 
 Following establishment, we update the set of classes in the bulk and calculate solutions to their time-dependent growth.
-Initially, classes with less than one individual are regarded as having gone extinct and removed from the bulk.
-Abundances of the remaining class and the newly established one change at a rate of  $\dot{n}_{i,j}(t) =(r_{i,j}-\bar{r})n_{i,j}(t)$ due to selection.
+Initially, low-fitness classes with less than one individual are regarded as having gone extinct and removed from the bulk, and the newly established class is added to the bulk.
+This will cause $N(t)=\sum n_{i,j}(t)$ to deviate slightly from $N$.
+To  maintain a total population size at $N$, we assume that selection in the bulk is governed by \cite[p.133]{Crow1970}
+\[ \dot{n}_{i,j} = (r_{i,j}-\bar{r})n_{i,j} - \bar{r}\left (\frac{N(t)}{N}-1 \right)n_{i,j}. \] 
+Since the removal and addition of rare types does not change $N$ much, this equation behaves almost identically to the standard selection equation $\dot{n}_{i,j}(t) =(r_{i,j}-\bar{r})n_{i,j}(t)$.
 Initial values are the size of each class at the time of establishment.
-Changes in the set of frequencies $p_{i,j}$ obey the same ordinary differential equation because $N$ is constant.
-Solutions of the coupled system of ODE's are then obtained using an ODE solver (NDSolve in Mathematica) for a period of one thousand generations, and are then used to determine the next establishment event.
-The process then continues for the duration of the simulation.
-Note that in our implementation the population size $N(t)=\sum n_{i,j}(t)$ can deviate slightly from the value $N$ as establishments take place and classes in the bulk become extinct.
-To account for this and maintain a total population size as close to $N$ as possible, we instead use the ODE
-\[ \dot{n}_{i,j} = (r_{i,j}-\bar{r})n_{i,j} - \bar{r}\left (\frac{N(t)}{N}-1 \right)n_{i,j} \] 
-given by \cite[p.133]{Crow1970} in our simulations.\par
+This coupled system is solved one thousand generations into the future, which is then used to determine the next establishment event.
+\par
 
@@
 cerevisiae \citep{desai2007speed,Levy2015,Perfeito2007}.
-We allowed the simulation to run for 5,000 generations, where one generation  This allowed the population to achieve beneficial mutation-selection balance before collecting data, and ensured that the transient effects from the initializing would not distort the statistics collected for the distribution.
+We allowed the simulation to run for 5,000 generations, where one generation is given by $1/s$ [flag].
+This allowed the population to achieve beneficial mutation-selection balance before collecting data, and ensured that the transient effects from the initializing would not distort the statistics collected for the distribution.
 Following the burn-in period of the simulation, we recorded the existing classes in the bulk and their respective abundances.
@@
 
-The data was used to compute the marginal mean fitness and variances of each trait, as well as their covariance.
-From these quantities we also computed the instantaneous rate of adaptation (Equation \eqref{eq:5}) for each trait and overall.
-The evolution of the population can be broken down into the evolution of the two traits, according to Lande's equation:
-\begin{equation}\label{eq:5}
+HERE: The frequencies $p_{i,j}$ of the bulk are used to calculate all of the quantitative genetic parameters that characterize the evolution of traits.
+Classes in the front are too small to affect the means, variances and covariances of the distribution  ($1/s \ll N$).
+The marginal mean fitnesses are  $\bar{r_1}=\sum_{i,j} p_{i,j} is_1$ and $\bar{r_2}=\sum_{i,j} p_{i,j} j s_2$.
+These  yield the mean contributions to fitness in the average individual from each trait.
+The genetic variances and covariance are $\sigma_1^2 = \sum_{i,j} p_{i,j} (r_i-\bar{r_1})^2$, $\sigma_2^2 = \sum_{i,j} p_{i,j} (r_j-\bar{r_2})^2$, and  $\sigma_{1,2} =\sum_{i,j} p_{i,j} (r_i-\bar{r_1})(r_j-\bar{r_2})$.
+According to Lande's equation, the evolution of the population can be broken down into the evolution of the two traits:
+\begin{equation}\label{eq:4}
 \left( \begin{array}{c}
@@
 \end{equation}
-where  $\bar{r_1}=\sum_{i,j} p_{i,j} is_1$, $\bar{r_2}=\sum_{i,j} p_{i,j} j s_2$, $\sigma_1^2 = \sum_{i,j} p_{i,j} (r_i-\bar{r_1})^2$, $\sigma_2^2 = \sum_{i,j} p_{i,j} (r_j-\bar{r_2})^2$, and  $\sigma_{1,2} =\sum_{i,j} p_{i,j} (r_i-\bar{r_1})(r_j-\bar{r_2})$.
-Here, it is the frequencies corresponding to the bulk that are relevant, since classes in the front are assumed to be too small to affect the means, variances and covariances of the distribution  ($1/s \ll N$).
-Lande's equation above is expressed as the change in the population's mean trait values over time intervals of size $s$.
-\par
+ expressed. Because there is no epistasis, $\bar{r} = \bar{r}_1 +\bar{r}_2$, and the instantaneous rate of adaptation of the population is $v(2U,N,s) = \sigma_1^2 +\sigma_2^2 + 2 \sigma_{1,2}$.  \par
 
@@
 \label{sec:results}
-
 The reduction in trait-specific adaptation $v_1$ from $v(U,N,s)$ down to $0.5 \hspace{.02 in} v(2U,N,s)$ (Figure \ref{fig:2}, extent to which solid lines are below 1), caused by clonal interference with the second trait, is driven by high levels of negative covariance (Figure \ref{fig:2}, closed circles).
 This is partly compensated for by a substantial increase in additive genetic variance (Figure \ref{fig:2}, open diamonds), because the negative covariance slows the removal of additive variance from the population.
+\par
 
@@
 \par
+
+\begin{figure*}[!ht]
+\centering
+\includegraphics[width=1\linewidth]{fig2.pdf} 
+\caption{Mean values in variance (open diamonds) of one trait and its covariance (closed circles) with the other, with their expected sum equal to the analytically calculated adaptation rate (solid line).
+Each average was calculated from one simulation that ran 1,000,000 generations.
+Variance of the focal trait is always higher than it would be in the absence of the second trait ($\sigma_1^2 >1$); negative covariance more than cancels this out ($|\sigma_{1,2}| > \sigma_1^2-1$) to reduce the adaptation rate.
+Mean variance and covariance are insensitive to changes in the population size $N$ (c), but exhibit some dependence on the values of $s$ (a) and $U$ (b).
+For the parameter values not being varied on the x-axis, $s=0.02$, $U=10^{-5}$ and $N=10^9$.
+}\label{fig:2}
+\end{figure*}
 
@@
 
-Under beneficial mutation-selection balance, the components of the \G matrix fluctuate far more than the instantaneous rate of adaptation, with transient spikes in additive genetic variance $\sigma_1^2$ mostly canceled out by transient spikes in negative covariance $\sigma_{1,2}$ (representative simulation shown in Figure \ref{fig:4}).
-As a result, fluctuations in variance and covariance can be substantially larger than those of total variance in fitness.\par
-
-The two-dimensional traveling wave of Figure \ref{fig:1} can be projected onto a one-dimension traveling wave of fitness, where classes along a fitness isocline such as those along the red line in Figure \ref{fig:1}, are aggregated into fitness classes (Figure \ref{fig:4}).
-The shape of the one-dimensional fitness distribution is Gaussian and highly peaked near the population's mean fitness \citep{desai2007beneficial}.
-This means that the \G matrix will dominated by the distribution of genotypes in this fitness class (red line in Figure \ref{fig:1}, mode in (Figure \ref{fig:4}).
-As adaptation continues, a fitter class becomes the most abundant that will have a different distribution of genotypes and hence different trait variances and covariance.
-On average, the time required for this shift is the mean time between establishments at the nose of the of the traveling wave, denoted $\tau_q$ in \citet{desai2007beneficial}.
-
-\begin{equation}\label{eq:8}
-\tau_q = \frac{\log^2(s/U)}{s(2\log(Ns)-\log(s/U))}
-\end{equation}\par
-
-Genotypes frequencies within the peak of the traveling wave are set during the stochastic growth phase of the fitness class prior to establishing \citep{Desai2013}.
-As previously discussed, beneficial mutations that give rise to mutants in the nose produce distinct but equally fit mutant lineages.
-Many of lineages become extinct, but those that manage to survive are subsequently and equally amplified by selection following establishment.
-Since the relative sizes of the surviving lineages with respect to one do not change once they begin growing deterministically, the frequencies of genotypes in a fitness class become ``frozen'' at the time of establishment.
-Beneficial mutations that occur after that time contribute very little to the size and composition of fitness class, since their corresponding lineages will be exponentially smaller than those which initially appeared.
-After a fitness class has established, it eventually goes on to become the dominant group of the fitness distribution, or rather, it forms the peak of the traveling wave.
-The time between its establishment and its succession as the dominant fitness class is called the sweep time (denoted $\tau_{sw}$).
-Consequently, the frequencies of genotypes in the nose determine the values of variance and covariances approximately $\tau_{sw}$ generations later.
-Figure \ref{fig:6} plots the correlation between trait covariances of the bulk and nose, as a function of a time offset.
-The signal peaks at approximately $\bar{\tau}_{fix}$ (mean time to fixation).
-It shows that covariances between traits are largely set by the behavior at the nose of the one dimensional traveling wave, which consists of genotypes along the lead fitness isocline through the stochastic front.
-The values of variances and covariance will be correlated from one transition to the next, because mutational influx depends on the abundances of adjacent less-fit genotypes.
-
-% (Dynamics of structures withing the \G matrix)
-Principle components can be used to examine the structural features of our \G matrix.
-Although Fluctuations in the individual components $\sigma_1^2$, $\sigma_1^2$, $\sigma_{1,2}$ can be large relative to the rate of adaptation, but there sum has lower overall variance (Figure \ref{fig:5}, black line).
-The
-\[
-E\left[
-\left( \begin{array}{cc}
-\sigma_1^2 & \sigma_{1,2} \\
-\sigma_{1,2} & \sigma_2^2 
-\end{array}\right)
-\right] = 
-\]
-
-\begin{figure}[!ht]
-\includegraphics[width=1\linewidth]{fig1a.pdf}
-\caption{Representative two-dimensional genotype distribution.
-Individuals with equal numbers of beneficial mutations in each trait are combined into classes (boxes).
-Abundances of the bulk (solid boxes) behave deterministically, while abundance of the stochastic front (patterned boxes) behave stochastically.
-The red dot marks the average number of beneficial mutations in each trait, and equally fit classes are found along the red fitness isocline).
-In asexuals, beneficial mutations must occur on the fittest genetic backgrounds in order to contribute to the adaptive process.
-These mutations are represented by arrows from the bulk into the stochastic front.
-Classes in the stochastic front become part of the bulk once their abundances have become sufficiently large, pushing the stochastic front to include new classes.
-As classes below the population mean fitness isocline decline and those above increase exponentially, a two-dimensional traveling wave is produced.
-(Simulation parameters: $N=10^9$, $s=0.02$, and $U=10^{-5}$)}\label{fig:1}
-\end{figure}
-
-\begin{figure*}[!ht]
-\centering
-\includegraphics[width=1\linewidth]{fig2.pdf} 
-\caption{Variance (open diamonds) of one trait and its covariance (closed circles) with the other, averaged across simulations, whose expected sum is equal to the analytically calculated adaptation rate (solid line).
-Variance of the focal trait is always higher than it would be in the absence of the second trait ($\sigma_1^2 >1$); negative covariance more than cancels this out ($|\sigma_{1,2}| > \sigma_1^2-1$) to reduce the adaptation rate.
-Mean variance and covariance are insensitive to changes in the population size $N$ (c), but exhibit some dependence on the values of $s$ (a) and $U$ (b).
-For the parameter values not being varied on the x-axis, $s=0.02$, $U=10^{-5}$ and $N=10^9$.
-}\label{fig:2}
-% * <masel@email.arizona.edu> 2018-01-22T19:36:50.877Z:
-% 
-% > averaged across simulations
-% How many simulations were averaged?
-% 
-% ^.
-\end{figure*}
-
 \begin{figure}[!ht]
@@
 \includegraphics[width=1\linewidth]{fig3.pdf}
-\caption{Rate of adaptation of a focal trait as a function of the number of total traits undergoing adaptation.
-Each additional trait reduces the rate of adaptation, but by marginally less as the trait number grows.
-The decrease decays to zero in the Equation \eqref{eq:6}, but in practice the expression is only valid when $kU/s \ll 1$.
+\caption{The rate of adaptation of a focal trait as a function of the number of total traits undergoing adaptation.
+The decrease from each additional trait ($n>2$)is marginally less and decays to zero in the Equation \eqref{eq:6}, but in practice the expression is only valid when $kU/s \ll 1$.
 The effects of clonal interference on trait evolution are most significant in the case of two traits.
-The values of the parameters are taken as $s=0.02$, $U=10^{-5}$ and $N=10^9$.}
-% * <masel@email.arizona.edu> 2017-10-28T22:59:26.669Z:
-% 
-% > Each plot includes the measured values produced by simulations with either varying $s$, $U$, or $N$. 
-% Aren't var and covar from simulations and v from the equation?
-% 
-% ^.
+The parameters are fixed at $s=0.02$, $U=10^{-5}$ and $N=10^9$.}
 \label{fig:3}
 \end{figure}
+
+Under beneficial mutation-selection balance, the components of the \G matrix fluctuate far more than the instantaneous rate of adaptation, with transient spikes in additive genetic variance $\sigma_1^2$ mostly canceled out by transient spikes in negative covariance $\sigma_{1,2}$ (representative simulation shown in Figure \ref{fig:4}).
+As a result, fluctuations in variance and covariance can be substantially larger than those of total variance in fitness.\par
 
@@
 % ^.
-\caption{Trajectories of the variance of trait one (dashed line), covariance (dotted line), and the total variance in fitness of the population (solid line), plotted over 10,000 generations.
+\caption{Trajectories of the variance of trait one (dashed line), covariance (dotted line), and the instantaneous rate of adaptation (solid line), plotted over 10,000 generations.
 The mean values of each are included as lines with matching line-styles.
@@
 % 
+% ^ <kgomez81@math.arizona.edu> 2018-02-11T06:41:44.720Z:
+% 
+% The solid line is the instantaneous rate of adaptation in fitness.
+%
 % ^.
 \end{figure}
+
+The two-dimensional traveling wave of Figure \ref{fig:1} can be projected onto a one-dimension traveling wave of fitness, where classes along a fitness isocline such as those along the red line in Figure \ref{fig:1}, are aggregated into fitness classes (Figure \ref{fig:4}).
+The shape of the one-dimensional fitness distribution is Gaussian and highly peaked near the population's mean fitness \citep{desai2007beneficial}.
+This means that the \G matrix will dominated by the distribution of genotypes in this fitness class (red line in Figure \ref{fig:1}, mode in (Figure \ref{fig:4}).
+As adaptation continues, a fitter class becomes the most abundant that will have a different distribution of genotypes and hence different trait variances and covariance.
+On average, the time required for this shift is the mean time between establishments at the nose of the of the traveling wave, denoted $\tau_q$ in \citet{desai2007beneficial}.
+\begin{equation}\label{eq:8}
+\tau_q = \frac{\log^2(s/U)}{s(2\log(Ns)-\log(s/U))}
+\end{equation}\par
 
@@
 
+Genotypes frequencies within the peak of the traveling wave are set during the stochastic growth phase of the fitness class prior to establishing \citep{Desai2013}.
+As previously discussed, beneficial mutations that give rise to mutants in the nose produce distinct but equally fit mutant lineages.
+Many of lineages become extinct, but those that manage to survive are subsequently and equally amplified by selection following establishment.
+Since the relative sizes of the surviving lineages with respect to one do not change once they begin growing deterministically, the frequencies of genotypes in a fitness class become ``frozen'' at the time of establishment.
+Beneficial mutations that occur after that time contribute very little to the size and composition of fitness class, since their corresponding lineages will be exponentially smaller than those which initially appeared.
+After a fitness class has established, it eventually goes on to become the dominant group of the fitness distribution, or rather, it forms the peak of the traveling wave.
+The time between its establishment and its succession as the dominant fitness class is called the sweep time (denoted $\tau_{sw}$).
+Consequently, the frequencies of genotypes in the nose determine the values of variance and covariances approximately $\tau_{sw}$ generations later.
+Figure \ref{fig:6} plots the correlation between trait covariances of the bulk and nose, as a function of a time offset.
+The signal peaks at approximately $\bar{\tau}_{fix}$ (mean time to fixation).
+It shows that covariances between traits are largely set by the behavior at the nose of the one dimensional traveling wave, which consists of genotypes along the lead fitness isocline through the stochastic front.
+The values of variances and covariance will be correlated from one transition to the next, because mutational influx depends on the abundances of adjacent less-fit genotypes.\par
+
 \begin{figure}[!ht]
@@
 \caption{Correlation between covariance of stochastic front and bulk.
-The correlation between the two peaks when their signals are offset by the the mean time to fixation $\bar{\tau}_{fix}$ (marked by the blue line), indicating that the dynamics of the stochastic front explains 69\% of the covariance lated detected in the bulk $\bar{\tau}_{fix}$ generation after.}\label{fig:6}
+The correlation between the two peaks when their signals are offset by the mean sweep time of the nose $\bar{\tau}_{SW}$ (dashed line), indicating that the dynamics of the stochastic front explains 69\% of the covariance lated detected in the bulk $\bar{\tau}_{SW}$ generation after.}\label{fig:6}
 % * <masel@email.arizona.edu> 2017-10-28T23:57:21.626Z:
@@
 \end{figure}
+
+Fluctuations in the individual components $\sigma_1^2$, $\sigma_1^2$, and $\sigma_{1,2}$ are large relative to the expected rate of adaptation of each trait (Figure \ref{fig:4}), but the shape and orientation of the \G matrix are largely preserved throughout its evolution.
+The \G matrix expressed in Equation \ref{eq:4} for two symmetric traits ($E[\sigma_1^2] = E[ \sigma_2^2]$ and $v_1 = v_2 $ ) has eigenvalues
+\[ \lambda = \frac{1}{2}\left( \sigma_1^2+ \sigma_2^2 \pm \sqrt{ (\sigma_1^2 -\sigma_2^2)^2 +4 \sigma_{1,2}^2 }\right) \approx \frac{1}{2}\left(\sigma_1^2+ \sigma_2^2\right) \pm  \sigma_{1,2}.  \]
+The first eigenvalue is the rate of adaptation in one trait, while the second is the variance in the difference between the two traits.
+The principle components are $\vec{x}_1 = \frac{1}{\sqrt{2}}(1,1)^T$ and $\vec{x}_2 =\frac{1}{\sqrt{2}}(1,-1)^T$ for the first and second eigenvalues, respectively.
 

### 2018-02-21 plots.py
added plot for angles and eigenvalues of the G matrix

@@ -333,3 +333,93 @@ ax.yaxis.set_ticks_position('left')
 ax.legend()
 plt.show()
 fig.savefig('./figures/fig6b.pdf')
+
+
+#-----------------------------------------------------------------------------------------
+#--------------FIGURE 7: Dynamics of G matrix shape & orientation-------------------------
+#-----------------------------------------------------------------------------------------
+
+# figure 3: plot of rate of adaptation, variances, covariance and their means
+[N,s,U] = [1e9,1e-2,1e-5]
+[sim_start,sim_end,snapshot] = [5e3,4e4,1.313e4]
+vUNs = pltfun.get_vNsU(N,s,U)
+
+# load time series data of distrStats from plotdata.py output
+pickle_file_name = './data/pythondata/distrStats_N-10p09_c1-0d01_c2-0d01_U1-1x10pn5_U2-1x10pn5_exp1.pickle'
+pickle_file = open(pickle_file_name,'rb') 
+[times,mean_fit,fit_var,fit_cov,pop_load,dcov_dt,vU_thry,v2U_thry] = pickle.load(pickle_file)
+pickle_file.close()
+
+# select interval of simulation data that will be used for plot
+# reduce loaded data to subset corresponding to selected interval
+[start_indx,end_indx] = pltfun.get_sample_window(times,sim_start,sim_end)
+times = times[start_indx:end_indx]
+fit_var = fit_var[start_indx:end_indx]
+fit_cov = fit_cov[start_indx:end_indx]
+var_diff = (fit_var[:,0]-fit_var[:,1])
+n1 = len(fit_cov)
+
+trG = fit_var[:,0]+fit_var[:,1] 
+detG =  np.asarray([fit_var[i,0]*fit_var[i,1]-fit_cov[i]**2 for i in range(n1)])
+Gmatr = [np.asarray([[fit_var[i,0],fit_cov[i]],[fit_cov[i],fit_var[i,1]]]) for i in range(n1)]
+Xmatr = np.asarray([[1/sqrt(2),1/sqrt(2)],[1/sqrt(2),-1/sqrt(2)]])
+
[…]

### 2018-02-21 plots.py
made some additional modification to the plot for the angle and eignvalues of the G matrix

@@ -388,9 +388,9 @@ for i in range(n1):
     Gang = Gang + [Ang1*2/pi]
 
 # test plots
-plt.plot(times,var_diff/vUNs)
-plt.plot(times,4*fit_cov/vUNs)
-plt.plot(times,10*np.asarray(Gang))
+#plt.plot(times,var_diff/vUNs)
+#plt.plot(times,4*fit_cov/vUNs)
+#plt.plot(times,10*np.asarray(Gang))
 
 my_xticks = [10000+i*1000 for i in range(11)]
 my_xlabel = ['10', '', '12', '', '14', '', '16', '', '18', '', '20']
@@ -401,10 +401,10 @@ my_xlabel = ['10', '', '12', '', '14', '', '16', '', '18', '', '20']
 
 fig, ax1 = plt.subplots(1,1,figsize=[8,8])
 ax2 = plt.twinx(ax1)
-ax1.plot(times,(1/vUNs)*lambda1,c="blue",linewidth=2.0,label='$\lambda_1$')
-ax1.plot(times,(1/vUNs)*lambda2,c="red",linewidth=2.0,label='$\lambda_2$')
-#ax1.plot(times,(1/vUNs)*var_diff,c="blue",linewidth=2.0,linestyle="-.",label='$\sigma_1^2-\sigma_2^2$')
-ax1.plot(times,(1/vUNs)*fit_cov,c="black",linewidth=2.0,linestyle="-.",label='$\sigma_{1,2}$')
+ax1.plot(times,(1/vUNs)*lambda1,c="blue",linewidth=2.0,linestyle="-",label='$\lambda_1$')
+ax1.plot(times,(1/vUNs)*lambda2,c="green",linewidth=2.0,linestyle="-",label='$\lambda_2$')
+ax1.plot(times,(1/vUNs)*var_diff,c="magenta",linewidth=2.0,linestyle="-",label='$\sigma_1^2-\sigma_2^2$')
+ax1.plot(times,(1/vUNs)*fit_cov,c="cyan",linewidth=2.0,linestyle="-",label='$\sigma_{1,2}$')
 ax1.set_ylabel('Eigenvalues of G matrix (multiples of v(U,N,s))',fontsize=20)
 ax1.set_xlabel('Time (Thousands of Generations)',fontsize=20)
 ax1.axhline(linewidth=0.5, color = 'k')
@@ -416,7 +416,7 @@ ax1.grid(b='off', which='both', axis='both')
[…]

### 2018-03-06 plots.py
updated figure 4 with jason's recommendations

@@ -267,13 +267,13 @@ my_ylabel = ['-4', '', '-2', '', '0', '', '2', '', '4', '', '6']
 
 ax=plt.subplot(311)
 ax.plot(times,(1/vU_thry)*rate_adpt_t1_avg,c="black",label=r'$v_1$',linewidth=2.0,linestyle = '-')                
-ax.plot(times,(1/vU_thry)*var1_avg,c="black",label=r'$\sigma_1^2$',linewidth=3.0,linestyle = '--')
-ax.plot(times,(1/vU_thry)*cov_avg,c="black",label=r'$\sigma_{1,2}$',linewidth=3.0,linestyle = ':')
-ax.plot(times,(1/vU_thry)*rate_adpt_t1,c="black",linestyle="-",linewidth=3.0)        
-ax.plot(times,(1/vU_thry)*fit_var[:,0],c="black",linestyle="--",linewidth=3.0)
-ax.plot(times,(1/vU_thry)*fit_cov[:],c="black",linestyle=":",linewidth=3.0)
+ax.plot(times,(1/vU_thry)*var1_avg,c="black",label=r'$\sigma_1^2$',linewidth=2.0,linestyle = '--')
+ax.plot(times,(1/vU_thry)*cov_avg,c="black",label=r'$\sigma_{1,2}$',linewidth=2.0,linestyle = ':')
+ax.plot(times,(1/vU_thry)*rate_adpt_t1,c="black",linestyle="-",linewidth=2.0)        
+ax.plot(times,(1/vU_thry)*fit_var[:,0],c="black",linestyle="--",linewidth=2.0)
+ax.plot(times,(1/vU_thry)*fit_cov[:],c="black",linestyle=":",linewidth=2.0)
 ax.axhline(linewidth=0.5, color = 'k')
-ax.set_ylabel(r'$\sigma_1^2$ & $\sigma_{1,2}$ / $v(U,N,s)$',fontsize=lsize1)
+ax.set_ylabel(r'$\sigma_1^2$ and $\sigma_{1,2}$ / $v(U,N,s)$',fontsize=lsize1)
 ax.legend(loc=4,ncol=3,fontsize=lsize1)
 ax.set_ylim((np.min(cov_avg),np.max(var1_avg)))
 ax.set_xlim((1e4,2e4))
@@ -292,11 +292,11 @@ my_ylabel = ['', '-2', '', '0', '', '2', '', '4', '', '6', '', '8', '', '10']
 ax=plt.subplot(312)
[…]

### 2018-03-06 plotdata.py
fixed code for getting data needed to create figure 4

@@ -529,7 +529,7 @@ n1 = len(fit_cov)
 trG = fit_var[:,0]+fit_var[:,1] 
 detG =  np.asarray([fit_var[i,0]*fit_var[i,1]-fit_cov[i]**2 for i in range(n1)])
 Gmatr = [np.asarray([[fit_var[i,0],fit_cov[i]],[fit_cov[i],fit_var[i,1]]]) for i in range(n1)]
-Xmatr = np.asarray([[1/sqrt(2),1/sqrt(2)],[1/sqrt(2),-1/sqrt(2)]])
+Xmatr = np.asarray([[1/np.sqrt(2),1/np.sqrt(2)],[1/np.sqrt(2),-1/np.sqrt(2)]])
 
 lambda1 = np.asarray([0.5*(trG[i]-np.sqrt(trG[i]**2-4*detG[i])) for i in range(n1)])
 lambda2 = np.asarray([0.5*(trG[i]+np.sqrt(trG[i]**2-4*detG[i])) for i in range(n1)])
@@ -547,9 +547,9 @@ for i in range(n1):
         Gvec = Gvec+[A[1]]
 
 for i in range(n1):
-    Ang1 = np.arccos((sign(Gvec[i][0,0])*Gvec[i][0,0]*Xmatr[0,0]+sign(Gvec[i][1,0])*Gvec[i][1,0]*Xmatr[1,0])/np.linalg.norm(Gvec[i][:,0]))
-    Ang1 = sign(sign(Gvec[i][1,0])*Gvec[i][1,0]-sign(Gvec[i][0,0])*Gvec[i][0,0])*Ang1
-    Gang = Gang + [Ang1*2/pi]
+    Ang1 = np.arccos((np.sign(Gvec[i][0,0])*Gvec[i][0,0]*Xmatr[0,0]+np.sign(Gvec[i][1,0])*Gvec[i][1,0]*Xmatr[1,0])/np.linalg.norm(Gvec[i][:,0]))
+    Ang1 = np.sign(np.sign(Gvec[i][1,0])*Gvec[i][1,0]-np.sign(Gvec[i][0,0])*Gvec[i][0,0])*Ang1
+    Gang = Gang + [Ang1*2/np.pi]
 
 # convert list to array  
 Gval = np.asarray(Gval)

### 2018-03-07 manuscript.tex
updated with overleaf code to debug

@@
 \usepackage[nameinlink,capitalise]{cleveref}
-\graphicspath{{figures/}}
+% \graphicspath{{figures/}}
 \newcommand{\G}{\textbf{G }}
@@
 
-Genetic variances and covariances also evolve over time, and as result, the long term evolution of correlated traits will depend on both selection gradients and the dynamics of the \G matrix.
-As the mean phenotype of a populations changes, mean fitness increases.
-The process can be visualized as movement along a fitness surface (Simpson's Adaptive Landscape) defined over the space of phenotypes \cite{phillips1989visualizing}.
-Lande's expression characterizes how the \G matrix influences the local behavior of trajectories on this surface.
-The trajectories followed by populations depict their macroevolution, and as result, interest in the \G matrix has grown among evolutionary biologist \cite{Arnold2001,Arnold2008}.
+The evolutionary trajectories of traits across many generations depend on the dynamics of the \G matrix.
+When the \G matrix is stable, it is possible to infer past selection gradients \cite{lande1979quantitative} and forecast future trait evolution (both direction and rate) \citep{via1985genotype,arnold1992constraints,bjorklund1996importance,Schluter1996,teplitsky2011quantitative,teplitsky2014assessing}.
+However, it has not come into widespread use for this purpose, perhaps in part because its stability cannot be assumed a priori.
+Theoretical models \citep{Turelli1988,Burger1994,Jones2003,Jones2004}, comparative studies \citep{bjorklund2013rapid,Waldmann2000}, and experimental evolution \citep{wilkinson1990resistance,Shaw1995,Phillips2001} have all demonstrated that rapid change in the \G matrix is possible.
+How stable \G is in natural populations remains an open question \citep{Steppan2002,Arnold2008}.
 
-When the \G matrix is stable, it is possible to infer past selection gradients (need citations) and forecast future trait evolution (both direction and rate) \citep{teplitsky2014assessing,teplitsky2011quantitative,Schluter1996}.
-However, it has not come into widespread use for this purpose, perhaps in part because its stability cannot be assumed a priori.
-\cite[p.407]{lande1979quantitative} suggested that the \G matrix might be near an equilibrium determined by a balance of mutation, selection and recombination \cite{lande1977influence, lande1980genetic}, and subsequent work has examined this claim in more detail.
-On one front, theoretical models based on quantitative genetics have developed to try and determine conditions under which the \G matrix is stable.
-On the other, empirical work focused on
+Comparative studies suggest that at least certain aspects of \G might be stable \citep{Arnold2008}.
+Early methods for comparing \G were inconclusive about its stability, but these relied on statistical approaches that compared the magnitudes of the individual variance and covariance elements \citep{atchley1992evolutionary,lofsvold1986quantitative,Brodie1993,paulsen1996quantitative,roff1999does,Roff1999}.
+In contrast, more recent methods that examine common principal components of \G matrices have shown that certain structures related to their shape and orientation are preserved between closely related populations \citep{Arnold2008}.
 
-A large body of work measuring and comparing it between populations.
-There are several approaches to comparing the \G matrices of different populations \citep{Steppan2002}.
-Some compare the magnitudes of the individual variance and covariance elements \citep{atchley1992evolutionary,lofsvold1986quantitative}.
-However, this might not capture important characteristics of \G matrices that are shared between populations.
-Methods based on the use of principal components, such as common principle component analysis (CPCA), in contrast, can compare features between \G matrices such as shape, size, and orientation.
-These latter methods have gained popularity, and have the potential to reveal which evolutionary forces significantly influenced the past dynamics of the population's \G matrix \cite{Phillips1999,Arnold2008}.
-% * <masel@email.arizona.edu> 2018-02-14T18:34:10.127Z:
-% 
-% > Some compare the magnitudes of the individual variance and covariance elements \citep{atchley1992evolutionary,lofsvold1986quantitative}.
-% Add more citations beyond the statistically flawed element by element.
-% 
-% 
-% ^.
-
-% (Lande Notes): Lande (1979) indicated that if the \G matrix were stable, it would be possible to measure selective forces leading to differentiation between two contemporary species, assuming other effects polyploidy, macro-mutation, and the direct effects of the environment can be ignored.
-Consequently, phyletic changes caused by responses to natural selection; assuming that the appearance of new forms is due to the accumulation of genetic variation, and ignoring polyploidy, macromutation, and the direct effects of the environment.
-Lande later points out that if genetic variances and covariances remain constant in a lineage, its net selection gradient can be obtained using \citep[Equation 9]{lande1979quantitative}.
-
-%Work by \cite{price1984genetic} examined this in two small populations of Darwin's ground finches (Geospiza), where he used measurements of genetic and ecological parameters obtained during a 6-year study of their morphological evolution.
-This included data on variances and covariances for weight, beak length, beak width, beak depth, population size, and estimates of the selection surface for local shape of the adaptive landscape(directional beta and quadratic term gamma using in Lande and Arnold's 1983 approach with regression).
-Observations of sexual selection indicated a connection between morphological evolution and reproductive isolation, hence speciation.
-Studies by Ratcliffe and Grant (1983) have shown that body size and beak characteristics are used by sexes to distinguish between con- and heterospecifics, and these may have been brought about by differences in sex ratios (observed in 1977 drought).
-The coupling of morphological evolution and reproductive isolation is unique in these species.
-The use of Lande's ideas involving selection leading to species transition are discussed page 58 \citep{lande1979quantitative}.
-
-%Lande's 1977 paper demonstrated that mating systems had no effect on the genetic variation of additive genes in large populations, and ``found'' that genetic variance is determined entirely by mutation and selection parameters, as well as environmental variance (Equation 9).
-The implication from this work is that in natural populations mutation and selection can maintain constant genetic variances (also depends on environmental variance).
-Lande's 1980 paper deals with finding expressions for the dynamics of the \G matrix under selection, pleiotropic mutations, and recombination.
-With selection weak relative to recombination (eq 18), the dynamics of the \G matrix depend entirely on mutations, selection and environmental variances (eq 21c, 22a); this assumption of weak selection is made clear below Equation 17.
-\par
-
-% * <masel@email.arizona.edu> 2018-02-14T18:17:58.918Z:
-% 
-% > When the \G matrix is stable, it is possible to infer past selection gradients
-% Need citation
-% 
-% ^.
-
-%For example, studies of past macro-evolution in populations and species have used net selection gradients to draw connections between morphological evolution and reproductive isolation \citep{price1984genetic}, infer morphological and phylogenetic relations \citep{schluter1984morphological}, and determine the extent of local adaptation between populations \citep{dudley1996}.
-Similar work has looked the impact of genetic constraints on rates of adaptation \citep{teplitsky2014assessing,teplitsky2011quantitative}, their ability to bias phenotypic evolution over millions of years \cite{Schluter1996}.
-
-% (Needs Work)
-At least some features of the \G matrix can be relatively stable.
-Over 70\% of the \G matrices comparisons between conspecific populations or between related species find similar shape and orientation \cite{Arnold2008}, suggesting that aspects of genetic constraints can be maintained over significant time scales.
-However, comparative studies of \G also indicate that selection and drift can produce instabilities even over very short time scales \cite{bjorklund2013rapid,Siren2017,Phillips2001,wilkinson1990resistance,roff1999does,Steppan1997,Waldmann2000}.
-% \cite{roff1999does} \cite{Begin2003,Begin2004} 
-5
-%(Needs Work)
-Quantitative genetic models confirm that the \G matrix can be changed by these and other factors, and they have been used to identify other conditions that contribute stability of the \G matrix \citep{Turelli1988,barton1989evolutionary,reeve2000predicting}.
-
-Environmental change affects genotype $\times$ environment interactions and hence changes the \G matrix stability, so to assume a constant \G matrix, one must assume the absence of GxE interactions.
-Simulations have revealed that large population sizes and strong correlational selection can induce stability in \G's dynamics \cite{Jones2003}.
-When mutational covariances can respond to selection, the \M matrix and hence the \G matrix is more stable \cite{Jones2007}.
-
-Historically, pleiotropic effects have been emphasized over linkage disequilibrium in their contribution to \G's dynamics, since recombination is assumed to remove the latter \citep{lande1979quantitative,lande1980genetic,Arnold2008} (more).
-As a result, we know relatively little about the possible effects that linkage disequilibrium might have on the long-term evolution of genetic variances and covariances.
-% * <masel@email.arizona.edu> 2018-02-14T18:56:02.901Z:
-% 
-% > When mutational covariances can respond to selection, the \M matrix and hence the \G matrix is more stable \cite{Jones2007}.
-% Look up this paper. This might not be the mechanism. If the paper is important, spell out the mechanism. Otherwise, leave it out.
-% 
-% 
-% ^.
-% * <masel@email.arizona.edu> 2018-02-14T18:53:18.109Z:
-% 
-% > large population sizes and strong correlational selection can induce stability in \G's dynamics \cite{Jones2003}
-% Rephrase to be about how drift changes G, not how large N promotes G stability.
-% 
-% ^.
-% * <masel@email.arizona.edu> 2018-02-14T18:42:01.952Z:
-% 
-% > However, comparative studies of \G also indicate that selection and drift can produce instabilities even over very short time scales \cite{bjorklund2013rapid,Siren2017,Phillips2001,wilkinson1990resistance,roff1999does,Steppan1997,Waldmann2000}.
-% > % \cite{roff1999does} \cite{Begin2003,Begin2004} 
-% > Quantitative genetic models confirm that the \G matrix can be changed by these and other factors, and they have been used to identify other conditions that contribute stability of the \G matrix \citep{Turelli1988,barton1989evolutionary,reeve2000predicting}.
-% Not specific enough: unpack. Can also be done in the Discussion instead. Following text does make more specific points: add citations of who made them to these.
-% 
-% ^.
-
-% * <masel@email.arizona.edu> 2018-02-12T00:03:04.529Z:
-% 
-% > such as the distribution of allelic effects \cite{burger2000mathematical}. Simulations have been used to bypass this problem
-% It seems unlikely that simulations would bypass the problem of not knowing the distribution of allelic effects.
-% 
-% ^ <kgomez81@math.arizona.edu> 2018-02-14T17:55:06.818Z:
-% 
-% I removed this sentence.
-I know that quantitative genetic models depend on frequencies of alleles and the distribution of their effects, and this has limited the use of theory to gain further insights on G's stability.
-Simulations, such as those of Jones (mentioned later), have been used instead, but I believe that they tackle mostly the issue of unknown frequencies.
-I am not sure that they address the distribution of allelic effects in any way that solves the issues identified in the quantitative genetic theory.
-I would have to fish out the details from Bruce's book.
-%
-% ^.
-% * <masel@email.arizona.edu> 2018-02-12T00:01:20.366Z:
-% 
-% > Other critical stability conditions involve the strength of selection and mutational covariances.
-% This sentence doesn't tell me what the stability condition is. You don't need to go into details, but you do need to give a heuristic.
-% 
-% ^ <kgomez81@math.arizona.edu> 2018-02-14T17:57:51.147Z:
-% 
-% I added a heuristic explanation, but I will work on improving it.  I think I can do better.
-%
-% ^.
-% * <masel@email.arizona.edu> 2018-02-12T00:00:05.266Z:
-% 
-% > changes in genotype $\times$ environment interactions directly affect genetic variances and covariances
-% I can't easily follow the logic of this just by reading; more explanation is in order.
-% 
-% 
-% ^ <kgomez81@math.arizona.edu> 2018-02-13T23:33:02.198Z:
-% 
-% For example, genotype $\times$ environment interactions must be absent since they can lead to changes in genetic variances and covariances with new environments.
-%
-% ^.
-
-The genetic correlations described by \G are due to some combination of pleiotropy and linkage disequilibrium \cite{Saltz2017}.
-Historically, pleiotropy was favored by the Edinburgh school, who argued that recombination would quickly eliminate linkage disequilibrium \citep[Chapter~20]{fox2006evolutionary}, particularly in the randomly mating animal populations that the Edinburgh school focused on.
+Stable aspects of the \G matrix have been interpreted as stable pleiotropic constraints \citep{lande1979quantitative,Arnold2008}.
+However, the genetic correlations described by \G are due to the combined forces of \M (mutation matrix) and selection, with much attention on \M, whose covariances derive from both pleiotropy and linkage disequilibrium \citep{Saltz2017}.
+A dominant role for pleiotropy was favored by the historically influential Edinburgh school of quantitative genetics, who argued that recombination would quickly eliminate linkage disequilibrium \citep[Chapter~20]{fox2006evolutionary}, particularly in the randomly mating animal populations that the Edinburgh school focused on \citep{lande1979quantitative,lande1980genetic,Arnold2008} (more).
 In contrast, quantitative geneticists in the alternative, Birmingham school largely worked on inbred lines of plants, where the effects of linkage disequilibrium were impossible to ignore.
 The Edinburgh view came to be widely accepted, and as a result, most models for the evolution of the \G  matrix assume that genetic correlations are pleiotropic and go on to examine how evolutionary processes influence them \citep{Turelli1988}.
-We know far less about the about the dynamics of genetic correlations arising from linkage disequilibrium.\par
+We know far less about the about the dynamics of genetic correlations arising from linkage disequilibrium.
 
-To study how linkage disequilibrium changes genetic variances and covariance over time [this is not what you're studying, need to define a purpose here related to your quesiton(s)], we have to consider explicit alleles or genotypes, not just quantitative traits.
+To study how linkage disequilibrium changes genetic variances and covariance over time, we have to consider explicit alleles or genotypes, not just quantitative traits.
 Beneficial mutations are discrete not infinitesimal, and appear on distinct genetic backgrounds.
 Selective sweeps compete with one another, and so many beneficial mutations are lost in a process known as ``clonal interference''.
-This will slow down adaptation \citep{hill1966effect} unless recombination brings them together, reducing the negative linkage disequilibrium between competing beneficial mutations \cite{fisher1930genetical,muller1932some}.
+This will slow down adaptation \citep{hill1966effect} unless recombination brings them together, reducing the negative linkage disequilibrium between competing beneficial mutations \citep{fisher1930genetical,muller1932some}.
 Classical models of clonal interference considered only two loci each with two alleles.
@@
 These models have documented that clonal interference can have an enormous impact on adaptation rates.
-However, to date, most of these models have treated only the evolution of ``fitness'' rather than the evolution of individual traits and their correlations \cite{}.
-\par
+However, to date, most of these models have treated only the evolution of ``fitness'' rather than the evolution of individual traits and their correlations.
 
@@
 % ^.
-% * <masel@email.arizona.edu> 2018-01-24T22:37:04.241Z:
-% 
-% >  \citep{rouzine2003solitary,desai2007beneficial,park2010speed,good2012distribution}
-% You can include sexual models here too, eg by Neher
-% 
-% ^ <kgomez81@math.arizona.edu> 2018-02-09T23:38:56.909Z:
-% 
-% I added the Neher paper, and other references. I'll conduct another search for other sources. 
-% 
-% ^.
 
 Here, we focus on asexual populations, because they have the strongest clonal interference and resulting linkage disequilibrium.
-Moreover, biology today has come to recognize the numerical prevalence and ubiquitous significance of relatively asexual microbes \cite{mcfall2013animals}.\par
+Moreover, biology today has come to recognize the numerical prevalence and ubiquitous significance of relatively asexual microbes \citep{mcfall2013animals}.
 
@@
 When there is only one adaptive trait, $\sigma^2$ is that trait's additive variance.
-\par
 
@@
 The reduction in trait-specific adaptation rate must be distributed between the two terms of Lande's equation $v_1 =\sigma_1^2 +\sigma_{1,2}$.
-More importantly, the contributions of each term will over time change if the \G matrix is not constant.\par
+More importantly, the contributions of each term will over time change if the \G matrix is not constant.
 
 Here we analyze a two-dimensional traveling wave model of an asexual population, and find that clonal interference alone, in the absence of pleiotropy, leads on average to greatly elevated variance of fitness-associated traits and to negative covariance between them.
-However, the \G matrix is highly unstable over the timescale of selective sweeps.\par
+However, the \G matrix is highly unstable over the timescale of selective sweeps.
 
@@
 We refer to each such set of individuals with the same $i$ and $j$ as a ``class'' and denote their abundances and frequencies by $n_{i,j}$ and $p_{i,j}$ respectively.
-The population's mean fitness is denoted $\bar{r}$.\par
+The population's mean fitness is denoted $\bar{r}$.
 
@@
 Beneficial mutations within the bulk are ignored because these are lost to clonal interference.
-\par
 
@@
 The steady-state patterns of genetic variation are determined by beneficial mutation-selection balance: while selection in the bulk continually acts to eliminate genetic variation, beneficial mutations create new genetic variation.
-However, the resulting patterns of genetic variation are not simple to analyze mathematically, particularly because we are interested in the way that genetic variation is  partitioned between two traits \cite{pearce2017rapid,Desai2013}.
-Moreover, the fluctuations around this steady state due to the stochasticity of the front have important consequences for patterns of genetic variation, and these are even harder to analyze \cite{fisher2013asexual} [cite hallatschek].
+However, the resulting patterns of genetic variation are not simple to analyze mathematically, particularly because we are interested in the way that genetic variation is  partitioned between two traits \citep{pearce2017rapid,Desai2013}.
+Moreover, the fluctuations around this steady state due to the stochasticity of the front have important consequences for patterns of genetic variation, and these are even harder to analyze \citep{hallatschek2011noisy,fisher2013asexual}.
 Thus, we performed numerical simulations.
@@
 The process then repeats.
-\par
 
@@
 (this is derived from \cite[Equation 16]{desai2007beneficial}).
-[Check i+1] The sojourn time $\tau$ for the appearance of the first mutation destined to establish can be transformed into the random variable $x=\int_0^\tau \pi \lambda_{i,j}(t) dt$ that has a standard exponential distribution.
+The sojourn time $\tau$ for the appearance of the first mutation destined to establish can be transformed into the random variable $x=\int_0^\tau \pi \lambda_{i,j}(t) dt$ that has a standard exponential distribution.
 The sampling of $\tau$ is be done by generating a value for  $x$ and numerically solving for $\tau$ in the integral equation.
@@
 In reality, an establishing class consists of multiple mutant lineages \cite{Desai2013}, but in practice this makes little quantitative difference (see \cite[p.1773]{desai2007beneficial}).
-\par
 
@@
 \end{equation}
-which is obtained from the probability density function of  $T$ in \cite[Equation 12]{desai2007beneficial}.\par
+which is obtained from the probability density function of  $T$ in \cite[Equation 12]{desai2007beneficial}.
 
@@
 This coupled system is solved one thousand generations into the future, which is then used to determine the next establishment event.
-\par
 
@@
 Following the burn-in period of the simulation, we recorded the existing classes in the bulk and their respective abundances.
-\par
 
@@
 \end{equation}
- expressed. Because there is no epistasis, $\bar{r} = \bar{r}_1 +\bar{r}_2$, and the instantaneous rate of adaptation of the population is $v(2U,N,s) = \sigma_1^2 +\sigma_2^2 + 2 \sigma_{1,2}$.  \par
+ expressed. Because there is no epistasis, $\bar{r} = \bar{r}_1 +\bar{r}_2$, and the instantaneous rate of adaptation of the population is $v(2U,N,s) = \sigma_1^2 +\sigma_2^2 + 2 \sigma_{1,2}$.
 
@@
 This is partly compensated for by a substantial increase in additive genetic variance (Figure \ref{fig:2}, open diamonds), because the negative covariance slows the removal of additive variance from the population.
-\par
 
@@
 This is striking, because this kind of evidence is often interpreted as evidence for trade-offs.
-\par
 
@@
 Each additional trait increases clonal interference on the focal trait by a diminishing amount, with curvature apparent even with respect to the logarithm of the number of traits (Figure \ref{fig:3}).
-This suggests that much can be learned even from the simplest case of clonal interference between only two fitness-associated traits.\par
+This suggests that much can be learned even from the simplest case of clonal interference between only two fitness-associated traits.
 
@@
 Under beneficial mutation-selection balance, the components of the \G matrix fluctuate far more than the instantaneous rate of adaptation, with transient spikes in additive genetic variance $\sigma_1^2$ mostly canceled out by transient spikes in negative covariance $\sigma_{1,2}$ (representative simulation shown in Figure \ref{fig:4}).
-As a result, fluctuations in variance and covariance can be substantially larger than those of total variance in fitness.\par
+As a result, fluctuations in variance and covariance can be substantially larger than those of total variance in fitness.
 
 \begin{figure}[!ht]
-\includegraphics[width=1\linewidth]{fig4.pdf}
+\includegraphics[width=0.905\linewidth]{fig4.pdf}
 % * <masel@email.arizona.edu> 2017-10-28T23:28:47.376Z:
@@
 Its two components, trait one variance and covariance, fluctuate significantly relative to the variance in total fitness over identical time periods.
-The trajectories were sampled from one simulation with parameters $s=0.02$, $U=10^{-5}$ and $N=10^9$.}\label{fig:4}
+The trajectories were sampled from one simulation with parameters $s=0.02$, $U=10^{-5}$ and $N=10^9$.
+The first eigenvalue is stable relative to the second, and the coefficient of variation (COV) of $\lambda_1$ is 44.1\%, while the COV of lambda 2 is 79.3\%.
+Eigenvalues of the \G matrix ($\lambda_1$ and $\lambda_2$) plotted across 10,000 generations, along with the angle $\theta$ between the first eigenvector and selection gradient (solid black line).
+The first eigenvector defines the direction that is aligned most closely with the selection gradient $\beta$, and as a result, its eigenvalue $\lambda_1$ is smaller in magnitude due to selection degrading most genetic variance along this direction.
+In contrast, genetic variance along the direction of the second eigenvector $\lambda_2$ which is not under selection varies closely with covariance.
+Changes in the angle $\theta$ between the first eigenvector and $\beta$ depend largely on the magnitude of the covariance.
+Smaller covariances indicate increases the efficiency of selection in trait with most variance, and as a result, the angle $\theta$   changes to reflect which of the traits has largest variance; trait 1 if $\theta$ positive or trait 2 if $\theta$ negative.
+}\label{fig:4}
 % * <masel@email.arizona.edu> 2017-10-28T23:32:29.075Z:
@@
 \tau_q = \frac{\log^2(s/U)}{s(2\log(Ns)-\log(s/U))}
-\end{equation}\par
+\end{equation}
 
@@
 It shows that covariances between traits are largely set by the behavior at the nose of the one dimensional traveling wave, which consists of genotypes along the lead fitness isocline through the stochastic front.
-The values of variances and covariance will be correlated from one transition to the next, because mutational influx depends on the abundances of adjacent less-fit genotypes.\par
+The values of variances and covariance will be correlated from one transition to the next, because mutational influx depends on the abundances of adjacent less-fit genotypes.
 
@@
 \end{figure}
-
 Fluctuations in the individual components $\sigma_1^2$, $\sigma_1^2$, and $\sigma_{1,2}$ are large relative to the expected rate of adaptation of each trait (Figure \ref{fig:4}), but the shape and orientation of the \G matrix are largely preserved throughout its evolution.
 The \G matrix expressed in Equation \ref{eq:4} for two symmetric traits ($E[\sigma_1^2] = E[ \sigma_2^2]$ and $v_1 = v_2 $ ) has eigenvalues
-\[ \lambda = \frac{1}{2}\left( \sigma_1^2+ \sigma_2^2 \pm \sqrt{ (\sigma_1^2 -\sigma_2^2)^2 +4 \sigma_{1,2}^2 }\right) \approx \frac{1}{2}\left(\sigma_1^2+ \sigma_2^2\right) \pm  \sigma_{1,2}.  \]
+\[ \lambda =  \frac{\left(\sigma_1^2+ \sigma_2^2\right)}{2}\pm \frac{1}{2} \sqrt{ (\sigma_1^2 -\sigma_2^2)^2 +4 \sigma_{1,2}^2 } \approx \frac{\left(\sigma_1^2+ \sigma_2^2\right)}{2} \pm  |\sigma_{1,2}|.
+\]
 The first eigenvalue is the rate of adaptation in one trait, while the second is the variance in the difference between the two traits.
 The principle components are $\vec{x}_1 = \frac{1}{\sqrt{2}}(1,1)^T$ and $\vec{x}_2 =\frac{1}{\sqrt{2}}(1,-1)^T$ for the first and second eigenvalues, respectively.
+
+The total variance in fitness $\sigma^2 = \beta^T \textbf{G} \beta = \lambda_1 \langle \beta,x_1 \rangle^2+ \lambda_2 \langle \beta,x_2 \rangle^2  = \sqrt{2}[\lambda_1 + (\lambda_2 -\lambda_1)\sin^2(\phi)]$, where $\phi $ is the angle between the selection gradient and the first eigenvector.
 
@@
 % ^.
+
 \label{sec:discussion}
-% (Genetic Constraints on the Rate of Adaptation)
-% \cite{phillips1989visualizing} discusses a unifying framework for micro- and macro-evolution based on a unification of Lande's quantitative genetics theory \cite{lande1979quantitative} and Simpson's adaptive landscapes \cite{simpson1944tempo}.
-His discussion centers on the role of the |G matrix, which encodes genetic constraints, in defining the local behavior of evolutionary trajectories on the fitness landscape.
-Aspects of this framework has been tested emprically by many.
-\cite{Schluter1996} analysis of empirical data on threespine stickleback populations showed that phenotypic responses over the past four million were biased towards \G's leading eigenvector ($g_{max}$); similar patterns were found in data on mice, sparrows, and finches.
-Studies like these have prompted subsequent investigations of the impact that genetic constraints might have on the rate of adaptation.
-\cite{teplitsky2014assessing} for example, estimated population-specific \G matrices and multivariate selection coefficients form long-term data for seven bird species (10 populations).
-He found that genetic correlations reduced the rate of adaptation by 28\% on average, and multivariate evolvability tended to be systematically lower in the direction of selection compared to others (see also Teplitsky et al.
-2011).
-\cite{kopp2014rapid} provides a review of this subject in relation to the potential for rapid adaptation to environmental changes given genetic constraints, phenotypic plasticity, and evolvability, based on different models.
-Their discussion of genetic constraints in quantitative genetic models review most of the finding expressed in past papers (see Schluter 1996, Phillips et al 1989, Hansen and Houle 2008).
-It is still unclear whether genetic constraints do in general limit the rate of adaptation, since other related studies have not been so conclusive on the issue \citep{Agrawal2009,hine2014evolutionary}.
-In addition, different metrics have been used measure the impact of genetic constraints on rates of adaptation using \G \citep{HANSEN2008,Kirkpatrick2009,Agrawal2009}, and there subtle differences in what they indicate about constraints (see Kopp et al.
-2014).
-Some of the issues may in fact be related to the use of \G itself \citep{chebib2017affects}, which may not truly measure genetic constraints at all.
-The role of genetic constraints must be addressed.
-\cite{matuszewski2015catch} showed in their work that standing genetic variation has greater potential for facilitating adaptation to environmental change than de novo beneficial mutations.
-In their approach, which combined those of adaptive walks models (discrete substitutions from new mutations) and quantitative genetic models (adaptation from standing genetic variation), and they found that populations that adapt from standing genetic variation are capable of significantly more phenotype change over time, and thus, have a higher potential to adapt when faced with environmental change that is fast, rather than slow.
+Although we have not included deleterious mutations in our model, their effects on the dynamics of the \G matrix will be negligible in most situations involving rapidly adapting populations \citep{desai2007beneficial,fogle2008clonal,rouzine2008traveling,good2012distribution,yu2010asymptotic}.
+In large populations such as ours, evolution is dominated by the influx of beneficial mutations as long as the fitness effect of deleterious mutations $s_d$ are on the order of $s$ or larger ($s_d \geq s$).
+Clonal interference prevents deleterious lineages from growing to any significant size that might change the behavior of two-dimensional distribution, just as it does with beneficial mutations occurring within the bulk.
+This will hold as long as the deleterious mutation rate isn't too large relative to the beneficial one.
+Still, prior work by \cite{goyal2012dynamic} shows that significantly less than a half of all mutations must be beneficial in order to ensure that adaptation.
+Deleterious mutations with fitness effects $s_d \leq s$ present much more subtle issues due to their ability to hitchhike, and their effects of on the adaptation of asexuals is still not relatively well understood \cite{good2014deleterious}.
+However, these mutations could potentially influence the evolution of the \G matrix in several important ways and should be explored in future work.
+% * <masel@email.arizona.edu> 2018-03-07T18:09:01.393Z:
+% 
+% > Although we have not included deleterious mutations in our model
+% Start with discussing what background selection is, how it fits into the paper, why it might matter, before arguing that it doesn't.
+% 
+% ^.
 
-% ((General theory for changes in G))
-Quantitative genetic models for the evolution of \G indicate that selection, drift, mutation and recombination each induce changes in genetic variances and covariances over time.
-Lande's Gaussian framework in particular is often used to specify a decomposition of changes in the \G matrix into three parts represented in the expression:
+The effects of selection, drift, mutation and recombination on the evolution of \G is given by the expression:
 \[\Delta G = G(\gamma - \beta \cdot \beta^T)G + 2 M + 2\sum_{i,j=1}^n r_{i,j}(C_{i,j}-C_{i,j}'),\]
-which is given in \cite[Chapter 20, Box 3]{fox2006evolutionary}.
-The first term describes the action of selection, which includes directional selection eroding genetic variation ($\beta$ terms), as well as stabilizing and correlational selection which reorienting the G matrix ($\gamma$ term).
+given in \cite[Chapter 20, Box 3]{fox2006evolutionary}.
+The first term includes directional selection ($\beta$ terms) and the effects of stabilizing and correlational selection which reorienting the G matrix ($\gamma$ term).
 The second term involves the mutational covariance matrix \textbf{M}, and gives the sum of mutational covariances generated by the effects of pleiotropic mutations at each locus.
@@
 when a mutation at a single locus simultaneously changes the value of the two traits (due to pleiotropy) and these effects are correlated, mutation can generate extreme stability of G.
-\cite{Jones2007} examine a two trait system and find that mutational correlation does evolve in response to selection on the bivariate phenotype (mutation is third quant-trait).\par
+\cite{Jones2007} examine a two trait system and find that mutational correlation does evolve in response to selection on the bivariate phenotype (mutation is third quant-trait).
+
+\cite{Johnson2005} has a discussion of the effects of mutation on quantitative genetic variation that should be discussed here.
 
@@
 % \end{equation} 
-% As the population adaptsAcknowledgments, mutations and selection  increase and decrease the amounts LD which ultimately determine how traits interact in their evolution.\par
+% As the population adaptsAcknowledgments, mutations and selection  increase and decrease the amounts LD which ultimately determine how traits interact in their evolution.
 
@@
 in asexuals; this fact has, for example, been the basis of explanations for the evolutionary advantages of sex and recombination \citep{Barton2005,Otto2009}
+
+Environmental change affects genotype $\times$ environment interactions and hence changes the \G matrix stability, so to assume a constant \G matrix, one must assume the absence of GxE interactions.
+Simulations have revealed that large population sizes and strong correlational selection can induce stability in \G's dynamics \cite{Jones2003}.
+When mutational covariances can respond to selection, the \M matrix and hence the \G matrix is more stable \cite{Jones2007}.
+
 \end{document}Acknowledgments

### 2018-03-08 plots.py
finished updating figure 4 and created figure 7

@@ -286,16 +286,19 @@ ax.axhline(linewidth=0.5, color = 'k')
 # *********************************************************************************
 # plots for figure 4b - The eigenvalues of the G matrix
 
-my_yticks = [(i-3) for i in range(14)]
-my_ylabel = ['', '-2', '', '0', '', '2', '', '4', '', '6', '', '8', '', '10']
+my_yticks = [(i-2) for i in range(8)]
+my_ylabel = ['-2', '', '0', '', '2', '', '4','']
 
 ax=plt.subplot(312)
-ax.plot(times,(1/vU_thry)*lambda1,c="black",linewidth=2.0,linestyle="-",label='$\lambda_1$')
-ax.plot(times,(1/vU_thry)*lambda2,c="black",linewidth=2.0,linestyle=":",label='$\lambda_2$')
-ax.set_ylabel(r'$\lambda$ / $v(U,N,s)$',fontsize=lsize1)
+ax.plot(times,(1/np.mean(lambda1))*lambda1,c="black",linewidth=2.0,linestyle="-",label='$\lambda_1$ / $\overline{\lambda}_1$')
+ax.plot(times,(1/np.mean(lambda2))*lambda2,c="black",linewidth=2.0,linestyle=":",label='$\lambda_2$ / $\overline{\lambda}_2$')
+ax.annotate('',xy=(1.05e4,3.9), xycoords='data',xytext=(1.05e4,4.9),arrowprops=dict(facecolor='black', shrink=0.01, width=1, headwidth=8))
+ax.annotate('',xy=(1.12e4,2.1), xycoords='data',xytext=(1.12e4,3.1),arrowprops=dict(facecolor='black', shrink=0.01, width=1, headwidth=8))
+ax.annotate('',xy=(1.2e4,0.2), xycoords='data',xytext=(1.2e4,-0.8),arrowprops=dict(facecolor='black', shrink=0.01, width=1, headwidth=8))
+ax.set_ylabel(r'Normalized eignvalues of G',fontsize=lsize1)
 ax.axhline(linewidth=0.5, color = 'k')
 ax.set_xlim((1e4,2e4))
-ax.set_ylim((-3,10))
+ax.set_ylim((-2,5))
 ax.legend(loc=4,ncol=2,fontsize=lsize1)
 ax.grid(b='off', which='both', axis='both')
 ax.tick_params(labelbottom='on',labelleft='on',labelright='off',axis='both',labelsize=lsize2)
[…]
